### 2022-01-13 manuscript/manuscript.tex
exploration plus manuscript init

\documentclass[]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
\usepackage[hyphens]{url}

  \journal{An awesome journal} % Sets Journal name

\usepackage{lineno} % add
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{graphicx}
%%%%%%%%%%%%%%%% end my additions to header

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Short Paper},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}

% Pandoc citation processing
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

% Pandoc header

\begin{document}
\begin{frontmatter}

  \title{Short Paper}
    \author[Some Institute of Technology]{Alice Anonymous\corref{1}}
   \ead{alice@example.com} 
    \author[Another University]{Bob Security}
   \ead{bob@example.com} 
    \author[Another University]{Cat Memes\corref{2}}
   \ead{cat@example.com} 
    \author[Some Institute of Technology]{Derek Zoolander\corref{2}}
   \ead{derek@example.com} 
      \address[Some Institute of Technology]{Department, Street, City,
State, Zip}
    \address[Another University]{Department, Street, City, State, Zip}
      \cortext[1]{Corresponding Author}
    \cortext[2]{Equal contribution}

  \begin{abstract}
  This is the abstract.

  It consists of two paragraphs.
  \end{abstract}

 \end{frontmatter}

\emph{Text based on elsarticle sample manuscript, see
\url{http://www.elsevier.com/author-schemas/latex-instructions\#elsarticle}}

\hypertarget{the-elsevier-article-class}{%
\section{The Elsevier article class}\label{the-elsevier-article-class}}

\hypertarget{installation}{%
\paragraph{Installation}\label{installation}}

If the document class \emph{elsarticle} is not available on your
computer, you can download and install the system package
\emph{texlive-publishers} (Linux) or install the LaTeX package
\emph{elsarticle} using the package manager of your TeX installation,
which is typically TeX Live or MikTeX.

\hypertarget{usage}{%
\paragraph{Usage}\label{usage}}

Once the package is properly installed, you can use the document class
\emph{elsarticle} to create a manuscript. Please make sure that your
manuscript follows the guidelines in the Guide for Authors of the
relevant journal. It is not necessary to typeset your manuscript in
exactly the same way as an article, unless you are submitting to a
camera-ready copy (CRC) journal.

\hypertarget{functionality}{%
\paragraph{Functionality}\label{functionality}}

The Elsevier article class is based on the standard article class and
supports almost all of the functionality of that class. In addition, it
features commands and options to format the

\begin{itemize}
\item
  document style
\item
  baselineskip
\item
  front matter
\item
  keywords and MSC codes
\item
  theorems, definitions and proofs
\item
  lables of enumerations
\item
  citation style and labeling.
\end{itemize}

\hypertarget{front-matter}{%
\section{Front matter}\label{front-matter}}

The author names and affiliations could be formatted in two ways:

\begin{enumerate}
\def\labelenumi{(\arabic{enumi})}
\item
  Group the authors per affiliation.
\item
  Use footnotes to indicate the affiliations.
\end{enumerate}

See the front matter of this document for examples. You are recommended
to conform your choice to the journal you are submitting to.

\hypertarget{bibliography-styles}{%
\section{Bibliography styles}\label{bibliography-styles}}

There are various bibliography styles available. You can select the
style of your choice in the preamble of this document. These styles are
Elsevier styles based on standard styles like Harvard and Vancouver.
Please use BibTeX~to generate your bibliography and include DOIs
whenever available.

Here are two sample references: (Dirac, 1953; Feynman and Vernon Jr.;
1963).

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-Dirac1953888}{}%
Dirac, P.A.M., 1953. The lorentz transformation and absolute time.
Physica 19, 888--896.
doi:\href{https://doi.org/10.1016/S0031-8914(53)80099-6}{10.1016/S0031-8914(53)80099-6}

\leavevmode\hypertarget{ref-Feynman1963118}{}%
Feynman, R.P., Vernon Jr., F.L., 1963. The theory of a general quantum
system interacting with a linear dissipative system. Annals of Physics
24, 118--173.
doi:\href{https://doi.org/10.1016/0003-4916(63)90068-X}{10.1016/0003-4916(63)90068-X}

\end{CSLReferences}

\end{document}

### 2022-01-13 analysis/exploration.R

@@ -7,6 +7,28 @@
 ## ts causality
 "https://towardsdatascience.com/inferring-causality-in-time-series-data-b8b75fe52c46"
 "https://jofam.github.io/covidBE_analysis/Overdispersion.html"
+"https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7558856/"
+"https://github.com/rollerb/cods"
+"https://www.sciencedirect.com/science/article/pii/S105752192030140X?casa_token=Rm23h3yv9y4AAAAA:-spSI7y0xg4yH_-JLt_KsAzwIX1WNvM8-2QE6nup83Dpx1LiJI0hTTFOCRVhL2GnWSigRcHGMA
+https://www.tandfonline.com/doi/full/10.1080/02692171.2020.1864301?casa_token=dhB_NdvreHgAAAAA%3AZRV6FGrikueeIQCBlx_ahepZ4J1oH84r0FOdBBQCDMnQOb7d5bzTdNph3U7ZTeSfxY8p-g5n8m_fQZDOMg
+https://www.sciencedirect.com/science/article/pii/S1544612320308497?casa_token=UQ0NiuBKKfUAAAAA:FBGglVAA5LQZlalUaDLqXicL0AjDBMjMd5WiuFFJdS9zwDNQk5iMANBa8H28GjbOZ-F0_aSirA
+https://link.springer.com/article/10.1007/s11356-021-12938-2
+https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0244920
+https://www.emerald.com/insight/content/doi/10.1108/SEF-09-2020-0385/full/html?casa_token=v7AqtRlwgDAAAAAA:bA-cYq18HmVxWAOW3bKUKfSx_FJ-dr5WrQ1WYdgsqbDJHJJ1uZN1h2GxOVWdnM_HT9Kj9ZCz6kW1tDphWm6YGs3vc4x9dkEPP3Dd8gp5BE-t-RHwtemv
+https://www.sciencedirect.com/science/article/pii/S0301420720309296?casa_token=Gf4xUdEoi_YAAAAA:PapQN5Tvb94DVnUFUAOEyDdOnQAN8buDMUm8cgO0fRtwdIbJ7ud78x2Ah5c5CLpcY2c52WrYsA
[…]

### 2022-01-13 manuscript/manuscript.Rmd
literature review

---
title: Influence of School Closure on COVID-19 Contaminations in Ireland and Repercussions across Age Groups
author:
  - name: Edgar Morgenroth
    email: edgar.morgenroth@dcu.ie
    affiliation: Dublin City University
    footnote: 1
  - name: Damien Dupré
    email: damien.dupre@dcu.ie
    affiliation: Dublin City University
    footnote: 2
address:
  - code: Dublin City University
    address: 	Glasnevin, Dublin 9, Ireland
footnote:
  - code: 1
    text: "Corresponding Author"
  - code: 2
    text: "Equal contribution"
abstract: |
  TBD

journal: "TBD"
date: "`r Sys.Date()`"

csl: ["csl/elsevier-harvard.csl"]
bibliography: ["bib/reference_list.bib"]

#linenumbers: true
#numbersections: true

output: rticles::elsevier_article
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("output"),
      output_file = glue::glue("{xfun::sans_ext(input)}_{Sys.Date()}.pdf"),
      envir = globalenv()
    )
  })
---

Introduction
==========================

Since the onset of Covid-19, contradicting messages are circulating about the spread of the virus in schools. Sometimes presented as a safe environment [@falk2021covid; @walger2020children], at other times designed as an aggravating factor of the pandemic [@lopez2020transmission; @meuris2021transmission], child care facilities and schools have been closed multiple times in most countries to protect the public health. However, the efficacy of school closure is still unknown and little is know about how Covid-19 Cases spread across age cohort and especially from the younger cohort being at school to older cohorts including parents, grand parents and relatives. Some evidence show that school is an influential environment where the virus is likely to spread. Once contaminated at school, a common spread believe is that the children and teenagers will then contaminate their parents. Therefore, among the political actions to stop the spread of Covid-19, the closure of primary and secondary schools was widely adopted worldwide. This research aims to investigate 1) the effect of school closure on the evolution of Covid-19 cases and 2) the temporal relationship between the rise of cases in children and teenagers with the rise of cases of older age group.

## Influence of School Closure

One of the strongest assumption of government actions is that by reducing social contact will reduce the spread of the virus. Therefore, it is legitimate to believe that by closing schools, a reduction of the contaminations would be observed in the younger age groups. However, the efficiency of school closure on the reduction of Covid-19 cases is still questioned. While some research have observed that school closures contribute to limit or to reduce the growth rate of confirmed cases after implementation [@yoshiyuki2020effects; @stage2021shut], other did not observe a change in the evolution of Covid-19 cases [@iwata2020school; @chang2020modelling]. For instance, a controlled comparison between similar localities in Japan with schools closed and school open did not revealed any evidence that school closures reduced the spread of Covid-19 [@fukumoto2021no]. If the school closure had a real impact on the evolution of confirmed Covid-19 cases, it should be possible to observe a decrease or a least an inflection in the trend of its evolution among younger age groups.

## Causal Relationship Between Age Groups

A second implicit believe regarding the effect of school closure on the spread of Covid-19 is that school not only has an effect on to the spread of the virus in children and teenagers but also has a knock-on effect on the spread of the virus in older age groups also called Secondary Attack Rate (SAR). The contaminated children and teenagers would bring the virus back at home and, then, they will contaminate their parents and relatives. Despite being reported in several clinical and epidemiological studies [@zhendong2020clinical; @siebach2021childhood], multiple research have shown that the SAR from children to household members was, in fact, lower than expected [@kim2021role; @ludvigsson2020children; @vanderhoek2020role; @heavey2020evidence]. However, the SAR of children and teenagers to the household member is likely to be age-dependent, with difference between infants, primary and secondary school children, and  college students [@grasleguen2021reopening]. A research investigating the contamination in the household network not only revealed an exceptional high rate of secondary contamination but also that these contamination happened when the school were closed [@sorianoarandes2021household].

Method
============

Contrary to Granger Causality, Transfer Entropy is robust to non-stationary time series [@silini2021fast]. The Transfer Entropy analysis of younger age groups daily changes in Covid-19 cases on older age groups is done using the R package RTransferEntropy [@behrendt2019rtransferentropy].

Results
===================

Discussion
===================

Knowledge about the transmission of the virus significantly improved with the amount of studies performed, especially in the case of how the virus behaves with children. From the early analyses showing that the virus was instantness in children [@li2020role], the results has changed to more nuanced position which states that the spread of the virus in children is moderate.

Conclusion
===================

References {#references .unnumbered}
==========

### 2022-01-13 output/manuscript_2022-01-13.tex
literature review

\documentclass[]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
\usepackage[hyphens]{url}

  \journal{TBD} % Sets Journal name

\usepackage{lineno} % add
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{graphicx}
%%%%%%%%%%%%%%%% end my additions to header

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Influence of School Closure on COVID-19 Contaminations in Ireland and Repercussions across Age Groups},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}

% Pandoc citation processing
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

% Pandoc header

\begin{document}
\begin{frontmatter}

  \title{Influence of School Closure on COVID-19 Contaminations in
Ireland and Repercussions across Age Groups}
    \author[Dublin City University]{Edgar Morgenroth\corref{1}}
   \ead{edgar.morgenroth@dcu.ie} 
    \author[Dublin City University]{Damien Dupré\corref{2}}
   \ead{damien.dupre@dcu.ie} 
      \address[Dublin City University]{Glasnevin, Dublin 9, Ireland}
      \cortext[1]{Corresponding Author}
    \cortext[2]{Equal contribution}

  \begin{abstract}
  TBD
  \end{abstract}

 \end{frontmatter}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Since the onset of Covid-19, contradicting messages are circulating
about the spread of the virus in schools. Sometimes presented as a safe
environment (Falk et al., 2021; Walger et al., 2020), at other times
designed as an aggravating factor of the pandemic (Lopez et al., 2020;
Meuris et al., 2021), child care facilities and schools have been closed
multiple times in most countries to protect the public health. However,
the efficacy of school closure is still unknown and little is know about
how Covid-19 Cases spread across age cohort and especially from the
younger cohort being at school to older cohorts including parents, grand
parents and relatives. Some evidence show that school is an influential
environment where the virus is likely to spread. Once contaminated at
school, a common spread believe is that the children and teenagers will
then contaminate their parents. Therefore, among the political actions
to stop the spread of Covid-19, the closure of primary and secondary
schools was widely adopted worldwide. This research aims to investigate
1) the effect of school closure on the evolution of Covid-19 cases and
2) the temporal relationship between the rise of cases in children and
teenagers with the rise of cases of older age group.

\hypertarget{influence-of-school-closure}{%
\subsection{Influence of School
Closure}\label{influence-of-school-closure}}

One of the strongest assumption of government actions is that by
reducing social contact will reduce the spread of the virus. Therefore,
it is legitimate to believe that by closing schools, a reduction of the
contaminations would be observed in the younger age groups. However, the
efficiency of school closure on the reduction of Covid-19 cases is still
questioned. While some research have observed that school closures
contribute to limit or to reduce the growth rate of confirmed cases
after implementation (Stage et al., 2021; Sugishita, 2020), other did
not observe a change in the evolution of Covid-19 cases (Chang et al.,
2020; Iwata et al., 2020). For instance, a controlled comparison between
similar localities in Japan with schools closed and school open did not
revealed any evidence that school closures reduced the spread of
Covid-19 (Fukumoto et al., 2021). If the school closure had a real
impact on the evolution of confirmed Covid-19 cases, it should be
possible to observe a decrease or a least an inflection in the trend of
its evolution among younger age groups.

\hypertarget{causal-relationship-between-age-groups}{%
\subsection{Causal Relationship Between Age
Groups}\label{causal-relationship-between-age-groups}}

A second implicit believe regarding the effect of school closure on the
spread of Covid-19 is that school not only has an effect on to the
spread of the virus in children and teenagers but also has a knock-on
effect on the spread of the virus in older age groups also called
Secondary Attack Rate (SAR). The contaminated children and teenagers
would bring the virus back at home and, then, they will contaminate
their parents and relatives. Despite being reported in several clinical
and epidemiological studies (Siebach et al., 2021; Zhen-Dong et al.,
2020), multiple research have shown that the SAR from children to
household members was, in fact, lower than expected (Heavey et al.,
2020; Hoek et al., 2020; Kim et al., 2021; Ludvigsson, 2020). However,
the SAR of children and teenagers to the household member is likely to
be age-dependent, with difference between infants, primary and secondary
school children, and college students (Gras-Le Guen et al., 2021). A
research investigating the contamination in the household network not
only revealed an exceptional high rate of secondary contamination but
also that these contamination happened when the school were closed
(Soriano-Arandes et al., 2021).

\hypertarget{method}{%
\section{Method}\label{method}}

Contrary to Granger Causality, Transfer Entropy is robust to
non-stationary time series (Silini and Masoller, 2021). The Transfer
Entropy analysis of younger age groups daily changes in Covid-19 cases
on older age groups is done using the R package RTransferEntropy
(Behrendt et al., 2019).

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

Knowledge about the transmission of the virus significantly improved
with the amount of studies performed, especially in the case of how the
virus behaves with children. From the early analyses showing that the
virus was instantness in children (Li et al., 2020), the results has
changed to more nuanced position which states that the spread of the
virus in children is moderate.

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-behrendt2019rtransferentropy}{}%
Behrendt, S., Dimpfl, T., Peter, F.J., Zimmermann, D.J., 2019.
RTransferEntropy --- quantifying information flow between different time
series using effective transfer entropy. SoftwareX 10, 100265.
doi:\href{https://doi.org/10.1038/s41598-021-87818-3}{10.1038/s41598-021-87818-3}

\leavevmode\hypertarget{ref-chang2020modelling}{}%
Chang, S.L., Harding, N., Zachreson, C., Cliff, O.M., Prokopenko, M.,
2020. Modelling transmission and control of the COVID-19 pandemic in
australia. Nature communications 11, 1--13.
doi:\href{https://doi.org/10.1038/s41467-020-19393-6}{10.1038/s41467-020-19393-6}

\leavevmode\hypertarget{ref-falk2021covid}{}%
Falk, A., Benda, A., Falk, P., Steffen, S., Wallace, Z., Høeg, T.B.,
2021. COVID-19 cases and transmission in 17 k--12 schools---wood county,
wisconsin, august 31--november 29, 2020. Morbidity and Mortality Weekly
Report 70, 136--140.
doi:\href{https://doi.org/10.15585/mmwr.mm7004e3}{10.15585/mmwr.mm7004e3}

\leavevmode\hypertarget{ref-fukumoto2021no}{}%
Fukumoto, K., McClean, C.T., Nakagawa, K., 2021. No causal effect of
school closures in japan on the spread of COVID-19 in spring 2020.
Nature medicine 27, 2111--2119.
doi:\href{https://doi.org/10.1038/s41591-021-01571-8}{10.1038/s41591-021-01571-8}

\leavevmode\hypertarget{ref-grasleguen2021reopening}{}%
Gras-Le Guen, C., Cohen, R., Rozenberg, J., Launay, E., Levy-Bruhl, D.,
Delacourt, C., 2021. Reopening schools in the context of increasing
COVID-19 community transmission: The french experience. Archives de
Pédiatrie 28, 178--185.
doi:\href{https://doi.org/10.1016/j.arcped.2021.02.001}{10.1016/j.arcped.2021.02.001}

\leavevmode\hypertarget{ref-heavey2020evidence}{}%
Heavey, L., Casey, G., Kelly, C., Kelly, D., McDarby, G., 2020. No
evidence of secondary transmission of COVID-19 from children attending
school in ireland, 2020. Eurosurveillance 25.
doi:\href{https://doi.org/10.2807/1560-7917.ES.2020.25.21.2000903}{10.2807/1560-7917.ES.2020.25.21.2000903}

\leavevmode\hypertarget{ref-vanderhoek2020role}{}%
Hoek, W. van der, Backer, J.A., Bodewes, R., Friesema, I., Meijer, A.,
Pijnacker, R., Reukers, D.F.M., Reusken, C., Roof, I., Rots, N., Te
Wierik, M.J.M., Gageldonk-Lafeber, A. van, Waegemaekers, C., Hof, S. van
den, 2020. The role of children in the transmission of SARS-CoV-2.
Nederlands tijdschrift voor geneeskunde 164.

\leavevmode\hypertarget{ref-iwata2020school}{}%
Iwata, K., Doi, A., Miyakoshi, C., 2020. Was school closure effective in
mitigating coronavirus disease 2019 (COVID-19)? Time series analysis
using bayesian inference. International Journal of Infectious Diseases
99, 57--61.
doi:\href{https://doi.org/10.1016/j.ijid.2020.07.052}{10.1016/j.ijid.2020.07.052}

\leavevmode\hypertarget{ref-kim2021role}{}%
Kim, J., Choe, Y.J., Lee, J., Park, Y.J., Park, O., Han, M.S., Kim,
J.-H., Choi, E.H., 2021. Role of children in household transmission of
COVID-19. Archives of Disease in Childhood 106, 709--711.
doi:\href{https://doi.org/10.1136/archdischild-2020-319910}{10.1136/archdischild-2020-319910}

\leavevmode\hypertarget{ref-li2020role}{}%
Li, X., Xu, W., Dozier, M., He, Y., Kirolos, A., Theodoratou, E.,
others, 2020. The role of children in transmission of SARS-CoV-2: A
rapid review. Journal of global health 10.
doi:\href{https://doi.org/10.7189/jogh.10.011101}{10.7189/jogh.10.011101}

\leavevmode\hypertarget{ref-lopez2020transmission}{}%
Lopez, A.S., Hill, M., Antezano, J., Vilven, D., Rutner, T., Bogdanow,
L., Claflin, C., Kracalik, I.T., Fields, V.L., Dunn, A., others, 2020.
Transmission dynamics of COVID-19 outbreaks associated with child care
facilities---salt lake city, utah, april--july 2020. Morbidity and
Mortality Weekly Report 69, 1319.
doi:\href{https://doi.org/10.15585/mmwr.mm6937e3}{10.15585/mmwr.mm6937e3}

\leavevmode\hypertarget{ref-ludvigsson2020children}{}%
Ludvigsson, J.F., 2020. Children are unlikely to be the main drivers of
the COVID-19 pandemic -- a systematic review. Acta Paediatrica 109,
1525--1530. doi:\url{https://doi.org/10.1111/apa.15371}

\leavevmode\hypertarget{ref-meuris2021transmission}{}%
Meuris, C., Kremer, C., Geerinck, A., Locquet, M., Bruyère, O., Defêche,
J., Meex, C., Hayette, M.-P., Duchene, L., Dellot, P., Azarzar, S.,
Maréchal, N., Sauvage, A.-S., Frippiat, F., Giot, J.-B., Léonard, P.,
Fombellida, K., Moutschen, M., Durkin, K., Artesi, M., Bours, V., Faes,
C., Hens, N., Darcis, G., 2021. Transmission of SARS-CoV-2 after
COVID-19 screening and mitigation measures for primary school children
attending school in liège, belgium. JAMA Network Open 4,
e2128757--e2128757.
doi:\href{https://doi.org/10.1001/jamanetworkopen.2021.28757}{10.1001/jamanetworkopen.2021.28757}

\leavevmode\hypertarget{ref-siebach2021childhood}{}%
Siebach, M.K., Piedimonte, G., Ley, S.H., 2021. COVID-19 in childhood:
Transmission, clinical presentation, complications and risk factors.
Pediatric Pulmonology 56, 1342--1356.
doi:\url{https://doi.org/10.1002/ppul.25344}

\leavevmode\hypertarget{ref-silini2021fast}{}%
Silini, R., Masoller, C., 2021. Fast and effective pseudo transfer
entropy for bivariate data-driven causal inference. Scientific Reports
11, 1--13.
doi:\href{https://doi.org/10.1016/j.softx.2019.100265}{10.1016/j.softx.2019.100265}

\leavevmode\hypertarget{ref-sorianoarandes2021household}{}%
Soriano-Arandes, A., Gatell, A., Serrano, P., Biosca, M., Campillo, F.,
Capdevila, R., Fàbrega, A., Lobato, Z., López, N., Moreno, A.M., Poblet,
M., Riera-Bosch, M.T., Rius, N., Ruiz, M., Sánchez, A., Valldepérez, C.,
Vilà, M., Pineda, V., Lazcano, U., Díaz, Y., Reyes-Urueña, J.,
Soler-Palacín, P., Catalonia Research Group, C.P.D. in, 2021. Household
severe acute respiratory syndrome coronavirus 2 transmission and
children: A network prospective study. Clinical Infectious Diseases 73,
e1261--e1269.
doi:\href{https://doi.org/10.1093/cid/ciab228}{10.1093/cid/ciab228}

\leavevmode\hypertarget{ref-stage2021shut}{}%
Stage, H.B., Shingleton, J., Ghosh, S., Scarabel, F., Pellis, L.,
Finnie, T., 2021. Shut and re-open: The role of schools in the spread of
COVID-19 in europe. Philosophical Transactions of the Royal Society B
376, 20200277.
doi:\href{https://doi.org/10.1098/rstb.2020.0277}{10.1098/rstb.2020.0277}

\leavevmode\hypertarget{ref-yoshiyuki2020effects}{}%
Sugishita, J.A.S., Yoshiyuki AND Kurita, 2020. Effects of voluntary
event cancellation and school closure as countermeasures against
COVID-19 outbreak in japan. PLOS ONE 15, 1--10.
doi:\href{https://doi.org/10.1371/journal.pone.0239455}{10.1371/journal.pone.0239455}

\leavevmode\hypertarget{ref-walger2020children}{}%
Walger, P., Heininger, U., Knuf, M., Exner, M., Popp, W., Fischbach, T.,
Trapp, S., Hübner, J., Herr, C., Simon, A., others, 2020. Children and
adolescents in the CoVid-19 pandemic: Schools and daycare centers are to
be opened again without restrictions. The protection of teachers,
educators, carers and parents and the general hygiene rules do not
conflict with this. GMS hygiene and infection control 15.
doi:\href{https://doi.org/10.3205/dgkh000346}{10.3205/dgkh000346}

\leavevmode\hypertarget{ref-zhendong2020clinical}{}%
Zhen-Dong, Y., Gao-Jun, Z., Run-Ming, J., Zhi-Sheng, L., Zong-Qi, D.,
Xiong, X., Guo-Wei, S., 2020. Clinical and transmission dynamics
characteristics of 406 children with coronavirus disease 2019 in china:
A review. Journal of Infection 81, e11--e15.
doi:\href{https://doi.org/10.1016/j.jinf.2020.04.030}{10.1016/j.jinf.2020.04.030}

\end{CSLReferences}

\end{document}

### 2022-01-13 analysis/exploration.R

@@ -29,11 +29,6 @@ https://www.scirp.org/journal/paperinformation.aspx?paperid=113444
 https://dl.acm.org/doi/pdf/10.1145/3441452
 https://www.nature.com/articles/s41598-019-49278-8
 "
-## school closure
-"https://www.nature.com/articles/s41591-021-01571-8"
-"https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8165592/pdf/rstb.2020.0277.pdf"
-"https://www.ijidonline.com/article/S1201-9712(20)30598-1/fulltext"
-"https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0239455"
 
 # libraries --------------------------------------------------------------------
 library(tidyverse)
@@ -392,7 +387,7 @@ NlinTS::causality.test(df_wide_changes$aged1to4, df_wide_changes$aged15to24, lag
 ### see corresponding paper https://www.sciencedirect.com/science/article/pii/S2352711019300779
 library(RTransferEntropy)
 df_model <- transfer_entropy(df_wide_7davg$aged1to4, df_wide_7davg$aged25to34);df_model
-df_model <- transfer_entropy(df_wide_7davg$aged1to4, df_wide_7davg$aged45to54);df_model
+df_model <- transfer_entropy(df_wide_7davg$aged1to4, df_wide_7davg$aged25to34);df_model
 df_model <- transfer_entropy(df_wide_7davg$aged15to24, df_wide_7davg$aged35to44);df_model
 df_model <- transfer_entropy(df_wide_7davg$aged15to24, df_wide_7davg$aged45to54);df_model
 df_model <- transfer_entropy(df$aged15to24, df$aged35to44);df_model

### 2022-01-13 manuscript/manuscript.Rmd
manuscript method

@@
 The contaminated children and teenagers would bring the virus back at home and, then, they will contaminate their parents and relatives.
+For example, a research investigating the contamination in the household network not only revealed an exceptional high rate of secondary contamination but also that these contamination happened when the school were closed [@sorianoarandes2021household].
+
 Despite being reported in several clinical and epidemiological studies [@zhendong2020clinical; @siebach2021childhood], multiple research have shown that the SAR from children to household members was, in fact, lower than expected [@kim2021role; @ludvigsson2020children; @vanderhoek2020role; @heavey2020evidence].
 However, the SAR of children and teenagers to the household member is likely to be age-dependent, with difference between infants, primary and secondary school children, and college students [@grasleguen2021reopening].
-A research investigating the contamination in the household network not only revealed an exceptional high rate of secondary contamination but also that these contamination happened when the school were closed [@sorianoarandes2021household].
+If a secondary transmission from children and teenagers to household member has a significant influence, then a temporal causality relationship between their evolution should be observed.
 
@@
 ============
+
+## Observations
+
+The data were collected from the official daily publication of Covid-19 cases by the Irish Department of Health.
+They include the amount of daily cases for 10 age groups from March 4th, 2020 to January 1st, 2022.
+For each day during this period, school closure either due to a government decision or a public holiday is determined.
+
+## Data Pre-processing
+
+A 7-day rolling average of the day-by-by change in Covid-19 cases reported (i.e., lag 1) is used to evaluate the efficiency of school closures.
+This rolling average approximate the real day-by-by change in Covid-19 cases reported which can be artificially increased or decreased due to days under- or over-reporting cases such as weekend or bank holidays for example.
+
+## Data Analysis
+
+Four periods of school closure longer that 14 days have been identified.
+These four periods are used to fit a Generalised Additive Model (or GAM) using the R package mgcv [@wood2017generalized] in order to test the hypothesis of a significant change in the evolution of cases among age groups from 1 to 4, from 5 to 14, and from 15 to 24.
+
+By estimating the degree of smoothness of a Bayesian spline smoothing using restricted maximum likelihood estimation [@wood2011fast], GAMs allow the identification of dynamic patterns underlying the day-by-day changes in Covid-19 cases reported while taking into account the random effect of different age groups as follows:
+
+\begin{align}
+Y_{is} = \alpha_{i} + f(X_{s}) + a_{is} + \epsilon_{is}
+\end{align}
+
+\noindent where $i$ is an age group among the 10 age groups investigated and $s$ is the date corresponding to the day-by-day change in Covid-19 cases.
+$Y_{is}$ represents the day-by-day change in Covid-19 cases assuming a negative binomial distribution for the fitting [@loader2006local].
+The response variable $Y_{is}$ includes a specific intercept for each age group ($\alpha_{i}$).
+A smooth effect over time $f(X_{s})$ is applied to model (Eq 1) to predict the nonlinear evolution of day-by-day change in Covid-19 cases.
+This smooth effect $f(X_{s})$ is built up in basic components, called the basis functions $b_{j}(X_{s})$, such that:
+
+\begin{align}
+f(X_{s})=\sum_{j=1}^{k}\beta_{j}\times b_{j}(X_{s})
+\end{align}
+
+\noindent where the regression parameters $\beta_{j}$ are estimated by penalized likelihood maximization.
+
+The model also includes the random effects term $a_{is}= Zb_{i}$ where $Z$ is a random effects matrix and $b_{i}$ is a vector of random effects described by $b_{i} \sim N(0, D)$.
+In this, $D$ represents a covariance matrix.
+The error term $\epsilon_{is}$ is assumed to be normally and independently distributed $\epsilon_{is} \sim N(0, \sigma^2)$.
 

### 2022-01-13 output/manuscript_2022-01-13.tex
manuscript method

@@
 would bring the virus back at home and, then, they will contaminate
-their parents and relatives. Despite being reported in several clinical
-and epidemiological studies (Siebach et al., 2021; Zhen-Dong et al.,
-2020), multiple research have shown that the SAR from children to
-household members was, in fact, lower than expected (Heavey et al.,
-2020; Hoek et al., 2020; Kim et al., 2021; Ludvigsson, 2020). However,
-the SAR of children and teenagers to the household member is likely to
-be age-dependent, with difference between infants, primary and secondary
-school children, and college students (Gras-Le Guen et al., 2021). A
-research investigating the contamination in the household network not
-only revealed an exceptional high rate of secondary contamination but
-also that these contamination happened when the school were closed
-(Soriano-Arandes et al., 2021).
+their parents and relatives. For example, a research investigating the
+contamination in the household network not only revealed an exceptional
+high rate of secondary contamination but also that these contamination
+happened when the school were closed (Soriano-Arandes et al., 2021).
+
+Despite being reported in several clinical and epidemiological studies
+(Siebach et al., 2021; Zhen-Dong et al., 2020), multiple research have
+shown that the SAR from children to household members was, in fact,
+lower than expected (Heavey et al., 2020; Hoek et al., 2020; Kim et al.,
+2021; Ludvigsson, 2020). However, the SAR of children and teenagers to
+the household member is likely to be age-dependent, with difference
+between infants, primary and secondary school children, and college
+students (Gras-Le Guen et al., 2021). If a secondary transmission from
+children and teenagers to household member has a significant influence,
+then a temporal causality relationship between their evolution should be
+observed.
 
@@
 \section{Method}\label{method}}
+
+\hypertarget{observations}{%
+\subsection{Observations}\label{observations}}
+
+The data were collected from the official daily publication of Covid-19
+cases by the Irish Department of Health. They include the amount of
+daily cases for 10 age groups from March 4th, 2020 to January 1st, 2022.
+For each day during this period, school closure either due to a
+government decision or a public holiday is determined.
+
+\hypertarget{data-pre-processing}{%
+\subsection{Data Pre-processing}\label{data-pre-processing}}
+
+A 7-day rolling average of the day-by-by change in Covid-19 cases
+reported (i.e., lag 1) is used to evaluate the efficiency of school
+closures. This rolling average approximate the real day-by-by change in
+Covid-19 cases reported which can be artificially increased or decreased
+due to days under- or over-reporting cases such as weekend or bank
+holidays for example.
+
+\hypertarget{data-analysis}{%
+\subsection{Data Analysis}\label{data-analysis}}
+
+Four periods of school closure longer that 14 days have been identified.
+These four periods are used to fit a Generalised Additive Model or GAM
+(Wood, 2017) in order to test the hypothesis of a significant change in
+the evolution of cases among age groups from 1 to 4, from 5 to 14, and
+from 15 to 24.
+
+By estimating the degree of smoothness of a Bayesian spline smoothing
+using restricted maximum likelihood estimation (Wood, 2011), GAMs allow
+the identification of dynamic patterns underlying the day-by-day changes
+in Covid-19 cases reported while taking into account the random effect
+of different age groups as follows:
+
+\begin{align}
+Y_{is} = \alpha_{i} + f(X_{s}) + a_{is} + \epsilon_{is}
+\end{align}
+
+\noindent where \(i\) is an age group among the 10 age groups
+investigated and \(s\) is the date corresponding to the day-by-day
+change in Covid-19 cases. \(Y_{is}\) represents the day-by-day change in
+Covid-19 cases assuming a negative binomial distribution for the fitting
+(Loader, 2006). The response variable \(Y_{is}\) includes a specific
+intercept for each age group (\(\alpha_{i}\)). A smooth effect over time
+\(f(X_{s})\) is applied to model (Eq 1) to predict the nonlinear
+evolution of day-by-day change in Covid-19 cases. This smooth effect
+\(f(X_{s})\) is built up in basic components, called the basis functions
+\(b_{j}(X_{s})\), such that:
+
+\begin{align}
+f(X_{s})=\sum_{j=1}^{k}\beta_{j}\times b_{j}(X_{s})
+\end{align}
+
+\noindent where the regression parameters \(\beta_{j}\) are estimated by
+penalized likelihood maximization.
+
+The model also includes the random effects term \(a_{is}= Zb_{i}\) where
+\(Z\) is a random effects matrix and \(b_{i}\) is a vector of random
+effects described by \(b_{i} \sim N(0, D)\). In this, \(D\) represents a
+covariance matrix. The error term \(\epsilon_{is}\) is assumed to be
+normally and independently distributed
+\(\epsilon_{is} \sim N(0, \sigma^2)\).
 
@@
 
+\leavevmode\hypertarget{ref-loader2006local}{}%
+Loader, C., 2006. Local regression and likelihood. Springer, Boston, MA.
+
 \leavevmode\hypertarget{ref-lopez2020transmission}{}%
@@
 
+\leavevmode\hypertarget{ref-wood2011fast}{}%
+Wood, S.N., 2011. Fast stable restricted maximum likelihood and marginal
+likelihood estimation of semiparametric generalized linear models.
+Journal of the Royal Statistical Society (B) 73, 3--36.
+
+\leavevmode\hypertarget{ref-wood2017generalized}{}%
+Wood, S.N., 2017. Generalized additive models: An introduction with r.
+CRC press.
+
 \leavevmode\hypertarget{ref-zhendong2020clinical}{}%

### 2022-01-14 manuscript/manuscript.Rmd
results

@@
 
-output: rticles::elsevier_article
+header-includes:
+  - \usepackage{booktabs}
+  - \usepackage{float}
+
+output:
+  bookdown::pdf_book:
+    base_format: rticles::elsevier_article
 knit: (function(input, ...) {
@@
 ---
+
+(ref:overall) Evolution of confirmed Covid-19 cases in Ireland alongside with school closures.
+(ref:descriptive) Day-by-day changes in Covid-19 cases by age groups in Ireland: raw data (bars) and 7 aay rolling average (lines).
+
+```{r setup, include = FALSE}
+# libraries --------------------------------------------------------------------
+library(here)
+library(janitor)
+library(knitr)
+library(tidyverse)
+library(zoo)
+
+# options ----------------------------------------------------------------------
+set.seed(123) # Seed for random number generation
+options(scipen = 999) # disable sci number format
+opts_chunk$set(
+  cache.extra = rand_seed, 
+  message = FALSE, 
+  warning = FALSE, 
+  error = FALSE,
+  echo = FALSE, 
+  fig.retina = 3
+  )
+
+# variables --------------------------------------------------------------------
+age_order <- c("aged1to4", "aged5to14", "aged15to24", "aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65to74", "aged75to84", "aged85up")
+age_name  <- c("1 to 4",   "5 to 14",   "15 to 24",   "25 to 34",   "35 to 44",   "45 to 54",   "55 to 64",   "65 to 74",   "75 to 84",   "85 and up")
+age_color <- c("#999999",  "#E69F00",   "#56B4E9",    "#009E73",    "#F0E442",    "#0072B2",    "#D55E00",    "#CC79A7",    "#000000",    "#CAB2D6")
+
+# data -------------------------------------------------------------------------
+df <- 
+  read_csv(
+    here("data/age_groups.csv"), 
+    col_types = list(
+      .default = col_double(), 
+      Date = col_date(format = "%d/%m/%Y")
+    )
+  ) |> 
+  clean_names() |> 
+  rename(total_cases = total) |> 
+  arrange(date) |> 
+  mutate(
+    timeline = date - lag(date),  # check if gap between dates
+    total_changes = total_cases - lag(total_cases)
+  )
+
+df_long <- df |> 
+  select(c(date, starts_with("aged"))) |> 
+  pivot_longer(-date, names_to = "age_group", values_to = "covid_cases") |> 
+  mutate(age_group = factor(age_group, levels = age_order)) |> 
+  group_by(age_group) |> 
+  mutate(
+    covid_cases_7davg = rollmean(covid_cases, k = 7, fill = NA),
+    covid_changes = covid_cases - lag(covid_cases),
+    covid_changes_7davg = rollmean(covid_changes, k = 7, fill = NA)
+  ) |> 
+  ungroup()
+```
 
@@
 The data were collected from the official daily publication of Covid-19 cases by the Irish Department of Health.
-They include the amount of daily cases for 10 age groups from March 4th, 2020 to January 1st, 2022.
+They include the amount of daily cases for 10 age groups from `r format(min(df$date), "%B %d, %Y")` to `r format(max(df$date), "%B %d, %Y")`.
 For each day during this period, school closure either due to a government decision or a public holiday is determined.
@@
 
+### Generalised Additive Model
+
 Four periods of school closure longer that 14 days have been identified.
@@
 
-By estimating the degree of smoothness of a Bayesian spline smoothing using restricted maximum likelihood estimation [@wood2011fast], GAMs allow the identification of dynamic patterns underlying the day-by-day changes in Covid-19 cases reported while taking into account the random effect of different age groups as follows:
+By estimating the degree of smoothness of a Bayesian spline smoothing using restricted maximum likelihood estimation [@wood2011fast], GAM identifies dynamic patterns underlying the day-by-day changes in Covid-19 cases reported while taking into account the random effect of different age groups as follows:
 
@@
 
-Contrary to Granger Causality, Transfer Entropy is robust to non-stationary time series [@silini2021fast].
-The Transfer Entropy analysis of younger age groups daily changes in Covid-19 cases on older age groups is done using the R package RTransferEntropy [@behrendt2019rtransferentropy].
+### Transfer Entropy
+
+Transfer Entropy can be used to infer the temporal relationship between two time series $I$ and $J$.
+This measure indicates whether $J$ can be used to to reduce the uncertainty on the future of $I$ and, consequently, that $J$ causes $I$ [@schreiber2000measuring].
+Contrary to Wiener-Granger Causality, Transfer Entropy is robust to non-stationary time series [@wollstadt2014efficient; @papana2016detecting].
+The Transfer Entropy analysis of younger age groups daily changes in Covid-19 cases on older age groups using Shannon's calculation [@shannon1948mathematical] is done using the R package RTransferEntropy [@behrendt2019rtransferentropy].
+
+The influence of day-by-day changes in Covid-19 cases across all age groups is evaluated using Shannon's transfer entropy is given by:
+
+\begin{align}
+  T_{J \rightarrow I}(k,l) = \sum_{i,j} p\left(i_{t+1}, i_t^{(k)}, j_t^{(l)}\right) \cdot log \left(\frac{p\left(i_{t+1}| i_t^{(k)}, j_t^{(l)}\right)}{p\left(i_{t+1}|i_t^{(k)}\right)}\right),
+\end{align}
+
+\noindent where $T_{J\rightarrow I}$ consequently measures the influence of the change dynamic from an eage group $J$ to another age group $I$.
 
@@
 ===================
+
+Overall, the trend of comfirm Covid-19 cases in Ireland is similar to those of other European countries (Figure \@ref(fig:overall)).
+Indeed it is possible to observe a significant increase corresponding to each of the 4 waves of transmission (March-April 2020, December 2020-January 2021, September-October 2021 and November-December 2021).
+
+```{r overall, fig.width=4.5, fig.height=3, fig.cap="(ref:overall)", fig.pos = "!h"}
+overall_plot <- df |> 
+  ggplot(aes(date, total_cases)) +
+  geom_tile(aes(fill = factor(schools_closed), height = Inf), na.rm = TRUE, alpha = .4) +
+  geom_line(na.rm = TRUE) +
+  scale_x_date(date_breaks = "3 months", date_labels = "%b %y") +
+  scale_fill_manual(labels = c("Open", "Closed", "Na"), values = c("light blue", "red", "orange")) +
+  labs(
+    x = "Date", 
+    y = "Covid-19 Cases",
+    fill = "Schools"
+  ) +
+  theme_bw() +
+  theme(
+    legend.position = "bottom",
+    text = element_text(size = 10, family = "serif"),
+    axis.text = element_text(size = 10),
+    panel.background = element_rect(colour = "black", size = 0.1)
+  )
+
+overall_plot
+```
+
+The evolution of day-by-day changes in Covid-19 cases reveals some similarities across all age groups.
+However, the influence of each waves on the each groups has also some particularities (Figure \@ref(fig:descriptive)).
+For example, the first wave was more important among the oldest age groups whereas the third wave was more important among the youngest age groups.
+
+```{r descriptive, fig.width=4.5, fig.height=7, fig.cap="(ref:descriptive)"}
+descriptive_plot <- df_long |> 
+  ggplot() +
+  geom_col(aes(date, covid_changes), na.rm = TRUE) +
+  geom_line(aes(date, covid_changes_7davg, color = age_group), na.rm = TRUE) +
+  scale_x_date("Date", date_breaks = "3 months", date_labels = "%b %y") +
+  scale_y_continuous("Day-by-day Changes in Covid-19 Cases") +
+  scale_color_manual(labels = age_name, values = age_color) +
+  facet_grid(
+    age_group ~ ., 
+    scales = "free",
+    labeller = as_labeller(
+      c(
+        aged1to4 = "1 to 4",
+        aged5to14 = "5 to 14",
+        aged15to24 = "15 to 24", 
+        aged25to34 = "25 to 34", 
+        aged35to44 = "35 to 44", 
+        aged45to54 = "45 to 54", 
+        aged55to64 = "55 to 64", 
+        aged65to74 = "65 to 74", 
+        aged75to84 = "75 to 84", 
+        aged85up = "85 and up"
+        )
+      )
+  ) +
+  theme_bw() +
+  theme(
+    legend.position = "none",
+    strip.background = element_blank(),
+    strip.placement = "outside",
+    text = element_text(size = 10, family = "serif"),
+    axis.text = element_text(size = 10),
+    panel.background = element_rect(colour = "black", size = 0.1)
+  )
+
+descriptive_plot
+```
 

### 2022-01-14 output/manuscript_2022-01-14.tex
results

--- 2022-01-13 output/manuscript_2022-01-13.tex
@@
 \bibliographystyle{elsarticle-harv}
+\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
 \ifxetex
@@
 
-\setcounter{secnumdepth}{0}
+\setcounter{secnumdepth}{5}
 % Pandoc toggle for numbering sections (defaults to be off)
-\setcounter{secnumdepth}{0}
 
@@
 % Pandoc header
+\usepackage{booktabs}
+\usepackage{float}
 
@@
 
-  \title{Influence of School Closure on COVID-19 Contaminations in
-Ireland and Repercussions across Age Groups}
+  \title{Influence of School Closure on COVID-19 Contaminations in Ireland and Repercussions across Age Groups}
     \author[Dublin City University]{Edgar Morgenroth\corref{1}}
@@
 
-Since the onset of Covid-19, contradicting messages are circulating
-about the spread of the virus in schools. Sometimes presented as a safe
-environment (Falk et al., 2021; Walger et al., 2020), at other times
-designed as an aggravating factor of the pandemic (Lopez et al., 2020;
-Meuris et al., 2021), child care facilities and schools have been closed
-multiple times in most countries to protect the public health. However,
-the efficacy of school closure is still unknown and little is know about
-how Covid-19 Cases spread across age cohort and especially from the
-younger cohort being at school to older cohorts including parents, grand
-parents and relatives. Some evidence show that school is an influential
-environment where the virus is likely to spread. Once contaminated at
-school, a common spread believe is that the children and teenagers will
-then contaminate their parents. Therefore, among the political actions
-to stop the spread of Covid-19, the closure of primary and secondary
-schools was widely adopted worldwide. This research aims to investigate
-1) the effect of school closure on the evolution of Covid-19 cases and
-2) the temporal relationship between the rise of cases in children and
-teenagers with the rise of cases of older age group.
+Since the onset of Covid-19, contradicting messages are circulating about the spread of the virus in schools.
+Sometimes presented as a safe environment (Falk et al., 2021; Walger et al., 2020), at other times designed as an aggravating factor of the pandemic (Lopez et al., 2020; Meuris et al., 2021), child care facilities and schools have been closed multiple times in most countries to protect the public health.
+However, the efficacy of school closure is still unknown and little is know about how Covid-19 Cases spread across age cohort and especially from the younger cohort being at school to older cohorts including parents, grand parents and relatives.
+Some evidence show that school is an influential environment where the virus is likely to spread.
+Once contaminated at school, a common spread believe is that the children and teenagers will then contaminate their parents.
+Therefore, among the political actions to stop the spread of Covid-19, the closure of primary and secondary schools was widely adopted worldwide.
+This research aims to investigate 1) the effect of school closure on the evolution of Covid-19 cases and 2) the temporal relationship between the rise of cases in children and teenagers with the rise of cases of older age group.
 
 \hypertarget{influence-of-school-closure}{%
-\subsection{Influence of School
-Closure}\label{influence-of-school-closure}}
+\subsection{Influence of School Closure}\label{influence-of-school-closure}}
 
-One of the strongest assumption of government actions is that by
-reducing social contact will reduce the spread of the virus. Therefore,
-it is legitimate to believe that by closing schools, a reduction of the
-contaminations would be observed in the younger age groups. However, the
-efficiency of school closure on the reduction of Covid-19 cases is still
-questioned. While some research have observed that school closures
-contribute to limit or to reduce the growth rate of confirmed cases
-after implementation (Stage et al., 2021; Sugishita, 2020), other did
-not observe a change in the evolution of Covid-19 cases (Chang et al.,
-2020; Iwata et al., 2020). For instance, a controlled comparison between
-similar localities in Japan with schools closed and school open did not
-revealed any evidence that school closures reduced the spread of
-Covid-19 (Fukumoto et al., 2021). If the school closure had a real
-impact on the evolution of confirmed Covid-19 cases, it should be
-possible to observe a decrease or a least an inflection in the trend of
-its evolution among younger age groups.
+One of the strongest assumption of government actions is that by reducing social contact will reduce the spread of the virus.
+Therefore, it is legitimate to believe that by closing schools, a reduction of the contaminations would be observed in the younger age groups.
+However, the efficiency of school closure on the reduction of Covid-19 cases is still questioned.
+While some research have observed that school closures contribute to limit or to reduce the growth rate of confirmed cases after implementation (Stage et al., 2021; Sugishita, 2020), other did not observe a change in the evolution of Covid-19 cases (Chang et al., 2020; Iwata et al., 2020).
+For instance, a controlled comparison between similar localities in Japan with schools closed and school open did not revealed any evidence that school closures reduced the spread of Covid-19 (Fukumoto et al., 2021).
+If the school closure had a real impact on the evolution of confirmed Covid-19 cases, it should be possible to observe a decrease or a least an inflection in the trend of its evolution among younger age groups.
 
 \hypertarget{causal-relationship-between-age-groups}{%
-\subsection{Causal Relationship Between Age
-Groups}\label{causal-relationship-between-age-groups}}
+\subsection{Causal Relationship Between Age Groups}\label{causal-relationship-between-age-groups}}
 
-A second implicit believe regarding the effect of school closure on the
-spread of Covid-19 is that school not only has an effect on to the
-spread of the virus in children and teenagers but also has a knock-on
-effect on the spread of the virus in older age groups also called
-Secondary Attack Rate (SAR). The contaminated children and teenagers
-would bring the virus back at home and, then, they will contaminate
-their parents and relatives. For example, a research investigating the
-contamination in the household network not only revealed an exceptional
-high rate of secondary contamination but also that these contamination
-happened when the school were closed (Soriano-Arandes et al., 2021).
+A second implicit believe regarding the effect of school closure on the spread of Covid-19 is that school not only has an effect on to the spread of the virus in children and teenagers but also has a knock-on effect on the spread of the virus in older age groups also called Secondary Attack Rate (SAR).
+The contaminated children and teenagers would bring the virus back at home and, then, they will contaminate their parents and relatives.
+For example, a research investigating the contamination in the household network not only revealed an exceptional high rate of secondary contamination but also that these contamination happened when the school were closed (Soriano-Arandes et al., 2021).
 
-Despite being reported in several clinical and epidemiological studies
-(Siebach et al., 2021; Zhen-Dong et al., 2020), multiple research have
-shown that the SAR from children to household members was, in fact,
-lower than expected (Heavey et al., 2020; Hoek et al., 2020; Kim et al.,
-2021; Ludvigsson, 2020). However, the SAR of children and teenagers to
-the household member is likely to be age-dependent, with difference
-between infants, primary and secondary school children, and college
-students (Gras-Le Guen et al., 2021). If a secondary transmission from
-children and teenagers to household member has a significant influence,
-then a temporal causality relationship between their evolution should be
-observed.
+Despite being reported in several clinical and epidemiological studies (Siebach et al., 2021; Zhen-Dong et al., 2020), multiple research have shown that the SAR from children to household members was, in fact, lower than expected (Heavey et al., 2020; Hoek et al., 2020; Kim et al., 2021; Ludvigsson, 2020).
+However, the SAR of children and teenagers to the household member is likely to be age-dependent, with difference between infants, primary and secondary school children, and college students (Gras-Le Guen et al., 2021).
+If a secondary transmission from children and teenagers to household member has a significant influence, then a temporal causality relationship between their evolution should be observed.
 
@@
 
-The data were collected from the official daily publication of Covid-19
-cases by the Irish Department of Health. They include the amount of
-daily cases for 10 age groups from March 4th, 2020 to January 1st, 2022.
-For each day during this period, school closure either due to a
-government decision or a public holiday is determined.
+The data were collected from the official daily publication of Covid-19 cases by the Irish Department of Health.
+They include the amount of daily cases for 10 age groups from March 04, 2020 to January 01, 2022.
+For each day during this period, school closure either due to a government decision or a public holiday is determined.
 
@@
 
-A 7-day rolling average of the day-by-by change in Covid-19 cases
-reported (i.e., lag 1) is used to evaluate the efficiency of school
-closures. This rolling average approximate the real day-by-by change in
-Covid-19 cases reported which can be artificially increased or decreased
-due to days under- or over-reporting cases such as weekend or bank
-holidays for example.
+A 7-day rolling average of the day-by-by change in Covid-19 cases reported (i.e., lag 1) is used to evaluate the efficiency of school closures.
+This rolling average approximate the real day-by-by change in Covid-19 cases reported which can be artificially increased or decreased due to days under- or over-reporting cases such as weekend or bank holidays for example.
 
@@
 
+\hypertarget{generalised-additive-model}{%
+\subsubsection{Generalised Additive Model}\label{generalised-additive-model}}
+
 Four periods of school closure longer that 14 days have been identified.
-These four periods are used to fit a Generalised Additive Model or GAM
-(Wood, 2017) in order to test the hypothesis of a significant change in
-the evolution of cases among age groups from 1 to 4, from 5 to 14, and
-from 15 to 24.
+These four periods are used to fit a Generalised Additive Model (or GAM) using the R package mgcv (Wood, 2017) in order to test the hypothesis of a significant change in the evolution of cases among age groups from 1 to 4, from 5 to 14, and from 15 to 24.
 
-By estimating the degree of smoothness of a Bayesian spline smoothing
-using restricted maximum likelihood estimation (Wood, 2011), GAMs allow
-the identification of dynamic patterns underlying the day-by-day changes
-in Covid-19 cases reported while taking into account the random effect
-of different age groups as follows:
+By estimating the degree of smoothness of a Bayesian spline smoothing using restricted maximum likelihood estimation (Wood, 2011), GAM identifies dynamic patterns underlying the day-by-day changes in Covid-19 cases reported while taking into account the random effect of different age groups as follows:
 
@@
 
-\noindent where \(i\) is an age group among the 10 age groups
-investigated and \(s\) is the date corresponding to the day-by-day
-change in Covid-19 cases. \(Y_{is}\) represents the day-by-day change in
-Covid-19 cases assuming a negative binomial distribution for the fitting
-(Loader, 2006). The response variable \(Y_{is}\) includes a specific
-intercept for each age group (\(\alpha_{i}\)). A smooth effect over time
-\(f(X_{s})\) is applied to model (Eq 1) to predict the nonlinear
-evolution of day-by-day change in Covid-19 cases. This smooth effect
-\(f(X_{s})\) is built up in basic components, called the basis functions
-\(b_{j}(X_{s})\), such that:
+\noindent where \(i\) is an age group among the 10 age groups investigated and \(s\) is the date corresponding to the day-by-day change in Covid-19 cases.
+\(Y_{is}\) represents the day-by-day change in Covid-19 cases assuming a negative binomial distribution for the fitting (Loader, 2006).
+The response variable \(Y_{is}\) includes a specific intercept for each age group (\(\alpha_{i}\)).
+A smooth effect over time \(f(X_{s})\) is applied to model (Eq 1) to predict the nonlinear evolution of day-by-day change in Covid-19 cases.
+This smooth effect \(f(X_{s})\) is built up in basic components, called the basis functions \(b_{j}(X_{s})\), such that:
 
@@
 
-\noindent where the regression parameters \(\beta_{j}\) are estimated by
-penalized likelihood maximization.
+\noindent where the regression parameters \(\beta_{j}\) are estimated by penalized likelihood maximization.
 
-The model also includes the random effects term \(a_{is}= Zb_{i}\) where
-\(Z\) is a random effects matrix and \(b_{i}\) is a vector of random
-effects described by \(b_{i} \sim N(0, D)\). In this, \(D\) represents a
-covariance matrix. The error term \(\epsilon_{is}\) is assumed to be
-normally and independently distributed
-\(\epsilon_{is} \sim N(0, \sigma^2)\).
+The model also includes the random effects term \(a_{is}= Zb_{i}\) where \(Z\) is a random effects matrix and \(b_{i}\) is a vector of random effects described by \(b_{i} \sim N(0, D)\).
+In this, \(D\) represents a covariance matrix.
+The error term \(\epsilon_{is}\) is assumed to be normally and independently distributed \(\epsilon_{is} \sim N(0, \sigma^2)\).
 
-Contrary to Granger Causality, Transfer Entropy is robust to
-non-stationary time series (Silini and Masoller, 2021). The Transfer
-Entropy analysis of younger age groups daily changes in Covid-19 cases
-on older age groups is done using the R package RTransferEntropy
-(Behrendt et al., 2019).
+\hypertarget{transfer-entropy}{%
+\subsubsection{Transfer Entropy}\label{transfer-entropy}}
+
+Transfer Entropy can be used to infer the temporal relationship between two time series \(I\) and \(J\).
+This measure indicates whether \(J\) can be used to to reduce the uncertainty on the future of \(I\) and, consequently, that \(J\) causes \(I\) (Schreiber, 2000).
+Contrary to Wiener-Granger Causality, Transfer Entropy is robust to non-stationary time series (Papana et al., 2016; Wollstadt, 2014).
+The Transfer Entropy analysis of younger age groups daily changes in Covid-19 cases on older age groups using Shannon's calculation (Shannon, 1948) is done using the R package RTransferEntropy (Behrendt et al., 2019).
+
+The influence of day-by-day changes in Covid-19 cases across all age groups is evaluated using Shannon's transfer entropy is given by:
+
+\begin{align}
+  T_{J \rightarrow I}(k,l) = \sum_{i,j} p\left(i_{t+1}, i_t^{(k)}, j_t^{(l)}\right) \cdot log \left(\frac{p\left(i_{t+1}| i_t^{(k)}, j_t^{(l)}\right)}{p\left(i_{t+1}|i_t^{(k)}\right)}\right),
+\end{align}
+
+\noindent where \(T_{J\rightarrow I}\) consequently measures the influence of the change dynamic from an eage group \(J\) to another age group \(I\).
 
@@
 
+Overall, the trend of comfirm Covid-19 cases in Ireland is similar to those of other European countries (Figure \ref{fig:overall}).
+Indeed it is possible to observe a significant increase corresponding to each of the 4 waves of transmission (March-April 2020, December 2020-January 2021, September-October 2021 and November-December 2021).
+
+\begin{figure}[!h]
+\includegraphics[width=324px]{/Users/damiendcu/OneDrive/Projects/covid_ts_causality/output/manuscript_2022-01-14_files/figure-latex/overall-1} \caption{Evolution of confirmed Covid-19 cases in Ireland alongside with school closures.}\label{fig:overall}
+\end{figure}
+
+The evolution of day-by-day changes in Covid-19 cases reveals some similarities across all age groups.
+However, the influence of each waves on the each groups has also some particularities (Figure \ref{fig:descriptive}).
+For example, the first wave was more important among the oldest age groups whereas the third wave was more important among the youngest age groups.
+
+\begin{figure}
+\includegraphics[width=324px]{/Users/damiendcu/OneDrive/Projects/covid_ts_causality/output/manuscript_2022-01-14_files/figure-latex/descriptive-1} \caption{Day-by-day changes in Covid-19 cases by age groups in Ireland: raw data (bars) and 7 aay rolling average (lines).}\label{fig:descriptive}
+\end{figure}
+
 \hypertarget{discussion}{%
@@
 
-Knowledge about the transmission of the virus significantly improved
-with the amount of studies performed, especially in the case of how the
-virus behaves with children. From the early analyses showing that the
-virus was instantness in children (Li et al., 2020), the results has
-changed to more nuanced position which states that the spread of the
-virus in children is moderate.
+Knowledge about the transmission of the virus significantly improved with the amount of studies performed, especially in the case of how the virus behaves with children.
+From the early analyses showing that the virus was instantness in children (Li et al., 2020), the results has changed to more nuanced position which states that the spread of the virus in children is moderate.
 
@@
 Behrendt, S., Dimpfl, T., Peter, F.J., Zimmermann, D.J., 2019.
-RTransferEntropy --- quantifying information flow between different time
-series using effective transfer entropy. SoftwareX 10, 100265.
-doi:\href{https://doi.org/10.1038/s41598-021-87818-3}{10.1038/s41598-021-87818-3}
+RTransferEntropy --- quantifying information flow between different time series using effective transfer entropy.
+SoftwareX 10, 100265.
+doi:\href{https://doi.org/10.1016/j.softx.2019.100265}{10.1016/j.softx.2019.100265}
 
 \leavevmode\hypertarget{ref-chang2020modelling}{}%
-Chang, S.L., Harding, N., Zachreson, C., Cliff, O.M., Prokopenko, M.,
-2020. Modelling transmission and control of the COVID-19 pandemic in
-australia. Nature communications 11, 1--13.
+Chang, S.L., Harding, N., Zachreson, C., Cliff, O.M., Prokopenko, M., 2020.
+Modelling transmission and control of the COVID-19 pandemic in australia.
+Nature communications 11, 1--13.
 doi:\href{https://doi.org/10.1038/s41467-020-19393-6}{10.1038/s41467-020-19393-6}
@@
 \leavevmode\hypertarget{ref-falk2021covid}{}%
-Falk, A., Benda, A., Falk, P., Steffen, S., Wallace, Z., Høeg, T.B.,
-2021. COVID-19 cases and transmission in 17 k--12 schools---wood county,
-wisconsin, august 31--november 29, 2020. Morbidity and Mortality Weekly
-Report 70, 136--140.
+Falk, A., Benda, A., Falk, P., Steffen, S., Wallace, Z., Høeg, T.B., 2021.
+COVID-19 cases and transmission in 17 k--12 schools---wood county, wisconsin, august 31--november 29, 2020.
+Morbidity and Mortality Weekly Report 70, 136--140.
 doi:\href{https://doi.org/10.15585/mmwr.mm7004e3}{10.15585/mmwr.mm7004e3}
@@
 \leavevmode\hypertarget{ref-fukumoto2021no}{}%
-Fukumoto, K., McClean, C.T., Nakagawa, K., 2021. No causal effect of
-school closures in japan on the spread of COVID-19 in spring 2020.
+Fukumoto, K., McClean, C.T., Nakagawa, K., 2021.
+No causal effect of school closures in japan on the spread of COVID-19 in spring 2020.
 Nature medicine 27, 2111--2119.
@@
 \leavevmode\hypertarget{ref-grasleguen2021reopening}{}%
-Gras-Le Guen, C., Cohen, R., Rozenberg, J., Launay, E., Levy-Bruhl, D.,
-Delacourt, C., 2021. Reopening schools in the context of increasing
-COVID-19 community transmission: The french experience. Archives de
-Pédiatrie 28, 178--185.
+Gras-Le Guen, C., Cohen, R., Rozenberg, J., Launay, E., Levy-Bruhl, D., Delacourt, C., 2021.
+Reopening schools in the context of increasing COVID-19 community transmission: The french experience.
+Archives de Pédiatrie 28, 178--185.
 doi:\href{https://doi.org/10.1016/j.arcped.2021.02.001}{10.1016/j.arcped.2021.02.001}
@@
 \leavevmode\hypertarget{ref-heavey2020evidence}{}%
-Heavey, L., Casey, G., Kelly, C., Kelly, D., McDarby, G., 2020. No
-evidence of secondary transmission of COVID-19 from children attending
-school in ireland, 2020. Eurosurveillance 25.
+Heavey, L., Casey, G., Kelly, C., Kelly, D., McDarby, G., 2020.
+No evidence of secondary transmission of COVID-19 from children attending school in ireland, 2020.
+Eurosurveillance 25.
 doi:\href{https://doi.org/10.2807/1560-7917.ES.2020.25.21.2000903}{10.2807/1560-7917.ES.2020.25.21.2000903}
@@
 \leavevmode\hypertarget{ref-vanderhoek2020role}{}%
-Hoek, W. van der, Backer, J.A., Bodewes, R., Friesema, I., Meijer, A.,
-Pijnacker, R., Reukers, D.F.M., Reusken, C., Roof, I., Rots, N., Te
-Wierik, M.J.M., Gageldonk-Lafeber, A. van, Waegemaekers, C., Hof, S. van
-den, 2020. The role of children in the transmission of SARS-CoV-2.
+Hoek, W.
+van der, Backer, J.A., Bodewes, R., Friesema, I., Meijer, A., Pijnacker, R., Reukers, D.F.M., Reusken, C., Roof, I., Rots, N., Te Wierik, M.J.M., Gageldonk-Lafeber, A.
+van, Waegemaekers, C., Hof, S.
+van den, 2020.
+The role of children in the transmission of SARS-CoV-2.
 Nederlands tijdschrift voor geneeskunde 164.
@@
 \leavevmode\hypertarget{ref-iwata2020school}{}%
-Iwata, K., Doi, A., Miyakoshi, C., 2020. Was school closure effective in
-mitigating coronavirus disease 2019 (COVID-19)? Time series analysis
-using bayesian inference. International Journal of Infectious Diseases
-99, 57--61.
+Iwata, K., Doi, A., Miyakoshi, C., 2020.
+Was school closure effective in mitigating coronavirus disease 2019 (COVID-19)?
+Time series analysis using bayesian inference.
+International Journal of Infectious Diseases 99, 57--61.
 doi:\href{https://doi.org/10.1016/j.ijid.2020.07.052}{10.1016/j.ijid.2020.07.052}
@@
 \leavevmode\hypertarget{ref-kim2021role}{}%
-Kim, J., Choe, Y.J., Lee, J., Park, Y.J., Park, O., Han, M.S., Kim,
-J.-H., Choi, E.H., 2021. Role of children in household transmission of
-COVID-19. Archives of Disease in Childhood 106, 709--711.
+Kim, J., Choe, Y.J., Lee, J., Park, Y.J., Park, O., Han, M.S., Kim, J.-H., Choi, E.H., 2021.
+Role of children in household transmission of COVID-19.
+Archives of Disease in Childhood 106, 709--711.
 doi:\href{https://doi.org/10.1136/archdischild-2020-319910}{10.1136/archdischild-2020-319910}
@@
 \leavevmode\hypertarget{ref-li2020role}{}%
-Li, X., Xu, W., Dozier, M., He, Y., Kirolos, A., Theodoratou, E.,
-others, 2020. The role of children in transmission of SARS-CoV-2: A
-rapid review. Journal of global health 10.
+Li, X., Xu, W., Dozier, M., He, Y., Kirolos, A., Theodoratou, E., others, 2020.
+The role of children in transmission of SARS-CoV-2: A rapid review.
+Journal of global health 10.
 doi:\href{https://doi.org/10.7189/jogh.10.011101}{10.7189/jogh.10.011101}
@@
 \leavevmode\hypertarget{ref-lopez2020transmission}{}%
-Lopez, A.S., Hill, M., Antezano, J., Vilven, D., Rutner, T., Bogdanow,
-L., Claflin, C., Kracalik, I.T., Fields, V.L., Dunn, A., others, 2020.
-Transmission dynamics of COVID-19 outbreaks associated with child care
-facilities---salt lake city, utah, april--july 2020. Morbidity and
-Mortality Weekly Report 69, 1319.
+Lopez, A.S., Hill, M., Antezano, J., Vilven, D., Rutner, T., Bogdanow, L., Claflin, C., Kracalik, I.T., Fields, V.L., Dunn, A., others, 2020.
+Transmission dynamics of COVID-19 outbreaks associated with child care facilities---salt lake city, utah, april--july 2020.
+Morbidity and Mortality Weekly Report 69, 1319.
 doi:\href{https://doi.org/10.15585/mmwr.mm6937e3}{10.15585/mmwr.mm6937e3}
@@
 \leavevmode\hypertarget{ref-ludvigsson2020children}{}%
-Ludvigsson, J.F., 2020. Children are unlikely to be the main drivers of
-the COVID-19 pandemic -- a systematic review. Acta Paediatrica 109,
-1525--1530. doi:\url{https://doi.org/10.1111/apa.15371}
+Ludvigsson, J.F., 2020. Children are unlikely to be the main drivers of the COVID-19 pandemic -- a systematic review. Acta Paediatrica 109, 1525--1530. doi:\url{https://doi.org/10.1111/apa.15371}
 
 \leavevmode\hypertarget{ref-meuris2021transmission}{}%
-Meuris, C., Kremer, C., Geerinck, A., Locquet, M., Bruyère, O., Defêche,
-J., Meex, C., Hayette, M.-P., Duchene, L., Dellot, P., Azarzar, S.,
-Maréchal, N., Sauvage, A.-S., Frippiat, F., Giot, J.-B., Léonard, P.,
-Fombellida, K., Moutschen, M., Durkin, K., Artesi, M., Bours, V., Faes,
-C., Hens, N., Darcis, G., 2021. Transmission of SARS-CoV-2 after
-COVID-19 screening and mitigation measures for primary school children
-attending school in liège, belgium. JAMA Network Open 4,
-e2128757--e2128757.
+Meuris, C., Kremer, C., Geerinck, A., Locquet, M., Bruyère, O., Defêche, J., Meex, C., Hayette, M.-P., Duchene, L., Dellot, P., Azarzar, S., Maréchal, N., Sauvage, A.-S., Frippiat, F., Giot, J.-B., Léonard, P., Fombellida, K., Moutschen, M., Durkin, K., Artesi, M., Bours, V., Faes, C., Hens, N., Darcis, G., 2021.
+Transmission of SARS-CoV-2 after COVID-19 screening and mitigation measures for primary school children attending school in liège, belgium.
+JAMA Network Open 4, e2128757--e2128757.
 doi:\href{https://doi.org/10.1001/jamanetworkopen.2021.28757}{10.1001/jamanetworkopen.2021.28757}
 
+\leavevmode\hypertarget{ref-papana2016detecting}{}%
+Papana, A., Kyrtsou, C., Kugiumtzis, D., Diks, C., 2016.
+Detecting causality in non-stationary time series using partial symbolic transfer entropy: Evidence in financial data.
+Computational economics 47, 341--365.
+doi:\href{https://doi.org/10.1007/s10614-015-9491-x}{10.1007/s10614-015-9491-x}
+
+\leavevmode\hypertarget{ref-schreiber2000measuring}{}%
+Schreiber, T., 2000. Measuring information transfer. Phys. Rev. Lett. 85, 461--464. doi:\href{https://doi.org/10.1103/PhysRevLett.85.461}{10.1103/PhysRevLett.85.461}
+
+\leavevmode\hypertarget{ref-shannon1948mathematical}{}%
+Shannon, C.E., 1948. A mathematical theory of communication. The Bell system technical journal 27, 379--423.
+
 \leavevmode\hypertarget{ref-siebach2021childhood}{}%
-Siebach, M.K., Piedimonte, G., Ley, S.H., 2021. COVID-19 in childhood:
-Transmission, clinical presentation, complications and risk factors.
+Siebach, M.K., Piedimonte, G., Ley, S.H., 2021.
+COVID-19 in childhood: Transmission, clinical presentation, complications and risk factors.
 Pediatric Pulmonology 56, 1342--1356.
@@
 
-\leavevmode\hypertarget{ref-silini2021fast}{}%
-Silini, R., Masoller, C., 2021. Fast and effective pseudo transfer
-entropy for bivariate data-driven causal inference. Scientific Reports
-11, 1--13.
-doi:\href{https://doi.org/10.1016/j.softx.2019.100265}{10.1016/j.softx.2019.100265}
-
 \leavevmode\hypertarget{ref-sorianoarandes2021household}{}%
-Soriano-Arandes, A., Gatell, A., Serrano, P., Biosca, M., Campillo, F.,
-Capdevila, R., Fàbrega, A., Lobato, Z., López, N., Moreno, A.M., Poblet,
-M., Riera-Bosch, M.T., Rius, N., Ruiz, M., Sánchez, A., Valldepérez, C.,
-Vilà, M., Pineda, V., Lazcano, U., Díaz, Y., Reyes-Urueña, J.,
-Soler-Palacín, P., Catalonia Research Group, C.P.D. in, 2021. Household
-severe acute respiratory syndrome coronavirus 2 transmission and
-children: A network prospective study. Clinical Infectious Diseases 73,
-e1261--e1269.
+Soriano-Arandes, A., Gatell, A., Serrano, P., Biosca, M., Campillo, F., Capdevila, R., Fàbrega, A., Lobato, Z., López, N., Moreno, A.M., Poblet, M., Riera-Bosch, M.T., Rius, N., Ruiz, M., Sánchez, A., Valldepérez, C., Vilà, M., Pineda, V., Lazcano, U., Díaz, Y., Reyes-Urueña, J., Soler-Palacín, P., Catalonia Research Group, C.P.D.
+in, 2021.
+Household severe acute respiratory syndrome coronavirus 2 transmission and children: A network prospective study.
+Clinical Infectious Diseases 73, e1261--e1269.
 doi:\href{https://doi.org/10.1093/cid/ciab228}{10.1093/cid/ciab228}
@@
 \leavevmode\hypertarget{ref-stage2021shut}{}%
-Stage, H.B., Shingleton, J., Ghosh, S., Scarabel, F., Pellis, L.,
-Finnie, T., 2021. Shut and re-open: The role of schools in the spread of
-COVID-19 in europe. Philosophical Transactions of the Royal Society B
-376, 20200277.
+Stage, H.B., Shingleton, J., Ghosh, S., Scarabel, F., Pellis, L., Finnie, T., 2021.
+Shut and re-open: The role of schools in the spread of COVID-19 in europe.
+Philosophical Transactions of the Royal Society B 376, 20200277.
 doi:\href{https://doi.org/10.1098/rstb.2020.0277}{10.1098/rstb.2020.0277}
@@
 \leavevmode\hypertarget{ref-yoshiyuki2020effects}{}%
-Sugishita, J.A.S., Yoshiyuki AND Kurita, 2020. Effects of voluntary
-event cancellation and school closure as countermeasures against
-COVID-19 outbreak in japan. PLOS ONE 15, 1--10.
+Sugishita, J.A.S., Yoshiyuki AND Kurita, 2020.
+Effects of voluntary event cancellation and school closure as countermeasures against COVID-19 outbreak in japan.
+PLOS ONE 15, 1--10.
 doi:\href{https://doi.org/10.1371/journal.pone.0239455}{10.1371/journal.pone.0239455}
@@
 \leavevmode\hypertarget{ref-walger2020children}{}%
-Walger, P., Heininger, U., Knuf, M., Exner, M., Popp, W., Fischbach, T.,
-Trapp, S., Hübner, J., Herr, C., Simon, A., others, 2020. Children and
-adolescents in the CoVid-19 pandemic: Schools and daycare centers are to
-be opened again without restrictions. The protection of teachers,
-educators, carers and parents and the general hygiene rules do not
-conflict with this. GMS hygiene and infection control 15.
+Walger, P., Heininger, U., Knuf, M., Exner, M., Popp, W., Fischbach, T., Trapp, S., Hübner, J., Herr, C., Simon, A., others, 2020.
+Children and adolescents in the CoVid-19 pandemic: Schools and daycare centers are to be opened again without restrictions.
+The protection of teachers, educators, carers and parents and the general hygiene rules do not conflict with this.
+GMS hygiene and infection control 15.
 doi:\href{https://doi.org/10.3205/dgkh000346}{10.3205/dgkh000346}
 
+\leavevmode\hypertarget{ref-wollstadt2014efficient}{}%
+Wollstadt, M.A.V., Patricia AND Martínez-Zarzuela, 2014.
+Efficient transfer entropy analysis of non-stationary neural time series.
+PLOS ONE 9, 1--21.
+doi:\href{https://doi.org/10.1371/journal.pone.0102833}{10.1371/journal.pone.0102833}
+
 \leavevmode\hypertarget{ref-wood2011fast}{}%
-Wood, S.N., 2011. Fast stable restricted maximum likelihood and marginal
-likelihood estimation of semiparametric generalized linear models.
-Journal of the Royal Statistical Society (B) 73, 3--36.
+Wood, S.N., 2011. Fast stable restricted maximum likelihood and marginal likelihood estimation of semiparametric generalized linear models. Journal of the Royal Statistical Society (B) 73, 3--36.
 
 \leavevmode\hypertarget{ref-wood2017generalized}{}%
-Wood, S.N., 2017. Generalized additive models: An introduction with r.
-CRC press.
+Wood, S.N., 2017. Generalized additive models: An introduction with r. CRC press.
 
 \leavevmode\hypertarget{ref-zhendong2020clinical}{}%
-Zhen-Dong, Y., Gao-Jun, Z., Run-Ming, J., Zhi-Sheng, L., Zong-Qi, D.,
-Xiong, X., Guo-Wei, S., 2020. Clinical and transmission dynamics
-characteristics of 406 children with coronavirus disease 2019 in china:
-A review. Journal of Infection 81, e11--e15.
+Zhen-Dong, Y., Gao-Jun, Z., Run-Ming, J., Zhi-Sheng, L., Zong-Qi, D., Xiong, X., Guo-Wei, S., 2020.
+Clinical and transmission dynamics characteristics of 406 children with coronavirus disease 2019 in china: A review.
+Journal of Infection 81, e11--e15.
 doi:\href{https://doi.org/10.1016/j.jinf.2020.04.030}{10.1016/j.jinf.2020.04.030}

### 2022-01-14 analysis/exploration.R

@@ -80,7 +80,7 @@ df_long <- df |>
   mutate(
     covid_cases_7davg = rollmean(covid_cases, k = 7, fill = NA),
     covid_changes = covid_cases - lag(covid_cases),
-    covid_changes_7davg = rollmean(covid_cases, k = 7, fill = NA)
+    covid_changes_7davg = rollmean(covid_changes, k = 7, fill = NA)
   ) |> 
   ungroup()
 
@@ -222,15 +222,16 @@ df_long |>
   theme(legend.position = "none")
 
 df_long |> 
-  ggplot(aes(date, covid_changes_7davg, color = age_group)) +
-  geom_line(na.rm = TRUE) +
+  ggplot() +
+  geom_col(aes(date, covid_changes), na.rm = TRUE) +
+  geom_line(aes(date, covid_changes_7davg, color = age_group), na.rm = TRUE) +
   scale_x_date("Date") +
-  scale_y_continuous("Rolling Average of Day-by-day Changes in Covid Cases") +
+  scale_y_continuous("Day-by-day Changes in Covid Cases") +
   scale_color_manual(labels = age_name, values = age_color) +
   facet_grid(age_group ~ ., scales = "free") +
   labs(
-    title = "Rolling Average of Day-by-day Changes in COVID-19 Cases by Age Groups in Ireland",
-    subtitle = "7 Day Rolling Average of Data Lag 1",
+    title = "Day-by-day Changes in COVID-19 Cases by Age Groups in Ireland",
+    subtitle = "Raw Data (Bars) and 7 Day Rolling Average of Data Lag 1 (Lines)",
     caption = "Source: Irish Department of Health (last update 2022-01-01)",
     x = "Date", 
     y = "Covid Cases",

### 2022-01-17 manuscript/manuscript.Rmd
gam + te

--- 2022-01-13 manuscript/manuscript.Rmd
@@
       output_dir = here::here("output"),
-      output_file = glue::glue("{xfun::sans_ext(input)}_{Sys.Date()}.pdf"),
+      output_file = glue::glue("{xfun::sans_ext(input)}.pdf"),
       envir = globalenv()
@@
 ---
+
+(ref:overall) Evolution of confirmed Covid-19 cases in Ireland alongside with school closures.
+(ref:descriptive) Evolution of confirmed Covid-19 cases by age groups in Ireland: raw data (bars) and 7 day rolling average (lines).
+(ref:gam) Standardized effect of the smooth term in Generalized Additive Model by age group.
+(ref:te) Matrix of Transfer Entropy coefficient and their corresponding p-value according every age group combination.
+Age groups on the x-axis are influencing the age groups on the y-axis ($X \rightarrow Y$).
+* $p < .05$, ** $p < .01$, *** $p < .001$.
+
+```{r setup, include = FALSE}
+# libraries --------------------------------------------------------------------
+library(here)
+library(janitor)
+library(knitr)
+library(mgcv)
+library(patchwork)
+library(RTransferEntropy)
+library(tidyverse)
+library(zoo)
+
+# options ----------------------------------------------------------------------
+set.seed(123) # Seed for random number generation
+options(scipen = 999) # disable sci number format
+opts_chunk$set(
+  cache.extra = rand_seed, 
+  message = FALSE, 
+  warning = FALSE, 
+  error = FALSE,
+  echo = FALSE, 
+  fig.retina = 3
+  )
+
+# variables --------------------------------------------------------------------
+age_order <- c("aged1to4", "aged5to14", "aged15to24", "aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65to74", "aged75to84", "aged85up")
+age_name  <- c("1 to 4",   "5 to 14",   "15 to 24",   "25 to 34",   "35 to 44",   "45 to 54",   "55 to 64",   "65 to 74",   "75 to 84",   "85 and up")
+age_color <- c("#999999",  "#E69F00",   "#56B4E9",    "#009E73",    "#F0E442",    "#0072B2",    "#D55E00",    "#CC79A7",    "#000000",    "#CAB2D6")
+
+# data -------------------------------------------------------------------------
+df <- 
+  read_csv(
+    here("data/age_groups.csv"), 
+    col_types = list(
+      .default = col_double(), 
+      Date = col_date(format = "%d/%m/%Y")
+    )
+  ) |> 
+  clean_names() |> 
+  rename(total_cases = total) |> 
+  arrange(date) |> 
+  mutate(
+    timeline = date - lag(date),  # check if gap between dates
+    total_changes = total_cases - lag(total_cases)
+  )
+
+df_long <- df |> 
+  select(c(date, starts_with("aged"))) |> 
+  pivot_longer(-date, names_to = "age_group", values_to = "covid_cases") |> 
+  mutate(age_group = factor(age_group, levels = age_order)) |> 
+  group_by(age_group) |> 
+  mutate(
+    covid_cases_7davg = rollmean(covid_cases, k = 7, fill = NA),
+    covid_changes = covid_cases - lag(covid_cases),
+    covid_changes_7davg = rollmean(covid_changes, k = 7, fill = NA)
+  ) |> 
+  ungroup()
+
+df_wide_changes <- df_long |> 
+  select(date, age_group, covid_changes) |> 
+  pivot_wider(names_from = age_group, values_from = covid_changes) |> 
+  drop_na()
+
+df_school <- df |> 
+  select(date, schools_closed) |> 
+  mutate(schools_closure_day = schools_closed - lag(schools_closed)) |> 
+  filter(schools_closure_day == 1) |> 
+  rowid_to_column("schools_closure_wave") |> 
+  full_join(df, by = c("date", "schools_closed")) |> 
+  arrange(date) |> 
+  fill(schools_closure_wave, .direction = "down") |> 
+  mutate(schools_closure_wave = replace_na(schools_closure_wave, 0)) |> 
+  group_by(schools_closure_wave) |> 
+  mutate(schools_time_since_closure = row_number()) |> 
+  ungroup()
+
+# df_school |> 
+#   filter(schools_closed == 1) |> 
+#   count(schools_closure_wave) |> 
+#   rename(schools_closure_wave_n = n)
+
+df_school_long <- df_school |> 
+  filter(schools_closed == 1) |> 
+  count(schools_closure_wave) |> 
+  rename(schools_closure_wave_n = n) |> 
+  right_join(df_school, by = "schools_closure_wave") |> 
+  filter(schools_closure_wave_n > 20 & schools_time_since_closure < 40) |> # keep only closure waves longer than 20 days but display only the first 39 days
+  pivot_longer(c("aged1to4", "aged5to14", "aged15to24"), names_to = "age_group", values_to = "covid_cases") |> 
+  mutate(
+    schools_closure_wave = factor(schools_closure_wave),
+    age_group = factor(age_group, levels = c("aged1to4", "aged5to14", "aged15to24"))
+  ) |> 
+  group_by(age_group) |> 
+  mutate(
+    covid_changes = covid_cases - lag(covid_cases),
+    covid_cases_7davg = rollmean(covid_cases, k = 7, fill = NA),
+    covid_changes_7davg = rollmean(covid_changes, k = 7, fill = NA)
+  ) |> 
+  ungroup()
+
+# df_school_long |> 
+#   filter(schools_closure_day == 1 & age_group == "aged1to4") |> 
+#   select(date, schools_closure_wave, schools_closure_wave_n)
+
+# analyses ---------------------------------------------------------------------
+gam_model <- 
+  gam(
+    covid_cases_7davg ~ age_group +
+      s(schools_time_since_closure, by = age_group) +
+      s(schools_time_since_closure, schools_closure_wave, bs = "fs"),
+    data = df_school_long,
+    method = "REML",
+    family = nb()
+  )
+
+# analyses ---------------------------------------------------------------------
+source(here("analysis/gam_analysis.R"))
+source(here("analysis/te_analysis.R"))
+
+```
 
@@
 The data were collected from the official daily publication of Covid-19 cases by the Irish Department of Health.
-They include the amount of daily cases for 10 age groups from March 4th, 2020 to January 1st, 2022.
+They include the amount of daily cases for 10 age groups from `r format(min(df$date), "%B %d, %Y")` to `r format(max(df$date), "%B %d, %Y")`.
 For each day during this period, school closure either due to a government decision or a public holiday is determined.
-
-## Data Pre-processing
-
-A 7-day rolling average of the day-by-by change in Covid-19 cases reported (i.e., lag 1) is used to evaluate the efficiency of school closures.
-This rolling average approximate the real day-by-by change in Covid-19 cases reported which can be artificially increased or decreased due to days under- or over-reporting cases such as weekend or bank holidays for example.
 
@@
 
+### Generalised Additive Model
+
 Four periods of school closure longer that 14 days have been identified.
-These four periods are used to fit a Generalised Additive Model (or GAM) using the R package mgcv [@wood2017generalized] in order to test the hypothesis of a significant change in the evolution of cases among age groups from 1 to 4, from 5 to 14, and from 15 to 24.
+These four periods are used to fit a Generalised Additive Model (or GAM) using the R package *mgcv* [@wood2017generalized] in order to test the hypothesis of a significant change in the evolution of cases among age groups from 1 to 4, from 5 to 14, and from 15 to 24 (Eq 1).
 
-By estimating the degree of smoothness of a Bayesian spline smoothing using restricted maximum likelihood estimation [@wood2011fast], GAMs allow the identification of dynamic patterns underlying the day-by-day changes in Covid-19 cases reported while taking into account the random effect of different age groups as follows:
+By estimating the degree of smoothness of a Bayesian spline smoothing using restricted maximum likelihood estimation [@wood2011fast], GAM identifies dynamic patterns underlying the evolution of Covid-19 cases reported while taking into account the random effect of different age groups as follows:
 
@@
 
-\noindent where $i$ is an age group among the 10 age groups investigated and $s$ is the date corresponding to the day-by-day change in Covid-19 cases.
-$Y_{is}$ represents the day-by-day change in Covid-19 cases assuming a negative binomial distribution for the fitting [@loader2006local].
+\noindent where $i$ is an age group among the 10 age groups investigated and $s$ is the date corresponding to the confirmed Covid-19 cases.
+$Y_{is}$ represents the confirmed Covid-19 cases assuming a negative binomial distribution for the fitting [@loader2006local].
 The response variable $Y_{is}$ includes a specific intercept for each age group ($\alpha_{i}$).
-A smooth effect over time $f(X_{s})$ is applied to model (Eq 1) to predict the nonlinear evolution of day-by-day change in Covid-19 cases.
+A smooth effect over time $f(X_{s})$ is applied to model (Eq 2) to predict the nonlinear evolution of Covid-19 cases.
 This smooth effect $f(X_{s})$ is built up in basic components, called the basis functions $b_{j}(X_{s})$, such that:
@@
 
-Contrary to Granger Causality, Transfer Entropy is robust to non-stationary time series [@silini2021fast].
-The Transfer Entropy analysis of younger age groups daily changes in Covid-19 cases on older age groups is done using the R package RTransferEntropy [@behrendt2019rtransferentropy].
+### Transfer Entropy
+
+Transfer Entropy (TE) can be used to infer the temporal relationship between two time series $I$ and $J$.
+This measure indicates whether $J$ can be used to to reduce the uncertainty on the future of $I$ and, consequently, that $J$ causes $I$ [@schreiber2000measuring].
+As such, Granger causality is a special case of TE applied to times series that are jointly Gaussian distributed [@barnett2009granger].
+Therefore, TE is a more robust analysis of times series especially when applied to the impact of age cohort in pandemic transmission [@kissler2020symbolic].
+
+The influence of the evolution in Covid-19 cases across all age groups is evaluated using Shannon's transfer entropy is given by:
+
+\begin{align}
+  T_{J \rightarrow I}(k,l) = \sum_{i,j} p\left(i_{t+1}, i_t^{(k)}, j_t^{(l)}\right) \cdot log \left(\frac{p\left(i_{t+1}| i_t^{(k)}, j_t^{(l)}\right)}{p\left(i_{t+1}|i_t^{(k)}\right)}\right),
+\end{align}
+
+\noindent where $T_{J\rightarrow I}$ consequently measures the influence of the change dynamic from an age group $J$ to another age group $I$ (Eq 3).
+
+In order to observe how increase in daily numbers reported from a specific age group affects the other age groups, the day-by-day difference in Covid-19 confirmed cases (i.e., differences with a lag 1) is calculated.
+The use of day-by-day difference in Covid-19 confirmed cases satisfies the requirement of stationary for the calculation of Shannon's Transfer Entropy [@shannon1948mathematical].
+The Transfer Entropy analysis of younger age groups daily changes in Covid-19 cases on older age groups is done using the R package *RTransferEntropy* [@behrendt2019rtransferentropy].
 
@@
 ===================
+
+Overall, the trend of confirmed Covid-19 cases in Ireland is similar to those of other European countries (Figure \@ref(fig:overall)).
+Indeed it is possible to observe a significant increase corresponding to each of the 4 waves of transmission (March-April 2020, December 2020-January 2021, September-October 2021 and November-December 2021).
+
+```{r overall, fig.width=4.5, fig.height=3, fig.cap="(ref:overall)", fig.pos = "!h"}
+overall_plot <- df |> 
+  ggplot(aes(date, total_cases)) +
+  geom_tile(aes(fill = factor(schools_closed), height = Inf), na.rm = TRUE, alpha = .4) +
+  geom_line(na.rm = TRUE) +
+  scale_x_date(date_breaks = "3 months", date_labels = "%b %y") +
+  scale_fill_manual(labels = c("Open", "Closed", "Na"), values = c("light blue", "red", "orange")) +
+  labs(
+    x = "Date", 
+    y = "Covid-19 Cases in Ireland",
+    fill = "Schools"
+  ) +
+  theme_bw() +
+  theme(
+    legend.position = "bottom",
+    text = element_text(size = 10, family = "serif"),
+    axis.text = element_text(size = 10),
+    panel.background = element_rect(colour = "black", size = 0.1)
+  )
+
+overall_plot
+```
+
+The evolution of Covid-19 cases reveals some similarities across all age groups.
+However, the influence of each waves on the each groups has also some particularities (Figure \@ref(fig:descriptive)).
+For example, the first wave was more important among the oldest age groups whereas the third wave was more important among the youngest age groups.
+
+```{r descriptive, fig.width=4.5, fig.height=7, fig.cap="(ref:descriptive)"}
+descriptive_plot <- df_long |> 
+  ggplot() +
+  geom_col(aes(date, covid_cases), na.rm = TRUE) +
+  geom_line(aes(date, covid_cases_7davg, color = age_group), na.rm = TRUE) +
+  scale_x_date("Date", date_breaks = "3 months", date_labels = "%b %y") +
+  scale_y_continuous("Confirmed Covid-19 Cases in Ireland") +
+  scale_color_manual(labels = age_name, values = age_color) +
+  facet_grid(
+    age_group ~ ., 
+    scales = "free",
+    labeller = as_labeller(
+      c(
+        aged1to4 = "1 to 4",
+        aged5to14 = "5 to 14",
+        aged15to24 = "15 to 24", 
+        aged25to34 = "25 to 34", 
+        aged35to44 = "35 to 44", 
+        aged45to54 = "45 to 54", 
+        aged55to64 = "55 to 64", 
+        aged65to74 = "65 to 74", 
+        aged75to84 = "75 to 84", 
+        aged85up = "85 and up"
+        )
+      )
+  ) +
+  theme_bw() +
+  theme(
+    legend.position = "none",
+    strip.background = element_blank(),
+    strip.placement = "outside",
+    text = element_text(size = 10, family = "serif"),
+    axis.text = element_text(size = 10),
+    panel.background = element_rect(colour = "black", size = 0.1)
+  )
+
+descriptive_plot
+```
+
+In order to evaluate the shape of the trend in the numbers of Covid-19 cases reported after the three school closure longer than 14 consecutive days, a GAM was fitted taking into account numbers for 1 to 4, 5 to 14, and 15 to 24 year old.
+The obtained results satisfy the requirements to fit this model which explains `r scales::percent(gam_summary$dev.expl, accuracy = 0.1)` of the deviance in Covid-19 cases.
+These results show no significant change in the evolution of cases for 1 to 4 year old (`r gam_results["s(date):1to4"]`) as well as for 5 to 14 year old (`r gam_results["s(date):5to14"]`).
+However, the smooth term of the GAM is significant for 15 to 24 year old (`r gam_results["s(date):15to24"]`), which indicates that the evolution of cases is not linear and, therefore, the school closure had a significant impact on the evolution of cases.
+
+```{r gam, fig.width=4.5, fig.height=3, fig.cap="(ref:gam)"}
+gam_plot_1 + gam_plot_2 + gam_plot_3
+```
+
+Before the Transfer Entropy analysis, an Augmented Dickey-Fuller has been applied on each age group in order to insure that the daily changes in Covid-19 cases are stationary.
+
+```{r}
+# adf_res <- df_long |> 
+#   group_by(age_group) |> 
+#   summarise(
+#     aTSA::adf.test(covid_changes, output = FALSE) |> 
+#       map_df(as_tibble)
+#   )
+```
+
+```{r te, fig.width=4.5, fig.height=4.5, fig.cap="(ref:te)"}
+te_plot
+```
 
@@
 
+This indicate mix results, while the school closure was ineffective in child care settings and primary school, the closure of secondary schools and colleges had an impact on Covid-19 transmission.
+
 Conclusion

### 2022-01-17 output/manuscript.tex
gam + te

--- 2022-01-14 output/manuscript_2022-01-14.tex
@@
 
-\hypertarget{data-pre-processing}{%
-\subsection{Data Pre-processing}\label{data-pre-processing}}
-
-A 7-day rolling average of the day-by-by change in Covid-19 cases reported (i.e., lag 1) is used to evaluate the efficiency of school closures.
-This rolling average approximate the real day-by-by change in Covid-19 cases reported which can be artificially increased or decreased due to days under- or over-reporting cases such as weekend or bank holidays for example.
-
 \hypertarget{data-analysis}{%
@@
 Four periods of school closure longer that 14 days have been identified.
-These four periods are used to fit a Generalised Additive Model (or GAM) using the R package mgcv (Wood, 2017) in order to test the hypothesis of a significant change in the evolution of cases among age groups from 1 to 4, from 5 to 14, and from 15 to 24.
+These four periods are used to fit a Generalised Additive Model (or GAM) using the R package \emph{mgcv} (Wood, 2017) in order to test the hypothesis of a significant change in the evolution of cases among age groups from 1 to 4, from 5 to 14, and from 15 to 24 (Eq 1).
 
-By estimating the degree of smoothness of a Bayesian spline smoothing using restricted maximum likelihood estimation (Wood, 2011), GAM identifies dynamic patterns underlying the day-by-day changes in Covid-19 cases reported while taking into account the random effect of different age groups as follows:
+By estimating the degree of smoothness of a Bayesian spline smoothing using restricted maximum likelihood estimation (Wood, 2011), GAM identifies dynamic patterns underlying the evolution of Covid-19 cases reported while taking into account the random effect of different age groups as follows:
 
@@
 
-\noindent where \(i\) is an age group among the 10 age groups investigated and \(s\) is the date corresponding to the day-by-day change in Covid-19 cases.
-\(Y_{is}\) represents the day-by-day change in Covid-19 cases assuming a negative binomial distribution for the fitting (Loader, 2006).
+\noindent where \(i\) is an age group among the 10 age groups investigated and \(s\) is the date corresponding to the confirmed Covid-19 cases.
+\(Y_{is}\) represents the confirmed Covid-19 cases assuming a negative binomial distribution for the fitting (Loader, 2006).
 The response variable \(Y_{is}\) includes a specific intercept for each age group (\(\alpha_{i}\)).
-A smooth effect over time \(f(X_{s})\) is applied to model (Eq 1) to predict the nonlinear evolution of day-by-day change in Covid-19 cases.
+A smooth effect over time \(f(X_{s})\) is applied to model (Eq 2) to predict the nonlinear evolution of Covid-19 cases.
 This smooth effect \(f(X_{s})\) is built up in basic components, called the basis functions \(b_{j}(X_{s})\), such that:
@@
 
-Transfer Entropy can be used to infer the temporal relationship between two time series \(I\) and \(J\).
+Transfer Entropy (TE) can be used to infer the temporal relationship between two time series \(I\) and \(J\).
 This measure indicates whether \(J\) can be used to to reduce the uncertainty on the future of \(I\) and, consequently, that \(J\) causes \(I\) (Schreiber, 2000).
-Contrary to Wiener-Granger Causality, Transfer Entropy is robust to non-stationary time series (Papana et al., 2016; Wollstadt, 2014).
-The Transfer Entropy analysis of younger age groups daily changes in Covid-19 cases on older age groups using Shannon's calculation (Shannon, 1948) is done using the R package RTransferEntropy (Behrendt et al., 2019).
+As such, Granger causality is a special case of TE applied to times series that are jointly Gaussian distributed (Barnett et al., 2009).
+Therefore, TE is a more robust analysis of times series especially when applied to the impact of age cohort in pandemic transmission (Kissler et al., 2020).
 
-The influence of day-by-day changes in Covid-19 cases across all age groups is evaluated using Shannon's transfer entropy is given by:
+The influence of the evolution in Covid-19 cases across all age groups is evaluated using Shannon's transfer entropy is given by:
 
@@
 
-\noindent where \(T_{J\rightarrow I}\) consequently measures the influence of the change dynamic from an eage group \(J\) to another age group \(I\).
+\noindent where \(T_{J\rightarrow I}\) consequently measures the influence of the change dynamic from an age group \(J\) to another age group \(I\) (Eq 3).
+
+In order to observe how increase in daily numbers reported from a specific age group affects the other age groups, the day-by-day difference in Covid-19 confirmed cases (i.e., differences with a lag 1) is calculated.
+The use of day-by-day difference in Covid-19 confirmed cases satisfies the requirement of stationary for the calculation of Shannon's Transfer Entropy (Shannon, 1948).
+The Transfer Entropy analysis of younger age groups daily changes in Covid-19 cases on older age groups is done using the R package \emph{RTransferEntropy} (Behrendt et al., 2019).
 
@@
 
-Overall, the trend of comfirm Covid-19 cases in Ireland is similar to those of other European countries (Figure \ref{fig:overall}).
+Overall, the trend of confirmed Covid-19 cases in Ireland is similar to those of other European countries (Figure \ref{fig:overall}).
 Indeed it is possible to observe a significant increase corresponding to each of the 4 waves of transmission (March-April 2020, December 2020-January 2021, September-October 2021 and November-December 2021).
@@
 \begin{figure}[!h]
-\includegraphics[width=324px]{/Users/damiendcu/OneDrive/Projects/covid_ts_causality/output/manuscript_2022-01-14_files/figure-latex/overall-1} \caption{Evolution of confirmed Covid-19 cases in Ireland alongside with school closures.}\label{fig:overall}
+\includegraphics[width=324px]{/Users/damiendcu/OneDrive/Projects/covid_ts_causality/output/manuscript_files/figure-latex/overall-1} \caption{Evolution of confirmed Covid-19 cases in Ireland alongside with school closures.}\label{fig:overall}
 \end{figure}
 
-The evolution of day-by-day changes in Covid-19 cases reveals some similarities across all age groups.
+The evolution of Covid-19 cases reveals some similarities across all age groups.
 However, the influence of each waves on the each groups has also some particularities (Figure \ref{fig:descriptive}).
@@
 \begin{figure}
-\includegraphics[width=324px]{/Users/damiendcu/OneDrive/Projects/covid_ts_causality/output/manuscript_2022-01-14_files/figure-latex/descriptive-1} \caption{Day-by-day changes in Covid-19 cases by age groups in Ireland: raw data (bars) and 7 aay rolling average (lines).}\label{fig:descriptive}
+\includegraphics[width=324px]{/Users/damiendcu/OneDrive/Projects/covid_ts_causality/output/manuscript_files/figure-latex/descriptive-1} \caption{Evolution of confirmed Covid-19 cases by age groups in Ireland: raw data (bars) and 7 day rolling average (lines).}\label{fig:descriptive}
+\end{figure}
+
+In order to evaluate the shape of the trend in the numbers of Covid-19 cases reported after the three school closure longer than 14 consecutive days, a GAM was fitted taking into account numbers for 1 to 4, 5 to 14, and 15 to 24 year old.
+The obtained results satisfy the requirements to fit this model which explains 99.6\% of the deviance in Covid-19 cases.
+These results show no significant change in the evolution of cases for 1 to 4 year old (\(\chi^2(1) = 2.15\), \(p = 0.143\)) as well as for 5 to 14 year old (\(\chi^2(1) = 1.61\), \(p = 0.204\)).
+However, the smooth term of the GAM is significant for 15 to 24 year old (\(\chi^2(5.02) = 201.27\), \(p < 0.001\)), which indicates that the evolution of cases is not linear and, therefore, the school closure had a significant impact on the evolution of cases.
+
+\begin{figure}
+\includegraphics[width=324px]{/Users/damiendcu/OneDrive/Projects/covid_ts_causality/output/manuscript_files/figure-latex/gam-1} \caption{Standardized effect of the smooth term in Generalized Additive Model by age group.}\label{fig:gam}
+\end{figure}
+
+Before the Transfer Entropy analysis, an Augmented Dickey-Fuller has been applied on each age group in order to insure that the daily changes in Covid-19 cases are stationary.
+
+\begin{figure}
+\includegraphics[width=324px]{/Users/damiendcu/OneDrive/Projects/covid_ts_causality/output/manuscript_files/figure-latex/te-1} \caption{Matrix of Transfer Entropy coefficient and their corresponding p-value according every age group combination.
+Age groups on the x-axis are influencing the age groups on the y-axis (\(X \rightarrow Y\)).
+* \(p < .05\), ** \(p < .01\), *** \(p < .001\).}\label{fig:te}
 \end{figure}
@@
 From the early analyses showing that the virus was instantness in children (Li et al., 2020), the results has changed to more nuanced position which states that the spread of the virus in children is moderate.
+
+This indicate mix results, while the school closure was ineffective in child care settings and primary school, the closure of secondary schools and colleges had an impact on Covid-19 transmission.
 
@@
 \begin{CSLReferences}{1}{0}
+\leavevmode\hypertarget{ref-barnett2009granger}{}%
+Barnett, L., Barrett, A.B., Seth, A.K., 2009.
+Granger causality and transfer entropy are equivalent for gaussian variables.
+Phys.
+Rev.
+Lett.
+103, 238701.
+doi:\href{https://doi.org/10.1103/PhysRevLett.103.238701}{10.1103/PhysRevLett.103.238701}
+
 \leavevmode\hypertarget{ref-behrendt2019rtransferentropy}{}%
@@
 
+\leavevmode\hypertarget{ref-kissler2020symbolic}{}%
+Kissler, S.M., Viboud, C., Grenfell, B.T., Gog, J.R., 2020.
+Symbolic transfer entropy reveals the age structure of pandemic influenza transmission from high-volume influenza-like illness data.
+Journal of the Royal Society Interface 17, 20190628.
+doi:\href{https://doi.org/10.1098/rsif.2019.0628}{10.1098/rsif.2019.0628}
+
 \leavevmode\hypertarget{ref-li2020role}{}%
@@
 doi:\href{https://doi.org/10.1001/jamanetworkopen.2021.28757}{10.1001/jamanetworkopen.2021.28757}
-
-\leavevmode\hypertarget{ref-papana2016detecting}{}%
-Papana, A., Kyrtsou, C., Kugiumtzis, D., Diks, C., 2016.
-Detecting causality in non-stationary time series using partial symbolic transfer entropy: Evidence in financial data.
-Computational economics 47, 341--365.
-doi:\href{https://doi.org/10.1007/s10614-015-9491-x}{10.1007/s10614-015-9491-x}
 
@@
 
-\leavevmode\hypertarget{ref-wollstadt2014efficient}{}%
-Wollstadt, M.A.V., Patricia AND Martínez-Zarzuela, 2014.
-Efficient transfer entropy analysis of non-stationary neural time series.
-PLOS ONE 9, 1--21.
-doi:\href{https://doi.org/10.1371/journal.pone.0102833}{10.1371/journal.pone.0102833}
-
 \leavevmode\hypertarget{ref-wood2011fast}{}%

### 2022-01-17 analysis/exploration.R

@@ -36,7 +36,7 @@ library(here)
 library(janitor)
 library(zoo)
 library(mgcv)
-library(gratia)
+#  library(gratia)
 # variables --------------------------------------------------------------------
 age_order <- c("aged1to4", "aged5to14", "aged15to24", "aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65to74", "aged75to84", "aged85up")
 age_name  <- c("1 to 4",   "5 to 14",   "15 to 24",   "25 to 34",   "35 to 44",   "45 to 54",   "55 to 64",   "65 to 74",   "75 to 84",   "85 and up")
@@ -101,7 +101,12 @@ df_school_long <- df_school |>
     age_group = factor(age_group, levels = c("aged1to4", "aged5to14", "aged15to24"))
   ) |> 
   group_by(age_group) |> 
-  mutate(covid_changes = covid_cases - lag(covid_cases)) |> 
+  arrange(date) |> 
+  mutate(
+    covid_changes = covid_cases - lag(covid_cases),
+    covid_cases_7davg = rollmean(covid_cases, k = 7, fill = NA),
+    covid_changes_7davg = rollmean(covid_changes, k = 7, fill = NA)
+  ) |> 
   ungroup()
 
 df_school_long |> 
@@ -304,8 +309,39 @@ df_school_long |>
   theme_bw() +
   theme(legend.position = "bottom")
 
+df_wide_7davg <- df_long |> 
+  select(date, age_group, covid_cases_7davg) |> 
+  pivot_wider(names_from = age_group, values_from = covid_cases_7davg) |> 
+  drop_na() |> 
+  left_join(df_school |> select(!starts_with("aged")), by = "date")
+
+
 # modelling --------------------------------------------------------------------
 ## Generalised Additive Model ##################################################
+#99%
+gam_model <- 
+  gam(
+    covid_cases_7davg ~ age_group +
+      s(schools_time_since_closure, by = age_group) +
+      s(schools_time_since_closure, schools_closure_wave, bs = "fs"),
+    data = df_school_long,
+    method = "REML",
+    family = nb()
+  ); summary(gam_model)
+
+gam_results <- broom::tidy(gam_model) |> 
+  mutate(
+    term = c("s(date):1to4", "s(date):5to14", "s(date):15to24", "s(date*wave"),
[…]

### 2022-01-17 analysis/gam_analysis.R

@@ -1,0 +1,84 @@
+# model ------------------------------------------------------------------------
+gam_summary <- 
+  gam(
+    covid_cases ~ age_group +
+      s(schools_time_since_closure, by = age_group) +
+      s(schools_time_since_closure, schools_closure_wave, bs = "fs"),
+    data = df_school_long,
+    family = nb(), 
+    method = "REML"
+  ); summary(gam_summary)
+
+# statistics -------------------------------------------------------------------
+gam_summary <- summary(gam_model)
+
+gam_results <- 
+  broom::tidy(gam_model) |> 
+  mutate(
+    term = c("s(date):1to4", "s(date):5to14", "s(date):15to24", "s(date*wave"),
+    p.value = scales::pvalue(p.value, accuracy = 0.001, add_p = FALSE, prefix = c("< ", "= ", "> ")),
+    across(where(is.numeric), round, 2),
+    print = glue::glue("$\\chi^2({edf}) = {statistic}$, $p {p.value}$")
+  ) |> 
+  select(term, print) |> 
+  deframe()
+
+# visualisation ----------------------------------------------------------------
+gam_plot <- plot(gam_model, residuals = TRUE)
+
+gam_plot_indiv <- gam_plot[[1]]
+gam_plot_indiv_df <- as.data.frame(gam_plot_indiv[c("x", "se", "fit")])
+#gam_plot_indiv_point <- as.data.frame(gam_plot_indiv[c("raw", "p.resid")])
+gam_plot_1 <- ggplot(gam_plot_indiv_df, aes(x = x, y = fit)) +
+  # geom_rug(data = gam_plot_indiv_point, mapping = aes(x = raw, y = NULL), sides = "b") +
+  # geom_point(data = gam_plot_indiv_point, mapping = aes(x = raw, y = p.resid)) +
+  geom_ribbon(aes(ymin = fit - se, ymax = fit + se, y = NULL), alpha = 0.3) +
+  geom_line() +
+  labs(
+    x = "Days since School Closure", 
+    y = "Standardized Effect in Age 1 to 4"
+  ) +
+  theme_bw() +
+  theme(
+    text = element_text(size = 10, family = "serif"),
+    axis.text = element_text(size = 10),
+    panel.background = element_rect(colour = "black", size = 0.1)
+  )
+
+gam_plot_indiv <- gam_plot[[2]]
[…]

### 2022-01-17 analysis/te_analysis.R

@@ -1,0 +1,102 @@
+
+# libraries --------------------------------------------------------------------
+library(gtools)
+library(tidyverse)
+library(RTransferEntropy)
+
+# variables --------------------------------------------------------------------
+age_order <- c("aged1to4", "aged5to14", "aged15to24", "aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65to74", "aged75to84", "aged85up")
+
+# solution 1 -------------------------------------------------------------------
+# te_agegroup <- function(var_x, var_y){
+#   
+#   var_x <- select(df_wide_changes, all_of(var_x))
+#   var_y <- select(df_wide_changes, all_of(var_y))
+#   
+#   calc_te(var_x, var_y)
+# }
+# 
+# te_results <-
+#   expand_grid(
+#     X = age_order, 
+#     Y = age_order
+#   ) |> 
+#   as_tibble() |> 
+#   group_by(X, Y) |> 
+#   summarise(te = te_agegroup(X, Y))
+
+# te_matrix <- te_results |> 
+#   mutate(
+#     X = factor(X, levels = age_order),
+#     Y = factor(Y, levels = age_order),
+#   ) |> 
+#   arrange(X, Y) |> 
+#   pivot_wider(names_from = Y, values_from = te) |> 
+#   column_to_rownames("X") |> 
+#   as_tibble()
+# 
+# ggplot(te_results, aes(X, Y)) + 
+#   geom_raster(aes(fill=te)) + 
+#   scale_fill_gradient(low="grey90", high="red") +
+#   theme_bw()
+
+# solution 2 -------------------------------------------------------------------
+library(future) # enable parallel processing for all future transfer_entropy calls
+plan(multiprocess)
+
+te_agegroup <- function(var_x, var_y){
+  
+  var_x <- select(df_wide_changes, all_of(var_x))
+  var_y <- select(df_wide_changes, all_of(var_y))
+  
+  transfer_entropy(var_x, var_y)[["coef"]] |> 
+    as_tibble() |> 
+    mutate(direction = c("X->Y", "Y->X"))
+}
+
+te_results <-
+  combinations(
+    n = length(age_order), 
+    r = 2, 
+    v = age_order, 
+    repeats.allowed = FALSE
+  ) |> 
+  as_tibble() |> 
+  rename(
+    X = V1,
+    Y = V2
+  ) |> 
+  group_by(X, Y) |> 
[…]

### 2022-01-18 manuscript/manuscript.Rmd
sent to edgar

@@
     base_format: rticles::elsevier_article
+    keep_tex: false
+  bookdown::word_document2:
+    default
 knit: (function(input, ...) {
@@
       output_dir = here::here("output"),
-      output_file = glue::glue("{xfun::sans_ext(input)}.pdf"),
+      output_file = c(glue::glue("{xfun::sans_ext(input)}.pdf"), glue::glue("{xfun::sans_ext(input)}.docx")), 
+      output_format = "all",
       envir = globalenv()
@@
   cache.extra = rand_seed, 
+  cache = TRUE,
   message = FALSE, 
@@
 \begin{align}
-  T_{J \rightarrow I}(k,l) = \sum_{i,j} p\left(i_{t+1}, i_t^{(k)}, j_t^{(l)}\right) \cdot log \left(\frac{p\left(i_{t+1}| i_t^{(k)}, j_t^{(l)}\right)}{p\left(i_{t+1}|i_t^{(k)}\right)}\right),
+  T_{X \rightarrow Y}(k,l) = \sum_{x,y} p\left(x_{t+1}, x_t^{(k)}, y_t^{(l)}\right) \cdot log \left(\frac{p\left(x_{t+1}| x_t^{(k)}, y_t^{(l)}\right)}{p\left(x_{t+1}|x_t^{(k)}\right)}\right),
 \end{align}
 
-\noindent where $T_{J\rightarrow I}$ consequently measures the influence of the change dynamic from an age group $J$ to another age group $I$ (Eq 3).
+\noindent where $T_{X\rightarrow Y}$ consequently measures the influence of the change dynamic from an age group $X$ to another age group $Y$ (Eq 3).
 
@@
 These results show no significant change in the evolution of cases for 1 to 4 year old (`r gam_results["s(date):1to4"]`) as well as for 5 to 14 year old (`r gam_results["s(date):5to14"]`).
-However, the smooth term of the GAM is significant for 15 to 24 year old (`r gam_results["s(date):15to24"]`), which indicates that the evolution of cases is not linear and, therefore, the school closure had a significant impact on the evolution of cases.
+However, the smooth term of the GAM is significant for 15 to 24 year old (`r gam_results["s(date):15to24"]`), which indicates that the evolution of cases is not linear and, therefore, the school closure had a significant impact on the evolution of cases (Figure \@ref(fig:gam)).
 
@@
 
-Discussion
+The results of the Transfer Entropy calculations between age groups are reported Figure \@ref(fig:te).
+The first observation is the absence of symmetry between influencing age groups (i.e., $X$) and influenced age groups (i.e., $Y$).
+Indeed, the change in Covid-19 cases in some age groups are influenced by other age groups but they are not reciprocally influencing these age groups.
+This is the case of 15 to 24 years old who appears to be contaminated by all age groups until 84 but there is no significant support that they are contaminating groups older than 34 which should include their parents and relatives.
+
+Discussion & Conclusion
 ===================
@@
 This indicate mix results, while the school closure was ineffective in child care settings and primary school, the closure of secondary schools and colleges had an impact on Covid-19 transmission.
-
-Conclusion
-===================
+This impact is probably responsible for the absence of reciprocity in the transmission of 15 to 24 age group.
+By drastically reducing their social contacts in secondary schools and colleges, this cohort has more been the target of Covid-19 contaminations rather than the vector of Covid-19.
 

### 2022-01-18 analysis/te_analysis.R

@@ -85,15 +85,17 @@ te_plot <- bind_rows(te_results_XY,te_results_YX) |>
   ) |>
   arrange(X, Y) |>
   ggplot(aes(X, Y)) + 
-  geom_raster(aes(fill = te)) + 
+  geom_tile(aes(fill = te)) + 
   geom_text(aes(label = p_stars)) +
   scale_x_discrete("Influencing Age Groups", labels = age_name) +
   scale_y_discrete("Influenced Age Groups", labels = age_name) +
   scale_fill_gradient("Transfer Entropy", low = "grey90", high = "red") +
   theme_bw() +
   theme(
+    legend.position = "bottom",
     text = element_text(size = 10, family = "serif"),
     axis.text = element_text(size = 10),
+    axis.text.x = element_text(angle = 45, hjust = 1),
     panel.background = element_rect(colour = "black", size = 0.1)
   )
 

### 2022-01-18 manuscript/manuscript.Rmd
upgrade

@@
 # libraries --------------------------------------------------------------------
+library(gratia)
 library(here)
@@
 library(mgcv)
-library(patchwork)
 library(RTransferEntropy)
@@
 
-# df_school |> 
-#   filter(schools_closed == 1) |> 
-#   count(schools_closure_wave) |> 
-#   rename(schools_closure_wave_n = n)
+df_school |>
+  filter(schools_closed == 1) |>
+  count(schools_closure_wave) |>
+  rename(schools_closure_wave_n = n)
 
@@
 
-# df_school_long |> 
-#   filter(schools_closure_day == 1 & age_group == "aged1to4") |> 
-#   select(date, schools_closure_wave, schools_closure_wave_n)
+df_school_long |>
+  filter(schools_closure_day == 1 & age_group == "aged1to4") |>
+  select(date, schools_closure_wave, schools_closure_wave_n)
+```
 
-# analyses ---------------------------------------------------------------------
-gam_model <- 
-  gam(
-    covid_cases_7davg ~ age_group +
-      s(schools_time_since_closure, by = age_group) +
-      s(schools_time_since_closure, schools_closure_wave, bs = "fs"),
-    data = df_school_long,
-    method = "REML",
-    family = nb()
-  )
-
-# analyses ---------------------------------------------------------------------
+```{r analyses, cache=TRUE}
 source(here("analysis/gam_analysis.R"))
 source(here("analysis/te_analysis.R"))
-
 ```
@@
 ```{r gam, fig.width=4.5, fig.height=3, fig.cap="(ref:gam)"}
-gam_plot_1 + gam_plot_2 + gam_plot_3
+gam_plot
 ```

### 2022-01-18 analysis/exploration.R

@@ -1,8 +1,3 @@
-################################################################################
-#            Influence of School Closure on COVID-19 Contaminations 
-#                     and Repercussions across Age Groups  
-################################################################################
-
 # references -------------------------------------------------------------------
 ## ts causality
 "https://towardsdatascience.com/inferring-causality-in-time-series-data-b8b75fe52c46"
@@ -31,12 +26,26 @@ https://www.nature.com/articles/s41598-019-49278-8
 "
 
 # libraries --------------------------------------------------------------------
-library(tidyverse)
 library(here)
+library(httr2)
+library(gratia)
 library(janitor)
-library(zoo)
+library(magrittr)
 library(mgcv)
-#  library(gratia)
+library(tidyverse)
+library(zoo)
+
+# api access -------------------------------------------------------------------
+req <- request("https://services1.arcgis.com/eNO7HHeQ3rUcBllm/arcgis/rest/services/CovidStatisticsProfileHPSCIrelandOpenData/FeatureServer/0/query?where=1%3D1&outFields=*&returnGeometry=false&outSR=4326&f=json") |> 
+  req_headers("Accept" = "application/json")
+resp <- req_perform(req) |> 
+  resp_body_json() |> 
+  use_series(features) |> 
+  flatten_dfr() |> 
+  clean_names() |> 
+  mutate(date = as.Date(as.POSIXct(date/1000, origin = "1970-01-01"))) |> 
+  select(c(date, starts_with("aged")))
+
 # variables --------------------------------------------------------------------
 age_order <- c("aged1to4", "aged5to14", "aged15to24", "aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65to74", "aged75to84", "aged85up")
 age_name  <- c("1 to 4",   "5 to 14",   "15 to 24",   "25 to 34",   "35 to 44",   "45 to 54",   "55 to 64",   "65 to 74",   "75 to 84",   "85 and up")
@@ -387,90 +396,4 @@ draw(df_model)
 appraise(df_model)
 k.check(df_model)
 
-# df_model <- 
-#   gamm(
-#     covid_cases ~ s(schools_time_since_closure, by = age_group) + age_group,
-#     data = df_school_long,
[…]

### 2022-01-18 analysis/gam_analysis.R

@@ -1,5 +1,5 @@
 # model ------------------------------------------------------------------------
-gam_summary <- 
+gam_model <- 
   gam(
     covid_cases ~ age_group +
       s(schools_time_since_closure, by = age_group) +
@@ -7,7 +7,7 @@ gam_summary <-
     data = df_school_long,
     family = nb(), 
     method = "REML"
-  ); summary(gam_summary)
+  ); summary(gam_model)
 
 # statistics -------------------------------------------------------------------
 gam_summary <- summary(gam_model)
@@ -24,61 +24,29 @@ gam_results <-
   deframe()
 
 # visualisation ----------------------------------------------------------------
-gam_plot <- plot(gam_model, residuals = TRUE)
-
-gam_plot_indiv <- gam_plot[[1]]
-gam_plot_indiv_df <- as.data.frame(gam_plot_indiv[c("x", "se", "fit")])
-#gam_plot_indiv_point <- as.data.frame(gam_plot_indiv[c("raw", "p.resid")])
-gam_plot_1 <- ggplot(gam_plot_indiv_df, aes(x = x, y = fit)) +
-  # geom_rug(data = gam_plot_indiv_point, mapping = aes(x = raw, y = NULL), sides = "b") +
-  # geom_point(data = gam_plot_indiv_point, mapping = aes(x = raw, y = p.resid)) +
-  geom_ribbon(aes(ymin = fit - se, ymax = fit + se, y = NULL), alpha = 0.3) +
-  geom_line() +
+gam_plot <- 
+  confint(gam_model, "s(schools_time_since_closure):age_groupaged15to24") |> 
+  mutate(age_group = factor(age_group, levels = c("aged1to4", "aged5to14", "aged15to24"))) |> 
+  ggplot(aes(x = schools_time_since_closure)) +
+  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.25) + 
+  geom_line(aes(y = est)) +
+  facet_wrap(
+    ~ age_group,
+    labeller = as_labeller(
+      c(
+        aged1to4 = "1 to 4",
+        aged5to14 = "5 to 14",
+        aged15to24 = "15 to 24"
+      )
+    )) +
   labs(
     x = "Days since School Closure", 
-    y = "Standardized Effect in Age 1 to 4"
+    y = "Standardized Effect"
   ) +
   theme_bw() +
   theme(
[…]

### 2022-01-18 analysis/te_analysis.R

@@ -49,7 +49,7 @@ te_agegroup <- function(var_x, var_y){
   var_x <- select(df_wide_changes, all_of(var_x))
   var_y <- select(df_wide_changes, all_of(var_y))
   
-  transfer_entropy(var_x, var_y)[["coef"]] |> 
+  transfer_entropy(var_x, var_y, quiet = TRUE)[["coef"]] |> 
     as_tibble() |> 
     mutate(direction = c("X->Y", "Y->X"))
 }
@@ -98,7 +98,3 @@ te_plot <- bind_rows(te_results_XY,te_results_YX) |>
     axis.text.x = element_text(angle = 45, hjust = 1),
     panel.background = element_rect(colour = "black", size = 0.1)
   )
-
-
-
-

### 2022-01-18 temp.R

@@ -1,0 +1,26 @@
+test <- df |> 
+  select(date, schools_closed) |> 
+  mutate(schools_closure_day = schools_closed - lag(schools_closed)) 
+
+df_school_long |>
+  filter(schools_closure_day %in% c(1,-1) & age_group == "aged1to4") |>
+  select(date, schools_closure_wave, schools_closure_wave_n)
+
+overall_plot <- df |> 
+  ggplot(aes(date, total_cases)) +
+  geom_co(aes(fill = factor(schools_closed), height = Inf), na.rm = TRUE, alpha = .4) +
+  geom_line(na.rm = TRUE) +
+  scale_x_date(date_breaks = "3 months", date_labels = "%b %y") +
+  scale_fill_manual(labels = c("Open", "Closed", "Na"), values = c("light blue", "red", "orange")) +
+  labs(
+    x = "Date", 
+    y = "Covid-19 Cases in Ireland",
+    fill = "Schools"
+  ) +
+  theme_bw() +
+  theme(
+    legend.position = "bottom",
+    text = element_text(size = 10, family = "serif"),
+    axis.text = element_text(size = 10),
+    panel.background = element_rect(colour = "black", size = 0.1)
+  ); overall_plot

### 2022-01-19 manuscript/manuscript.Rmd
upgrade figures

@@
 
-(ref:overall) Evolution of confirmed Covid-19 cases in Ireland alongside with school closures.
+(ref:overall) Evolution of confirmed Covid-19 cases in Ireland alongside with school closures periods in grey areas (SC).
 (ref:descriptive) Evolution of confirmed Covid-19 cases by age groups in Ireland: raw data (bars) and 7 day rolling average (lines).
@@
 
-df_school |>
-  filter(schools_closed == 1) |>
-  count(schools_closure_wave) |>
-  rename(schools_closure_wave_n = n)
-
 df_school_long <- df_school |> 
@@
 
-df_school_long |>
-  filter(schools_closure_day == 1 & age_group == "aged1to4") |>
-  select(date, schools_closure_wave, schools_closure_wave_n)
+df_school_closure <- df_school |> 
+  filter(schools_closed == 1) |> 
+  count(schools_closure_wave) |> 
+  rename(schools_closure_wave_n = n) |> 
+  right_join(df_school, by = "schools_closure_wave") |>
+  filter(schools_closure_day == 1) |>
+  select(schools_closure_start_date = date, schools_closure_wave, schools_closure_wave_n) |> 
+  mutate(
+    schools_closure_end_date = schools_closure_start_date + schools_closure_wave_n,
+    schools_closure_name = glue::glue("SC {schools_closure_wave}"),
+    x_text = schools_closure_start_date + (schools_closure_end_date - schools_closure_start_date)/2
+    )
 ```
@@
 ```{r overall, fig.width=4.5, fig.height=3, fig.cap="(ref:overall)", fig.pos = "!h"}
-overall_plot <- df |> 
-  ggplot(aes(date, total_cases)) +
-  geom_tile(aes(fill = factor(schools_closed), height = Inf), na.rm = TRUE, alpha = .4) +
-  geom_line(na.rm = TRUE) +
+overall_plot <-
+  ggplot() +
+  geom_rect(data = df_school_closure, aes(xmin = schools_closure_start_date, xmax = schools_closure_end_date, ymin = -Inf, ymax = Inf), alpha = 0.25) +
+  geom_text(data = df_school_closure, aes(label = schools_closure_name, x = x_text, y = Inf), vjust = 2, family = "serif", fontface = "bold", size = 3) +
+  geom_line(data = df, aes(date, total_cases), na.rm = TRUE) +
   scale_x_date(date_breaks = "3 months", date_labels = "%b %y") +
-  scale_fill_manual(labels = c("Open", "Closed", "Na"), values = c("light blue", "red", "orange")) +
   labs(
@@
   ggplot() +
+  geom_rect(data = df_school_closure, aes(xmin = schools_closure_start_date, xmax = schools_closure_end_date, ymin = -Inf, ymax = Inf), alpha = 0.25) +
   geom_col(aes(date, covid_cases), na.rm = TRUE) +
@@
 
-```{r te, fig.width=4.5, fig.height=4.5, fig.cap="(ref:te)"}
-te_plot
-```
-
 The results of the Transfer Entropy calculations between age groups are reported Figure \@ref(fig:te).
@@
 This is the case of 15 to 24 years old who appears to be contaminated by all age groups until 84 but there is no significant support that they are contaminating groups older than 34 which should include their parents and relatives.
+
+```{r te, fig.width=4.5, fig.height=4.5, fig.cap="(ref:te)", fig.pos = "!h"}
+te_plot
+```
 

### 2022-01-19 analysis/te_analysis.R

@@ -42,7 +42,7 @@ age_order <- c("aged1to4", "aged5to14", "aged15to24", "aged25to34", "aged35to44"
 
 # solution 2 -------------------------------------------------------------------
 library(future) # enable parallel processing for all future transfer_entropy calls
-plan(multiprocess)
+plan(multisession)
 
 te_agegroup <- function(var_x, var_y){
   
@@ -61,11 +61,7 @@ te_results <-
     v = age_order, 
     repeats.allowed = FALSE
   ) |> 
-  as_tibble() |> 
-  rename(
-    X = V1,
-    Y = V2
-  ) |> 
+  as_tibble(.name_repair = ~ c("X", "Y")) |> 
   group_by(X, Y) |> 
   summarise(te_agegroup(X, Y)) |> 
   ungroup()

### 2022-01-19 temp.R

@@ -1,26 +1,0 @@
-test <- df |> 
-  select(date, schools_closed) |> 
-  mutate(schools_closure_day = schools_closed - lag(schools_closed)) 
-
-df_school_long |>
-  filter(schools_closure_day %in% c(1,-1) & age_group == "aged1to4") |>
-  select(date, schools_closure_wave, schools_closure_wave_n)
-
-overall_plot <- df |> 
-  ggplot(aes(date, total_cases)) +
-  geom_co(aes(fill = factor(schools_closed), height = Inf), na.rm = TRUE, alpha = .4) +
-  geom_line(na.rm = TRUE) +
-  scale_x_date(date_breaks = "3 months", date_labels = "%b %y") +
-  scale_fill_manual(labels = c("Open", "Closed", "Na"), values = c("light blue", "red", "orange")) +
-  labs(
-    x = "Date", 
-    y = "Covid-19 Cases in Ireland",
-    fill = "Schools"
-  ) +
-  theme_bw() +
-  theme(
-    legend.position = "bottom",
-    text = element_text(size = 10, family = "serif"),
-    axis.text = element_text(size = 10),
-    panel.background = element_rect(colour = "black", size = 0.1)
-  ); overall_plot

### 2022-02-10 manuscript/appendix_1.Rmd
europe

# Augmented Dickey-Fuller test for stationay timeseries

```{r}
adf_res <- df_long |>
  group_by(age_group) |>
  summarise(
    aTSA::adf.test(covid_changes, output = FALSE) |>
      map_df(as_tibble)
  )
```

### 2022-02-10 manuscript/manuscript.Rmd
europe

--- 2022-01-18 manuscript/manuscript.Rmd
@@
 ---
-title: Influence of School Closure on COVID-19 Contaminations in Ireland and Repercussions across Age Groups
+title: Influence of School Closure on COVID-19 Contaminations in Europe and Repercussions across Age Groups
 author:
@@
 header-includes:
+  - \usepackage{float}
   - \usepackage{booktabs}
-  - \usepackage{float}
+  - \usepackage{longtable}
+  - \usepackage{threeparttable}
+  - \floatplacement{figure}{H}
+  - \floatplacement{table}{H}
+  - \usepackage{caption}
 
@@
 
-(ref:overall) Evolution of confirmed Covid-19 cases in Ireland alongside with school closures.
-(ref:descriptive) Evolution of confirmed Covid-19 cases by age groups in Ireland: raw data (bars) and 7 day rolling average (lines).
+(ref:overall) Cumulative COVID-19 cases number for selected European countries since the beginning of the pandemic. Source: COVerAGE-DB [@riffe2021data].
+(ref:descriptive) Periods of schools closure since the beginning of the COVID-19 pandemic for selected European countries and their reason: regular academic break vs.
+closure due to government decisions.
+Source: @unesco2022https.
 (ref:gam) Standardized effect of the smooth term in Generalized Additive Model by age group.
-(ref:te) Matrix of Transfer Entropy coefficient and their corresponding p-value according every age group combination.
+(ref:te) Matrix of Transfer Entropy coefficients according every age group combination for each of the selected European country.
 Age groups on the x-axis are influencing the age groups on the y-axis ($X \rightarrow Y$).
-* $p < .05$, ** $p < .01$, *** $p < .001$.
+The significance of each Transfer Entropy coefficient is provided in the Appendix 2.
 
@@
 # libraries --------------------------------------------------------------------
-library(gratia)
+library(ggrepel)
 library(here)
@@
 library(knitr)
-library(mgcv)
-library(RTransferEntropy)
+library(osfr)
+library(scales)
 library(tidyverse)
-library(zoo)
 
@@
   echo = FALSE, 
-  fig.retina = 3
+  fig.retina = 3,
+  dev = "ragg_png"
   )
 
-# variables --------------------------------------------------------------------
-age_order <- c("aged1to4", "aged5to14", "aged15to24", "aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65to74", "aged75to84", "aged85up")
-age_name  <- c("1 to 4",   "5 to 14",   "15 to 24",   "25 to 34",   "35 to 44",   "45 to 54",   "55 to 64",   "65 to 74",   "75 to 84",   "85 and up")
-age_color <- c("#999999",  "#E69F00",   "#56B4E9",    "#009E73",    "#F0E442",    "#0072B2",    "#D55E00",    "#CC79A7",    "#000000",    "#CAB2D6")
+# raw data ---------------------------------------------------------------------
+# osf_retrieve_file("7tnfh") |>
+#   osf_download(conflicts = "overwrite", progress = TRUE, verbose = TRUE)
+# 
+# df <- "Output_5.zip" |>
+#   read_csv(skip = 3, col_types = "ccccciiddd") |>
+#   mutate(Date = as.Date(Date, format = "%d.%m.%Y")) |>
+#   clean_names()
+# 
+# file.remove("Output_5.zip")
+# 
+# write_csv(df, here("data/df.csv"))
+# 
+# df <- read_csv(here("data/df.csv"))
+# 
+# df_school_closure <- 
+#   read_csv(
+#     "https://en.unesco.org/sites/default/files/covid_impact_education_full.csv",
+#     col_types = list(.default = col_character(), Date = col_date(format = "%d/%m/%Y"))) |> 
+#   clean_names() |>
+#   select(date, country, status) |> 
+#   mutate(country = case_when(
+#     country == "Bolivia (Plurinational State of)" ~ "Bolivia",  
+#     country == "Central African republic" ~ "Central African Republic",
+#     country == "Iran (Islamic Republic of)" ~ "Iran",
+#     country == "Republic of Moldova" ~ "Moldova",
+#     country == "Republic of Korea" ~ "South Korea",
+#     country == "United Kingdom of Great Britain and Northern Ireland" ~ "United Kingdom",
+#     country == "United States of America" ~ "USA",
+#     country == "Viet Nam" ~ "Vietnam",
+#     TRUE ~ country
+#     )
+#   )
+# 
+# df_check <- 
+#   read_csv("https://covid19.who.int/WHO-COVID-19-global-data.csv") |> 
+#   clean_names() |> 
+#   rename(date = date_reported) |> 
+#   mutate(country = case_when(
+#     country == "Bolivia (Plurinational State of)" ~ "Bolivia",  
+#     country == "Iran (Islamic Republic of)" ~ "Iran",
+#     country == "occupied Palestinian territory, including east Jerusalem" ~ "Palestine",
+#     country == "Republic of Moldova" ~ "Moldova",
+#     country == "Republic of Korea" ~ "South Korea",
+#     country == "United States of America" ~ "USA",
+#     country == "Venezuela (Bolivarian Republic of)" ~ "Venezuela",
+#     country == "Viet Nam" ~ "Vietnam",
+#     TRUE ~ country
+#     )
+#   )
+# 
+# # variables --------------------------------------------------------------------
+# ## keep only closure waves longer than `days_min_closure` but display only the first `days_after_closure`
+# days_min_closure <- 21
+# days_after_closure <- 28
+# ## type of school closure to investigate
+# period_closure_to_keep <- c(
+#   #"Partially open",
+#   "Academic break",
+#   "Closed due to COVID-19"
+#   )
+# 
+# ## countries of interest
+# countries_of_interest <- c(
+#   "Austria", "Belgium", "Bulgaria", "Croatia", "Estonia", "France", 
+#   "Germany", "Greece", "Netherlands", "Portugal", "Slovakia", "Spain"
+#   )
+# 
+# # transformed data -------------------------------------------------------------
+# df_global <- df |> 
+#   filter(
+#     country %in% countries_of_interest,
+#     date > as.Date("2020-01-01"),
+#     region == "All",
+#     sex == "b"
+#   ) |> 
+#   group_by(country, date) |> 
+#   summarise(cumulative_cases_cal = sum(cases)) |> 
+#   group_by(country) |> 
+#   complete(date = full_seq(date, 1)) |> 
+#   mutate(
+#     timeline = date - lag(date),  # check if gap between dates
+#     daily_cases = round(cumulative_cases_cal - lag(cumulative_cases_cal), 0)
+#   ) |> 
+#   ungroup() |> 
+#   left_join(df_school_closure, by = c("date", "country")) |> 
+#   group_by(country) |> 
+#   fill(status) |> 
+#   ungroup() |> 
+#   mutate(schools_closed = if_else(status %in% period_closure_to_keep, 1, 0)) |> 
+#   drop_na(status) |> 
+#   left_join(df_check, by = c("date", "country"))
+# 
+# write_csv(df_global, here("data/df_global.csv"))
+# 
+# df_global_age <- df |> 
+#   filter(
+#     country %in% countries_of_interest,
+#     date > as.Date("2020-01-01"),
+#     region == "All",
+#     sex == "b"
+#   ) |> 
+#   mutate(
+#     age_grp = case_when(
+#       age == 0  ~ "aged0to4",
+#       age == 5  ~ "aged5to9",
+#       age == 10 ~ "aged10to14",
+#       age == 15 ~ "aged15to19",
+#       age == 20 ~ "aged20to24",
+#       between(age, 25, 34) ~ "aged25to34",
+#       between(age, 35, 44) ~ "aged35to44",
+#       between(age, 45, 54) ~ "aged45to54",
+#       between(age, 55, 64) ~ "aged55to64",
+#       age > 64 ~ "aged65+"
+#     )
+#   ) |> 
+#   group_by(country, date, age_grp) |> 
+#   summarise(cumulative_cases_cal = sum(cases)) |> 
+#   group_by(country, age_grp) |> 
+#   complete(date = full_seq(date, 1)) |> 
+#   mutate(
+#     timeline = date - lag(date),  # check if gap between dates
+#     daily_cases = round(cumulative_cases_cal - lag(cumulative_cases_cal), 0),
+#     daily_changes = daily_cases - lag(daily_cases)
+#   ) |> 
+#   ungroup()
+# 
+# write_csv(df_global_age, here("data/df_global_age.csv"))
 
-# data -------------------------------------------------------------------------
-df <- 
-  read_csv(
-    here("data/age_groups.csv"), 
-    col_types = list(
-      .default = col_double(), 
-      Date = col_date(format = "%d/%m/%Y")
-    )
-  ) |> 
-  clean_names() |> 
-  rename(total_cases = total) |> 
-  arrange(date) |> 
-  mutate(
-    timeline = date - lag(date),  # check if gap between dates
-    total_changes = total_cases - lag(total_cases)
-  )
-
-df_long <- df |> 
-  select(c(date, starts_with("aged"))) |> 
-  pivot_longer(-date, names_to = "age_group", values_to = "covid_cases") |> 
-  mutate(age_group = factor(age_group, levels = age_order)) |> 
-  group_by(age_group) |> 
-  mutate(
-    covid_cases_7davg = rollmean(covid_cases, k = 7, fill = NA),
-    covid_changes = covid_cases - lag(covid_cases),
-    covid_changes_7davg = rollmean(covid_changes, k = 7, fill = NA)
-  ) |> 
-  ungroup()
-
-df_wide_changes <- df_long |> 
-  select(date, age_group, covid_changes) |> 
-  pivot_wider(names_from = age_group, values_from = covid_changes) |> 
-  drop_na()
-
-df_school <- df |> 
-  select(date, schools_closed) |> 
-  mutate(schools_closure_day = schools_closed - lag(schools_closed)) |> 
-  filter(schools_closure_day == 1) |> 
-  rowid_to_column("schools_closure_wave") |> 
-  full_join(df, by = c("date", "schools_closed")) |> 
-  arrange(date) |> 
-  fill(schools_closure_wave, .direction = "down") |> 
-  mutate(schools_closure_wave = replace_na(schools_closure_wave, 0)) |> 
-  group_by(schools_closure_wave) |> 
-  mutate(schools_time_since_closure = row_number()) |> 
-  ungroup()
-
-df_school |>
-  filter(schools_closed == 1) |>
-  count(schools_closure_wave) |>
-  rename(schools_closure_wave_n = n)
-
-df_school_long <- df_school |> 
-  filter(schools_closed == 1) |> 
-  count(schools_closure_wave) |> 
-  rename(schools_closure_wave_n = n) |> 
-  right_join(df_school, by = "schools_closure_wave") |> 
-  filter(schools_closure_wave_n > 20 & schools_time_since_closure < 40) |> # keep only closure waves longer than 20 days but display only the first 39 days
-  pivot_longer(c("aged1to4", "aged5to14", "aged15to24"), names_to = "age_group", values_to = "covid_cases") |> 
-  mutate(
-    schools_closure_wave = factor(schools_closure_wave),
-    age_group = factor(age_group, levels = c("aged1to4", "aged5to14", "aged15to24"))
-  ) |> 
-  group_by(age_group) |> 
-  mutate(
-    covid_changes = covid_cases - lag(covid_cases),
-    covid_cases_7davg = rollmean(covid_cases, k = 7, fill = NA),
-    covid_changes_7davg = rollmean(covid_changes, k = 7, fill = NA)
-  ) |> 
-  ungroup()
-
-df_school_long |>
-  filter(schools_closure_day == 1 & age_group == "aged1to4") |>
-  select(date, schools_closure_wave, schools_closure_wave_n)
+df_global <- read_csv(here("data/df_global.csv"))
+df_global_age <- read_csv(here("data/df_global_age.csv"))
 ```
 
-```{r analyses, cache=TRUE}
-source(here("analysis/gam_analysis.R"))
-source(here("analysis/te_analysis.R"))
+```{r analyses, cache=FALSE}
+source(here("analysis/gam_analysis_europe.R"))
+source(here("analysis/te_analysis_europe.R"))
 ```
@@
 
-Since the onset of Covid-19, contradicting messages are circulating about the spread of the virus in schools.
+Since the onset of COVID-19, contradicting messages are circulating about the spread of the virus in schools.
 Sometimes presented as a safe environment [@falk2021covid; @walger2020children], at other times designed as an aggravating factor of the pandemic [@lopez2020transmission; @meuris2021transmission], child care facilities and schools have been closed multiple times in most countries to protect the public health.
-However, the efficacy of school closure is still unknown and little is know about how Covid-19 Cases spread across age cohort and especially from the younger cohort being at school to older cohorts including parents, grand parents and relatives.
+However, the efficacy of school closure is still unknown and little is know about how COVID-19 Cases spread across age cohort and especially from the younger cohort being at school to older cohorts including parents, grand parents and relatives.
 Some evidence show that school is an influential environment where the virus is likely to spread.
 Once contaminated at school, a common spread believe is that the children and teenagers will then contaminate their parents.
-Therefore, among the political actions to stop the spread of Covid-19, the closure of primary and secondary schools was widely adopted worldwide.
-This research aims to investigate 1) the effect of school closure on the evolution of Covid-19 cases and 2) the temporal relationship between the rise of cases in children and teenagers with the rise of cases of older age group.
+Therefore, among the political actions to stop the spread of COVID-19, the closure of primary and secondary schools was widely adopted worldwide.
+This research aims to investigate 1) the effect of school closure on the evolution of COVID-19 cases and 2) the temporal relationship between the rise of cases in children and teenagers with the rise of cases of older age group.
 
@@
 Therefore, it is legitimate to believe that by closing schools, a reduction of the contaminations would be observed in the younger age groups.
-However, the efficiency of school closure on the reduction of Covid-19 cases is still questioned.
-While some research have observed that school closures contribute to limit or to reduce the growth rate of confirmed cases after implementation [@yoshiyuki2020effects; @stage2021shut], other did not observe a change in the evolution of Covid-19 cases [@iwata2020school; @chang2020modelling].
-For instance, a controlled comparison between similar localities in Japan with schools closed and school open did not revealed any evidence that school closures reduced the spread of Covid-19 [@fukumoto2021no].
-If the school closure had a real impact on the evolution of confirmed Covid-19 cases, it should be possible to observe a decrease or a least an inflection in the trend of its evolution among younger age groups.
+However, the efficiency of school closure on the reduction of COVID-19 cases is still questioned.
+While some research have observed that school closures contribute to limit or to reduce the growth rate of confirmed cases after implementation [@yoshiyuki2020effects; @stage2021shut], other did not observe a change in the evolution of COVID-19 cases [@iwata2020school; @chang2020modelling].
+For instance, a controlled comparison between similar localities in Japan with schools closed and school open did not revealed any evidence that school closures reduced the spread of COVID-19 [@fukumoto2021no].
+If the school closure had a real impact on the evolution of confirmed COVID-19 cases, it should be possible to observe a decrease or a least an inflection in the trend of its evolution among younger age groups.
 
@@
 
-A second implicit believe regarding the effect of school closure on the spread of Covid-19 is that school not only has an effect on to the spread of the virus in children and teenagers but also has a knock-on effect on the spread of the virus in older age groups also called Secondary Attack Rate (SAR).
+A second implicit believe regarding the effect of school closure on the spread of COVID-19 is that school not only has an effect on to the spread of the virus in children and teenagers but also has a knock-on effect on the spread of the virus in older age groups also called Secondary Attack Rate (SAR).
 The contaminated children and teenagers would bring the virus back at home and, then, they will contaminate their parents and relatives.
@@
 
-The data were collected from the official daily publication of Covid-19 cases by the Irish Department of Health.
-They include the amount of daily cases for 10 age groups from `r format(min(df$date), "%B %d, %Y")` to `r format(max(df$date), "%B %d, %Y")`.
-For each day during this period, school closure either due to a government decision or a public holiday is determined.
+### COVID-19 cases per age group
+
+The number of COVID-19 cases per age group has been taken from the COVerAGE-DB project [@riffe2021data].
+The COVerAGE-DB project consists in 3 data files: an "Input" data file which collects the official COVID-19 cases from 117 countries, the "Output_5" data file which is a projection of COVID-19 cases by group of 5 years, and the "Output_10" data file which is a projection of COVID-19 cases by group of 10 years.
+For our analyses we have used the "Output_5" which deals with the heterogeneity of countries' reporting format by using spline approximations when the data for this age bracket is not available for a country.
+Therefore, it allows a precise analysis between age groups under 24 while age groups of 25 and above are reduced to 10 year brackets and groups above 64 concatenated all together.
+
+In order to cross validate the data of obtained after spline approximations, a comparison with the data published by the @who2022https reveals perfect similarities.
+
+The original data consist in 14089320 observations of 10 variables (117 distinct country, region within the country, an unique observation code, the date of the observation, the gender which can be male, female or both, the age bracket by 5 years from 0 to 100, a confirmation of the age interval for each bracket, the total number of cases so far, the total number of deaths and the total number of tests performed) from `r format(min(df_global$date), "%B %d, %Y")` to `r format(max(df_global$date), "%B %d, %Y")`.
+
+After removing countries with missing and inconstant values, only 22 are suitable for the data analyses.
+However, in order to focus this analysis to geographically and culturally comparable countries, only 12 European countries are kept: Austria, Belgium, Bulgaria, Croatia, Estonia, France, Germany, Greece, Netherlands, Portugal, Slovakia, and Spain.
+
+The observations are reported in terms of total number of COVID-19 cases from the start to the pandemic.
+Consequently, the daily number of cases at a specific date $n\prime_{t}$ is calculated with the difference between the total cases at a date $t$ and the total cases at a date $t-1$ (i.e., derivative 1).
+In addition the change in daily number of cases $n\prime\prime_{t}$ between $n\prime_{t}$ and $n\prime_{t-1}$ has also been calculated (i.e., derivative 2).
+
+### School closure
+
+The information regarding the school closure on a day by day basis is provided by the @unesco2022https.
+For each day, in each country, the status of the schools is indicated as fully open, partially open, closed due to COVID-19, or closed due to an academic break.
+Because it would be difficult to measure the effect of schools closure at a country level when schools are partially closed, on closures due to COVID-19, or due to an academic break are considered.
+Indeed, both are considered as closure at a country wide level.
 
@@
 
-Four periods of school closure longer that 14 days have been identified.
-These four periods are used to fit a Generalised Additive Model (or GAM) using the R package *mgcv* [@wood2017generalized] in order to test the hypothesis of a significant change in the evolution of cases among age groups from 1 to 4, from 5 to 14, and from 15 to 24 (Eq 1).
+The effect of school closure on the trend of daily COVID-19 cases is analysed for each country.
+Periods of closure longer than 21 days are compared for 28 days regardless if the schools opened or not after 21 days.
 
-By estimating the degree of smoothness of a Bayesian spline smoothing using restricted maximum likelihood estimation [@wood2011fast], GAM identifies dynamic patterns underlying the evolution of Covid-19 cases reported while taking into account the random effect of different age groups as follows:
+A Generalised Additive Model (or GAM) using the R package *mgcv* [@wood2017generalized] is fitted on the daily COVID-19 cases to test the hypothesis of a significant non-linear evolution of cases among age groups from 0 to 4, from 5 to 9, from 10 to 14, from 15 to 19, and from 20 to 24.
+The model also estimates the overall non-linear effect by country taking into account the interaction between age groups and countries as random intercepts and the interaction between time, countries and period of closure as random effects (Eq 1).
+
+By estimating the degree of smoothness of a Bayesian spline smoothing using restricted fast maximum likelihood estimation [@wood2011fast], GAM identifies dynamic patterns underlying the evolution of COVID-19 cases reported while taking into account the random effect of different age groups and countries as follows:
 
 \begin{align}
-Y_{is} = \alpha_{i} + f(X_{s}) + a_{is} + \epsilon_{is}
+n\prime_{tgc} = \alpha_{gc} + f(closure_{tg}) + f(closure_{tc}) + a_{tcp} + \epsilon_{tgc}
 \end{align}
 
-\noindent where $i$ is an age group among the 10 age groups investigated and $s$ is the date corresponding to the confirmed Covid-19 cases.
-$Y_{is}$ represents the confirmed Covid-19 cases assuming a negative binomial distribution for the fitting [@loader2006local].
-The response variable $Y_{is}$ includes a specific intercept for each age group ($\alpha_{i}$).
-A smooth effect over time $f(X_{s})$ is applied to model (Eq 2) to predict the nonlinear evolution of Covid-19 cases.
-This smooth effect $f(X_{s})$ is built up in basic components, called the basis functions $b_{j}(X_{s})$, such that:
+\noindent where $t$ is the date corresponding to the confirmed COVID-19 cases, $g$ is an age group among the 10 age groups investigated, $c$ is a country among the 12 European countries selected, and $p$ is a specific period of closure.
+$n\prime_{tgc}$ represents the confirmed COVID-19 cases assuming a negative binomial distribution for the fitting [@loader2006local].
+The response variable $n\prime_{tgc}$ includes a specific intercept for each age group and each country ($\alpha_{gc}$).
+A smooth effect over time $f(closure_{tg})$ is applied to model for each age group and a smooth effect over time $f(closure_{tc})$ is applied to model for each country (Eq 2 and 3).
+These smooth effect $f(closure_{tg})$ and $f(closure_{tc})$ are built up in basic components, called the basis functions $b_{i}(X_{tg})$ and $b_{j}(X_{tg})$, such that:
 
 \begin{align}
-f(X_{s})=\sum_{j=1}^{k}\beta_{j}\times b_{j}(X_{s})
+f(closure_{tg})=\sum_{i=1}^{k}\beta_{i}\times b_{i}(X_{tg})
+\\
+f(closure_{tc})=\sum_{j=1}^{k}\beta_{j}\times b_{j}(X_{tc})
 \end{align}
 
-\noindent where the regression parameters $\beta_{j}$ are estimated by penalized likelihood maximization.
+\noindent where the regression parameters $\beta_{i}$ and $\beta_{j}$ are estimated by penalized likelihood maximization.
 
-The model also includes the random effects term $a_{is}= Zb_{i}$ where $Z$ is a random effects matrix and $b_{i}$ is a vector of random effects described by $b_{i} \sim N(0, D)$.
+The model also includes the random effects term $a_{tcp}= Zb_{i}$ where $Z$ is a random effects matrix and $b_{i}$ is a vector of random effects described by $b_{i} \sim N(0, D)$.
 In this, $D$ represents a covariance matrix.
-The error term $\epsilon_{is}$ is assumed to be normally and independently distributed $\epsilon_{is} \sim N(0, \sigma^2)$.
+The error term $\epsilon_{tcp}$ is assumed to be normally and independently distributed $\epsilon_{tcp} \sim N(0, \sigma^2)$.
 
@@
 
-Transfer Entropy (TE) can be used to infer the temporal relationship between two time series $I$ and $J$.
-This measure indicates whether $J$ can be used to to reduce the uncertainty on the future of $I$ and, consequently, that $J$ causes $I$ [@schreiber2000measuring].
+Transfer Entropy (TE) can be used to infer the temporal relationship between two time series $X$ and $Y$.
+This measure indicates whether $X$ can be used to to reduce the uncertainty on the future of $Y$ and, consequently, that $X$ causes $Y$ [@schreiber2000measuring].
 As such, Granger causality is a special case of TE applied to times series that are jointly Gaussian distributed [@barnett2009granger].
@@
 
-The influence of the evolution in Covid-19 cases across all age groups is evaluated using Shannon's transfer entropy is given by:
+The influence of the evolution in COVID-19 cases across all age groups is evaluated using Shannon's transfer entropy is given by:
 
 \begin{align}
-  T_{X \rightarrow Y}(k,l) = \sum_{x,y} p\left(x_{t+1}, x_t^{(k)}, y_t^{(l)}\right) \cdot log \left(\frac{p\left(x_{t+1}| x_t^{(k)}, y_t^{(l)}\right)}{p\left(x_{t+1}|x_t^{(k)}\right)}\right),
+  T_{X \rightarrow Y}(k,l) = \sum_{x,y} p\left(x_{t+1}, x_t^{(k)}, y_t^{(l)}\right) log \left(\frac{p\left(x_{t+1}| x_t^{(k)}, y_t^{(l)}\right)}{p\left(x_{t+1}|x_t^{(k)}\right)}\right)
 \end{align}
 
-\noindent where $T_{X\rightarrow Y}$ consequently measures the influence of the change dynamic from an age group $X$ to another age group $Y$ (Eq 3).
+\noindent where $T_{X\rightarrow Y}$ consequently measures the influence of the change dynamic from an age group $X$ to another age group $Y$ for every country (Eq 4).
 
-In order to observe how increase in daily numbers reported from a specific age group affects the other age groups, the day-by-day difference in Covid-19 confirmed cases (i.e., differences with a lag 1) is calculated.
-The use of day-by-day difference in Covid-19 confirmed cases satisfies the requirement of stationary for the calculation of Shannon's Transfer Entropy [@shannon1948mathematical].
-The Transfer Entropy analysis of younger age groups daily changes in Covid-19 cases on older age groups is done using the R package *RTransferEntropy* [@behrendt2019rtransferentropy].
+The day-by-day difference in COVID-19 confirmed cases $n\prime\prime_t$ is used to satisfy the stationary requirement for the calculation of Shannon's Transfer Entropy [@shannon1948mathematical].
+The Transfer Entropy analysis acrross all age group is done using the R package *RTransferEntropy* [@behrendt2019rtransferentropy].
 
@@
 
-Overall, the trend of confirmed Covid-19 cases in Ireland is similar to those of other European countries (Figure \@ref(fig:overall)).
-Indeed it is possible to observe a significant increase corresponding to each of the 4 waves of transmission (March-April 2020, December 2020-January 2021, September-October 2021 and November-December 2021).
+Overall, the trend of confirmed COVID-19 cases follows similar patterns across the selected European countries with scales respecting the size of population in these countries (Figure \@ref(fig:overall)).
+Indeed, the bigger the country, the higher the total number of COVID-19 cases.
 
-```{r overall, fig.width=4.5, fig.height=3, fig.cap="(ref:overall)", fig.pos = "!h"}
-overall_plot <- df |> 
-  ggplot(aes(date, total_cases)) +
-  geom_tile(aes(fill = factor(schools_closed), height = Inf), na.rm = TRUE, alpha = .4) +
-  geom_line(na.rm = TRUE) +
-  scale_x_date(date_breaks = "3 months", date_labels = "%b %y") +
-  scale_fill_manual(labels = c("Open", "Closed", "Na"), values = c("light blue", "red", "orange")) +
-  labs(
-    x = "Date", 
-    y = "Covid-19 Cases in Ireland",
-    fill = "Schools"
+```{r overall, fig.height=3, out.width="\\textwidth", fig.cap="(ref:overall)"}
+overall_plot <- df_global |> 
+  mutate(
+    country = fct_rev(country)
+  ) |> 
+  ggplot(aes(date, cumulative_cases_cal, group = country)) +
+  geom_line(na.rm = TRUE, color = "gray40", alpha = 0.4) +
+  geom_text_repel(
+    data = df_global |>
+      group_by(country) |>
+      slice_max(cumulative_cases_cal, with_ties = FALSE),
+    aes(label = country),
+    hjust = 0,
+    direction = "y",
+    segment.color = NA,
+    seed = 123,
+    size = 3,
+    force = 0.5,
+    family = "serif"
+  ) +
+  scale_x_date(
+    "Date", 
+    date_breaks = "3 months", 
+    date_labels = "%b %y", 
+    limits = c(as.Date("2020-02-01"), as.Date("2022-05-01"))
+  ) +
+  scale_y_continuous(
+    "Total COVID-19 Cases (Million)", 
+    labels = unit_format(unit = "M", scale = 1e-6)
   ) +
@@
   theme(
-    legend.position = "bottom",
-    text = element_text(size = 10, family = "serif"),
-    axis.text = element_text(size = 10),
-    panel.background = element_rect(colour = "black", size = 0.1)
+    legend.position = "none",
+    text = element_text(size = 10, family = "serif")
   )
@@
 
-The evolution of Covid-19 cases reveals some similarities across all age groups.
+The evolution of COVID-19 cases reveals some similarities across all age groups.
 However, the influence of each waves on the each groups has also some particularities (Figure \@ref(fig:descriptive)).
@@
 
-```{r descriptive, fig.width=4.5, fig.height=7, fig.cap="(ref:descriptive)"}
-descriptive_plot <- df_long |> 
-  ggplot() +
-  geom_col(aes(date, covid_cases), na.rm = TRUE) +
-  geom_line(aes(date, covid_cases_7davg, color = age_group), na.rm = TRUE) +
+```{r descriptive, fig.height=3, out.width="\\textwidth", fig.cap="(ref:descriptive)"}
+descriptive_plot <- df_global |> 
+  filter(schools_closed == 1) |> 
+  mutate(
+    status = status |> 
+      str_remove("Closed due to ") |> 
+      fct_rev(),
+    country = country |> 
+      fct_rev()
+  ) |> 
+  ggplot(aes(date, country, color = status)) +
+  geom_point(na.rm = TRUE) +
   scale_x_date("Date", date_breaks = "3 months", date_labels = "%b %y") +
-  scale_y_continuous("Confirmed Covid-19 Cases in Ireland") +
-  scale_color_manual(labels = age_name, values = age_color) +
-  facet_grid(
-    age_group ~ ., 
-    scales = "free",
-    labeller = as_labeller(
-      c(
-        aged1to4 = "1 to 4",
-        aged5to14 = "5 to 14",
-        aged15to24 = "15 to 24", 
-        aged25to34 = "25 to 34", 
-        aged35to44 = "35 to 44", 
-        aged45to54 = "45 to 54", 
-        aged55to64 = "55 to 64", 
-        aged65to74 = "65 to 74", 
-        aged75to84 = "75 to 84", 
-        aged85up = "85 and up"
-        )
-      )
-  ) +
+  scale_y_discrete("Selected European Countries") +
+  scale_colour_manual("Schools Closure Reason", values = c("gray10", "gray60")) +
   theme_bw() +
   theme(
-    legend.position = "none",
-    strip.background = element_blank(),
-    strip.placement = "outside",
-    text = element_text(size = 10, family = "serif"),
-    axis.text = element_text(size = 10),
-    panel.background = element_rect(colour = "black", size = 0.1)
+    legend.position = "bottom",
+    text = element_text(size = 10, family = "serif")
   )
@@
 
-In order to evaluate the shape of the trend in the numbers of Covid-19 cases reported after the three school closure longer than 14 consecutive days, a GAM was fitted taking into account numbers for 1 to 4, 5 to 14, and 15 to 24 year old.
-The obtained results satisfy the requirements to fit this model which explains `r scales::percent(gam_summary$dev.expl, accuracy = 0.1)` of the deviance in Covid-19 cases.
-These results show no significant change in the evolution of cases for 1 to 4 year old (`r gam_results["s(date):1to4"]`) as well as for 5 to 14 year old (`r gam_results["s(date):5to14"]`).
-However, the smooth term of the GAM is significant for 15 to 24 year old (`r gam_results["s(date):15to24"]`), which indicates that the evolution of cases is not linear and, therefore, the school closure had a significant impact on the evolution of cases (Figure \@ref(fig:gam)).
+In order to evaluate the shape of the trend in the numbers of COVID-19 cases reported after the three school closure longer than 21 consecutive days, a GAM was fitted taking into account numbers for 0 to 4, 5 to 9, 10 to 14, 15 to 19, and 20 to 24 year old across all the selected European countries.
+The obtained results satisfy the requirements to fit this model which explains `r percent(gam_summary$dev.expl, accuracy = 0.1)` of the deviance in COVID-19 cases.
 
-```{r gam, fig.width=4.5, fig.height=3, fig.cap="(ref:gam)"}
-gam_plot
+Overall, the results did not revealed a main effect of school closure at a country level for most of the selected European countries (Austria: `r gam_results["Austria"]`; Belgium: `r gam_results["Belgium"]`; Bulgaria: `r gam_results["Bulgaria"]`; Croatia: `r gam_results["Croatia"]`; Estonia: `r gam_results["Estonia"]`; France: `r gam_results["France"]`; Greece: `r gam_results["Greece"]`; Netherlands: `r gam_results["Netherlands"]`; Portugal: `r gam_results["Portugal"]`; Spain: `r gam_results["Spain"]`) except for Germany (`r gam_results["Germany"]`) and for Slovakia (`r gam_results["Slovakia"]`).
+
+When looking at the difference between age groups, results show no significant change in the evolution of cases until 19 year old (0 to 4: `r gam_results["0to4"]`; 5 to 9: `r gam_results["5to9"]`; 10 to 14: `r gam_results["10to14"]`; 15 to 19: `r gam_results["15to19"]`).
+However, the smooth term of the GAM is significant for 20 to 24 year old (`r gam_results["20to24"]`), which indicates that the evolution of cases is not linear and, therefore, the school closure had a significant impact on the evolution of cases (Figure \@ref(fig:gam)).
+
+```{r gam, fig.height=3, out.width="\\textwidth", fig.cap="(ref:gam)"}
+gam_plot_age
 ```
 
-Before the Transfer Entropy analysis, an Augmented Dickey-Fuller has been applied on each age group in order to insure that the daily changes in Covid-19 cases are stationary.
+Before the Transfer Entropy analysis, an Augmented Dickey-Fuller has been applied on each age group in order to insure that the daily changes in COVID-19 cases are stationary (see Appendix 1).
 
-```{r}
-# adf_res <- df_long |> 
-#   group_by(age_group) |> 
-#   summarise(
-#     aTSA::adf.test(covid_changes, output = FALSE) |> 
-#       map_df(as_tibble)
-#   )
-```
+The results of the Transfer Entropy calculations between age groups for each of the 12 selected European countries are reported Figure \@ref(fig:te).
+The first observation is the absence of symmetry between influencing age groups (i.e., $X$) and influenced age groups (i.e., $Y$).
+Indeed, the change in COVID-19 cases in some age groups are influenced by other age groups but they are not reciprocally influencing these age groups.
 
-```{r te, fig.width=4.5, fig.height=4.5, fig.cap="(ref:te)"}
+Additionally, to highlight differences and similarities between countries, a grid has been added to Figure \@ref(fig:te): the upper left quadrant indicates how younger age cohorts are influencing older age cohorts, the bottom right quadrant indicates how older age cohorts are influencing younger age cohorts, finally lower left and and upper right are indicating how younger or older age cohorts are influencing themselves.
+By analysing these quadrant it is possible to identify similar patterns across multiple countries.
+Indeed, it appears that Austria, Germany and the Netherlands have significantly higher TE coefficients in the upper left quadrant of the matrix, which indicate that the daily changes in COVID-19 cases number in younger cohorts is predicting the the daily changes in COVID-19 cases number in older cohorts.
+Alternatively, it appears that Austria, the Netherlands, Portugal and Spain have significantly higher TE coefficients in the lower right quadrant of the matrix, which indicate that the daily changes in COVID-19 cases number in older cohorts is predicting the the daily changes in COVID-19 cases number in younger cohorts.
+
+```{r te, fig.height=8, out.width="\\textwidth", fig.cap="(ref:te)"}
 te_plot
 ```
-
-The results of the Transfer Entropy calculations between age groups are reported Figure \@ref(fig:te).
-The first observation is the absence of symmetry between influencing age groups (i.e., $X$) and influenced age groups (i.e., $Y$).
-Indeed, the change in Covid-19 cases in some age groups are influenced by other age groups but they are not reciprocally influencing these age groups.
-This is the case of 15 to 24 years old who appears to be contaminated by all age groups until 84 but there is no significant support that they are contaminating groups older than 34 which should include their parents and relatives.
 
@@
 
-This indicate mix results, while the school closure was ineffective in child care settings and primary school, the closure of secondary schools and colleges had an impact on Covid-19 transmission.
-This impact is probably responsible for the absence of reciprocity in the transmission of 15 to 24 age group.
-By drastically reducing their social contacts in secondary schools and colleges, this cohort has more been the target of Covid-19 contaminations rather than the vector of Covid-19.
+This indicate mix results, while the school closure was ineffective in pre-primary, primary and secondary levels (i.e., ISCED 0 to 3), the closure of tertiary levels (i.e., ISCED 5 to 8) had an impact on COVID-19 transmission.
+However, different pattern emerges according the selected European countries.
+Some have shown asymmetric patterns of influence between age groups while other did.
 

### 2022-02-10 report/preprocessing_report.Rmd
europe

---
title: "Data Preprocessing Report"
output:
  bookdown::pdf_book:
    toc: false
---

(ref:data-collected) Disparities between countries regarding the reporting of Covid-19 cases. Each dots is a valid observation.
(ref:final-list) List of remaining countries and the corresponding number of target periods that can be analysed from each country.

```{r setup, include=FALSE}
# libraries --------------------------------------------------------------------
library(janitor)
library(kableExtra)
library(knitr)
library(osfr)
library(tidyverse)

# options ----------------------------------------------------------------------
set.seed(123) # Seed for random number generation
options(scipen = 999) # disable sci number format
opts_chunk$set(
  cache.extra = rand_seed, 
  cache = TRUE,
  message = FALSE, 
  warning = FALSE, 
  error = FALSE,
  echo = FALSE, 
  fig.retina = 3
  )
# variables --------------------------------------------------------------------
## keep only closure waves longer than `days_min_closure` but display only the first `days_after_closure`
days_min_closure <- 21
days_after_closure <- 28
## type of school closure to investigate
period_closure_to_keep <- c(
  #"Partially open",
  "Academic break",
  "Closed due to COVID-19"
  )

```

```{r data, include=FALSE, cache=TRUE}
osf_retrieve_file("7tnfh") |>
  osf_download(conflicts = "overwrite", progress = TRUE, verbose = TRUE)

df <- "Output_5.zip" |>
  read_csv(skip = 3, col_types = "ccccciiddd") |>
  mutate(Date = as.Date(Date, format = "%d.%m.%Y")) |>
  clean_names()

file.remove("Output_5.zip")

df_school_closure <- 
  read_csv(
    "https://en.unesco.org/sites/default/files/covid_impact_education_full.csv",
    col_types = list(.default = col_character(), Date = col_date(format = "%d/%m/%Y"))) |> 
  clean_names() |>
  select(date, country, status) |> 
  mutate(country = case_when(
    country == "Bolivia (Plurinational State of)" ~ "Bolivia",  
    country == "Central African republic" ~ "Central African Republic",
    country == "Iran (Islamic Republic of)" ~ "Iran",
    country == "Republic of Moldova" ~ "Moldova",
    country == "Republic of Korea" ~ "South Korea",
    country == "United Kingdom of Great Britain and Northern Ireland" ~ "United Kingdom",
    country == "United States of America" ~ "USA",
    country == "Viet Nam" ~ "Vietnam",
    TRUE ~ country
    )
  )

df_check <- 
  read_csv("https://covid19.who.int/WHO-COVID-19-global-data.csv") |> 
  clean_names() |> 
  rename(date = date_reported) |> 
  mutate(country = case_when(
    country == "Bolivia (Plurinational State of)" ~ "Bolivia",  
    country == "Iran (Islamic Republic of)" ~ "Iran",
    country == "occupied Palestinian territory, including east Jerusalem" ~ "Palestine",
    country == "Republic of Moldova" ~ "Moldova",
    country == "Republic of Korea" ~ "South Korea",
    country == "United States of America" ~ "USA",
    country == "Venezuela (Bolivarian Republic of)" ~ "Venezuela",
    country == "Viet Nam" ~ "Vietnam",
    TRUE ~ country
    )
  )
```

```{r transformed-data}
df_global <- df |> 
  filter(
    date > as.Date("2019-12-31"),
    region == "All",
    sex == "b"
  )
df_global_cases <- df_global |> 
  drop_na(cases)

df_m_and_sd <- df_global_cases |> 
  count(country, age) |> 
  summarise(avg = round(mean(n), 0), sd = round(sd(n), 0))

df_global_daily <- df_global_cases |> 
  group_by(country, date) |> 
  summarise(cumulative_cases_cal = sum(cases)) |> 
  group_by(country) |> 
  complete(date = full_seq(date, 1)) |> 
  mutate(
    timeline = date - lag(date),  # check if gap between dates
    daily_cases = round(cumulative_cases_cal - lag(cumulative_cases_cal), 0)
  ) |> 
  ungroup()

df_global_final <- df_global_daily |> 
  left_join(df_school_closure, by = c("date", "country")) |> 
  group_by(country) |> 
  fill(status) |> 
  ungroup() |> 
  mutate(schools_closed = if_else(status %in% period_closure_to_keep, 1, 0)) |> 
  drop_na(status) |> 
  left_join(df_check, by = c("date", "country"))

df_global_age <- df |> 
  filter(
    date > as.Date("2020-03-01"),
    region == "All",
    sex == "b"
  ) |> 
  mutate(
    age_grp = case_when(
      age == 0  ~ "aged0to4",
      age == 5  ~ "aged5to9",
      age == 10 ~ "aged10to14",
      age == 15 ~ "aged15to19",
      age == 20 ~ "aged20to24",
      between(age, 25, 34) ~ "aged25to34",
      between(age, 35, 44) ~ "aged35to44",
      between(age, 45, 54) ~ "aged45to54",
      between(age, 55, 64) ~ "aged55to64",
      age > 64 ~ "aged64+"
    )
  ) |> 
  group_by(country, date, age_grp) |> 
  summarise(cumulative_cases_cal = sum(cases)) |> 
  group_by(country, age_grp) |> 
  complete(date = full_seq(date, 1)) |> 
  mutate(
    timeline = date - lag(date),  # check if gap between dates
    daily_cases = round(cumulative_cases_cal - lag(cumulative_cases_cal), 0)
  ) |> 
  ungroup()

df_school_closure_wave <- df_global_final |> 
  select(date, country, schools_closed) |> 
  group_by(country) |> 
  mutate(schools_closure_day = schools_closed - lag(schools_closed)) |> 
  filter(schools_closure_day == 1) |> 
  mutate(schools_closure_wave = row_number())

df_global_school <- df_school_closure_wave |> 
  full_join(df_global_final, by = c("date", "country", "schools_closed")) |> 
  group_by(country) |> 
  arrange(date) |> 
  fill(schools_closure_wave, .direction = "down") |> 
  drop_na(schools_closure_wave) |> 
  group_by(country, schools_closure_wave) |> 
  mutate(schools_time_since_closure = row_number()) |> 
  ungroup()

df_global_school_age <- df_global_school |> 
  select(date, country, who_region, starts_with("schools_")) |> 
  full_join(df_global_age, by = c("date", "country"))

df_global_school_length <- df_global_school |> 
  filter(schools_closed == 1) |> 
  count(country, schools_closure_wave) |> 
  rename(schools_closure_wave_n = n)

negative_daily_case_wave <- df_global_school_age |> 
  filter(daily_cases < 0) |> 
  distinct(country, schools_closure_wave)

df_global_school_age_long <- df_global_school_length |> 
  anti_join(negative_daily_case_wave, by = c("country", "schools_closure_wave")) |> 
  right_join(df_global_school_age, by = c("country", "schools_closure_wave")) |> 
  filter(
    schools_closure_wave_n >= days_min_closure,
    schools_time_since_closure <= days_after_closure,
    age_grp %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")
  )

missing_daily_case_wave <- df_global_school_age_long |> 
  group_by(country, age_grp, schools_closure_wave) |> 
  summarise(n_missing = sum(is.na(daily_cases))) |> 
  filter(n_missing != 0) |> 
  distinct(country, schools_closure_wave)

df_global_school_age_long_clean <- df_global_school_age_long |> 
  anti_join(missing_daily_case_wave, by = c("country", "schools_closure_wave"))
```

## The COVerAGE-DB Project

The COVerAGE-DB project (https://osf.io/mpwjq/) contains 3 data files:

- the "Input" data file which consist in the accumulation of all the raw data communicated by governments
- the "Output_5" data file which is a projection of Covid-19 cases by group of 5 years
- the "Output_10" data file which is a projection of Covid-19 cases by group of 10 years

Given the necessity to distinguish finely between the age groups, the use of the "Output_5" data file appears as the best option. Indeed, contrary to the "Input" data file in which data can be given in different age brackets according to each countries' reporting standard, the "Output_5" data file provides an homogeneous analysis of Covid-19 cases by age brackets of 5 years using spline approximations when the data for this age bracket is not available for a country.

## The "Output_5" Dataset

The original data consist in `r nrow(df)` observations of 10 variables (`r nrow(distinct(df,country))` distinct *country*, *region* within the country, an unique observation *code*, the *date* of the observation, the *gender* which can be male, female or both, the *age* bracket by 5 years from 0 to 100, a confirmation of the *age interval* for each bracket, the total number of *cases* so far, the total number of *deaths* and the total number of *tests* performed).

However, all the data are not available for each observation. For instance, out of the `r nrow(df)` observations, `r sum(is.na(df$cases))` are missing from the *cases* variable, `r sum(is.na(df$deaths))` are missing from the *deaths* variable, and `r sum(is.na(df$tests))` are missing from the *tests* variable.

After having kept observation for whole countries, including both male and female cases and removing all artifact observations recorded before December 31st, 2019, date of the first official case reported in Wuhan, China, the data from `r nrow(distinct(df_global,country))` are then available.

Among the covid numbers reported, only the total number of cases is relevant for our analysis. However, after removing the missing values in the *cases* variable `r df_global |> drop_na(cases) |> pull(country) |> n_distinct()` countries are remaining.

## Disparities Between Countries

Even if each of the remaining `r df_global |> drop_na(cases) |> pull(country) |> n_distinct()` countries have exploitable *cases* number reported, the data file is not homogeneous in term of the period covered and regularity in reported data (Figure \@ref(fig:data-collected)).

```{r data-collected, fig.height=8, fig.width=5, fig.cap="(ref:data-collected)"}
df_global_cases |> 
  ggplot(aes(date, country)) +
  geom_point(size = 0.5) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b %y") +
  labs(x = "Date", y = "") +
  theme_bw() +
  theme(text = element_text(size = 10, family = "serif"))
```

Regarding the period covered by the data collected, it appears that some countries have started their data reporting very late in the pandemic and other have ended their reporting very early. In average, each country have a total of `r df_m_and_sd[["avg"]]` days reported with a standard deviation of `r df_m_and_sd[["sd"]]`. However the most important problem is the continuity in the reporting of data. Some countries are missing days even weeks without any regularity that would explain this phenomenon. As a result, many more countries will be doped out from the pre-processing.

## Changes in Covid-19 Cases After School Closures

Despite the data of the remaining `r df_global |> drop_na(cases) |> pull(country) |> n_distinct()` countries are not homogeneous, none of them has been dropped so far for this reason. However, the following steps in the data pre-processing will lead to significantly reduce the amount of data analysed.

First, the data reported are the total number of Covid-19 cases from the start to the pandemic to a specific date, also called total cases, but the analysis needs to be done using the daily number of cases at a specific date. The *daily number of cases* at a specific date $n_x$ is calculated with the difference between the total cases at a date $t_x$ and the total cases at a date $t_{x-1}$ (i.e., lag 1). If $t_x$ or $t_{x-1}$ is missing then $n_x$ will be missing as well. This brings the amount of countries to `r df_global_daily |> drop_na(daily_cases) |> pull(country) |> n_distinct()`.

Second, our aim is to evaluate the change in daily cases after school closure. Therefore, countries time series are cut in multiple Target Period from the first day of the school closure to 28 days (4 weeks) after the closure whether the schools have reopen or not. However, to be taken into account, a Target Period should include a closure for at least 21 days after the closure. As a result, `r n_distinct(df_global_school_age_long$country)` countries have at least one of the research Target Period.

Finally, due to the non-linear nature of spline approximation to calculate the daily cases, the difference between $t_x$ and $t_{x-1}$ can lead to negative values. However, because negative daily cases are not plausible, if a target period contains a negative value then the entire period is removed. In addition, ensure the quality of this analysis, Target Periods with missing values are also removed.

As a result, the exploitable final number of countries is `r n_distinct(df_global_school_age_long_clean$country)`, each having a different amount of target period (Table \@ref(tab:final-list)).

```{r final-list}
df_global_school_age_long_clean |> 
  distinct(country, schools_closure_wave) |> 
  group_by(country) |> 
  slice_max(schools_closure_wave) |> 
  rename(Countries = country, `Nb of Target Periods` = schools_closure_wave) |> 
  kable(
    caption = "(ref:final-list)", 
    booktabs = TRUE,
    linesep = ""
  ) |> 
  kable_styling(
    font_size = 8,
    latex_options = c("hold_position")
  )
```

### 2022-02-10 analysis/exploration_global_input.R

@@ -1,311 +1,0 @@
-
-# options ----------------------------------------------------------------------
-options(scipen = 999)
-future::plan(multisession)
-
-# libraries --------------------------------------------------------------------
-library(janitor)
-library(mgcv)
-library(osfr)
-library(tidyverse)
-
-# raw data ---------------------------------------------------------------------
-  
-  osf_retrieve_file("9dsfk") |> 
-  osf_download(conflicts = "overwrite", progress = TRUE, verbose = TRUE)
-
-df <- 
-  read_csv(
-    "inputDB.zip", 
-    skip = 1, col_types = "cccccciccdc"
-  ) |> 
-  mutate(Date = as.Date(Date, format = "%d.%m.%Y")) |> 
-  clean_names()
-
-file.remove("inputDB.zip")
-
-df_school_closure <- 
-  read_csv(
-    "https://en.unesco.org/sites/default/files/covid_impact_education_full.csv",
-    col_types = list(.default = col_character(), Date = col_date(format = "%d/%m/%Y"))) |> 
-  clean_names() |>
-  select(date, country, status) |> 
-  mutate(country = case_when(
-    country == "Bolivia (Plurinational State of)" ~ "Bolivia",  
-    country == "Central African republic" ~ "Central African Republic",
-    country == "Iran (Islamic Republic of)" ~ "Iran",
-    country == "Republic of Moldova" ~ "Moldova",
-    country == "Republic of Korea" ~ "South Korea",
-    country == "United Kingdom of Great Britain and Northern Ireland" ~ "United Kingdom",
-    country == "United States of America" ~ "USA",
-    country == "Viet Nam" ~ "Vietnam",
-    TRUE ~ country
-    )
-  )
-
-df_check <- 
-  read_csv("https://covid19.who.int/WHO-COVID-19-global-data.csv") |> 
-  clean_names() |> 
-  rename(date = date_reported) |> 
-  mutate(country = case_when(
-    country == "Bolivia (Plurinational State of)" ~ "Bolivia",  
-    country == "Iran (Islamic Republic of)" ~ "Iran",
-    country == "occupied Palestinian territory, including east Jerusalem" ~ "Palestine",
-    country == "Republic of Moldova" ~ "Moldova",
-    country == "Republic of Korea" ~ "South Korea",
-    country == "United States of America" ~ "USA",
[…]

### 2022-02-10 analysis/exploration_global_output_5_breaks+covid.R

@@ -1,240 +1,0 @@
-
-# options ----------------------------------------------------------------------
-options(scipen = 999)
-future::plan("future::multisession")
-
-# libraries --------------------------------------------------------------------
-library(janitor)
-library(mgcv)
-library(osfr)
-library(tidyverse)
-
-# raw data ---------------------------------------------------------------------
-osf_retrieve_file("7tnfh") |>
-  osf_download(conflicts = "overwrite", progress = TRUE, verbose = TRUE)
-
-df <- "Output_5.zip" |>
-  read_csv(skip = 3, col_types = "ccccciiddd") |>
-  mutate(Date = as.Date(Date, format = "%d.%m.%Y")) |>
-  clean_names()
-
-file.remove("Output_5.zip")
-
-df_school_closure <- 
-  read_csv(
-    "https://en.unesco.org/sites/default/files/covid_impact_education_full.csv",
-    col_types = list(.default = col_character(), Date = col_date(format = "%d/%m/%Y"))) |> 
-  clean_names() |>
-  select(date, country, status) |> 
-  mutate(country = case_when(
-    country == "Bolivia (Plurinational State of)" ~ "Bolivia",  
-    country == "Central African republic" ~ "Central African Republic",
-    country == "Iran (Islamic Republic of)" ~ "Iran",
-    country == "Republic of Moldova" ~ "Moldova",
-    country == "Republic of Korea" ~ "South Korea",
-    country == "United Kingdom of Great Britain and Northern Ireland" ~ "United Kingdom",
-    country == "United States of America" ~ "USA",
-    country == "Viet Nam" ~ "Vietnam",
-    TRUE ~ country
-    )
-  )
-
-df_check <- 
-  read_csv("https://covid19.who.int/WHO-COVID-19-global-data.csv") |> 
-  clean_names() |> 
-  rename(date = date_reported) |> 
-  mutate(country = case_when(
-    country == "Bolivia (Plurinational State of)" ~ "Bolivia",  
-    country == "Iran (Islamic Republic of)" ~ "Iran",
-    country == "occupied Palestinian territory, including east Jerusalem" ~ "Palestine",
-    country == "Republic of Moldova" ~ "Moldova",
-    country == "Republic of Korea" ~ "South Korea",
-    country == "United States of America" ~ "USA",
-    country == "Venezuela (Bolivarian Republic of)" ~ "Venezuela",
[…]

### 2022-02-10 analysis/exploration_ireland.R

@@ -1,387 +1,0 @@
-
-# libraries --------------------------------------------------------------------
-library(here)
-library(httr2)
-library(gratia)
-library(janitor)
-library(magrittr)
-library(mgcv)
-library(tidyverse)
-library(zoo)
-
-# api access -------------------------------------------------------------------
-req <- request("https://services1.arcgis.com/eNO7HHeQ3rUcBllm/arcgis/rest/services/CovidStatisticsProfileHPSCIrelandOpenData/FeatureServer/0/query?where=1%3D1&outFields=*&returnGeometry=false&outSR=4326&f=json") |> 
-  req_headers("Accept" = "application/json")
-resp <- req_perform(req) |> 
-  resp_body_json() |> 
-  use_series(features) |> 
-  flatten_dfr() |> 
-  clean_names() |> 
-  mutate(date = as.Date(as.POSIXct(date/1000, origin = "1970-01-01"))) |> 
-  select(c(date, starts_with("aged")))
-
-# variables --------------------------------------------------------------------
-age_order <- c("aged1to4", "aged5to14", "aged15to24", "aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65to74", "aged75to84", "aged85up")
-age_name  <- c("1 to 4",   "5 to 14",   "15 to 24",   "25 to 34",   "35 to 44",   "45 to 54",   "55 to 64",   "65 to 74",   "75 to 84",   "85 and up")
-age_color <- c("#999999",  "#E69F00",   "#56B4E9",    "#009E73",    "#F0E442",    "#0072B2",    "#D55E00",    "#CC79A7",    "#000000",    "#CAB2D6")
-
-# data -------------------------------------------------------------------------
-df_school_closure <- read_csv(
-  "https://en.unesco.org/sites/default/files/covid_impact_education.csv", 
-  col_types = list(.default = col_character(), Date = col_date(format = "%d/%m/%Y"))) |> 
-  clean_names() |>
-  select(date, country, status) |> 
-  filter(country == "Ireland")
-
-df <- 
-  read_csv(
-    here("data/age_groups.csv"), 
-    col_types = list(
-      .default = col_double(), 
-      Date = col_date(format = "%d/%m/%Y")
-    )
-  ) |> 
-  clean_names() |> 
-  rename(total_cases = total) |> 
-  arrange(date) |> 
-  mutate(
[…]

### 2022-02-10 analysis/exploration_school_closure.R

@@ -1,60 +1,0 @@
-
-# libraries --------------------------------------------------------------------
-library(janitor)
-library(tidyverse)
-
-# variables --------------------------------------------------------------------
-period_closure_to_keep <- c(
-  #"Partially open",
-  #"Academic break",
-  "Closed due to COVID-19"
-)
-
-# data -------------------------------------------------------------------------
-covid_impact_education <- 
-  read_csv(
-    "https://en.unesco.org/sites/default/files/covid_impact_education.csv",
-    col_types = list(.default = col_character(), Date = col_date(format = "%d/%m/%Y"))) |> 
-  clean_names() |>
-  select(date, country, status) |> 
-  mutate(country = case_when(
-    country == "Bolivia (Plurinational State of)" ~ "Bolivia",  
-    country == "Central African republic" ~ "Central African Republic",
-    country == "Iran (Islamic Republic of)" ~ "Iran",
-    country == "Republic of Moldova" ~ "Moldova",
-    country == "Republic of Korea" ~ "South Korea",
-    country == "United Kingdom of Great Britain and Northern Ireland" ~ "United Kingdom",
-    country == "United States of America" ~ "USA",
-    country == "Viet Nam" ~ "Vietnam",
-    TRUE ~ country
-  )
-  )
-
-covid_impact_education_full <- read_csv(
-  "https://en.unesco.org/sites/default/files/covid_impact_education_full.csv",
-  col_types = list(.default = col_character(), Date = col_date(format = "%d/%m/%Y"))) |> 
-  clean_names() |>
-  select(date, country, status_test = status) |> 
-  mutate(country = case_when(
-    country == "Bolivia (Plurinational State of)" ~ "Bolivia",  
-    country == "Central African republic" ~ "Central African Republic",
-    country == "Iran (Islamic Republic of)" ~ "Iran",
-    country == "Republic of Moldova" ~ "Moldova",
-    country == "Republic of Korea" ~ "South Korea",
-    country == "United Kingdom of Great Britain and Northern Ireland" ~ "United Kingdom",
-    country == "United States of America" ~ "USA",
-    country == "Viet Nam" ~ "Vietnam",
-    TRUE ~ country
-  )
-  )
-
-# analysis ---------------------------------------------------------------------
-## for some reasons covid_impact_education is shorter than covid_impact_education_full
-comparison_closure <- 
-  full_join(
-    covid_impact_education, 
[…]

### 2022-02-10 analysis/gam_analysis.R

@@ -1,53 +1,0 @@
-# model ------------------------------------------------------------------------
-gam_model <- 
-  gam(
-    covid_cases ~ age_group +
-      s(schools_time_since_closure, by = age_group) +
-      s(schools_time_since_closure, schools_closure_wave, bs = "fs"),
-    data = df_school_long,
-    family = nb(), 
-    method = "REML"
-  ); summary(gam_model)
-
-# statistics -------------------------------------------------------------------
-gam_summary <- summary(gam_model)
-
-gam_results <- 
-  broom::tidy(gam_model) |> 
-  mutate(
-    term = c("s(date):1to4", "s(date):5to14", "s(date):15to24", "s(date*wave"),
-    p.value = scales::pvalue(p.value, accuracy = 0.001, add_p = FALSE, prefix = c("< ", "= ", "> ")),
-    across(where(is.numeric), round, 2),
-    print = glue::glue("$\\chi^2({edf}) = {statistic}$, $p {p.value}$")
-  ) |> 
-  select(term, print) |> 
-  deframe()
-
-#plot(modelbased::estimate_relation(gam_model, length = 100, preserve_range = FALSE))
-# visualisation ----------------------------------------------------------------
-gam_plot <- 
-  confint(gam_model, "s(schools_time_since_closure):age_groupaged15to24") |> 
-  mutate(age_group = factor(age_group, levels = c("aged1to4", "aged5to14", "aged15to24"))) |> 
-  ggplot(aes(x = schools_time_since_closure)) +
-  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.25) + 
-  geom_line(aes(y = est)) +
-  facet_wrap(
-    ~ age_group,
-    labeller = as_labeller(
-      c(
-        aged1to4 = "1 to 4",
-        aged5to14 = "5 to 14",
-        aged15to24 = "15 to 24"
-      )
-    )) +
-  labs(
-    x = "Days since School Closure", 
-    y = "Standardized Effect"
-  ) +
-  theme_bw() +
-  theme(
-    text = element_text(size = 10, family = "serif"),
-    axis.text = element_text(size = 10),
-    strip.background = element_blank(),
-    panel.background = element_rect(colour = "black", size = 0.1)
-  )

### 2022-02-10 analysis/gam_analysis_europe.R

@@ -1,0 +1,151 @@
+# libraries --------------------------------------------------------------------
+library(broom)
+library(glue)
+library(gratia)
+library(here)
+library(scales)
+library(tidyverse)
+
+# # libraries --------------------------------------------------------------------
+# library(broom)
+# library(future)
+# library(glue)
+# library(gratia)
+# library(here)
+# library(mgcv)
+# library(parallel)
+# library(scales)
+# library(tidyverse)
+# 
+# # options ----------------------------------------------------------------------
+# options(scipen = 999)
+# plan("future::multisession")
+# 
+# # data -------------------------------------------------------------------------
+# df_school_closure_wave <- df_global |> 
+#   select(date, country, schools_closed) |> 
+#   group_by(country) |> 
+#   mutate(schools_closure_day = schools_closed - lag(schools_closed)) |> 
+#   filter(schools_closure_day == 1) |> 
+#   mutate(schools_closure_wave = row_number())
+# 
+# df_global_school <- df_school_closure_wave |> 
+#   full_join(df_global, by = c("date", "country", "schools_closed")) |> 
+#   group_by(country) |> 
+#   arrange(date) |> 
+#   fill(schools_closure_wave, .direction = "down") |> 
+#   drop_na(schools_closure_wave) |> 
+#   group_by(country, schools_closure_wave) |> 
+#   mutate(schools_time_since_closure = row_number()) |> 
+#   ungroup()
+# 
+# df_global_school_age <- df_global_school |> 
+#   select(date, country, schools_closed, schools_closure_day, schools_closure_wave, schools_time_since_closure, who_region) |> 
+#   full_join(df_global_age, by = c("date", "country"))
+# 
+# df_global_school_length <- df_global_school |> 
+#   filter(schools_closed == 1) |> 
+#   count(country, schools_closure_wave) |> 
+#   rename(schools_closure_wave_n = n)
+# 
+# df_global_school_age_long <- df_global_school_length |> 
+#   right_join(df_global_school_age, by = c("country", "schools_closure_wave")) |> 
+#   filter(
+#     schools_closure_wave_n >= days_min_closure,
+#     schools_time_since_closure <= days_after_closure,
[…]

### 2022-02-10 analysis/te_analysis.R

@@ -1,96 +1,0 @@
-
-# libraries --------------------------------------------------------------------
-library(gtools)
-library(tidyverse)
-library(RTransferEntropy)
-
-# variables --------------------------------------------------------------------
-age_order <- c("aged1to4", "aged5to14", "aged15to24", "aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65to74", "aged75to84", "aged85up")
-
-# solution 1 -------------------------------------------------------------------
-# te_agegroup <- function(var_x, var_y){
-#   
-#   var_x <- select(df_wide_changes, all_of(var_x))
-#   var_y <- select(df_wide_changes, all_of(var_y))
-#   
-#   calc_te(var_x, var_y)
-# }
-# 
-# te_results <-
-#   expand_grid(
-#     X = age_order, 
-#     Y = age_order
-#   ) |> 
-#   as_tibble() |> 
-#   group_by(X, Y) |> 
-#   summarise(te = te_agegroup(X, Y))
-
-# te_matrix <- te_results |> 
-#   mutate(
-#     X = factor(X, levels = age_order),
-#     Y = factor(Y, levels = age_order),
-#   ) |> 
-#   arrange(X, Y) |> 
-#   pivot_wider(names_from = Y, values_from = te) |> 
-#   column_to_rownames("X") |> 
-#   as_tibble()
-# 
-# ggplot(te_results, aes(X, Y)) + 
-#   geom_raster(aes(fill=te)) + 
-#   scale_fill_gradient(low="grey90", high="red") +
-#   theme_bw()
-
-# solution 2 -------------------------------------------------------------------
-library(future) # enable parallel processing for all future transfer_entropy calls
-plan(multisession)
-
-te_agegroup <- function(var_x, var_y){
-  
-  var_x <- select(df_wide_changes, all_of(var_x))
-  var_y <- select(df_wide_changes, all_of(var_y))
-  
-  transfer_entropy(var_x, var_y, quiet = TRUE)[["coef"]] |> 
-    as_tibble() |> 
-    mutate(direction = c("X->Y", "Y->X"))
-}
-
-te_results <-
-  combinations(
-    n = length(age_order), 
-    r = 2, 
-    v = age_order, 
-    repeats.allowed = FALSE
-  ) |> 
-  as_tibble(.name_repair = ~ c("X", "Y")) |> 
-  group_by(X, Y) |> 
[…]

### 2022-02-10 analysis/te_analysis_europe.R

@@ -1,0 +1,98 @@
+# libraries --------------------------------------------------------------------
+library(here)
+library(tidyverse)
+
+# # variables --------------------------------------------------------------------
+age_order <- c(
+  "aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24",
+  "aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+"
+)
+age_name <- c(
+  "0 to 4", "5 to 9", "10 to 14", "15 to 19", "20 to 24", "25 to 34",
+  "35 to 44",   "45 to 54",   "55 to 64",   "65 and up"
+)
+
+countries_of_interest <- c(
+  "Austria", "Belgium", "Bulgaria", "Croatia", "Estonia", "France",
+  "Germany", "Greece", "Netherlands", "Portugal", "Slovakia", "Spain"
+)
+
+# # libraries --------------------------------------------------------------------
+# library(future)
+# library(here)
+# library(tidyverse)
+# library(RTransferEntropy)
+# 
+# # options ----------------------------------------------------------------------
+# options(scipen = 999)
+# plan("future::multisession")
+# 
+# # data -------------------------------------------------------------------------
+# df_wide_changes <- df_global_age |>
+#   select(date, country, age_grp, daily_changes) |>
+#   pivot_wider(names_from = age_grp, values_from = daily_changes)
+# 
+# # transfer entropy -------------------------------------------------------------
+# te_agegroup <- function(var_x, var_y, country_select){
+#   df <- df_wide_changes |>
+#     filter(country == country_select)
+#   var_x <- select(df, all_of(var_x))
+#   var_y <- select(df, all_of(var_y))
+#   transfer_entropy(var_x, var_y, quiet = TRUE)[["coef"]] |>
+#     as_tibble() |>
+#     mutate(direction = c("X->Y", "Y->X"))
+# }
+# 
+# te_results <-
+#   combinations(
+#     n = length(age_order),
+#     r = 2,
+#     v = age_order,
+#     repeats.allowed = FALSE
+#   ) |>
+#   as_tibble(.name_repair = ~ c("X", "Y")) |>
+#   expand_grid(country = countries_of_interest) |>
+#   group_by(X, Y, country) |>
+#   summarise(te_agegroup(X, Y, country)) |>
+#   ungroup()
+# 
[…]

### 2022-02-10 exploration/exploration_england.R

@@ -1,0 +1,44 @@
+
+# libraries --------------------------------------------------------------------
+library(janitor)
+library(tidyverse)
+
+# options ----------------------------------------------------------------------
+options(scipen = 999)
+
+# data -------------------------------------------------------------------------
+df <- read_csv("https://api.coronavirus.data.gov.uk/v2/data?areaType=nation&areaCode=E92000001&metric=newCasesBySpecimenDateAgeDemographics&format=csv")
+
+df_clean <- df |> 
+  clean_names() |> 
+  select(date, age, cases)
+
+# visual exploration -----------------------------------------------------------
+df |> 
+  group_by(date) |> 
+  summarise(total_cases = sum(cases)) |> 
+  ggplot(aes(total_cases)) +
+  geom_density(na.rm = TRUE) +
+  labs(
+    title = "Distribution of COVID-19 Daily Cases in Ireland",
+    subtitle = "",
+    caption = "Source: Irish Department of Health (last update 2022-01-01)",
+    x = "Covid Cases", 
+    y = "Density"
+  ) +
+  theme_bw()
+
+df |> 
+  group_by(date) |> 
+  summarise(total_cases = sum(cases)) |> 
+  ggplot(aes(date, total_cases)) +
+  geom_line(na.rm = TRUE) +
+  labs(
+    title = "Evolution of COVID-19 Cases in England",
+    subtitle = "Raw data",
+    caption = "Source: UK Department of Health",
+    x = "Date", 
+    y = "Covid Cases"
+  ) +
+  theme_bw() +
+  theme(legend.position = "bottom")

### 2022-02-10 exploration/exploration_global_input.R

@@ -1,0 +1,311 @@
+
+# options ----------------------------------------------------------------------
+options(scipen = 999)
+future::plan(multisession)
+
+# libraries --------------------------------------------------------------------
+library(janitor)
+library(mgcv)
+library(osfr)
+library(tidyverse)
+
+# raw data ---------------------------------------------------------------------
+  
+  osf_retrieve_file("9dsfk") |> 
+  osf_download(conflicts = "overwrite", progress = TRUE, verbose = TRUE)
+
+df <- 
+  read_csv(
+    "inputDB.zip", 
+    skip = 1, col_types = "cccccciccdc"
+  ) |> 
+  mutate(Date = as.Date(Date, format = "%d.%m.%Y")) |> 
+  clean_names()
+
+file.remove("inputDB.zip")
+
+df_school_closure <- 
+  read_csv(
+    "https://en.unesco.org/sites/default/files/covid_impact_education_full.csv",
+    col_types = list(.default = col_character(), Date = col_date(format = "%d/%m/%Y"))) |> 
+  clean_names() |>
+  select(date, country, status) |> 
+  mutate(country = case_when(
+    country == "Bolivia (Plurinational State of)" ~ "Bolivia",  
+    country == "Central African republic" ~ "Central African Republic",
+    country == "Iran (Islamic Republic of)" ~ "Iran",
+    country == "Republic of Moldova" ~ "Moldova",
+    country == "Republic of Korea" ~ "South Korea",
+    country == "United Kingdom of Great Britain and Northern Ireland" ~ "United Kingdom",
+    country == "United States of America" ~ "USA",
+    country == "Viet Nam" ~ "Vietnam",
+    TRUE ~ country
+    )
+  )
+
+df_check <- 
+  read_csv("https://covid19.who.int/WHO-COVID-19-global-data.csv") |> 
+  clean_names() |> 
+  rename(date = date_reported) |> 
+  mutate(country = case_when(
+    country == "Bolivia (Plurinational State of)" ~ "Bolivia",  
+    country == "Iran (Islamic Republic of)" ~ "Iran",
+    country == "occupied Palestinian territory, including east Jerusalem" ~ "Palestine",
+    country == "Republic of Moldova" ~ "Moldova",
+    country == "Republic of Korea" ~ "South Korea",
+    country == "United States of America" ~ "USA",
[…]

### 2022-02-10 exploration/exploration_global_output_5.R

@@ -1,0 +1,306 @@
+
+# options ----------------------------------------------------------------------
+options(scipen = 999)
+future::plan(multisession)
+
+# libraries --------------------------------------------------------------------
+library(janitor)
+library(mgcv)
+library(osfr)
+library(tidyverse)
+
+# raw data ---------------------------------------------------------------------
+osf_retrieve_file("7tnfh") |>
+  osf_download(conflicts = "overwrite", progress = TRUE, verbose = TRUE)
+
+df <- "Output_5.zip" |>
+  read_csv(skip = 3, col_types = "ccccciiddd") |>
+  mutate(Date = as.Date(Date, format = "%d.%m.%Y")) |>
+  clean_names()
+
+file.remove("Output_5.zip")
+
+df_school_closure <- 
+  read_csv(
+    "https://en.unesco.org/sites/default/files/covid_impact_education_full.csv",
+    col_types = list(.default = col_character(), Date = col_date(format = "%d/%m/%Y"))) |> 
+  clean_names() |>
+  select(date, country, status) |> 
+  mutate(country = case_when(
+    country == "Bolivia (Plurinational State of)" ~ "Bolivia",  
+    country == "Central African republic" ~ "Central African Republic",
+    country == "Iran (Islamic Republic of)" ~ "Iran",
+    country == "Republic of Moldova" ~ "Moldova",
+    country == "Republic of Korea" ~ "South Korea",
+    country == "United Kingdom of Great Britain and Northern Ireland" ~ "United Kingdom",
+    country == "United States of America" ~ "USA",
+    country == "Viet Nam" ~ "Vietnam",
+    TRUE ~ country
+    )
+  )
+
+df_check <- 
+  read_csv("https://covid19.who.int/WHO-COVID-19-global-data.csv") |> 
+  clean_names() |> 
+  rename(date = date_reported) |> 
+  mutate(country = case_when(
+    country == "Bolivia (Plurinational State of)" ~ "Bolivia",  
+    country == "Iran (Islamic Republic of)" ~ "Iran",
+    country == "occupied Palestinian territory, including east Jerusalem" ~ "Palestine",
+    country == "Republic of Moldova" ~ "Moldova",
+    country == "Republic of Korea" ~ "South Korea",
+    country == "United States of America" ~ "USA",
[…]

### 2022-02-10 exploration/exploration_global_output_5_breaks+covid.R

@@ -1,0 +1,248 @@
+
+# options ----------------------------------------------------------------------
+options(scipen = 999)
+future::plan("future::multisession")
+
+# libraries --------------------------------------------------------------------
+library(janitor)
+library(mgcv)
+library(osfr)
+library(tidyverse)
+
+# raw data ---------------------------------------------------------------------
+osf_retrieve_file("7tnfh") |>
+  osf_download(conflicts = "overwrite", progress = TRUE, verbose = TRUE)
+
+df <- "Output_5.zip" |>
+  read_csv(skip = 3, col_types = "ccccciiddd") |>
+  mutate(Date = as.Date(Date, format = "%d.%m.%Y")) |>
+  clean_names()
+
+file.remove("Output_5.zip")
+
+df_school_closure <- 
+  read_csv(
+    "https://en.unesco.org/sites/default/files/covid_impact_education_full.csv",
+    col_types = list(.default = col_character(), Date = col_date(format = "%d/%m/%Y"))) |> 
+  clean_names() |>
+  select(date, country, status) |> 
+  mutate(country = case_when(
+    country == "Bolivia (Plurinational State of)" ~ "Bolivia",  
+    country == "Central African republic" ~ "Central African Republic",
+    country == "Iran (Islamic Republic of)" ~ "Iran",
+    country == "Republic of Moldova" ~ "Moldova",
+    country == "Republic of Korea" ~ "South Korea",
+    country == "United Kingdom of Great Britain and Northern Ireland" ~ "United Kingdom",
+    country == "United States of America" ~ "USA",
+    country == "Viet Nam" ~ "Vietnam",
+    TRUE ~ country
+    )
+  )
+
+df_check <- 
+  read_csv("https://covid19.who.int/WHO-COVID-19-global-data.csv") |> 
+  clean_names() |> 
+  rename(date = date_reported) |> 
+  mutate(country = case_when(
+    country == "Bolivia (Plurinational State of)" ~ "Bolivia",  
+    country == "Iran (Islamic Republic of)" ~ "Iran",
+    country == "occupied Palestinian territory, including east Jerusalem" ~ "Palestine",
+    country == "Republic of Moldova" ~ "Moldova",
+    country == "Republic of Korea" ~ "South Korea",
+    country == "United States of America" ~ "USA",
+    country == "Venezuela (Bolivarian Republic of)" ~ "Venezuela",
[…]

### 2022-02-10 exploration/exploration_global_output_5_covid.R

@@ -1,0 +1,226 @@
+
+# options ----------------------------------------------------------------------
+options(scipen = 999)
+future::plan(multisession)
+
+# libraries --------------------------------------------------------------------
+library(janitor)
+library(mgcv)
+library(osfr)
+library(tidyverse)
+
+# raw data ---------------------------------------------------------------------
+osf_retrieve_file("7tnfh") |>
+  osf_download(conflicts = "overwrite", progress = TRUE, verbose = TRUE)
+
+df <- "Output_5.zip" |>
+  read_csv(skip = 3, col_types = "ccccciiddd") |>
+  mutate(Date = as.Date(Date, format = "%d.%m.%Y")) |>
+  clean_names()
+
+file.remove("Output_5.zip")
+
+df_school_closure <- 
+  read_csv(
+    "https://en.unesco.org/sites/default/files/covid_impact_education_full.csv",
+    col_types = list(.default = col_character(), Date = col_date(format = "%d/%m/%Y"))) |> 
+  clean_names() |>
+  select(date, country, status) |> 
+  mutate(country = case_when(
+    country == "Bolivia (Plurinational State of)" ~ "Bolivia",  
+    country == "Central African republic" ~ "Central African Republic",
+    country == "Iran (Islamic Republic of)" ~ "Iran",
+    country == "Republic of Moldova" ~ "Moldova",
+    country == "Republic of Korea" ~ "South Korea",
+    country == "United Kingdom of Great Britain and Northern Ireland" ~ "United Kingdom",
+    country == "United States of America" ~ "USA",
+    country == "Viet Nam" ~ "Vietnam",
+    TRUE ~ country
+    )
+  )
+
+df_check <- 
+  read_csv("https://covid19.who.int/WHO-COVID-19-global-data.csv") |> 
+  clean_names() |> 
+  rename(date = date_reported) |> 
+  mutate(country = case_when(
+    country == "Bolivia (Plurinational State of)" ~ "Bolivia",  
+    country == "Iran (Islamic Republic of)" ~ "Iran",
+    country == "occupied Palestinian territory, including east Jerusalem" ~ "Palestine",
+    country == "Republic of Moldova" ~ "Moldova",
+    country == "Republic of Korea" ~ "South Korea",
+    country == "United States of America" ~ "USA",
+    country == "Venezuela (Bolivarian Republic of)" ~ "Venezuela",
+    country == "Viet Nam" ~ "Vietnam",
[…]

### 2022-02-10 exploration/exploration_ireland.R

@@ -1,0 +1,387 @@
+
+# libraries --------------------------------------------------------------------
+library(here)
+library(httr2)
+library(gratia)
+library(janitor)
+library(magrittr)
+library(mgcv)
+library(tidyverse)
+library(zoo)
+
+# api access -------------------------------------------------------------------
+req <- request("https://services1.arcgis.com/eNO7HHeQ3rUcBllm/arcgis/rest/services/CovidStatisticsProfileHPSCIrelandOpenData/FeatureServer/0/query?where=1%3D1&outFields=*&returnGeometry=false&outSR=4326&f=json") |> 
+  req_headers("Accept" = "application/json")
+resp <- req_perform(req) |> 
+  resp_body_json() |> 
+  use_series(features) |> 
+  flatten_dfr() |> 
+  clean_names() |> 
+  mutate(date = as.Date(as.POSIXct(date/1000, origin = "1970-01-01"))) |> 
+  select(c(date, starts_with("aged")))
+
+# variables --------------------------------------------------------------------
+age_order <- c("aged1to4", "aged5to14", "aged15to24", "aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65to74", "aged75to84", "aged85up")
+age_name  <- c("1 to 4",   "5 to 14",   "15 to 24",   "25 to 34",   "35 to 44",   "45 to 54",   "55 to 64",   "65 to 74",   "75 to 84",   "85 and up")
+age_color <- c("#999999",  "#E69F00",   "#56B4E9",    "#009E73",    "#F0E442",    "#0072B2",    "#D55E00",    "#CC79A7",    "#000000",    "#CAB2D6")
+
+# data -------------------------------------------------------------------------
+df_school_closure <- read_csv(
+  "https://en.unesco.org/sites/default/files/covid_impact_education.csv", 
+  col_types = list(.default = col_character(), Date = col_date(format = "%d/%m/%Y"))) |> 
+  clean_names() |>
+  select(date, country, status) |> 
+  filter(country == "Ireland")
+
+df <- 
+  read_csv(
+    here("data/age_groups.csv"), 
+    col_types = list(
+      .default = col_double(), 
+      Date = col_date(format = "%d/%m/%Y")
+    )
+  ) |> 
+  clean_names() |> 
+  rename(total_cases = total) |> 
+  arrange(date) |> 
+  mutate(
[…]

### 2022-02-10 exploration/exploration_school_closure.R

@@ -1,0 +1,60 @@
+
+# libraries --------------------------------------------------------------------
+library(janitor)
+library(tidyverse)
+
+# variables --------------------------------------------------------------------
+period_closure_to_keep <- c(
+  #"Partially open",
+  #"Academic break",
+  "Closed due to COVID-19"
+)
+
+# data -------------------------------------------------------------------------
+covid_impact_education <- 
+  read_csv(
+    "https://en.unesco.org/sites/default/files/covid_impact_education.csv",
+    col_types = list(.default = col_character(), Date = col_date(format = "%d/%m/%Y"))) |> 
+  clean_names() |>
+  select(date, country, status) |> 
+  mutate(country = case_when(
+    country == "Bolivia (Plurinational State of)" ~ "Bolivia",  
+    country == "Central African republic" ~ "Central African Republic",
+    country == "Iran (Islamic Republic of)" ~ "Iran",
+    country == "Republic of Moldova" ~ "Moldova",
+    country == "Republic of Korea" ~ "South Korea",
+    country == "United Kingdom of Great Britain and Northern Ireland" ~ "United Kingdom",
+    country == "United States of America" ~ "USA",
+    country == "Viet Nam" ~ "Vietnam",
+    TRUE ~ country
+  )
+  )
+
+covid_impact_education_full <- read_csv(
+  "https://en.unesco.org/sites/default/files/covid_impact_education_full.csv",
+  col_types = list(.default = col_character(), Date = col_date(format = "%d/%m/%Y"))) |> 
+  clean_names() |>
+  select(date, country, status_test = status) |> 
+  mutate(country = case_when(
+    country == "Bolivia (Plurinational State of)" ~ "Bolivia",  
+    country == "Central African republic" ~ "Central African Republic",
+    country == "Iran (Islamic Republic of)" ~ "Iran",
+    country == "Republic of Moldova" ~ "Moldova",
+    country == "Republic of Korea" ~ "South Korea",
+    country == "United Kingdom of Great Britain and Northern Ireland" ~ "United Kingdom",
+    country == "United States of America" ~ "USA",
+    country == "Viet Nam" ~ "Vietnam",
+    TRUE ~ country
+  )
+  )
+
+# analysis ---------------------------------------------------------------------
+## for some reasons covid_impact_education is shorter than covid_impact_education_full
+comparison_closure <- 
+  full_join(
+    covid_impact_education, 
[…]

### 2022-02-10 exploration/gam_analysis.R

@@ -1,0 +1,53 @@
+# model ------------------------------------------------------------------------
+gam_model <- 
+  gam(
+    covid_cases ~ age_group +
+      s(schools_time_since_closure, by = age_group) +
+      s(schools_time_since_closure, schools_closure_wave, bs = "fs"),
+    data = df_school_long,
+    family = nb(), 
+    method = "REML"
+  ); summary(gam_model)
+
+# statistics -------------------------------------------------------------------
+gam_summary <- summary(gam_model)
+
+gam_results <- 
+  broom::tidy(gam_model) |> 
+  mutate(
+    term = c("s(date):1to4", "s(date):5to14", "s(date):15to24", "s(date*wave"),
+    p.value = scales::pvalue(p.value, accuracy = 0.001, add_p = FALSE, prefix = c("< ", "= ", "> ")),
+    across(where(is.numeric), round, 2),
+    print = glue::glue("$\\chi^2({edf}) = {statistic}$, $p {p.value}$")
+  ) |> 
+  select(term, print) |> 
+  deframe()
+
+#plot(modelbased::estimate_relation(gam_model, length = 100, preserve_range = FALSE))
+# visualisation ----------------------------------------------------------------
+gam_plot <- 
+  confint(gam_model, "s(schools_time_since_closure):age_groupaged15to24") |> 
+  mutate(age_group = factor(age_group, levels = c("aged1to4", "aged5to14", "aged15to24"))) |> 
+  ggplot(aes(x = schools_time_since_closure)) +
+  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.25) + 
+  geom_line(aes(y = est)) +
+  facet_wrap(
+    ~ age_group,
+    labeller = as_labeller(
+      c(
+        aged1to4 = "1 to 4",
+        aged5to14 = "5 to 14",
+        aged15to24 = "15 to 24"
+      )
+    )) +
+  labs(
+    x = "Days since School Closure", 
+    y = "Standardized Effect"
+  ) +
+  theme_bw() +
+  theme(
+    text = element_text(size = 10, family = "serif"),
+    axis.text = element_text(size = 10),
+    strip.background = element_blank(),
+    panel.background = element_rect(colour = "black", size = 0.1)
+  )

### 2022-02-10 exploration/te_analysis.R

@@ -1,0 +1,96 @@
+
+# libraries --------------------------------------------------------------------
+library(gtools)
+library(tidyverse)
+library(RTransferEntropy)
+
+# variables --------------------------------------------------------------------
+age_order <- c("aged1to4", "aged5to14", "aged15to24", "aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65to74", "aged75to84", "aged85up")
+
+# solution 1 -------------------------------------------------------------------
+# te_agegroup <- function(var_x, var_y){
+#   
+#   var_x <- select(df_wide_changes, all_of(var_x))
+#   var_y <- select(df_wide_changes, all_of(var_y))
+#   
+#   calc_te(var_x, var_y)
+# }
+# 
+# te_results <-
+#   expand_grid(
+#     X = age_order, 
+#     Y = age_order
+#   ) |> 
+#   as_tibble() |> 
+#   group_by(X, Y) |> 
+#   summarise(te = te_agegroup(X, Y))
+
+# te_matrix <- te_results |> 
+#   mutate(
+#     X = factor(X, levels = age_order),
+#     Y = factor(Y, levels = age_order),
+#   ) |> 
+#   arrange(X, Y) |> 
+#   pivot_wider(names_from = Y, values_from = te) |> 
+#   column_to_rownames("X") |> 
+#   as_tibble()
+# 
+# ggplot(te_results, aes(X, Y)) + 
+#   geom_raster(aes(fill=te)) + 
+#   scale_fill_gradient(low="grey90", high="red") +
+#   theme_bw()
+
+# solution 2 -------------------------------------------------------------------
+library(future) # enable parallel processing for all future transfer_entropy calls
+plan(multisession)
+
+te_agegroup <- function(var_x, var_y){
+  
+  var_x <- select(df_wide_changes, all_of(var_x))
+  var_y <- select(df_wide_changes, all_of(var_y))
+  
+  transfer_entropy(var_x, var_y, quiet = TRUE)[["coef"]] |> 
+    as_tibble() |> 
+    mutate(direction = c("X->Y", "Y->X"))
+}
+
+te_results <-
+  combinations(
+    n = length(age_order), 
+    r = 2, 
+    v = age_order, 
+    repeats.allowed = FALSE
+  ) |> 
+  as_tibble(.name_repair = ~ c("X", "Y")) |> 
+  group_by(X, Y) |> 
[…]

### 2023-05-18 manuscript/manuscript.Rmd
rss

@@
 ---
-title: Influence of School Closure on COVID-19 Contaminations in Europe and Repercussions across Age Groups
-author:
-  - name: Edgar Morgenroth
-    email: edgar.morgenroth@dcu.ie
-    affiliation: Dublin City University
+title: "Examining the Impact of School Closures on COVID-19 Infections in Europe and their Effects on Different Age Cohorts"
+short_title: "Impact of School Closures on COVID-19 by Age Cohorts"
+short_author: "Dupré & Morgenroth"
+authors: 
+- name: "Damien Dupré"
+  affiliation: "Business School, Dublin City University"
+  city: "Dublin"
+  country: "Ireland"
+  email: "damien.dupre\\@dcu.ie"
   footnote: 1
-  - name: Damien Dupré
-    email: damien.dupre@dcu.ie
-    affiliation: Dublin City University
+- name: "Edgar Morgenroth"
+  affiliation: "Business School, Dublin City University"
+  city: "Dublin"
+  country: "Ireland"
+  email: "edgar.morgenroth\\@dcu.ie"
   footnote: 2
-address:
-  - code: Dublin City University
-    address: 	Glasnevin, Dublin 9, Ireland
 footnote:
@@
 abstract: |
-  TBD
-
-journal: "TBD"
-date: "`r Sys.Date()`"
-
-csl: ["csl/elsevier-harvard.csl"]
-bibliography: ["bib/reference_list.bib"]
-
-#linenumbers: true
-#numbersections: true
-
+  This paper presents a comprehensive analysis of COVID-19 case trends and the influence of age groups in selected European countries.
+The study examines the relationship between population size and the total number of confirmed cases, revealing that larger countries tend to have higher case numbers.
+The impact of different waves on age groups is investigated, with the first wave affecting older age groups and the third wave primarily impacting the youngest age groups.
+The study also evaluates the effects of school closures on COVID-19 cases using a Generalized Additive Model (GAM).
+The findings demonstrate a decreasing non-linear effect of school closure across all countries studied.
+Moreover, the analysis of age groups reveals distinct patterns, with the 0 to 4 age group experiencing a downward trend after school closure, while age groups 5 to 14 exhibit a significant increase in cases.
+Age groups 15 to 24 show a surge immediately after closure, followed by a decline.
+Additionally, Transfer Entropy calculations highlight asymmetry in age group influences, indicating that changes in COVID-19 cases in certain age groups predict changes in other age groups but not vice versa.
+The study also identifies consistent patterns across countries, such as the predictive influence of younger cohorts on older cohorts in Austria, Germany, and the Netherlands.
+Conversely, Portugal and Spain show a predictive influence of older cohorts on younger cohorts.
+These findings contribute to a better understanding of COVID-19 dynamics in European countries and provide insights for public health strategies and interventions.
+keywords: COVID-19, School Closures, Causality, Generalised Additive Model, Transfer Entropy
+bibliography: "bibliography"
 header-includes:
@@
   bookdown::pdf_book:
-    base_format: rticles::elsevier_article
+    base_format: rticles::rss_article
     keep_tex: false
@@
     default
-knit: (function(input, ...) {
-    rmarkdown::render(
-      input,
-      output_dir = here::here("output"),
-      output_file = c(glue::glue("{xfun::sans_ext(input)}.pdf"), glue::glue("{xfun::sans_ext(input)}.docx")), 
-      output_format = "all",
-      envir = globalenv()
-    )
-  })
 ---
@@
 Source: @unesco2022https.
-(ref:gam) Standardized effect of the smooth term in Generalized Additive Model by age group.
+(ref:country) Standardized effect of the smooth term in Generalized Additive Model by country.
+(ref:age) Standardized effect of the smooth term in Generalized Additive Model by age group.
 (ref:te) Matrix of Transfer Entropy coefficients according every age group combination for each of the selected European country.
-Age groups on the x-axis are influencing the age groups on the y-axis ($X \rightarrow Y$).
+Age groups on the x-axis are influencing the age groups on the y-axis ($ag\,x \rightarrow\,ag\,y$).
 The significance of each Transfer Entropy coefficient is provided in the Appendix 2.
@@
 
-Since the onset of COVID-19, contradicting messages are circulating about the spread of the virus in schools.
-Sometimes presented as a safe environment [@falk2021covid; @walger2020children], at other times designed as an aggravating factor of the pandemic [@lopez2020transmission; @meuris2021transmission], child care facilities and schools have been closed multiple times in most countries to protect the public health.
-However, the efficacy of school closure is still unknown and little is know about how COVID-19 Cases spread across age cohort and especially from the younger cohort being at school to older cohorts including parents, grand parents and relatives.
-Some evidence show that school is an influential environment where the virus is likely to spread.
-Once contaminated at school, a common spread believe is that the children and teenagers will then contaminate their parents.
-Therefore, among the political actions to stop the spread of COVID-19, the closure of primary and secondary schools was widely adopted worldwide.
-This research aims to investigate 1) the effect of school closure on the evolution of COVID-19 cases and 2) the temporal relationship between the rise of cases in children and teenagers with the rise of cases of older age group.
+A wide range of non-pharmaceutical interventions (NPI) were enacted to control the spread of COVID-19, particularly in the period of the pandemic when no effective vaccine was available.
+Due to their social mixing patterns, children have been identified as an age group that can drive the spread of respiratory infections such as influenza [@moser2018estimating], and school closures were found to be effective in mitigating the spread of influenza H1N1in Japan in 2009 [@kawano2015substantial].
+It is therefore not surprising that school closures were one of the NPIs that were introduced in many countries to stop the spread of COVID-19.
+They were used particularly during the initial wave of the pandemic and during subsequent waves when case numbers were high.
+
+Recently a body of literature has shown various negative consequences on child development and health due to school closures.
+School closures have been found to have negatively affected child mental health [@moulin2022longitudinal], nutrition/obesity [@sugimoto2023temporal], and education [@lerkkanen2023reading].
+Some governments are now questioning their decisions to close schools [@de2021determines].
+For example the German health minister Karl Lauterbach, in an interview to one of Germany’s publicly funded television stations, admitted that “in retrospect it had been wrong to keep schools and childcare closed for so long” [@ard2023lauterbach].
+
+Even if school closures had negative consequences for children, they might nevertheless have been effective at stopping the spread of COVID-19.
+However, analysis on the effectiveness of school closures on the spread of COVID-19 remain inconclusive.
+@alfano2022effects in a study of a panel of European countries found that school closures were associated with a reduced COVID-19 incidence.
+In contrast, @walsh2021school in a review of 40 studies covering 150 countries concluded that the effectiveness of school closures was uncertain, with 60% having identified no impact and point to the potential for the analysis to be affected by confounding factors and collinearity.
+The latter shortcomings suggest that it is hard to draw firm inference as even a positive result may not imply causality.
+
+This paper investigates the dynamics of COVID-19 across age groups and how this is affected by school closures.
+This allows us to identify the effect of school closure on the evolution of COVID-19 cases older age groups including parents, grand parents and relatives about which little is known.
+In particular, we first use non-linear time series regression to identify the relationship between the school closures and case numbers in different age groups.
+Then, using the transfer entropy method, we evaluate the impact of case numbers changes between age groups in order to identify if school closure has secondary repercussions in the household.
+Transfer entropy can adapt to changes in the underlying system dynamics over time.
+It does not assume stationarity and can capture evolving causal relationships, making it suitable for studying dynamic systems and non-stationary processes such as COVID-19 contaminations.
 
@@
 
-One of the strongest assumption of government actions is that by reducing social contact will reduce the spread of the virus.
+Once it was clear that COVID-19 spreads from person to person, it was natural that governments were advised to enact measures to reduce social contact in order to reduce the spread of the virus.
 Therefore, it is legitimate to believe that by closing schools, a reduction of the contaminations would be observed in the younger age groups.
-However, the efficiency of school closure on the reduction of COVID-19 cases is still questioned.
-While some research have observed that school closures contribute to limit or to reduce the growth rate of confirmed cases after implementation [@yoshiyuki2020effects; @stage2021shut], other did not observe a change in the evolution of COVID-19 cases [@iwata2020school; @chang2020modelling].
+However, the efficiency of school closure on the reduction of COVID-19 cases is still questioned [@bayham2020impact; @esposito2021comprehensive].
+While some research has found that school closures contribute to limit or to reduce the growth rate of confirmed cases after implementation [@stage2021shut; @yoshiyuki2020effects], other did not observe a change in the evolution of COVID-19 cases [@chang2020modelling; @iwata2020school].
 For instance, a controlled comparison between similar localities in Japan with schools closed and school open did not revealed any evidence that school closures reduced the spread of COVID-19 [@fukumoto2021no].
@@
 
-A second implicit believe regarding the effect of school closure on the spread of COVID-19 is that school not only has an effect on to the spread of the virus in children and teenagers but also has a knock-on effect on the spread of the virus in older age groups also called Secondary Attack Rate (SAR).
+A second implicit belief regarding the effect of school closure on the spread of COVID-19 is that school not only has an effect on to the spread of the virus in children and teenagers but also has a knock-on effect on the spread of the virus in older age groups also called Secondary Attack Rate (SAR).
 The contaminated children and teenagers would bring the virus back at home and, then, they will contaminate their parents and relatives.
@@
 
-Despite being reported in several clinical and epidemiological studies [@zhendong2020clinical; @siebach2021childhood], multiple research have shown that the SAR from children to household members was, in fact, lower than expected [@kim2021role; @ludvigsson2020children; @vanderhoek2020role; @heavey2020evidence].
+Despite being reported in several clinical and epidemiological studies [@siebach2021childhood; @zhendong2020clinical], multiple research have shown that the SAR from children to household members was, in fact, lower than expected [@heavey2020evidence; @vanderhoek2020role; @kim2021role; @ludvigsson2020children].
 However, the SAR of children and teenagers to the household member is likely to be age-dependent, with difference between infants, primary and secondary school children, and college students [@grasleguen2021reopening].
 If a secondary transmission from children and teenagers to household member has a significant influence, then a temporal causality relationship between their evolution should be observed.
+
+To evaluate the impact of school closure on COVID-19 cases across different age groups, we first performed a non-linear time series regression to analyze the relationship between school closures and the number of COVID-19 cases in various age groups.
+In a second analysis, the transfer entropy method is employed to assess the influence of changes in case numbers between age groups, aiming to determine if school closures have secondary effects within households.
 
@@
 
-## Observations
+## Data
 
@@
 
+For the analysis data on covid case counts by age group are required.
 The number of COVID-19 cases per age group has been taken from the COVerAGE-DB project [@riffe2021data].
@@
 
-The observations are reported in terms of total number of COVID-19 cases from the start to the pandemic.
-Consequently, the daily number of cases at a specific date $n\prime_{t}$ is calculated with the difference between the total cases at a date $t$ and the total cases at a date $t-1$ (i.e., derivative 1).
+The observations are reported in terms of total number of COVID-19 cases per day from the start to the pandemic.
+The daily number of cases at a specific date $n\prime_{t}$ is calculated with the difference between the total cases at a date $t$ and the total cases at a date $t-1$ (i.e., derivative 1).
 In addition the change in daily number of cases $n\prime\prime_{t}$ between $n\prime_{t}$ and $n\prime_{t-1}$ has also been calculated (i.e., derivative 2).
@@
 
-The information regarding the school closure on a day by day basis is provided by the @unesco2022https.
+To identify the impact of school closures on the number of cases information regarding the school closures on a day by day basis is required and this is taken from the “UNESCO global education coalition” (2022).
 For each day, in each country, the status of the schools is indicated as fully open, partially open, closed due to COVID-19, or closed due to an academic break.
-Because it would be difficult to measure the effect of schools closure at a country level when schools are partially closed, on closures due to COVID-19, or due to an academic break are considered.
+Because it would be difficult to measure the effect of school closures at a country level when schools are partially closed, on closures due to COVID-19, or due to an academic break are considered.
 Indeed, both are considered as closure at a country wide level.
+Note that periods of closure longer than 21 days are compared for 28 days regardless if the schools opened or not after 21 days.
+Any COVID-19 case numbers beyond the 28-day period are not relevant for evaluating the impact of school closures.
 
-## Data Analysis
+## Data Analysis Methodology
 
@@
 
-The effect of school closure on the trend of daily COVID-19 cases is analysed for each country.
-Periods of closure longer than 21 days are compared for 28 days regardless if the schools opened or not after 21 days.
+The effect of school closure on the trend of daily COVID-19 cases is analysed for each country using a Generalised Additive Model (GAM).
+GAM is a flexible modelling approach that estimates non-linear relationships between variables.
+Compared to other methods such as Vector Autoregression, GAM can handle fixed and random smooth effects without making strict assumptions about linearity.
 
-A Generalised Additive Model (or GAM) using the R package *mgcv* [@wood2017generalized] is fitted on the daily COVID-19 cases to test the hypothesis of a significant non-linear evolution of cases among age groups from 0 to 4, from 5 to 9, from 10 to 14, from 15 to 19, and from 20 to 24.
+The GAM is fitted on the daily COVID-19 cases to test the hypothesis of a significant non-linear evolution of cases among age groups from 0 to 4, from 5 to 9, from 10 to 14, from 15 to 19, and from 20 to 24 [@wood2017generalized].
 The model also estimates the overall non-linear effect by country taking into account the interaction between age groups and countries as random intercepts and the interaction between time, countries and period of closure as random effects (Eq 1).
@@
 
-\begin{align}
-n\prime_{tgc} = \alpha_{gc} + f(closure_{tg}) + f(closure_{tc}) + a_{tcp} + \epsilon_{tgc}
-\end{align}
+\begin{align*}
+  \tag{1}
+  n\prime_t &\sim Poisson(\mu_t) \\
+  \log(\mu_t) &= \beta_0 + \beta_1\,wave_t + \beta_2\,country_t + \beta_3\,age\,group_t + \\
+              & f_1(closure_t) + f_2(closure_t,country_t) + f_3(closure_t,age\,group_t)
+\end{align*}
 
-\noindent where $t$ is the date corresponding to the confirmed COVID-19 cases, $g$ is an age group among the 10 age groups investigated, $c$ is a country among the 12 European countries selected, and $p$ is a specific period of closure.
-$n\prime_{tgc}$ represents the confirmed COVID-19 cases assuming a negative binomial distribution for the fitting [@loader2006local].
-The response variable $n\prime_{tgc}$ includes a specific intercept for each age group and each country ($\alpha_{gc}$).
-A smooth effect over time $f(closure_{tg})$ is applied to model for each age group and a smooth effect over time $f(closure_{tc})$ is applied to model for each country (Eq 2 and 3).
-These smooth effect $f(closure_{tg})$ and $f(closure_{tc})$ are built up in basic components, called the basis functions $b_{i}(X_{tg})$ and $b_{j}(X_{tg})$, such that:
-
-\begin{align}
-f(closure_{tg})=\sum_{i=1}^{k}\beta_{i}\times b_{i}(X_{tg})
-\\
-f(closure_{tc})=\sum_{j=1}^{k}\beta_{j}\times b_{j}(X_{tc})
-\end{align}
-
-\noindent where the regression parameters $\beta_{i}$ and $\beta_{j}$ are estimated by penalized likelihood maximization.
-
-The model also includes the random effects term $a_{tcp}= Zb_{i}$ where $Z$ is a random effects matrix and $b_{i}$ is a vector of random effects described by $b_{i} \sim N(0, D)$.
-In this, $D$ represents a covariance matrix.
-The error term $\epsilon_{tcp}$ is assumed to be normally and independently distributed $\epsilon_{tcp} \sim N(0, \sigma^2)$.
+\noindent where $n\prime_t$ represents the confirmed COVID-19 cases assuming a Poisson distribution for the fitting [@loader2006local] and $t$ is the date corresponding to the confirmed COVID-19 cases.
+The response variable includes a specific random effect taking into account variation within waves of school closure, countries, and age groups.
+The terms $f_1$ to $f_3$ are smooth functions of the time since closure, the time since closure for each country, and the time since closure for each age group.
+The restricted maximum likelihood (REML) was used to avoid overfitting while estimating smoothing parameters.
+In order to accurately account for the autocorrelation arising from the time series data, the residuals are modeled using an AR1 error model.
+By incorporating the autoregressive component, the AR1 model acknowledges the dependence of each residual on its previous value, thus providing a comprehensive representation of the data's temporal dynamics.
 
@@
 
-Transfer Entropy (TE) can be used to infer the temporal relationship between two time series $X$ and $Y$.
-This measure indicates whether $X$ can be used to to reduce the uncertainty on the future of $Y$ and, consequently, that $X$ causes $Y$ [@schreiber2000measuring].
-As such, Granger causality is a special case of TE applied to times series that are jointly Gaussian distributed [@barnett2009granger].
-Therefore, TE is a more robust analysis of times series especially when applied to the impact of age cohort in pandemic transmission [@kissler2020symbolic].
+Transfer entropy ($T$) can be used to infer the temporal relationship between two time series $X$ and $Y$.
+This measure indicates whether changes of in COVID-19 cases within a certain age group $X$ can be used to to reduce the uncertainty on the future COVID-19 cases within a another age group $Y$.
+If it does, $T$ is considered as evidence of a causal effect from the age group $X$ to the age group $Y$ [@schreiber2000measuring].
+As such, Granger causality is a special case of transfer entropy applied to times series that are jointly Gaussian distributed [@barnett2009granger].
+Therefore, transfer entropy is a more robust analysis of times series especially when applied to the impact of age cohort in pandemic transmission [@kissler2020symbolic].
 
@@
 \begin{align}
-  T_{X \rightarrow Y}(k,l) = \sum_{x,y} p\left(x_{t+1}, x_t^{(k)}, y_t^{(l)}\right) log \left(\frac{p\left(x_{t+1}| x_t^{(k)}, y_t^{(l)}\right)}{p\left(x_{t+1}|x_t^{(k)}\right)}\right)
+  \tag{2}
+  T_{ag\,x \rightarrow\,ag\,y}(k,l) = \sum_{ag\,x,ag\,y} p\left(ag\,x_{t+1}, ag\,x_t^{(k)}, ag\,y_t^{(l)}\right) \log \left(\frac{p\left(ag\,x_{t+1}| ag\,x_t^{(k)}, ag\,y_t^{(l)}\right)}{p\left(ag\,x_{t+1}|ag\,x_t^{(k)}\right)}\right)
 \end{align}
 
-\noindent where $T_{X\rightarrow Y}$ consequently measures the influence of the change dynamic from an age group $X$ to another age group $Y$ for every country (Eq 4).
+\noindent where $T_{ag\,x \rightarrow\,ag\,y}$ consequently measures the influence of the change dynamic from an age group $X$ (or $ag\,x$) to another age group $Y$ (or $ag\,y$) for every country (Eq 2).
 
-The day-by-day difference in COVID-19 confirmed cases $n\prime\prime_t$ is used to satisfy the stationary requirement for the calculation of Shannon's Transfer Entropy [@shannon1948mathematical].
-The Transfer Entropy analysis acrross all age group is done using the R package *RTransferEntropy* [@behrendt2019rtransferentropy].
+The day-by-day difference in COVID-19 confirmed cases $n\prime\prime_t$ is used to satisfy the stationary requirement for the calculation of Shannon's Transfer Entropy [@shannon1948mathematical; @behrendt2019rtransferentropy].
 
@@
     seed = 123,
-    size = 3,
+    size = 2,
     force = 0.5,
@@
 The evolution of COVID-19 cases reveals some similarities across all age groups.
-However, the influence of each waves on the each groups has also some particularities (Figure \@ref(fig:descriptive)).
+However, the influence of each waves on each age group has also some particularities (Figure \@ref(fig:descriptive)).
 For example, the first wave was more important among the oldest age groups whereas the third wave was more important among the youngest age groups.
@@
 
-In order to evaluate the shape of the trend in the numbers of COVID-19 cases reported after the three school closure longer than 21 consecutive days, a GAM was fitted taking into account numbers for 0 to 4, 5 to 9, 10 to 14, 15 to 19, and 20 to 24 year old across all the selected European countries.
+In order to evaluate the shape of the trend in the numbers of COVID-19 cases reported after the three school closure longer than 21 consecutive days, a GAM was fitted as described above, taking into account the overall effect across all the selected European countries as well as the effect for age groups: 0 to 4, 5 to 9, 10 to 14, 15 to 19, and 20 to 24 year old.
 The obtained results satisfy the requirements to fit this model which explains `r percent(gam_summary$dev.expl, accuracy = 0.1)` of the deviance in COVID-19 cases.
 
-Overall, the results did not revealed a main effect of school closure at a country level for most of the selected European countries (Austria: `r gam_results["Austria"]`; Belgium: `r gam_results["Belgium"]`; Bulgaria: `r gam_results["Bulgaria"]`; Croatia: `r gam_results["Croatia"]`; Estonia: `r gam_results["Estonia"]`; France: `r gam_results["France"]`; Greece: `r gam_results["Greece"]`; Netherlands: `r gam_results["Netherlands"]`; Portugal: `r gam_results["Portugal"]`; Spain: `r gam_results["Spain"]`) except for Germany (`r gam_results["Germany"]`) and for Slovakia (`r gam_results["Slovakia"]`).
+Overall, the results revealed a decreasing non-linear effect of school closure at a country level for the selected European countries (Austria: `r gam_results["Austria"]`; Belgium: `r gam_results["Belgium"]`; Bulgaria: `r gam_results["Bulgaria"]`; Croatia: `r gam_results["Croatia"]`; Estonia: `r gam_results["Estonia"]`; France: `r gam_results["France"]`; Germany (`r gam_results["Germany"]`); Greece: `r gam_results["Greece"]`; The Netherlands: `r gam_results["Netherlands"]`; Portugal: `r gam_results["Portugal"]`; Slovakia (`r gam_results["Slovakia"]`; Spain: `r gam_results["Spain"]`, see Figure \@ref(fig:age)).
 
-When looking at the difference between age groups, results show no significant change in the evolution of cases until 19 year old (0 to 4: `r gam_results["0to4"]`; 5 to 9: `r gam_results["5to9"]`; 10 to 14: `r gam_results["10to14"]`; 15 to 19: `r gam_results["15to19"]`).
-However, the smooth term of the GAM is significant for 20 to 24 year old (`r gam_results["20to24"]`), which indicates that the evolution of cases is not linear and, therefore, the school closure had a significant impact on the evolution of cases (Figure \@ref(fig:gam)).
+```{r country, fig.height=5, out.width="\\textwidth", fig.cap="(ref:country)"}
+gam_plot_country
+```
 
-```{r gam, fig.height=3, out.width="\\textwidth", fig.cap="(ref:gam)"}
+Considering all countries, the analysis of age groups reveals distinct patterns (Figure \@ref(fig:age)).
+The 0 to 4 age group experiences a downward trend during the initial two weeks following school closure, followed by stabilization (`r gam_results["0to4"]`).
+In contrast, age groups ranging from 5 to 14 exhibit a substantial increase in cases after 14 days of school closure (5 to 9 age group: `r gam_results["5to9"]` and 10 to 14 age group: `r gam_results["10to14"]`).
+Moving on, age groups between 15 and 24 demonstrate a notable surge immediately after school closure, followed by a decrease after 14 days (15 to 19 age group: `r gam_results["15to19"]` and 20 to 24 age group: `r gam_results["20to24"]`).
+
+```{r age, fig.height=3, out.width="\\textwidth", fig.cap="(ref:age)"}
 gam_plot_age
@@
 
-Before the Transfer Entropy analysis, an Augmented Dickey-Fuller has been applied on each age group in order to insure that the daily changes in COVID-19 cases are stationary (see Appendix 1).
+Before the Transfer Entropy analysis is carried out, an Augmented Dickey-Fuller has been applied on each age group in order to insure that the daily changes in COVID-19 cases are stationary (see Appendix 1).
 
-The results of the Transfer Entropy calculations between age groups for each of the 12 selected European countries are reported Figure \@ref(fig:te).
-The first observation is the absence of symmetry between influencing age groups (i.e., $X$) and influenced age groups (i.e., $Y$).
+The results of the transfer entropy calculations between age groups for each of the 12 selected European countries are reported Figure \@ref(fig:te).
+The absence of symmetry between influencing age groups (i.e., $ag\,x$) and influenced age groups (i.e., $ag\,y$) is found.
 Indeed, the change in COVID-19 cases in some age groups are influenced by other age groups but they are not reciprocally influencing these age groups.
 
-Additionally, to highlight differences and similarities between countries, a grid has been added to Figure \@ref(fig:te): the upper left quadrant indicates how younger age cohorts are influencing older age cohorts, the bottom right quadrant indicates how older age cohorts are influencing younger age cohorts, finally lower left and and upper right are indicating how younger or older age cohorts are influencing themselves.
+Figure \@ref(fig:te) shows how different age cohorts influence the COVID-19 case numbers of all age groups across countries.
+The upper left quadrant indicates how younger age cohorts are influencing older age cohorts, the bottom right quadrant indicates how older age cohorts are influencing younger age cohorts, finally lower left and and upper right are indicating how younger or older age cohorts are influencing themselves.
 By analysing these quadrant it is possible to identify similar patterns across multiple countries.
-Indeed, it appears that Austria, Germany and the Netherlands have significantly higher TE coefficients in the upper left quadrant of the matrix, which indicate that the daily changes in COVID-19 cases number in younger cohorts is predicting the the daily changes in COVID-19 cases number in older cohorts.
-Alternatively, it appears that Austria, the Netherlands, Portugal and Spain have significantly higher TE coefficients in the lower right quadrant of the matrix, which indicate that the daily changes in COVID-19 cases number in older cohorts is predicting the the daily changes in COVID-19 cases number in younger cohorts.
+Indeed, it appears that Austria, Germany and The Netherlands have significantly higher $T$ coefficients in the upper left quadrant of the matrix, which indicate that the daily changes in COVID-19 cases number in younger cohorts is predicting the the daily changes in COVID-19 cases number in older cohorts.
+Alternatively, it appears that Austria, the Netherlands, Portugal and Spain have significantly higher $T$ coefficients in the lower right quadrant of the matrix, which indicate that the daily changes in COVID-19 cases number in older cohorts is predicting the the daily changes in COVID-19 cases number in younger cohorts.
 
@@
 
-This indicate mix results, while the school closure was ineffective in pre-primary, primary and secondary levels (i.e., ISCED 0 to 3), the closure of tertiary levels (i.e., ISCED 5 to 8) had an impact on COVID-19 transmission.
-However, different pattern emerges according the selected European countries.
-Some have shown asymmetric patterns of influence between age groups while other did.
+Regarding the influence of school closures on age groups, the study reveals both similarities and specific characteristics.
+While there are commonalities in the evolution of COVID-19 cases across all European countries investigated, school closures exhibit distinct impacts on different age groups.
+Notably, the analysis shows that the 0 to 4 age group experiences a downward trend in COVID-19 cases during the initial two weeks following school closure, followed by stabilization.
+In contrast, age groups ranging from 5 to 14 exhibit a substantial increase in cases after 14 days of school closure.
+Additionally, age groups between 15 and 24 demonstrate a notable surge immediately after school closure, followed by a decrease after 14 days.
+These results are partially replicating observations from @alfano2022effects while providing a clearer picture of the structure of the effect of school closure.
 
-References {#references .unnumbered}
-==========
+Moreover, the paper includes Transfer Entropy calculations between age groups for each of the 12 selected European countries.
+The results reveal an absence of symmetry between influencing age groups (X) and influenced age groups (Y).
+Some age groups influence the COVID-19 case numbers of other age groups, but there is no reciprocal influence.
+This finding suggests that changes in COVID-19 cases in certain age groups can predict the changes in other age groups but not vice versa.
+The absence of reciprocity in the influence of age groups on each other suggests the presence of unidirectional relationships.
+These findings imply that changes in COVID-19 cases in certain age groups can serve as predictors for changes in other age groups, highlighting the importance of considering intergenerational interactions in designing effective control measures.
+
+Furthermore, the quadrant analysis reveals similar patterns across multiple countries.
+Austria, Germany, and the Netherlands exhibit significantly higher TE coefficients in the upper left quadrant, indicating that changes in COVID-19 cases in younger cohorts predict the changes in older cohorts.
+On the other hand, Austria, the Netherlands, Portugal, and Spain show higher TE coefficients in the lower right quadrant, indicating that changes in COVID-19 cases in older cohorts predict the changes in younger cohorts.
+
+In conclusion, this study sheds light on the trends and patterns of confirmed COVID-19 cases in selected European countries, with a particular focus on the influence of age groups and school closures.
+As it is the case for lockdowns [@alfano2020efficacy; @molefi2021impact], this finding suggests that prolonged school closures can contribute to mitigating the spread of the virus, particularly in the context of decreasing case numbers over time.
+However, it is crucial to consider the specific dynamics of different age groups, as the impact of school closures varies among them.
+The study reveals important insights that can contribute to our understanding of the dynamics of the pandemic and inform public health strategies.
+
+Data Availability
+===================
+
+All data extracted from the COVerAGE-DB project and from the UNESCO global education coalition that are used to produce these results are fully available here https://github.com/damien-dupre/covid_ts_causality.
+The repository also includes the R code that were used to produce these results.

### 2023-05-18 analysis/gam_analysis_europe.R

@@ -1,102 +1,105 @@
 # libraries --------------------------------------------------------------------
 library(broom)
+library(future)
 library(glue)
 library(gratia)
 library(here)
+library(mgcv)
+library(parallel)
 library(scales)
 library(tidyverse)
 
-# # libraries --------------------------------------------------------------------
-# library(broom)
-# library(future)
-# library(glue)
-# library(gratia)
-# library(here)
-# library(mgcv)
-# library(parallel)
-# library(scales)
-# library(tidyverse)
-# 
-# # options ----------------------------------------------------------------------
-# options(scipen = 999)
-# plan("future::multisession")
-# 
-# # data -------------------------------------------------------------------------
-# df_school_closure_wave <- df_global |> 
-#   select(date, country, schools_closed) |> 
-#   group_by(country) |> 
-#   mutate(schools_closure_day = schools_closed - lag(schools_closed)) |> 
-#   filter(schools_closure_day == 1) |> 
-#   mutate(schools_closure_wave = row_number())
-# 
-# df_global_school <- df_school_closure_wave |> 
-#   full_join(df_global, by = c("date", "country", "schools_closed")) |> 
-#   group_by(country) |> 
-#   arrange(date) |> 
-#   fill(schools_closure_wave, .direction = "down") |> 
-#   drop_na(schools_closure_wave) |> 
-#   group_by(country, schools_closure_wave) |> 
-#   mutate(schools_time_since_closure = row_number()) |> 
-#   ungroup()
-# 
-# df_global_school_age <- df_global_school |> 
-#   select(date, country, schools_closed, schools_closure_day, schools_closure_wave, schools_time_since_closure, who_region) |> 
-#   full_join(df_global_age, by = c("date", "country"))
-# 
-# df_global_school_length <- df_global_school |> 
-#   filter(schools_closed == 1) |> 
-#   count(country, schools_closure_wave) |> 
-#   rename(schools_closure_wave_n = n)
-# 
-# df_global_school_age_long <- df_global_school_length |> 
-#   right_join(df_global_school_age, by = c("country", "schools_closure_wave")) |> 
-#   filter(
-#     schools_closure_wave_n >= days_min_closure,
-#     schools_time_since_closure <= days_after_closure,
[…]

### 2023-05-18 manuscript/manuscript.Rmd
rss to edgar

@@
 keywords: COVID-19, School Closures, Causality, Generalised Additive Model, Transfer Entropy
-bibliography: "bibliography"
+bibliography: "bibliography.bib"
 header-includes:
@@
     default
+knit: (function(input, ...) {
+    rmarkdown::render(
+      input,
+      output_dir = here::here("output"),
+      output_file = c(glue::glue("{xfun::sans_ext(input)}.pdf"), glue::glue("{xfun::sans_ext(input)}.docx")),
+      output_format = "all",
+      envir = globalenv()
+    )
+  })
 ---
@@
   cache.extra = rand_seed, 
-  cache = TRUE,
+  cache = FALSE,
   message = FALSE, 
@@
 In conclusion, this study sheds light on the trends and patterns of confirmed COVID-19 cases in selected European countries, with a particular focus on the influence of age groups and school closures.
-As it is the case for lockdowns [@alfano2020efficacy; @molefi2021impact], this finding suggests that prolonged school closures can contribute to mitigating the spread of the virus, particularly in the context of decreasing case numbers over time.
+It suggests that like lockdowns [@alfano2020efficacy; @molefi2021impact], prolonged school closures can contribute to mitigating the spread of the virus, particularly in the context of decreasing case numbers over time.
 However, it is crucial to consider the specific dynamics of different age groups, as the impact of school closures varies among them.

### 2023-12-15 analysis/gam_analysis_europe.R
country figures + appendix 1

@@ -153,6 +153,46 @@ gam_plot_age <- gam_df |>
     text = element_text(size = 10, family = "serif")
   )
 
+## original visualisation
+# gam_plot_country <- gam_df |>
+#   filter(by == "country") |>
+#   mutate(
+#     smooth = smooth |>
+#       str_remove("s\\(schools_time_since_closure\\):country")
+#   ) |>
+#   ggplot(aes(x = schools_time_since_closure, y = est, group = smooth)) +
+#   geom_line(color = "gray40", alpha = 0.4) +
+#   geom_text_repel(
+#     data = gam_df |>
+#       group_by(country) |>
+#       slice_max(schools_time_since_closure, with_ties = FALSE),
+#     aes(label = country),
+#     hjust = 0,
+#     direction = "y",
+#     segment.color = NA,
+#     seed = 123,
+#     size = 4,
+#     force = 0.5,
+#     family = "serif"
+#   ) +
+#   scale_x_continuous(
+#     "Day Since Closure",
+#     breaks = seq(0, 28, by = 7),
+#     limits = c(0, 31)
+#   ) +
+#   scale_y_continuous(
+#     "Standardized Effect"
+#   ) +
+#   scale_color_discrete(
+#     "Country"
+#   ) +
+#   theme_bw() +
+#   theme(
+#     legend.position = "none",
+#     text = element_text(size = 10, family = "serif")
+#   )
+
+# regular facets
 gam_plot_country <- gam_df |>
   filter(by == "country") |>
   mutate(
@@ -160,20 +200,8 @@ gam_plot_country <- gam_df |>
       str_remove("s\\(schools_time_since_closure\\):country")
   ) |>
   ggplot(aes(x = schools_time_since_closure, y = est, group = smooth)) +
-  geom_line(color = "gray40", alpha = 0.4) +
-  geom_text_repel(
-    data = gam_df |>
-      group_by(country) |>
-      slice_max(schools_time_since_closure, with_ties = FALSE),
-    aes(label = country),
-    hjust = 0,
-    direction = "y",
-    segment.color = NA,
-    seed = 123,
-    size = 4,
-    force = 0.5,
-    family = "serif"
-  ) +
+  geom_line() +
+  facet_wrap(~country) +
   scale_x_continuous(
[…]

### 2023-12-18 manuscript_elsevier/appendix_1.Rmd
manuscript elsevier

---
title: "Appendix 1: Dickey-Fuller test for stationay timeseries"
toc: false

output:
  bookdown::pdf_book

header-includes:
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{threeparttable}
  - \floatplacement{figure}{H}
  - \floatplacement{table}{H}
  - \usepackage{caption}
---

Test for the null hypothesis of a unit root of a univarate time series (equivalently, non-stationary time series).

The Augmented Dickey-Fuller test incorporates three types of linear regression models. The first type is a linear model with no drift and linear trend with respect to time. The second type is a linear model with drift but no linear trend. The third type is a linear model with both drift and linear trend.

The p.value is calculated by interpolating the test statistics from the corresponding critical values tables (Fuller, 1996) for each type of linear models with given sample size.

```{r setup, include = FALSE}
# libraries --------------------------------------------------------------------
library(here)
library(tidyverse)
library(knitr)
library(kableExtra)

# options ----------------------------------------------------------------------
set.seed(123) # Seed for random number generation
options(scipen = 999) # disable sci number format
opts_chunk$set(
  cache.extra = rand_seed, 
  cache = FALSE,
  message = FALSE, 
  warning = FALSE, 
  error = FALSE,
  echo = FALSE, 
  fig.retina = 3,
  dev = "ragg_png"
  )
```

```{r}
df_global_age <- read_csv(here("data/df_global_age.csv"))

adf_res_type1 <- df_global_age |>
  group_by(age_grp, country) |>
  group_modify(~ aTSA::adf.test(.x$daily_changes, output = FALSE, nlag = 2) |>
                 bind_rows() |> 
                 pluck("type1") |> 
                 as_tibble()
  ) |> 
  rename(ADF_type1 = ADF, p.value_type1 = p.value)

adf_res_type2 <- df_global_age |>
  group_by(age_grp, country) |>
  group_modify(~ aTSA::adf.test(.x$daily_changes, output = FALSE, nlag = 2) |>
                 bind_rows() |> 
                 pluck("type2") |> 
                 as_tibble()
  ) |> 
  rename(ADF_type2 = ADF, p.value_type2 = p.value)

adf_res_type3 <- df_global_age |>
  group_by(age_grp, country) |>
  group_modify(~ aTSA::adf.test(.x$daily_changes, output = FALSE, nlag = 2) |>
                 bind_rows() |> 
                 pluck("type3") |> 
                 as_tibble()
  ) |> 
  rename(ADF_type3 = ADF, p.value_type3 = p.value)

adf_res <- adf_res_type1 |> 
  full_join(adf_res_type2) |> 
  full_join(adf_res_type3)
```

```{r}
adf_res |>
  filter(age_grp %in% c("aged0to4", "aged5to9")) |> 
  kbl(
    caption = "Dickey-Fuller test for all combinaisons of countries and age groups investigated",
    col.names = c("Age Group", "Country", "Lag", "ADF", "p", "ADF", "p", "ADF", "p")
  ) |> 
  kable_classic(full_width = F) |> 
  add_header_above(c(" " = 3, "Type 1" = 2, "Type 2" = 2, "type 3" = 2)) |>
  collapse_rows(columns = 1:2, valign = "middle")
```

\newpage

```{r}
adf_res |>
  filter(age_grp %in% c("aged10to14", "aged15to19")) |> 
  kbl(
    caption = "Dickey-Fuller test for all combinaisons of countries and age groups investigated",
    col.names = c("Age Group", "Country", "Lag", "ADF", "p", "ADF", "p", "ADF", "p")
  ) |> 
  kable_classic(full_width = F) |> 
  add_header_above(c(" " = 3, "Type 1" = 2, "Type 2" = 2, "type 3" = 2)) |>
  collapse_rows(columns = 1:2, valign = "middle")
```

\newpage

```{r}
adf_res |>
  filter(age_grp %in% c("aged20to24", "aged25to34")) |> 
  kbl(
    caption = "Dickey-Fuller test for all combinaisons of countries and age groups investigated",
    col.names = c("Age Group", "Country", "Lag", "ADF", "p", "ADF", "p", "ADF", "p")
  ) |> 
  kable_classic(full_width = F) |> 
  add_header_above(c(" " = 3, "Type 1" = 2, "Type 2" = 2, "type 3" = 2)) |>
  collapse_rows(columns = 1:2, valign = "middle")
```

\newpage

```{r}
adf_res |>
  filter(age_grp %in% c("aged35to44", "aged45to54")) |> 
  kbl(
    caption = "Dickey-Fuller test for all combinaisons of countries and age groups investigated",
    col.names = c("Age Group", "Country", "Lag", "ADF", "p", "ADF", "p", "ADF", "p")
  ) |> 
  kable_classic(full_width = F) |> 
  add_header_above(c(" " = 3, "Type 1" = 2, "Type 2" = 2, "type 3" = 2)) |>
  collapse_rows(columns = 1:2, valign = "middle")
```

\newpage

```{r}
adf_res |>
  filter(age_grp %in% c("aged55to64", "aged65+")) |> 
  kbl(
    caption = "Dickey-Fuller test for all combinaisons of countries and age groups investigated",
    col.names = c("Age Group", "Country", "Lag", "ADF", "p", "ADF", "p", "ADF", "p")
  ) |> 
  kable_classic(full_width = F) |> 
  add_header_above(c(" " = 3, "Type 1" = 2, "Type 2" = 2, "type 3" = 2)) |>
  collapse_rows(columns = 1:2, valign = "middle")
```

### 2023-12-18 manuscript_elsevier/appendix_2.Rmd
manuscript elsevier

---
title: "Appendix 2: Significance of Transfer Entropy coefficients"
toc: false

output:
  bookdown::pdf_book

header-includes:
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{threeparttable}
  - \floatplacement{figure}{H}
  - \floatplacement{table}{H}
  - \usepackage{caption}
---

Results for each permutation of age groups from each country. Tables include transfer entropy estimates, the effective transfer entropy estimates, standard errors and p-values based on bootstrap replications of the Markov chains under the null hypothesis of statistical independence.

```{r setup, include = FALSE}
# libraries --------------------------------------------------------------------
library(here)
library(tidyverse)
library(knitr)
library(kableExtra)

# options ----------------------------------------------------------------------
set.seed(123) # Seed for random number generation
options(scipen = 999) # disable sci number format
opts_chunk$set(
  cache.extra = rand_seed, 
  cache = FALSE,
  message = FALSE, 
  warning = FALSE, 
  error = FALSE,
  echo = FALSE, 
  fig.retina = 3,
  dev = "ragg_png"
  )

# variables --------------------------------------------------------------------
age_order <- c(
  "aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24",
  "aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+"
)

# data -------------------------------------------------------------------------
te_results <- read_csv(here("data/te_results.csv")) |> 
    mutate(
      X = factor(X, levels = age_order),
      Y = factor(Y, levels = age_order)
    )
```

```{r}
te_results |>
  filter(X == "aged0to4") |> 
  filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |> 
  arrange(Y) |> 
  kbl(
    caption = "Significance of Transfer Entropy coefficients",
    digits = 2
  ) |> 
  kable_classic(full_width = F) |> 
  collapse_rows(columns = 1:2, valign = "middle") |> 
  kable_styling(font_size = 7)
```

\newpage

```{r}
te_results |>
  filter(X == "aged0to4") |> 
  filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |> 
  arrange(Y) |> 
  kbl(
    caption = "Significance of Transfer Entropy coefficients",
    digits = 2
  ) |> 
  kable_classic(full_width = F) |> 
  collapse_rows(columns = 1:2, valign = "middle") |> 
  kable_styling(font_size = 7)
```

\newpage

```{r}
te_results |>
  filter(X == "aged5to9") |> 
  filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |> 
  arrange(Y) |> 
  kbl(
    caption = "Significance of Transfer Entropy coefficients",
    digits = 2
  ) |> 
  kable_classic(full_width = F) |> 
  collapse_rows(columns = 1:2, valign = "middle") |> 
  kable_styling(font_size = 7)
```

\newpage

```{r}
te_results |>
  filter(X == "aged5to9") |> 
  filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |> 
  arrange(Y) |> 
  kbl(
    caption = "Significance of Transfer Entropy coefficients",
    digits = 2
  ) |> 
  kable_classic(full_width = F) |> 
  collapse_rows(columns = 1:2, valign = "middle") |> 
  kable_styling(font_size = 7)
```

\newpage
```{r}
te_results |>
  filter(X == "aged10to14") |> 
  filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |> 
  arrange(Y) |> 
  kbl(
    caption = "Significance of Transfer Entropy coefficients",
    digits = 2
  ) |> 
  kable_classic(full_width = F) |> 
  collapse_rows(columns = 1:2, valign = "middle") |> 
  kable_styling(font_size = 7)
```

\newpage

```{r}
te_results |>
  filter(X == "aged10to14") |> 
  filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |> 
  arrange(Y) |> 
  kbl(
    caption = "Significance of Transfer Entropy coefficients",
    digits = 2
  ) |> 
  kable_classic(full_width = F) |> 
  collapse_rows(columns = 1:2, valign = "middle") |> 
  kable_styling(font_size = 7)
```

\newpage

```{r}
te_results |>
  filter(X == "aged15to19") |> 
  filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |> 
  arrange(Y) |> 
  kbl(
    caption = "Significance of Transfer Entropy coefficients",
    digits = 2
  ) |> 
  kable_classic(full_width = F) |> 
  collapse_rows(columns = 1:2, valign = "middle") |> 
  kable_styling(font_size = 7)
```

\newpage

```{r}
te_results |>
  filter(X == "aged15to19") |> 
  filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |> 
  arrange(Y) |> 
  kbl(
    caption = "Significance of Transfer Entropy coefficients",
    digits = 2
  ) |> 
  kable_classic(full_width = F) |> 
  collapse_rows(columns = 1:2, valign = "middle") |> 
  kable_styling(font_size = 7)
```

\newpage

```{r}
te_results |>
  filter(X == "aged20to24") |> 
  filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |> 
  arrange(Y) |> 
  kbl(
    caption = "Significance of Transfer Entropy coefficients",
    digits = 2
  ) |> 
  kable_classic(full_width = F) |> 
  collapse_rows(columns = 1:2, valign = "middle") |> 
  kable_styling(font_size = 7)
```

\newpage

```{r}
te_results |>
  filter(X == "aged20to24") |> 
  filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |> 
  arrange(Y) |> 
  kbl(
    caption = "Significance of Transfer Entropy coefficients",
    digits = 2
  ) |> 
  kable_classic(full_width = F) |> 
  collapse_rows(columns = 1:2, valign = "middle") |> 
  kable_styling(font_size = 7)
```

\newpage

```{r}
te_results |>
  filter(X == "aged25to34") |> 
  filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |> 
  arrange(Y) |> 
  kbl(
    caption = "Significance of Transfer Entropy coefficients",
    digits = 2
  ) |> 
  kable_classic(full_width = F) |> 
  collapse_rows(columns = 1:2, valign = "middle") |> 
  kable_styling(font_size = 7)
```

\newpage

```{r}
te_results |>
  filter(X == "aged25to34") |> 
  filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |> 
  arrange(Y) |> 
  kbl(
    caption = "Significance of Transfer Entropy coefficients",
    digits = 2
  ) |> 
  kable_classic(full_width = F) |> 
  collapse_rows(columns = 1:2, valign = "middle") |> 
  kable_styling(font_size = 7)
```

\newpage

```{r}
te_results |>
  filter(X == "aged35to44") |> 
  filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |> 
  arrange(Y) |> 
  kbl(
    caption = "Significance of Transfer Entropy coefficients",
    digits = 2
  ) |> 
  kable_classic(full_width = F) |> 
  collapse_rows(columns = 1:2, valign = "middle") |> 
  kable_styling(font_size = 7)
```

\newpage

```{r}
te_results |>
  filter(X == "aged35to44") |> 
  filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |> 
  arrange(Y) |> 
  kbl(
    caption = "Significance of Transfer Entropy coefficients",
    digits = 2
  ) |> 
  kable_classic(full_width = F) |> 
  collapse_rows(columns = 1:2, valign = "middle") |> 
  kable_styling(font_size = 7)
```

\newpage

```{r}
te_results |>
  filter(X == "aged45to54") |> 
  filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |> 
  arrange(Y) |> 
  kbl(
    caption = "Significance of Transfer Entropy coefficients",
    digits = 2
  ) |> 
  kable_classic(full_width = F) |> 
  collapse_rows(columns = 1:2, valign = "middle") |> 
  kable_styling(font_size = 7)
```

\newpage

```{r}
te_results |>
  filter(X == "aged45to54") |> 
  filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |> 
  arrange(Y) |> 
  kbl(
    caption = "Significance of Transfer Entropy coefficients",
    digits = 2
  ) |> 
  kable_classic(full_width = F) |> 
  collapse_rows(columns = 1:2, valign = "middle") |> 
  kable_styling(font_size = 7)
```

\newpage

```{r}
te_results |>
  filter(X == "aged55to64") |> 
  filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |> 
  arrange(Y) |> 
  kbl(
    caption = "Significance of Transfer Entropy coefficients",
    digits = 2
  ) |> 
  kable_classic(full_width = F) |> 
  collapse_rows(columns = 1:2, valign = "middle") |> 
  kable_styling(font_size = 7)
```

\newpage

```{r}
te_results |>
  filter(X == "aged55to64") |> 
  filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |> 
  arrange(Y) |> 
  kbl(
    caption = "Significance of Transfer Entropy coefficients",
    digits = 2
  ) |> 
  kable_classic(full_width = F) |> 
  collapse_rows(columns = 1:2, valign = "middle") |> 
  kable_styling(font_size = 7)
```

\newpage

```{r}
te_results |>
  filter(X == "aged65+") |> 
  filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |> 
  arrange(Y) |> 
  kbl(
    caption = "Significance of Transfer Entropy coefficients",
    digits = 2
  ) |> 
  kable_classic(full_width = F) |> 
  collapse_rows(columns = 1:2, valign = "middle") |> 
  kable_styling(font_size = 7)
```

\newpage

```{r}
te_results |>
  filter(X == "aged65+") |> 
  filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |> 
  arrange(Y) |> 
  kbl(
    caption = "Significance of Transfer Entropy coefficients",
    digits = 2
  ) |> 
  kable_classic(full_width = F) |> 
  collapse_rows(columns = 1:2, valign = "middle") |> 
  kable_styling(font_size = 7)
```

### 2023-12-18 manuscript_elsevier/manuscript_elsevier.Rmd
manuscript elsevier

--- 2023-05-18 manuscript/manuscript.Rmd
@@
 ---
-title: "Examining the Impact of School Closures on COVID-19 Infections in Europe and their Effects on Different Age Cohorts"
-short_title: "Impact of School Closures on COVID-19 by Age Cohorts"
-short_author: "Dupré & Morgenroth"
-authors: 
-- name: "Damien Dupré"
-  affiliation: "Business School, Dublin City University"
-  city: "Dublin"
-  country: "Ireland"
-  email: "damien.dupre\\@dcu.ie"
-  footnote: 1
-- name: "Edgar Morgenroth"
-  affiliation: "Business School, Dublin City University"
-  city: "Dublin"
-  country: "Ireland"
-  email: "edgar.morgenroth\\@dcu.ie"
-  footnote: 2
-footnote:
-  - code: 1
-    text: "Corresponding Author"
-  - code: 2
-    text: "Equal contribution"
+title: Examining the Impact of School Closures on COVID-19 Infections in Europe and their Effects on Different Age Cohorts
+author:
+  - name: Damien Dupré
+    email: damien.dupre@dcu.ie
+    affiliation: Business School, Dublin City University
+    correspondingauthor: true
+  - name: Edgar Morgenroth
+    email: edgar.morgenroth@dcu.ie
+    affiliation: Business School, Dublin City University
+address:
+  - code: Business School, Dublin City University
+    organization: Business School, Dublin City University
+    city: Dublin
+    country: Ireland
 abstract: |
-  This paper presents a comprehensive analysis of COVID-19 case trends and the influence of age groups in selected European countries.
-The study examines the relationship between population size and the total number of confirmed cases, revealing that larger countries tend to have higher case numbers.
-The impact of different waves on age groups is investigated, with the first wave affecting older age groups and the third wave primarily impacting the youngest age groups.
+  This paper analyzes the trends of COVID-19 cases in different age groups in selected European countries, assessing to what extent school closures helped reduce case numbers, how school closures affected case numbers in different age groups, and how COVID-19 spread between age groups.
 The study also evaluates the effects of school closures on COVID-19 cases using a Generalized Additive Model (GAM).
-The findings demonstrate a decreasing non-linear effect of school closure across all countries studied.
-Moreover, the analysis of age groups reveals distinct patterns, with the 0 to 4 age group experiencing a downward trend after school closure, while age groups 5 to 14 exhibit a significant increase in cases.
+The findings demonstrate a decreasing non-linear effect of school closure on the total COVID-19 case number across all countries studied.
+However, the analysis of age groups only reveals a consistent downward trend in infections in the 0 to 4 pre-school age group, while the school-going age groups 5 to 14 exhibit a significant increase in cases.
 Age groups 15 to 24 show a surge immediately after closure, followed by a decline.
-Additionally, Transfer Entropy calculations highlight asymmetry in age group influences, indicating that changes in COVID-19 cases in certain age groups predict changes in other age groups but not vice versa.
-The study also identifies consistent patterns across countries, such as the predictive influence of younger cohorts on older cohorts in Austria, Germany, and the Netherlands.
-Conversely, Portugal and Spain show a predictive influence of older cohorts on younger cohorts.
+Transfer Entropy calculations highlight asymmetry in age group influences, indicating that changes in COVID-19 cases in certain age groups predict changes in other age groups but not vice versa.
 These findings contribute to a better understanding of COVID-19 dynamics in European countries and provide insights for public health strategies and interventions.
-keywords: COVID-19, School Closures, Causality, Generalised Additive Model, Transfer Entropy
-bibliography: "bibliography.bib"
+keywords: 
+  - COVID-19
+  - School Closures
+  - Intergenerational Transmission.
+journal: "Social Science & Medicine"
+date: "`r Sys.Date()`"
+classoption: preprint, 3p, authoryear
+linenumbers: false
+numbersections: true
+bibliography: bibliography.bib
+csl: https://www.zotero.org/styles/elsevier-harvard
+output:
+  bookdown::pdf_book:
+    base_format: rticles::elsevier_article
+    keep_tex: false
+    citation_package: default
 header-includes:
@@
   - \usepackage{caption}
-
-output:
-  bookdown::pdf_book:
-    base_format: rticles::rss_article
-    keep_tex: false
-  bookdown::word_document2:
-    default
-knit: (function(input, ...) {
-    rmarkdown::render(
-      input,
-      output_dir = here::here("output"),
-      output_file = c(glue::glue("{xfun::sans_ext(input)}.pdf"), glue::glue("{xfun::sans_ext(input)}.docx")),
-      output_format = "all",
-      envir = globalenv()
-    )
-  })
 ---
@@
 (ref:overall) Cumulative COVID-19 cases number for selected European countries since the beginning of the pandemic. Source: COVerAGE-DB [@riffe2021data].
-(ref:descriptive) Periods of schools closure since the beginning of the COVID-19 pandemic for selected European countries and their reason: regular academic break vs.
+(ref:descriptive) Periods of school closure since the beginning of the COVID-19 pandemic for selected European countries and their reason: regular academic break vs.
 closure due to government decisions.
@@
 (ref:age) Standardized effect of the smooth term in Generalized Additive Model by age group.
-(ref:te) Matrix of Transfer Entropy coefficients according every age group combination for each of the selected European country.
+(ref:te) Matrix of Transfer Entropy coefficients according to every age group combination for each of the selected European country.
 Age groups on the x-axis are influencing the age groups on the y-axis ($ag\,x \rightarrow\,ag\,y$).
-The significance of each Transfer Entropy coefficient is provided in the Appendix 2.
+The significance of each Transfer Entropy coefficient is provided in Appendix 2.
 
@@
 
-Introduction
-==========================
+# Introduction
 
-A wide range of non-pharmaceutical interventions (NPI) were enacted to control the spread of COVID-19, particularly in the period of the pandemic when no effective vaccine was available.
-Due to their social mixing patterns, children have been identified as an age group that can drive the spread of respiratory infections such as influenza [@moser2018estimating], and school closures were found to be effective in mitigating the spread of influenza H1N1in Japan in 2009 [@kawano2015substantial].
+A wide range of non-pharmaceutical interventions (NPI) were enacted to control the spread of COVID-19, especially before the availability of effective vaccines.
+Due to their social mixing patterns, children have been identified as an age group that can drive the spread of respiratory infections such as influenza [@moser2018estimating], and school closures were found to be effective in mitigating the spread of influenza H1N1 in Japan in 2009 [@kawano2015substantial].
 It is therefore not surprising that school closures were one of the NPIs that were introduced in many countries to stop the spread of COVID-19.
@@
 
-Recently a body of literature has shown various negative consequences on child development and health due to school closures.
+Recently, a body of literature has shown various negative consequences on child development and health due to school closures.
 School closures have been found to have negatively affected child mental health [@moulin2022longitudinal], nutrition/obesity [@sugimoto2023temporal], and education [@lerkkanen2023reading].
 Some governments are now questioning their decisions to close schools [@de2021determines].
-For example the German health minister Karl Lauterbach, in an interview to one of Germany’s publicly funded television stations, admitted that “in retrospect it had been wrong to keep schools and childcare closed for so long” [@ard2023lauterbach].
+For example, the German health minister Karl Lauterbach, in an interview with one of Germany’s publicly funded television stations, admitted that “in retrospect it had been wrong to keep schools and childcare closed for so long” [@ard2023lauterbach].
 
 Even if school closures had negative consequences for children, they might nevertheless have been effective at stopping the spread of COVID-19.
-However, analysis on the effectiveness of school closures on the spread of COVID-19 remain inconclusive.
+However, analysis on the effectiveness of school closures on the spread of COVID-19 remains inconclusive.
 @alfano2022effects in a study of a panel of European countries found that school closures were associated with a reduced COVID-19 incidence.
-In contrast, @walsh2021school in a review of 40 studies covering 150 countries concluded that the effectiveness of school closures was uncertain, with 60% having identified no impact and point to the potential for the analysis to be affected by confounding factors and collinearity.
-The latter shortcomings suggest that it is hard to draw firm inference as even a positive result may not imply causality.
+In contrast, @walsh2021school in a review of 40 studies covering 150 countries concluded that the effectiveness of school closures was uncertain, with 60% having identified no impact and pointing to the potential for the analysis to be affected by confounding factors and collinearity.
+The latter shortcomings suggest that it is hard to draw firm inferences, as even a positive result may not imply causality.
 
-This paper investigates the dynamics of COVID-19 across age groups and how this is affected by school closures.
-This allows us to identify the effect of school closure on the evolution of COVID-19 cases older age groups including parents, grand parents and relatives about which little is known.
-In particular, we first use non-linear time series regression to identify the relationship between the school closures and case numbers in different age groups.
-Then, using the transfer entropy method, we evaluate the impact of case numbers changes between age groups in order to identify if school closure has secondary repercussions in the household.
+This paper investigates the dynamics of COVID-19 across age groups and how this was affected by school closures.
+This allows us to identify the effect of school closure on the evolution of COVID-19 cases in other age groups and particularly on older and more vulnerable age groups, including parents, grandparents, and relatives about which little is known.
+
+We identify the relationship between the school closures and case numbers in different age groups using Generalized Additive Models (GAM), which are more flexible than conventional Vector Autoregressions, given that they allow for non-linear relationships between variables.
+Furthermore, compared to other methods such as Vector Autoregression, GAM can handle fixed and random smooth effects without making strict assumptions about linearity.
+Then, using the transfer entropy method, we evaluate the impact of case number changes between age groups in order to identify if school closure has secondary repercussions in the household.
 Transfer entropy can adapt to changes in the underlying system dynamics over time.
@@
 However, the efficiency of school closure on the reduction of COVID-19 cases is still questioned [@bayham2020impact; @esposito2021comprehensive].
-While some research has found that school closures contribute to limit or to reduce the growth rate of confirmed cases after implementation [@stage2021shut; @yoshiyuki2020effects], other did not observe a change in the evolution of COVID-19 cases [@chang2020modelling; @iwata2020school].
-For instance, a controlled comparison between similar localities in Japan with schools closed and school open did not revealed any evidence that school closures reduced the spread of COVID-19 [@fukumoto2021no].
-If the school closure had a real impact on the evolution of confirmed COVID-19 cases, it should be possible to observe a decrease or a least an inflection in the trend of its evolution among younger age groups.
+While some research has found that school closures contribute to limit or to reduce the growth rate of confirmed cases after implementation [@stage2021shut; @yoshiyuki2020effects], others did not observe a change in the evolution of COVID-19 cases [@chang2020modelling; @iwata2020school].
+For instance, a controlled comparison between similar localities in Japan with schools closed and schools open did not reveal any evidence that school closures reduced the spread of COVID-19 [@fukumoto2021no].
+If the school closure had a real impact on the evolution of confirmed COVID-19 cases, it should be possible to observe a decrease or at least an inflection in the trend of its evolution among younger age groups.
 
@@
 
-A second implicit belief regarding the effect of school closure on the spread of COVID-19 is that school not only has an effect on to the spread of the virus in children and teenagers but also has a knock-on effect on the spread of the virus in older age groups also called Secondary Attack Rate (SAR).
-The contaminated children and teenagers would bring the virus back at home and, then, they will contaminate their parents and relatives.
-For example, a research investigating the contamination in the household network not only revealed an exceptional high rate of secondary contamination but also that these contamination happened when the school were closed [@sorianoarandes2021household].
+A second implicit belief regarding the effect of school closure on the spread of COVID-19 is that school not only influences the spread of the virus among children and teenagers but also has a knock-on effect on the spread of the virus in older age groups, also called Secondary Attack Rate (SAR).
+The contaminated children and teenagers would bring the virus back home and would pass the virus on to their parents and relatives.
+For example, research investigating the contamination in the household network not only revealed an exceptionally high rate of secondary contamination but also that this contamination happened when the schools were closed [@sorianoarandes2021household].
 
-Despite being reported in several clinical and epidemiological studies [@siebach2021childhood; @zhendong2020clinical], multiple research have shown that the SAR from children to household members was, in fact, lower than expected [@heavey2020evidence; @vanderhoek2020role; @kim2021role; @ludvigsson2020children].
-However, the SAR of children and teenagers to the household member is likely to be age-dependent, with difference between infants, primary and secondary school children, and college students [@grasleguen2021reopening].
+Despite being reported in several clinical and epidemiological studies [@siebach2021childhood; @zhendong2020clinical], multiple research have shown that the SAR from children to household members was lower than expected [@heavey2020evidence; @vanderhoek2020role; @kim2021role; @ludvigsson2020children].
+However, the SAR of children and teenagers to the household member is likely to be age-dependent, with differences between infants, primary and secondary school children, and college students [@grasleguen2021reopening].
 If a secondary transmission from children and teenagers to household member has a significant influence, then a temporal causality relationship between their evolution should be observed.
@@
 
-Method
-============
+# Estimation Methodology
 
-## Data
+## Generalized Additive Model
 
-### COVID-19 cases per age group
-
-For the analysis data on covid case counts by age group are required.
-The number of COVID-19 cases per age group has been taken from the COVerAGE-DB project [@riffe2021data].
-The COVerAGE-DB project consists in 3 data files: an "Input" data file which collects the official COVID-19 cases from 117 countries, the "Output_5" data file which is a projection of COVID-19 cases by group of 5 years, and the "Output_10" data file which is a projection of COVID-19 cases by group of 10 years.
-For our analyses we have used the "Output_5" which deals with the heterogeneity of countries' reporting format by using spline approximations when the data for this age bracket is not available for a country.
-Therefore, it allows a precise analysis between age groups under 24 while age groups of 25 and above are reduced to 10 year brackets and groups above 64 concatenated all together.
-
-In order to cross validate the data of obtained after spline approximations, a comparison with the data published by the @who2022https reveals perfect similarities.
-
-The original data consist in 14089320 observations of 10 variables (117 distinct country, region within the country, an unique observation code, the date of the observation, the gender which can be male, female or both, the age bracket by 5 years from 0 to 100, a confirmation of the age interval for each bracket, the total number of cases so far, the total number of deaths and the total number of tests performed) from `r format(min(df_global$date), "%B %d, %Y")` to `r format(max(df_global$date), "%B %d, %Y")`.
-
-After removing countries with missing and inconstant values, only 22 are suitable for the data analyses.
-However, in order to focus this analysis to geographically and culturally comparable countries, only 12 European countries are kept: Austria, Belgium, Bulgaria, Croatia, Estonia, France, Germany, Greece, Netherlands, Portugal, Slovakia, and Spain.
-
-The observations are reported in terms of total number of COVID-19 cases per day from the start to the pandemic.
-The daily number of cases at a specific date $n\prime_{t}$ is calculated with the difference between the total cases at a date $t$ and the total cases at a date $t-1$ (i.e., derivative 1).
-In addition the change in daily number of cases $n\prime\prime_{t}$ between $n\prime_{t}$ and $n\prime_{t-1}$ has also been calculated (i.e., derivative 2).
-
-### School closure
-
-To identify the impact of school closures on the number of cases information regarding the school closures on a day by day basis is required and this is taken from the “UNESCO global education coalition” (2022).
-For each day, in each country, the status of the schools is indicated as fully open, partially open, closed due to COVID-19, or closed due to an academic break.
-Because it would be difficult to measure the effect of school closures at a country level when schools are partially closed, on closures due to COVID-19, or due to an academic break are considered.
-Indeed, both are considered as closure at a country wide level.
-Note that periods of closure longer than 21 days are compared for 28 days regardless if the schools opened or not after 21 days.
-Any COVID-19 case numbers beyond the 28-day period are not relevant for evaluating the impact of school closures.
-
-## Data Analysis Methodology
-
-### Generalised Additive Model
-
-The effect of school closure on the trend of daily COVID-19 cases is analysed for each country using a Generalised Additive Model (GAM).
-GAM is a flexible modelling approach that estimates non-linear relationships between variables.
+The effect of school closure on the trend of daily COVID-19 cases is analyzed using a Generalized Additive Model (GAM).
+GAM is a flexible modeling approach that estimates non-linear relationships between variables.
 Compared to other methods such as Vector Autoregression, GAM can handle fixed and random smooth effects without making strict assumptions about linearity.
@@
 The GAM is fitted on the daily COVID-19 cases to test the hypothesis of a significant non-linear evolution of cases among age groups from 0 to 4, from 5 to 9, from 10 to 14, from 15 to 19, and from 20 to 24 [@wood2017generalized].
-The model also estimates the overall non-linear effect by country taking into account the interaction between age groups and countries as random intercepts and the interaction between time, countries and period of closure as random effects (Eq 1).
+The model also estimates the overall non-linear effect by country, taking into account the interaction between age groups and countries as random intercepts and the interaction between time, countries, and the period of closure as random effects (Eq 1).
 
-By estimating the degree of smoothness of a Bayesian spline smoothing using restricted fast maximum likelihood estimation [@wood2011fast], GAM identifies dynamic patterns underlying the evolution of COVID-19 cases reported while taking into account the random effect of different age groups and countries as follows:
+By estimating the degree of smoothness of a Bayesian spline smoothing using restricted fast maximum likelihood estimation [@wood2011fast], GAM identifies dynamic patterns underlying the evolution of COVID-19 cases reported while including the random effect of different age groups and countries as follows:
 
@@
 
-\noindent where $n\prime_t$ represents the confirmed COVID-19 cases assuming a Poisson distribution for the fitting [@loader2006local] and $t$ is the date corresponding to the confirmed COVID-19 cases.
+\noindent where $n\prime_t$ represents the confirmed COVID-19 cases, assuming a Poisson distribution for the fitting [@loader2006local], and $t$ is the date corresponding to the confirmed COVID-19 cases.
 The response variable includes a specific random effect taking into account variation within waves of school closure, countries, and age groups.
@@
 In order to accurately account for the autocorrelation arising from the time series data, the residuals are modeled using an AR1 error model.
-By incorporating the autoregressive component, the AR1 model acknowledges the dependence of each residual on its previous value, thus providing a comprehensive representation of the data's temporal dynamics.
+By incorporating the autoregressive component, the AR1 model acknowledges the dependence of each residual on its previous value, thus providing a comprehensive representation of the data’s temporal dynamics.
 
-### Transfer Entropy
+## Transfer Entropy
 
 Transfer entropy ($T$) can be used to infer the temporal relationship between two time series $X$ and $Y$.
-This measure indicates whether changes of in COVID-19 cases within a certain age group $X$ can be used to to reduce the uncertainty on the future COVID-19 cases within a another age group $Y$.
-If it does, $T$ is considered as evidence of a causal effect from the age group $X$ to the age group $Y$ [@schreiber2000measuring].
-As such, Granger causality is a special case of transfer entropy applied to times series that are jointly Gaussian distributed [@barnett2009granger].
-Therefore, transfer entropy is a more robust analysis of times series especially when applied to the impact of age cohort in pandemic transmission [@kissler2020symbolic].
+This measure indicates whether changes in COVID-19 cases within a certain age group $X$ can be used to reduce the uncertainty on the future COVID-19 cases within another age group $Y$.
+If it does, $T$ is considered evidence of a causal effect from the age group $X$ to the age group $Y$ [@schreiber2000measuring].
+As such, Granger causality is a special case of transfer entropy applied to time series that are jointly Gaussian distributed [@barnett2009granger].
+Therefore, transfer entropy is a more robust analysis of time series, especially when applied to the impact of age cohorts on pandemic transmission [@kissler2020symbolic].
 
-The influence of the evolution in COVID-19 cases across all age groups is evaluated using Shannon's transfer entropy is given by:
+The influence of the evolution in COVID-19 cases across all age groups is evaluated using Shannon’s transfer entropy, given by:
 
@@
 
-The day-by-day difference in COVID-19 confirmed cases $n\prime\prime_t$ is used to satisfy the stationary requirement for the calculation of Shannon's Transfer Entropy [@shannon1948mathematical; @behrendt2019rtransferentropy].
+The day-by-day difference in COVID-19 confirmed cases $n\prime\prime_t$ is used to satisfy the stationary requirement for the calculation of Shannon’s Transfer Entropy [@shannon1948mathematical; @behrendt2019rtransferentropy].
 
-Results
-===================
+# Data
 
-Overall, the trend of confirmed COVID-19 cases follows similar patterns across the selected European countries with scales respecting the size of population in these countries (Figure \@ref(fig:overall)).
-Indeed, the bigger the country, the higher the total number of COVID-19 cases.
+## COVID-19 cases per age group
+
+For the analysis, data on COVID case counts by age group are required.
+The number of COVID-19 cases per age group has been taken from the COVerAGE-DB project [@riffe2021data].
+The COVerAGE-DB project consists of 3 data files: an “Input” data file that collects the official COVID-19 cases from 117 countries, the “Output_5” data file, which is a projection of COVID-19 cases by a group of 5 years, and the “Output_10” data file, which is a projection of COVID-19 cases by a group of 10 years.
+For our analyses, we have used the “Output_5,” which deals with the heterogeneity of countries’ reporting formats by using spline approximations when the data for this age bracket is not available for a country.
+Therefore, it allows a precise analysis between age groups under 24, while age groups of 25 and above are reduced to 10-year brackets, and groups above 64 are concatenated all together.
+
+In order to cross-validate the data obtained after spline approximations, a comparison with the data published by the @who2022https reveals perfect similarities.
+The original data consist of 14,089,320 observations of 10 variables (117 distinct countries, region within the country, a unique observation code, the date of the observation, the gender, which can be male, female, or both, the age bracket by 5 years from 0 to 100, a confirmation of the age interval for each bracket, the total number of cases so far, the total number of deaths, and the total number of tests performed) from `r format(min(df_global$date), "%B %d, %Y")` to `r format(max(df_global$date), "%B %d, %Y")`.
+After removing countries with missing and inconsistent values, only 22 are suitable for data analyses.
+However, to focus this analysis on geographically and culturally comparable countries, only 12 European countries are kept: Austria, Belgium, Bulgaria, Croatia, Estonia, France, Germany, Greece, Netherlands, Portugal, Slovakia, and Spain.
+The observations are reported in terms of the total number of COVID-19 cases per day from the start of the pandemic.
+The daily number of cases at a specific date $n\prime_{t}$ is calculated with the difference between the total cases at a date $t$ and the total cases at a date $t-1$ (i.e., derivative 1).
+In addition, the change in the daily number of cases $n\prime\prime_{t}$ between $n\prime_{t}$ and $n\prime_{t-1}$ has also been calculated (i.e., derivative 2).
+
+## School closure
+
+To identify the impact of school closures on the number of cases, information regarding the school closures on a day-by-day basis is required, and this is taken from the “UNESCO global education coalition” (2022).
+For each day, in each country, the status of the schools is indicated as fully open, partially open, closed due to COVID-19, or closed due to an academic break.
+Because it would be difficult to measure the effect of school closures at a country level when schools are partially closed, only closures due to COVID-19 or due to an academic break are considered.
+Indeed, both are considered as closure at a country-wide level.
+Note that periods of closure longer than 21 days are compared for 28 days, regardless of whether the schools opened or not after 21 days.
+Any COVID-19 case numbers beyond the 28-day period are not relevant for evaluating the impact of school closures.
+
+# Results
+
+The data show that the trend of confirmed COVID-19 cases follows similar patterns across the selected European countries, with scales following the size of the population in these countries (Figure \@ref(fig:overall)).
+Thus, the bigger the country, the higher the total number of COVID-19 cases.
 
@@
 The evolution of COVID-19 cases reveals some similarities across all age groups.
-However, the influence of each waves on each age group has also some particularities (Figure \@ref(fig:descriptive)).
-For example, the first wave was more important among the oldest age groups whereas the third wave was more important among the youngest age groups.
+However, the influence of each wave on individual age groups also has some particularities (Figure \@ref(fig:descriptive)).
+For example, the first wave was more important among the oldest age groups, whereas the third wave was more important among the youngest age groups.
 
@@
 
-In order to evaluate the shape of the trend in the numbers of COVID-19 cases reported after the three school closure longer than 21 consecutive days, a GAM was fitted as described above, taking into account the overall effect across all the selected European countries as well as the effect for age groups: 0 to 4, 5 to 9, 10 to 14, 15 to 19, and 20 to 24 year old.
-The obtained results satisfy the requirements to fit this model which explains `r percent(gam_summary$dev.expl, accuracy = 0.1)` of the deviance in COVID-19 cases.
+As outlined above, to evaluate the shape of the trend in the numbers of COVID-19 cases reported after the three school closures longer than 21 consecutive days, a GAM was fitted as described above, taking into account the overall effect across all the selected European countries, as well as the effect for age groups: 0 to 4, 5 to 9, 10 to 14, 15 to 19, and 20 to 24 years old.
+The obtained results satisfy the requirements to fit this model, which explains `r percent(gam_summary$dev.expl, accuracy = 0.1)` of the variation in COVID-19 cases.
 
@@
 Considering all countries, the analysis of age groups reveals distinct patterns (Figure \@ref(fig:age)).
-The 0 to 4 age group experiences a downward trend during the initial two weeks following school closure, followed by stabilization (`r gam_results["0to4"]`).
-In contrast, age groups ranging from 5 to 14 exhibit a substantial increase in cases after 14 days of school closure (5 to 9 age group: `r gam_results["5to9"]` and 10 to 14 age group: `r gam_results["10to14"]`).
-Moving on, age groups between 15 and 24 demonstrate a notable surge immediately after school closure, followed by a decrease after 14 days (15 to 19 age group: `r gam_results["15to19"]` and 20 to 24 age group: `r gam_results["20to24"]`).
+For the 0 to 4 age group, i.e., the pre-school group, there was a downward trend during the initial two weeks following school closure, followed by flattening off (`r gam_results["0to4"]`).
+In contrast, for the age group ranging from 5 to 14, a substantial increase in cases after 14 days of school closure is found (5 to 9 age group: `r gam_results["5to9"]` and 10 to 14 age group: `r gam_results["10to14"]`).
+For age groups between 15 and 24, there is a notable surge immediately after school closure, followed by a decrease after 14 days (15 to 19 age group: `r gam_results["15to19"]` and 20 to 24 age group: `r gam_results["20to24"]`).
+Thus, while the total case number declined following school closures, among younger age groups only the pre-school age group had a consistent reduction in cases, while for other age groups, there was even an increase in cases following school closures.
 
@@
 
-Before the Transfer Entropy analysis is carried out, an Augmented Dickey-Fuller has been applied on each age group in order to insure that the daily changes in COVID-19 cases are stationary (see Appendix 1).
+Given that school closures were not only aimed at reducing COVID-19 cases among school-going children but also age groups, it is important to test the degree to which there was transmission across age groups, which is done using transfer entropy analysis.
+Before this analysis is carried out, an Augmented Dickey-Fuller has been applied to each age group to ensure that the daily changes in COVID-19 cases are stationary (see Appendix 1).
 
-The results of the transfer entropy calculations between age groups for each of the 12 selected European countries are reported Figure \@ref(fig:te).
+The results of the transfer entropy calculations between age groups for each of the 12 selected European countries are reported in Figure \@ref(fig:te).
 The absence of symmetry between influencing age groups (i.e., $ag\,x$) and influenced age groups (i.e., $ag\,y$) is found.
-Indeed, the change in COVID-19 cases in some age groups are influenced by other age groups but they are not reciprocally influencing these age groups.
-
+Indeed, the change in COVID-19 cases in some age groups is influenced by other age groups, but they are not reciprocally influencing these age groups.
 Figure \@ref(fig:te) shows how different age cohorts influence the COVID-19 case numbers of all age groups across countries.
-The upper left quadrant indicates how younger age cohorts are influencing older age cohorts, the bottom right quadrant indicates how older age cohorts are influencing younger age cohorts, finally lower left and and upper right are indicating how younger or older age cohorts are influencing themselves.
-By analysing these quadrant it is possible to identify similar patterns across multiple countries.
-Indeed, it appears that Austria, Germany and The Netherlands have significantly higher $T$ coefficients in the upper left quadrant of the matrix, which indicate that the daily changes in COVID-19 cases number in younger cohorts is predicting the the daily changes in COVID-19 cases number in older cohorts.
-Alternatively, it appears that Austria, the Netherlands, Portugal and Spain have significantly higher $T$ coefficients in the lower right quadrant of the matrix, which indicate that the daily changes in COVID-19 cases number in older cohorts is predicting the the daily changes in COVID-19 cases number in younger cohorts.
+The upper left quadrant indicates how younger age cohorts are influencing older age cohorts, the bottom right quadrant indicates how older age cohorts are influencing younger age cohorts, finally, the lower left and upper right indicate how younger or older age cohorts are influencing themselves.
+By analyzing these quadrants, it is possible to identify similar patterns across multiple countries.
+Indeed, it appears that Austria, Germany, and The Netherlands have significantly higher $T$ coefficients in the upper left quadrant of the matrix, which indicates that the daily changes in COVID-19 cases number in younger cohorts are predicting the daily changes in COVID-19 cases number in older cohorts.
+Alternatively, it appears that Austria, the Netherlands, Portugal, and Spain have significantly higher $T$ coefficients in the lower right quadrant of the matrix, which indicates that the daily changes in COVID-19 cases number in older cohorts are predicting the daily changes in COVID-19 cases number in younger cohorts.
 
@@
 
-Discussion & Conclusion
-===================
+# Discussion & Conclusion
 
-Knowledge about the transmission of the virus significantly improved with the amount of studies performed, especially in the case of how the virus behaves with children.
-From the early analyses showing that the virus was instantness in children [@li2020role], the results has changed to more nuanced position which states that the spread of the virus in children is moderate.
-
-Regarding the influence of school closures on age groups, the study reveals both similarities and specific characteristics.
-While there are commonalities in the evolution of COVID-19 cases across all European countries investigated, school closures exhibit distinct impacts on different age groups.
+Knowledge about the transmission of the virus significantly improved over time as more studies have been published.
+While an early study found that children did not play an important role in the transmission of the virus [@li2020role], more recent results give a more nuanced position, stating that the spread of the virus in children is moderate.
+This study, using robust statistical methods, considered the effect of school closures on COVID-19 cases across age groups.
+While there are commonalities in the evolution of COVID-19 cases across all European countries that were included in our data, school closures exhibit distinct impacts on different age groups.
 Notably, the analysis shows that the 0 to 4 age group experiences a downward trend in COVID-19 cases during the initial two weeks following school closure, followed by stabilization.
@@
 Additionally, age groups between 15 and 24 demonstrate a notable surge immediately after school closure, followed by a decrease after 14 days.
-These results are partially replicating observations from @alfano2022effects while providing a clearer picture of the structure of the effect of school closure.
+Thus, the results do not support the hypothesis that school closures were effective.
+These results partially replicate observations from @alfano2022effects while providing a clearer picture of the structure of the effect of school closure.
 
-Moreover, the paper includes Transfer Entropy calculations between age groups for each of the 12 selected European countries.
-The results reveal an absence of symmetry between influencing age groups (X) and influenced age groups (Y).
+The Transfer Entropy calculations between age groups for each of the 12 selected European countries reveal an absence of symmetry between influencing age groups (X) and influenced age groups (Y).
 Some age groups influence the COVID-19 case numbers of other age groups, but there is no reciprocal influence.
-This finding suggests that changes in COVID-19 cases in certain age groups can predict the changes in other age groups but not vice versa.
-The absence of reciprocity in the influence of age groups on each other suggests the presence of unidirectional relationships.
-These findings imply that changes in COVID-19 cases in certain age groups can serve as predictors for changes in other age groups, highlighting the importance of considering intergenerational interactions in designing effective control measures.
+This finding suggests that changes in COVID-19 cases in certain age groups can predict the changes in other age groups but not vice versa, i.e., causality can be unidirectional.
+These findings highlight the importance of considering intergenerational interactions in designing effective control measures.
 
@@
 In conclusion, this study sheds light on the trends and patterns of confirmed COVID-19 cases in selected European countries, with a particular focus on the influence of age groups and school closures.
-It suggests that like lockdowns [@alfano2020efficacy; @molefi2021impact], prolonged school closures can contribute to mitigating the spread of the virus, particularly in the context of decreasing case numbers over time.
-However, it is crucial to consider the specific dynamics of different age groups, as the impact of school closures varies among them.
-The study reveals important insights that can contribute to our understanding of the dynamics of the pandemic and inform public health strategies.
+While a reduction in overall COVID-19 cases was found during school closures, which might suggest that like lockdowns [@alfano2020efficacy; @molefi2021impact], prolonged school closures could contribute to mitigating the spread of the virus, the results for the school-going age groups suggest that this is not the case.
 
-Data Availability
-===================
+# Data Availability
 
@@
 The repository also includes the R code that were used to produce these results.
+
+# References {-}

### 2023-12-18 analysis/te_analysis_europe.R

@@ -65,9 +65,9 @@ countries_of_interest <- c(
 # 
 # te_results <- bind_rows(te_results_XY,te_results_YX)
 # 
-# write_csv(te_results, here("results/te_results.csv"))
+# write_csv(te_results, here("data/te_results.csv"))
 
-te_results <- read_csv(here("results/te_results.csv"))
+te_results <- read_csv(here("data/te_results.csv"))
 
 te_plot <- te_results |> 
   mutate(

### 2023-12-18 manuscript_elsevier/abstract_elsevier.Rmd
Social Science & Medicine submission

--- 2023-12-18 manuscript_elsevier/manuscript_elsevier.Rmd
@@
 title: Examining the Impact of School Closures on COVID-19 Infections in Europe and their Effects on Different Age Cohorts
-author:
-  - name: Damien Dupré
-    email: damien.dupre@dcu.ie
-    affiliation: Business School, Dublin City University
-    correspondingauthor: true
-  - name: Edgar Morgenroth
-    email: edgar.morgenroth@dcu.ie
-    affiliation: Business School, Dublin City University
-address:
-  - code: Business School, Dublin City University
-    organization: Business School, Dublin City University
-    city: Dublin
-    country: Ireland
+# author:
+#   - name: Damien Dupré
+#     email: damien.dupre@dcu.ie
+#     affiliation: Business School, Dublin City University
+#     correspondingauthor: true
+#   - name: Edgar Morgenroth
+#     email: edgar.morgenroth@dcu.ie
+#     affiliation: Business School, Dublin City University
+# address:
+#   - code: Business School, Dublin City University
+#     organization: Business School, Dublin City University
+#     city: Dublin
+#     country: Ireland
 abstract: |
@@
 ---
-
-(ref:overall) Cumulative COVID-19 cases number for selected European countries since the beginning of the pandemic. Source: COVerAGE-DB [@riffe2021data].
-(ref:descriptive) Periods of school closure since the beginning of the COVID-19 pandemic for selected European countries and their reason: regular academic break vs.
-closure due to government decisions.
-Source: @unesco2022https.
-(ref:country) Standardized effect of the smooth term in Generalized Additive Model by country.
-(ref:age) Standardized effect of the smooth term in Generalized Additive Model by age group.
-(ref:te) Matrix of Transfer Entropy coefficients according to every age group combination for each of the selected European country.
-Age groups on the x-axis are influencing the age groups on the y-axis ($ag\,x \rightarrow\,ag\,y$).
-The significance of each Transfer Entropy coefficient is provided in Appendix 2.
-
-```{r setup, include = FALSE}
-# libraries --------------------------------------------------------------------
-library(ggrepel)
-library(here)
-library(janitor)
-library(knitr)
-library(osfr)
-library(scales)
-library(tidyverse)
-
-# options ----------------------------------------------------------------------
-set.seed(123) # Seed for random number generation
-options(scipen = 999) # disable sci number format
-opts_chunk$set(
-  cache.extra = rand_seed, 
-  cache = FALSE,
-  message = FALSE, 
-  warning = FALSE, 
-  error = FALSE,
-  echo = FALSE, 
-  fig.retina = 3,
-  dev = "ragg_png"
-  )
-
-# raw data ---------------------------------------------------------------------
-# osf_retrieve_file("7tnfh") |>
-#   osf_download(conflicts = "overwrite", progress = TRUE, verbose = TRUE)
-# 
-# df <- "Output_5.zip" |>
-#   read_csv(skip = 3, col_types = "ccccciiddd") |>
-#   mutate(Date = as.Date(Date, format = "%d.%m.%Y")) |>
-#   clean_names()
-# 
-# file.remove("Output_5.zip")
-# 
-# write_csv(df, here("data/df.csv"))
-# 
-# df <- read_csv(here("data/df.csv"))
-# 
-# df_school_closure <- 
-#   read_csv(
-#     "https://en.unesco.org/sites/default/files/covid_impact_education_full.csv",
-#     col_types = list(.default = col_character(), Date = col_date(format = "%d/%m/%Y"))) |> 
-#   clean_names() |>
-#   select(date, country, status) |> 
-#   mutate(country = case_when(
-#     country == "Bolivia (Plurinational State of)" ~ "Bolivia",  
-#     country == "Central African republic" ~ "Central African Republic",
-#     country == "Iran (Islamic Republic of)" ~ "Iran",
-#     country == "Republic of Moldova" ~ "Moldova",
-#     country == "Republic of Korea" ~ "South Korea",
-#     country == "United Kingdom of Great Britain and Northern Ireland" ~ "United Kingdom",
-#     country == "United States of America" ~ "USA",
-#     country == "Viet Nam" ~ "Vietnam",
-#     TRUE ~ country
-#     )
-#   )
-# 
-# df_check <- 
-#   read_csv("https://covid19.who.int/WHO-COVID-19-global-data.csv") |> 
-#   clean_names() |> 
-#   rename(date = date_reported) |> 
-#   mutate(country = case_when(
-#     country == "Bolivia (Plurinational State of)" ~ "Bolivia",  
-#     country == "Iran (Islamic Republic of)" ~ "Iran",
-#     country == "occupied Palestinian territory, including east Jerusalem" ~ "Palestine",
-#     country == "Republic of Moldova" ~ "Moldova",
-#     country == "Republic of Korea" ~ "South Korea",
-#     country == "United States of America" ~ "USA",
-#     country == "Venezuela (Bolivarian Republic of)" ~ "Venezuela",
-#     country == "Viet Nam" ~ "Vietnam",
-#     TRUE ~ country
-#     )
-#   )
-# 
-# # variables --------------------------------------------------------------------
-# ## keep only closure waves longer than `days_min_closure` but display only the first `days_after_closure`
-# days_min_closure <- 21
-# days_after_closure <- 28
-# ## type of school closure to investigate
-# period_closure_to_keep <- c(
-#   #"Partially open",
-#   "Academic break",
-#   "Closed due to COVID-19"
-#   )
-# 
-# ## countries of interest
-# countries_of_interest <- c(
-#   "Austria", "Belgium", "Bulgaria", "Croatia", "Estonia", "France", 
-#   "Germany", "Greece", "Netherlands", "Portugal", "Slovakia", "Spain"
-#   )
-# 
-# # transformed data -------------------------------------------------------------
-# df_global <- df |> 
-#   filter(
-#     country %in% countries_of_interest,
-#     date > as.Date("2020-01-01"),
-#     region == "All",
-#     sex == "b"
-#   ) |> 
-#   group_by(country, date) |> 
-#   summarise(cumulative_cases_cal = sum(cases)) |> 
-#   group_by(country) |> 
-#   complete(date = full_seq(date, 1)) |> 
-#   mutate(
-#     timeline = date - lag(date),  # check if gap between dates
-#     daily_cases = round(cumulative_cases_cal - lag(cumulative_cases_cal), 0)
-#   ) |> 
-#   ungroup() |> 
-#   left_join(df_school_closure, by = c("date", "country")) |> 
-#   group_by(country) |> 
-#   fill(status) |> 
-#   ungroup() |> 
-#   mutate(schools_closed = if_else(status %in% period_closure_to_keep, 1, 0)) |> 
-#   drop_na(status) |> 
-#   left_join(df_check, by = c("date", "country"))
-# 
-# write_csv(df_global, here("data/df_global.csv"))
-# 
-# df_global_age <- df |> 
-#   filter(
-#     country %in% countries_of_interest,
-#     date > as.Date("2020-01-01"),
-#     region == "All",
-#     sex == "b"
-#   ) |> 
-#   mutate(
-#     age_grp = case_when(
-#       age == 0  ~ "aged0to4",
-#       age == 5  ~ "aged5to9",
-#       age == 10 ~ "aged10to14",
-#       age == 15 ~ "aged15to19",
-#       age == 20 ~ "aged20to24",
-#       between(age, 25, 34) ~ "aged25to34",
-#       between(age, 35, 44) ~ "aged35to44",
-#       between(age, 45, 54) ~ "aged45to54",
-#       between(age, 55, 64) ~ "aged55to64",
-#       age > 64 ~ "aged65+"
-#     )
-#   ) |> 
-#   group_by(country, date, age_grp) |> 
-#   summarise(cumulative_cases_cal = sum(cases)) |> 
-#   group_by(country, age_grp) |> 
-#   complete(date = full_seq(date, 1)) |> 
-#   mutate(
-#     timeline = date - lag(date),  # check if gap between dates
-#     daily_cases = round(cumulative_cases_cal - lag(cumulative_cases_cal), 0),
-#     daily_changes = daily_cases - lag(daily_cases)
-#   ) |> 
-#   ungroup()
-# 
-# write_csv(df_global_age, here("data/df_global_age.csv"))
-
-df_global <- read_csv(here("data/df_global.csv"))
-df_global_age <- read_csv(here("data/df_global_age.csv"))
-```
-
-```{r analyses, cache=FALSE}
-source(here("analysis/gam_analysis_europe.R"))
-source(here("analysis/te_analysis_europe.R"))
-```
-
-# Introduction
-
-A wide range of non-pharmaceutical interventions (NPI) were enacted to control the spread of COVID-19, especially before the availability of effective vaccines.
-Due to their social mixing patterns, children have been identified as an age group that can drive the spread of respiratory infections such as influenza [@moser2018estimating], and school closures were found to be effective in mitigating the spread of influenza H1N1 in Japan in 2009 [@kawano2015substantial].
-It is therefore not surprising that school closures were one of the NPIs that were introduced in many countries to stop the spread of COVID-19.
-They were used particularly during the initial wave of the pandemic and during subsequent waves when case numbers were high.
-
-Recently, a body of literature has shown various negative consequences on child development and health due to school closures.
-School closures have been found to have negatively affected child mental health [@moulin2022longitudinal], nutrition/obesity [@sugimoto2023temporal], and education [@lerkkanen2023reading].
-Some governments are now questioning their decisions to close schools [@de2021determines].
-For example, the German health minister Karl Lauterbach, in an interview with one of Germany’s publicly funded television stations, admitted that “in retrospect it had been wrong to keep schools and childcare closed for so long” [@ard2023lauterbach].
-
-Even if school closures had negative consequences for children, they might nevertheless have been effective at stopping the spread of COVID-19.
-However, analysis on the effectiveness of school closures on the spread of COVID-19 remains inconclusive.
-@alfano2022effects in a study of a panel of European countries found that school closures were associated with a reduced COVID-19 incidence.
-In contrast, @walsh2021school in a review of 40 studies covering 150 countries concluded that the effectiveness of school closures was uncertain, with 60% having identified no impact and pointing to the potential for the analysis to be affected by confounding factors and collinearity.
-The latter shortcomings suggest that it is hard to draw firm inferences, as even a positive result may not imply causality.
-
-This paper investigates the dynamics of COVID-19 across age groups and how this was affected by school closures.
-This allows us to identify the effect of school closure on the evolution of COVID-19 cases in other age groups and particularly on older and more vulnerable age groups, including parents, grandparents, and relatives about which little is known.
-
-We identify the relationship between the school closures and case numbers in different age groups using Generalized Additive Models (GAM), which are more flexible than conventional Vector Autoregressions, given that they allow for non-linear relationships between variables.
-Furthermore, compared to other methods such as Vector Autoregression, GAM can handle fixed and random smooth effects without making strict assumptions about linearity.
-Then, using the transfer entropy method, we evaluate the impact of case number changes between age groups in order to identify if school closure has secondary repercussions in the household.
-Transfer entropy can adapt to changes in the underlying system dynamics over time.
-It does not assume stationarity and can capture evolving causal relationships, making it suitable for studying dynamic systems and non-stationary processes such as COVID-19 contaminations.
-
-## Influence of School Closure
-
-Once it was clear that COVID-19 spreads from person to person, it was natural that governments were advised to enact measures to reduce social contact in order to reduce the spread of the virus.
-Therefore, it is legitimate to believe that by closing schools, a reduction of the contaminations would be observed in the younger age groups.
-However, the efficiency of school closure on the reduction of COVID-19 cases is still questioned [@bayham2020impact; @esposito2021comprehensive].
-While some research has found that school closures contribute to limit or to reduce the growth rate of confirmed cases after implementation [@stage2021shut; @yoshiyuki2020effects], others did not observe a change in the evolution of COVID-19 cases [@chang2020modelling; @iwata2020school].
-For instance, a controlled comparison between similar localities in Japan with schools closed and schools open did not reveal any evidence that school closures reduced the spread of COVID-19 [@fukumoto2021no].
-If the school closure had a real impact on the evolution of confirmed COVID-19 cases, it should be possible to observe a decrease or at least an inflection in the trend of its evolution among younger age groups.
-
-## Causal Relationship Between Age Groups
-
-A second implicit belief regarding the effect of school closure on the spread of COVID-19 is that school not only influences the spread of the virus among children and teenagers but also has a knock-on effect on the spread of the virus in older age groups, also called Secondary Attack Rate (SAR).
-The contaminated children and teenagers would bring the virus back home and would pass the virus on to their parents and relatives.
-For example, research investigating the contamination in the household network not only revealed an exceptionally high rate of secondary contamination but also that this contamination happened when the schools were closed [@sorianoarandes2021household].
-
-Despite being reported in several clinical and epidemiological studies [@siebach2021childhood; @zhendong2020clinical], multiple research have shown that the SAR from children to household members was lower than expected [@heavey2020evidence; @vanderhoek2020role; @kim2021role; @ludvigsson2020children].
-However, the SAR of children and teenagers to the household member is likely to be age-dependent, with differences between infants, primary and secondary school children, and college students [@grasleguen2021reopening].
-If a secondary transmission from children and teenagers to household member has a significant influence, then a temporal causality relationship between their evolution should be observed.
-
-To evaluate the impact of school closure on COVID-19 cases across different age groups, we first performed a non-linear time series regression to analyze the relationship between school closures and the number of COVID-19 cases in various age groups.
-In a second analysis, the transfer entropy method is employed to assess the influence of changes in case numbers between age groups, aiming to determine if school closures have secondary effects within households.
-
-# Estimation Methodology
-
-## Generalized Additive Model
-
-The effect of school closure on the trend of daily COVID-19 cases is analyzed using a Generalized Additive Model (GAM).
-GAM is a flexible modeling approach that estimates non-linear relationships between variables.
-Compared to other methods such as Vector Autoregression, GAM can handle fixed and random smooth effects without making strict assumptions about linearity.
-
-The GAM is fitted on the daily COVID-19 cases to test the hypothesis of a significant non-linear evolution of cases among age groups from 0 to 4, from 5 to 9, from 10 to 14, from 15 to 19, and from 20 to 24 [@wood2017generalized].
-The model also estimates the overall non-linear effect by country, taking into account the interaction between age groups and countries as random intercepts and the interaction between time, countries, and the period of closure as random effects (Eq 1).
-
-By estimating the degree of smoothness of a Bayesian spline smoothing using restricted fast maximum likelihood estimation [@wood2011fast], GAM identifies dynamic patterns underlying the evolution of COVID-19 cases reported while including the random effect of different age groups and countries as follows:
-
-\begin{align*}
-  \tag{1}
-  n\prime_t &\sim Poisson(\mu_t) \\
-  \log(\mu_t) &= \beta_0 + \beta_1\,wave_t + \beta_2\,country_t + \beta_3\,age\,group_t + \\
-              & f_1(closure_t) + f_2(closure_t,country_t) + f_3(closure_t,age\,group_t)
-\end{align*}
-
-\noindent where $n\prime_t$ represents the confirmed COVID-19 cases, assuming a Poisson distribution for the fitting [@loader2006local], and $t$ is the date corresponding to the confirmed COVID-19 cases.
-The response variable includes a specific random effect taking into account variation within waves of school closure, countries, and age groups.
-The terms $f_1$ to $f_3$ are smooth functions of the time since closure, the time since closure for each country, and the time since closure for each age group.
-The restricted maximum likelihood (REML) was used to avoid overfitting while estimating smoothing parameters.
-In order to accurately account for the autocorrelation arising from the time series data, the residuals are modeled using an AR1 error model.
-By incorporating the autoregressive component, the AR1 model acknowledges the dependence of each residual on its previous value, thus providing a comprehensive representation of the data’s temporal dynamics.
-
-## Transfer Entropy
-
-Transfer entropy ($T$) can be used to infer the temporal relationship between two time series $X$ and $Y$.
-This measure indicates whether changes in COVID-19 cases within a certain age group $X$ can be used to reduce the uncertainty on the future COVID-19 cases within another age group $Y$.
-If it does, $T$ is considered evidence of a causal effect from the age group $X$ to the age group $Y$ [@schreiber2000measuring].
-As such, Granger causality is a special case of transfer entropy applied to time series that are jointly Gaussian distributed [@barnett2009granger].
-Therefore, transfer entropy is a more robust analysis of time series, especially when applied to the impact of age cohorts on pandemic transmission [@kissler2020symbolic].
-
-The influence of the evolution in COVID-19 cases across all age groups is evaluated using Shannon’s transfer entropy, given by:
-
-\begin{align}
-  \tag{2}
-  T_{ag\,x \rightarrow\,ag\,y}(k,l) = \sum_{ag\,x,ag\,y} p\left(ag\,x_{t+1}, ag\,x_t^{(k)}, ag\,y_t^{(l)}\right) \log \left(\frac{p\left(ag\,x_{t+1}| ag\,x_t^{(k)}, ag\,y_t^{(l)}\right)}{p\left(ag\,x_{t+1}|ag\,x_t^{(k)}\right)}\right)
-\end{align}
-
-\noindent where $T_{ag\,x \rightarrow\,ag\,y}$ consequently measures the influence of the change dynamic from an age group $X$ (or $ag\,x$) to another age group $Y$ (or $ag\,y$) for every country (Eq 2).
-
-The day-by-day difference in COVID-19 confirmed cases $n\prime\prime_t$ is used to satisfy the stationary requirement for the calculation of Shannon’s Transfer Entropy [@shannon1948mathematical; @behrendt2019rtransferentropy].
-
-# Data
-
-## COVID-19 cases per age group
-
-For the analysis, data on COVID case counts by age group are required.
-The number of COVID-19 cases per age group has been taken from the COVerAGE-DB project [@riffe2021data].
-The COVerAGE-DB project consists of 3 data files: an “Input” data file that collects the official COVID-19 cases from 117 countries, the “Output_5” data file, which is a projection of COVID-19 cases by a group of 5 years, and the “Output_10” data file, which is a projection of COVID-19 cases by a group of 10 years.
-For our analyses, we have used the “Output_5,” which deals with the heterogeneity of countries’ reporting formats by using spline approximations when the data for this age bracket is not available for a country.
-Therefore, it allows a precise analysis between age groups under 24, while age groups of 25 and above are reduced to 10-year brackets, and groups above 64 are concatenated all together.
-
-In order to cross-validate the data obtained after spline approximations, a comparison with the data published by the @who2022https reveals perfect similarities.
-The original data consist of 14,089,320 observations of 10 variables (117 distinct countries, region within the country, a unique observation code, the date of the observation, the gender, which can be male, female, or both, the age bracket by 5 years from 0 to 100, a confirmation of the age interval for each bracket, the total number of cases so far, the total number of deaths, and the total number of tests performed) from `r format(min(df_global$date), "%B %d, %Y")` to `r format(max(df_global$date), "%B %d, %Y")`.
-After removing countries with missing and inconsistent values, only 22 are suitable for data analyses.
-However, to focus this analysis on geographically and culturally comparable countries, only 12 European countries are kept: Austria, Belgium, Bulgaria, Croatia, Estonia, France, Germany, Greece, Netherlands, Portugal, Slovakia, and Spain.
-The observations are reported in terms of the total number of COVID-19 cases per day from the start of the pandemic.
-The daily number of cases at a specific date $n\prime_{t}$ is calculated with the difference between the total cases at a date $t$ and the total cases at a date $t-1$ (i.e., derivative 1).
-In addition, the change in the daily number of cases $n\prime\prime_{t}$ between $n\prime_{t}$ and $n\prime_{t-1}$ has also been calculated (i.e., derivative 2).
-
-## School closure
-
-To identify the impact of school closures on the number of cases, information regarding the school closures on a day-by-day basis is required, and this is taken from the “UNESCO global education coalition” (2022).
-For each day, in each country, the status of the schools is indicated as fully open, partially open, closed due to COVID-19, or closed due to an academic break.
-Because it would be difficult to measure the effect of school closures at a country level when schools are partially closed, only closures due to COVID-19 or due to an academic break are considered.
-Indeed, both are considered as closure at a country-wide level.
-Note that periods of closure longer than 21 days are compared for 28 days, regardless of whether the schools opened or not after 21 days.
-Any COVID-19 case numbers beyond the 28-day period are not relevant for evaluating the impact of school closures.
-
-# Results
-
-The data show that the trend of confirmed COVID-19 cases follows similar patterns across the selected European countries, with scales following the size of the population in these countries (Figure \@ref(fig:overall)).
-Thus, the bigger the country, the higher the total number of COVID-19 cases.
-
-```{r overall, fig.height=3, out.width="\\textwidth", fig.cap="(ref:overall)"}
-overall_plot <- df_global |> 
-  mutate(
-    country = fct_rev(country)
-  ) |> 
-  ggplot(aes(date, cumulative_cases_cal, group = country)) +
-  geom_line(na.rm = TRUE, color = "gray40", alpha = 0.4) +
-  geom_text_repel(
-    data = df_global |>
-      group_by(country) |>
-      slice_max(cumulative_cases_cal, with_ties = FALSE),
-    aes(label = country),
-    hjust = 0,
-    direction = "y",
-    segment.color = NA,
-    seed = 123,
-    size = 2,
-    force = 0.5,
-    family = "serif"
-  ) +
-  scale_x_date(
-    "Date", 
-    date_breaks = "3 months", 
-    date_labels = "%b %y", 
-    limits = c(as.Date("2020-02-01"), as.Date("2022-05-01"))
-  ) +
-  scale_y_continuous(
-    "Total COVID-19 Cases (Million)", 
-    labels = unit_format(unit = "M", scale = 1e-6)
-  ) +
-  theme_bw() +
-  theme(
-    legend.position = "none",
-    text = element_text(size = 10, family = "serif")
-  )
-
-overall_plot
-```
-
-The evolution of COVID-19 cases reveals some similarities across all age groups.
-However, the influence of each wave on individual age groups also has some particularities (Figure \@ref(fig:descriptive)).
-For example, the first wave was more important among the oldest age groups, whereas the third wave was more important among the youngest age groups.
-
-```{r descriptive, fig.height=3, out.width="\\textwidth", fig.cap="(ref:descriptive)"}
-descriptive_plot <- df_global |> 
-  filter(schools_closed == 1) |> 
-  mutate(
-    status = status |> 
-      str_remove("Closed due to ") |> 
-      fct_rev(),
-    country = country |> 
-      fct_rev()
-  ) |> 
-  ggplot(aes(date, country, color = status)) +
-  geom_point(na.rm = TRUE) +
-  scale_x_date("Date", date_breaks = "3 months", date_labels = "%b %y") +
-  scale_y_discrete("Selected European Countries") +
-  scale_colour_manual("Schools Closure Reason", values = c("gray10", "gray60")) +
-  theme_bw() +
-  theme(
-    legend.position = "bottom",
-    text = element_text(size = 10, family = "serif")
-  )
-
-descriptive_plot
-```
-
-As outlined above, to evaluate the shape of the trend in the numbers of COVID-19 cases reported after the three school closures longer than 21 consecutive days, a GAM was fitted as described above, taking into account the overall effect across all the selected European countries, as well as the effect for age groups: 0 to 4, 5 to 9, 10 to 14, 15 to 19, and 20 to 24 years old.
-The obtained results satisfy the requirements to fit this model, which explains `r percent(gam_summary$dev.expl, accuracy = 0.1)` of the variation in COVID-19 cases.
-
-Overall, the results revealed a decreasing non-linear effect of school closure at a country level for the selected European countries (Austria: `r gam_results["Austria"]`; Belgium: `r gam_results["Belgium"]`; Bulgaria: `r gam_results["Bulgaria"]`; Croatia: `r gam_results["Croatia"]`; Estonia: `r gam_results["Estonia"]`; France: `r gam_results["France"]`; Germany (`r gam_results["Germany"]`); Greece: `r gam_results["Greece"]`; The Netherlands: `r gam_results["Netherlands"]`; Portugal: `r gam_results["Portugal"]`; Slovakia (`r gam_results["Slovakia"]`; Spain: `r gam_results["Spain"]`, see Figure \@ref(fig:age)).
-
-```{r country, fig.height=5, out.width="\\textwidth", fig.cap="(ref:country)"}
-gam_plot_country
-```
-
-Considering all countries, the analysis of age groups reveals distinct patterns (Figure \@ref(fig:age)).
-For the 0 to 4 age group, i.e., the pre-school group, there was a downward trend during the initial two weeks following school closure, followed by flattening off (`r gam_results["0to4"]`).
-In contrast, for the age group ranging from 5 to 14, a substantial increase in cases after 14 days of school closure is found (5 to 9 age group: `r gam_results["5to9"]` and 10 to 14 age group: `r gam_results["10to14"]`).
-For age groups between 15 and 24, there is a notable surge immediately after school closure, followed by a decrease after 14 days (15 to 19 age group: `r gam_results["15to19"]` and 20 to 24 age group: `r gam_results["20to24"]`).
-Thus, while the total case number declined following school closures, among younger age groups only the pre-school age group had a consistent reduction in cases, while for other age groups, there was even an increase in cases following school closures.
-
-```{r age, fig.height=3, out.width="\\textwidth", fig.cap="(ref:age)"}
-gam_plot_age
-```
-
-Given that school closures were not only aimed at reducing COVID-19 cases among school-going children but also age groups, it is important to test the degree to which there was transmission across age groups, which is done using transfer entropy analysis.
-Before this analysis is carried out, an Augmented Dickey-Fuller has been applied to each age group to ensure that the daily changes in COVID-19 cases are stationary (see Appendix 1).
-
-The results of the transfer entropy calculations between age groups for each of the 12 selected European countries are reported in Figure \@ref(fig:te).
-The absence of symmetry between influencing age groups (i.e., $ag\,x$) and influenced age groups (i.e., $ag\,y$) is found.
-Indeed, the change in COVID-19 cases in some age groups is influenced by other age groups, but they are not reciprocally influencing these age groups.
-Figure \@ref(fig:te) shows how different age cohorts influence the COVID-19 case numbers of all age groups across countries.
-The upper left quadrant indicates how younger age cohorts are influencing older age cohorts, the bottom right quadrant indicates how older age cohorts are influencing younger age cohorts, finally, the lower left and upper right indicate how younger or older age cohorts are influencing themselves.
-By analyzing these quadrants, it is possible to identify similar patterns across multiple countries.
-Indeed, it appears that Austria, Germany, and The Netherlands have significantly higher $T$ coefficients in the upper left quadrant of the matrix, which indicates that the daily changes in COVID-19 cases number in younger cohorts are predicting the daily changes in COVID-19 cases number in older cohorts.
-Alternatively, it appears that Austria, the Netherlands, Portugal, and Spain have significantly higher $T$ coefficients in the lower right quadrant of the matrix, which indicates that the daily changes in COVID-19 cases number in older cohorts are predicting the daily changes in COVID-19 cases number in younger cohorts.
-
-```{r te, fig.height=8, out.width="\\textwidth", fig.cap="(ref:te)"}
-te_plot
-```
-
-# Discussion & Conclusion
-
-Knowledge about the transmission of the virus significantly improved over time as more studies have been published.
-While an early study found that children did not play an important role in the transmission of the virus [@li2020role], more recent results give a more nuanced position, stating that the spread of the virus in children is moderate.
-This study, using robust statistical methods, considered the effect of school closures on COVID-19 cases across age groups.
-While there are commonalities in the evolution of COVID-19 cases across all European countries that were included in our data, school closures exhibit distinct impacts on different age groups.
-Notably, the analysis shows that the 0 to 4 age group experiences a downward trend in COVID-19 cases during the initial two weeks following school closure, followed by stabilization.
-In contrast, age groups ranging from 5 to 14 exhibit a substantial increase in cases after 14 days of school closure.
-Additionally, age groups between 15 and 24 demonstrate a notable surge immediately after school closure, followed by a decrease after 14 days.
-Thus, the results do not support the hypothesis that school closures were effective.
-These results partially replicate observations from @alfano2022effects while providing a clearer picture of the structure of the effect of school closure.
-
-The Transfer Entropy calculations between age groups for each of the 12 selected European countries reveal an absence of symmetry between influencing age groups (X) and influenced age groups (Y).
-Some age groups influence the COVID-19 case numbers of other age groups, but there is no reciprocal influence.
-This finding suggests that changes in COVID-19 cases in certain age groups can predict the changes in other age groups but not vice versa, i.e., causality can be unidirectional.
-These findings highlight the importance of considering intergenerational interactions in designing effective control measures.
-
-Furthermore, the quadrant analysis reveals similar patterns across multiple countries.
-Austria, Germany, and the Netherlands exhibit significantly higher TE coefficients in the upper left quadrant, indicating that changes in COVID-19 cases in younger cohorts predict the changes in older cohorts.
-On the other hand, Austria, the Netherlands, Portugal, and Spain show higher TE coefficients in the lower right quadrant, indicating that changes in COVID-19 cases in older cohorts predict the changes in younger cohorts.
-
-In conclusion, this study sheds light on the trends and patterns of confirmed COVID-19 cases in selected European countries, with a particular focus on the influence of age groups and school closures.
-While a reduction in overall COVID-19 cases was found during school closures, which might suggest that like lockdowns [@alfano2020efficacy; @molefi2021impact], prolonged school closures could contribute to mitigating the spread of the virus, the results for the school-going age groups suggest that this is not the case.
-
-# Data Availability
-
-All data extracted from the COVerAGE-DB project and from the UNESCO global education coalition that are used to produce these results are fully available here https://github.com/damien-dupre/covid_ts_causality.
-The repository also includes the R code that were used to produce these results.
-
-# References {-}

### 2023-12-18 manuscript_elsevier/cover_letter.Rmd
Social Science & Medicine submission

---
title: "Submission to Social Science & Medicine"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  error = FALSE, 
  echo = FALSE)
journal <- "Social Science & Medicine"
```

Dear editor of `r journal`,

Please find enclosed our manuscript to be considered for publication entitled “Examining the Impact of School Closures on COVID-19 Infections in Europe and their Effects on Different Age Cohorts”.

This research endeavours to contribute valuable insights into the dynamics of COVID-19 in European countries, specifically focusing on the influence of school closures on different age groups. Our study employs a meticulous analysis of COVID-19 cases and their trends across selected European nations. Utilising a Generalized Additive Model (GAM) and Transfer Entropy (TE), our study seeks to unravel the multifaceted impact of school closures on the overall case numbers and, more importantly, within specific age cohorts.

The central findings of our research underscore a nuanced narrative regarding the effectiveness of school closures. Contrary to expectations, our analysis reveals a decreasing non-linear effect of school closure on the total COVID-19 case numbers across the studied countries. A particularly noteworthy revelation is the inconsistent impact on different age groups. While the 0 to 4 pre-school age group consistently exhibits a downward trend in infections, the school-going age groups of 5 to 14 show a significant increase in cases. Age groups 15 to 24, on the other hand, experience a surge immediately after closure, followed by a subsequent decline.

Moreover, our study employs Transfer Entropy calculations to highlight the asymmetry in age group influences, revealing that changes in COVID-19 cases in certain age groups predict changes in other age groups, but not vice versa. This unique contribution sheds light on the intricate dynamics of COVID-19 spread among various age cohorts, providing a deeper understanding of the virus's behaviour and offering critical insights for public health strategies and interventions.

The meticulous methodology and robust analysis presented in our manuscript make a compelling case for its inclusion in Social Science & Medicine. We believe that our findings have the potential to significantly contribute to the existing body of knowledge on COVID-19 dynamics in European countries and provide actionable insights for policymakers and public health professionals.

All authors disclose any actual or potential conflict of interest including any financial, personal or other relationships with other people or organizations that could inappropriately influence, or be perceived to influence, their work.

This manuscript has not been published elsewhere, accepted for publication elsewhere or under editorial review for publication elsewhere. My coauthors and I have complied with the ethical guidelines in the treatment of the data and preparation of the manuscript. We had no prior discussions with an Editorial Board Member about the work described in the manuscript.

The list of authors:

* Damien Dupré (corresponding author and main investigator of the study)
  * Dublin City University
  * damien.dupre@dcu.ie
* Edgar Morgenroth
  * Dublin City University
  * edgar.morgenroth@dcu.ie

All future correspondence can be directed to me at the e-mail address provided in this submission. If you have any questions or concerns pertaining to this manuscript, please do not hesitate to contact me. Thank you for your time and consideration, and we look forward to hearing back from you.

We look forward to the opportunity to share our research with the wider academic community through the esteemed platform of Social Science & Medicine. Thank you for considering our submission, and we are optimistic about the positive impact our study can have on the discourse surrounding COVID-19 and public health strategies.

Best regards,

Damien Dupré

### 2023-12-18 manuscript_elsevier/highlights.Rmd
Social Science & Medicine submission

---
title: "Submission to Social Science & Medicine"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  error = FALSE, 
  echo = FALSE)
journal <- "Social Science & Medicine"
```

The research paper, titled "Examining the Impact of School Closures on COVID-19 Infections in Europe and their Effects on Different Age Cohorts," analyzes COVID-19 trends in various age groups across selected European countries. The study employs a Generalized Additive Model (GAM) to evaluate the effects of school closures on COVID-19 cases and utilizes Transfer Entropy calculations to assess asymmetry in age group influences.

The findings reveal a decreasing non-linear effect of school closure on the total COVID-19 case numbers across all studied countries. Notably, a consistent downward trend in infections is observed in the 0 to 4 pre-school age group following school closures. In contrast, the school-going age groups (5 to 14) exhibit a significant increase in cases, while age groups 15 to 24 show a surge immediately after closure, followed by a decline. Transfer Entropy analysis indicates that changes in COVID-19 cases in certain age groups predict changes in others but not vice versa, highlighting the complexity of intergenerational interactions.

The paper discusses the broader context of the impact of school closures on child development, mental health, nutrition, and education, referencing conflicting opinions on their effectiveness in controlling COVID-19 spread. It explores the belief that school closures not only influence the spread among children but also have a knock-on effect on older age groups (Secondary Attack Rate), considering the age-dependent nature of secondary transmission.

The estimation methodology, involving Generalized Additive Models (GAM) and Transfer Entropy methods, reflects the paper's commitment to robust analysis. Leveraging data from the COVerAGE-DB project and school closure information from the UNESCO global education coalition adds a layer of credibility to the research, ensuring a solid foundation for its findings.

The paper concludes that, while overall COVID-19 cases decrease after school closures, school-going age groups exhibit an increase, emphasizing the need to consider intergenerational interactions in designing effective control measures. In summary, this research provides valuable insights into the intricate dynamics of COVID-19 transmission in different age groups and the nuanced effects of school closures in European countries.

### 2023-12-18 manuscript_elsevier/manuscript_elsevier.Rmd
Social Science & Medicine submission

@@
 library(knitr)
+library(kableExtra)
 library(osfr)
@@
 # References {-}
+
+<div id="refs"></div>
+
+\newpage
+
+# Appendix 1: Dickey-Fuller test for stationay timeseries {-}
+
+Test for the null hypothesis of a unit root of a univarate time series (equivalently, non-stationary time series).
+
+The Augmented Dickey-Fuller test incorporates three types of linear regression models.
+The first type is a linear model with no drift and linear trend with respect to time.
+The second type is a linear model with drift but no linear trend.
+The third type is a linear model with both drift and linear trend.
+
+The p.value is calculated by interpolating the test statistics from the corresponding critical values tables (Fuller, 1996) for each type of linear models with given sample size.
+
+```{r}
+adf_res_type1 <- df_global_age |>
+  group_by(age_grp, country) |>
+  group_modify(~ aTSA::adf.test(.x$daily_changes, output = FALSE, nlag = 2) |>
+                 bind_rows() |> 
+                 pluck("type1") |> 
+                 as_tibble()
+  ) |> 
+  rename(ADF_type1 = ADF, p.value_type1 = p.value)
+
+adf_res_type2 <- df_global_age |>
+  group_by(age_grp, country) |>
+  group_modify(~ aTSA::adf.test(.x$daily_changes, output = FALSE, nlag = 2) |>
+                 bind_rows() |> 
+                 pluck("type2") |> 
+                 as_tibble()
+  ) |> 
+  rename(ADF_type2 = ADF, p.value_type2 = p.value)
+
+adf_res_type3 <- df_global_age |>
+  group_by(age_grp, country) |>
+  group_modify(~ aTSA::adf.test(.x$daily_changes, output = FALSE, nlag = 2) |>
+                 bind_rows() |> 
+                 pluck("type3") |> 
+                 as_tibble()
+  ) |> 
+  rename(ADF_type3 = ADF, p.value_type3 = p.value)
+
+adf_res <- adf_res_type1 |> 
+  full_join(adf_res_type2) |> 
+  full_join(adf_res_type3)
+```
+
+\newpage
+
+```{r}
+adf_res |>
+  filter(age_grp %in% c("aged0to4", "aged5to9")) |> 
+  kbl(
+    caption = "Dickey-Fuller test for all combinaisons of countries and age groups investigated",
+    col.names = c("Age Group", "Country", "Lag", "ADF", "p", "ADF", "p", "ADF", "p")
+  ) |> 
+  kable_classic(full_width = F) |> 
+  add_header_above(c(" " = 3, "Type 1" = 2, "Type 2" = 2, "type 3" = 2)) |>
+  collapse_rows(columns = 1:2, valign = "middle")
+```
+
+\newpage
+
+```{r}
+adf_res |>
+  filter(age_grp %in% c("aged10to14", "aged15to19")) |> 
+  kbl(
+    caption = "Dickey-Fuller test for all combinaisons of countries and age groups investigated",
+    col.names = c("Age Group", "Country", "Lag", "ADF", "p", "ADF", "p", "ADF", "p")
+  ) |> 
+  kable_classic(full_width = F) |> 
+  add_header_above(c(" " = 3, "Type 1" = 2, "Type 2" = 2, "type 3" = 2)) |>
+  collapse_rows(columns = 1:2, valign = "middle")
+```
+
+\newpage
+
+```{r}
+adf_res |>
+  filter(age_grp %in% c("aged20to24", "aged25to34")) |> 
+  kbl(
+    caption = "Dickey-Fuller test for all combinaisons of countries and age groups investigated",
+    col.names = c("Age Group", "Country", "Lag", "ADF", "p", "ADF", "p", "ADF", "p")
+  ) |> 
+  kable_classic(full_width = F) |> 
+  add_header_above(c(" " = 3, "Type 1" = 2, "Type 2" = 2, "type 3" = 2)) |>
+  collapse_rows(columns = 1:2, valign = "middle")
+```
+
+\newpage
+
+```{r}
+adf_res |>
+  filter(age_grp %in% c("aged35to44", "aged45to54")) |> 
+  kbl(
+    caption = "Dickey-Fuller test for all combinaisons of countries and age groups investigated",
+    col.names = c("Age Group", "Country", "Lag", "ADF", "p", "ADF", "p", "ADF", "p")
+  ) |> 
+  kable_classic(full_width = F) |> 
+  add_header_above(c(" " = 3, "Type 1" = 2, "Type 2" = 2, "type 3" = 2)) |>
+  collapse_rows(columns = 1:2, valign = "middle")
+```
+
+\newpage
+
+```{r}
+adf_res |>
+  filter(age_grp %in% c("aged55to64", "aged65+")) |> 
+  kbl(
+    caption = "Dickey-Fuller test for all combinaisons of countries and age groups investigated",
+    col.names = c("Age Group", "Country", "Lag", "ADF", "p", "ADF", "p", "ADF", "p")
+  ) |> 
+  kable_classic(full_width = F) |> 
+  add_header_above(c(" " = 3, "Type 1" = 2, "Type 2" = 2, "type 3" = 2)) |>
+  collapse_rows(columns = 1:2, valign = "middle")
+```
+
+\newpage
+
+# Appendix 2: Significance of Transfer Entropy coefficients
+
+Results for each permutation of age groups from each country.
+Tables include transfer entropy estimates, the effective transfer entropy estimates, standard errors and p-values based on bootstrap replications of the Markov chains under the null hypothesis of statistical independence.
+
+```{r}
+# variables --------------------------------------------------------------------
+age_order <- c(
+  "aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24",
+  "aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+"
+)
+
+# data -------------------------------------------------------------------------
+te_results <- read_csv(here("data/te_results.csv")) |> 
+    mutate(
+      X = factor(X, levels = age_order),
+      Y = factor(Y, levels = age_order)
+    )
+```
+
+```{r}
+te_results |>
+  filter(X == "aged0to4") |> 
+  filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |> 
+  arrange(Y) |> 
+  kbl(
+    caption = "Significance of Transfer Entropy coefficients",
+    digits = 2
+  ) |> 
+  kable_classic(full_width = F) |> 
+  collapse_rows(columns = 1:2, valign = "middle") |> 
+  kable_styling(font_size = 7)
+```
+
+\newpage
+
+```{r}
+te_results |>
+  filter(X == "aged0to4") |> 
+  filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |> 
+  arrange(Y) |> 
+  kbl(
+    caption = "Significance of Transfer Entropy coefficients",
+    digits = 2
+  ) |> 
+  kable_classic(full_width = F) |> 
+  collapse_rows(columns = 1:2, valign = "middle") |> 
+  kable_styling(font_size = 7)
+```
+
+\newpage
+
+```{r}
+te_results |>
+  filter(X == "aged5to9") |> 
+  filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |> 
+  arrange(Y) |> 
+  kbl(
+    caption = "Significance of Transfer Entropy coefficients",
+    digits = 2
+  ) |> 
+  kable_classic(full_width = F) |> 
+  collapse_rows(columns = 1:2, valign = "middle") |> 
+  kable_styling(font_size = 7)
+```
+
+\newpage
+
+```{r}
+te_results |>
+  filter(X == "aged5to9") |> 
+  filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |> 
+  arrange(Y) |> 
+  kbl(
+    caption = "Significance of Transfer Entropy coefficients",
+    digits = 2
+  ) |> 
+  kable_classic(full_width = F) |> 
+  collapse_rows(columns = 1:2, valign = "middle") |> 
+  kable_styling(font_size = 7)
+```
+
+\newpage
+```{r}
+te_results |>
+  filter(X == "aged10to14") |> 
+  filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |> 
+  arrange(Y) |> 
+  kbl(
+    caption = "Significance of Transfer Entropy coefficients",
+    digits = 2
+  ) |> 
+  kable_classic(full_width = F) |> 
+  collapse_rows(columns = 1:2, valign = "middle") |> 
+  kable_styling(font_size = 7)
+```
+
+\newpage
+
+```{r}
+te_results |>
+  filter(X == "aged10to14") |> 
+  filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |> 
+  arrange(Y) |> 
+  kbl(
+    caption = "Significance of Transfer Entropy coefficients",
+    digits = 2
+  ) |> 
+  kable_classic(full_width = F) |> 
+  collapse_rows(columns = 1:2, valign = "middle") |> 
+  kable_styling(font_size = 7)
+```
+
+\newpage
+
+```{r}
+te_results |>
+  filter(X == "aged15to19") |> 
+  filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |> 
+  arrange(Y) |> 
+  kbl(
+    caption = "Significance of Transfer Entropy coefficients",
+    digits = 2
+  ) |> 
+  kable_classic(full_width = F) |> 
+  collapse_rows(columns = 1:2, valign = "middle") |> 
+  kable_styling(font_size = 7)
+```
+
+\newpage
+
+```{r}
+te_results |>
+  filter(X == "aged15to19") |> 
+  filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |> 
+  arrange(Y) |> 
+  kbl(
+    caption = "Significance of Transfer Entropy coefficients",
+    digits = 2
+  ) |> 
+  kable_classic(full_width = F) |> 
+  collapse_rows(columns = 1:2, valign = "middle") |> 
+  kable_styling(font_size = 7)
+```
+
+\newpage
+
+```{r}
+te_results |>
+  filter(X == "aged20to24") |> 
+  filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |> 
+  arrange(Y) |> 
+  kbl(
+    caption = "Significance of Transfer Entropy coefficients",
+    digits = 2
+  ) |> 
+  kable_classic(full_width = F) |> 
+  collapse_rows(columns = 1:2, valign = "middle") |> 
+  kable_styling(font_size = 7)
+```
+
+\newpage
+
+```{r}
+te_results |>
+  filter(X == "aged20to24") |> 
+  filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |> 
+  arrange(Y) |> 
+  kbl(
+    caption = "Significance of Transfer Entropy coefficients",
+    digits = 2
+  ) |> 
+  kable_classic(full_width = F) |> 
+  collapse_rows(columns = 1:2, valign = "middle") |> 
+  kable_styling(font_size = 7)
+```
+
+\newpage
+
+```{r}
+te_results |>
+  filter(X == "aged25to34") |> 
+  filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |> 
+  arrange(Y) |> 
+  kbl(
+    caption = "Significance of Transfer Entropy coefficients",
+    digits = 2
+  ) |> 
+  kable_classic(full_width = F) |> 
+  collapse_rows(columns = 1:2, valign = "middle") |> 
+  kable_styling(font_size = 7)
+```
+
+\newpage
+
+```{r}
+te_results |>
+  filter(X == "aged25to34") |> 
+  filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |> 
+  arrange(Y) |> 
+  kbl(
+    caption = "Significance of Transfer Entropy coefficients",
+    digits = 2
+  ) |> 
+  kable_classic(full_width = F) |> 
+  collapse_rows(columns = 1:2, valign = "middle") |> 
+  kable_styling(font_size = 7)
+```
+
+\newpage
+
+```{r}
+te_results |>
+  filter(X == "aged35to44") |> 
+  filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |> 
+  arrange(Y) |> 
+  kbl(
+    caption = "Significance of Transfer Entropy coefficients",
+    digits = 2
+  ) |> 
+  kable_classic(full_width = F) |> 
+  collapse_rows(columns = 1:2, valign = "middle") |> 
+  kable_styling(font_size = 7)
+```
+
+\newpage
+
+```{r}
+te_results |>
+  filter(X == "aged35to44") |> 
+  filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |> 
+  arrange(Y) |> 
+  kbl(
+    caption = "Significance of Transfer Entropy coefficients",
+    digits = 2
+  ) |> 
+  kable_classic(full_width = F) |> 
+  collapse_rows(columns = 1:2, valign = "middle") |> 
+  kable_styling(font_size = 7)
+```
+
+\newpage
+
+```{r}
+te_results |>
+  filter(X == "aged45to54") |> 
+  filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |> 
+  arrange(Y) |> 
+  kbl(
+    caption = "Significance of Transfer Entropy coefficients",
+    digits = 2
+  ) |> 
+  kable_classic(full_width = F) |> 
+  collapse_rows(columns = 1:2, valign = "middle") |> 
+  kable_styling(font_size = 7)
+```
+
+\newpage
+
+```{r}
+te_results |>
+  filter(X == "aged45to54") |> 
+  filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |> 
+  arrange(Y) |> 
+  kbl(
+    caption = "Significance of Transfer Entropy coefficients",
+    digits = 2
+  ) |> 
+  kable_classic(full_width = F) |> 
+  collapse_rows(columns = 1:2, valign = "middle") |> 
+  kable_styling(font_size = 7)
+```
+
+\newpage
+
+```{r}
+te_results |>
+  filter(X == "aged55to64") |> 
+  filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |> 
+  arrange(Y) |> 
+  kbl(
+    caption = "Significance of Transfer Entropy coefficients",
+    digits = 2
+  ) |> 
+  kable_classic(full_width = F) |> 
+  collapse_rows(columns = 1:2, valign = "middle") |> 
+  kable_styling(font_size = 7)
+```
+
+\newpage
+
+```{r}
+te_results |>
+  filter(X == "aged55to64") |> 
+  filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |> 
+  arrange(Y) |> 
+  kbl(
+    caption = "Significance of Transfer Entropy coefficients",
+    digits = 2
+  ) |> 
+  kable_classic(full_width = F) |> 
+  collapse_rows(columns = 1:2, valign = "middle") |> 
+  kable_styling(font_size = 7)
+```
+
+\newpage
+
+```{r}
+te_results |>
+  filter(X == "aged65+") |> 
+  filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |> 
+  arrange(Y) |> 
+  kbl(
+    caption = "Significance of Transfer Entropy coefficients",
+    digits = 2
+  ) |> 
+  kable_classic(full_width = F) |> 
+  collapse_rows(columns = 1:2, valign = "middle") |> 
+  kable_styling(font_size = 7)
+```
+
+\newpage
+
+```{r}
+te_results |>
+  filter(X == "aged65+") |> 
+  filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |> 
+  arrange(Y) |> 
+  kbl(
+    caption = "Significance of Transfer Entropy coefficients",
+    digits = 2
+  ) |> 
+  kable_classic(full_width = F) |> 
+  collapse_rows(columns = 1:2, valign = "middle") |> 
+  kable_styling(font_size = 7)
+```

### 2024-02-19 manuscript_rss/appendix_2.Rmd
submission RSS

--- 2023-12-18 manuscript_elsevier/appendix_2.Rmd
@@
   )
+```
 
+```{r}
 # variables --------------------------------------------------------------------
@@
       Y = factor(Y, levels = age_order)
-    )
+    ) |> 
+  select(-direction)
 ```

### 2024-02-19 manuscript_rss/cover_letter.Rmd
submission RSS

--- 2023-12-18 manuscript_elsevier/cover_letter.Rmd
@@
 ---
-title: "Submission to Social Science & Medicine"
+title: "Submission to the Journal of the Royal Statistical Society, Series C (Applied Statistics) "
 output: pdf_document
@@
   echo = FALSE)
-journal <- "Social Science & Medicine"
+journal <- "Journal of the Royal Statistical Society, Series C (Applied Statistics)"
 ```

### 2024-02-19 manuscript_rss/manuscript.Rmd
submission RSS

--- 2023-12-18 manuscript_elsevier/manuscript_elsevier.Rmd
@@
 ---
-title: Examining the Impact of School Closures on COVID-19 Infections in Europe and their Effects on Different Age Cohorts
-author:
-  - name: Damien Dupré
-    email: damien.dupre@dcu.ie
-    affiliation: Business School, Dublin City University
-    correspondingauthor: true
-  - name: Edgar Morgenroth
-    email: edgar.morgenroth@dcu.ie
-    affiliation: Business School, Dublin City University
-address:
-  - code: Business School, Dublin City University
-    organization: Business School, Dublin City University
-    city: Dublin
-    country: Ireland
+title: "Examining the Impact of School Closures on COVID-19 Infections in Europe and their Effects on Different Age Cohorts"
+short_title: "Impact of School Closures on COVID-19 by Age Cohorts"
+short_author: "Author 1 et. al."
+authors: 
+- name: "Author 1"
+  affiliation: "Address"
+  city: "City"
+  country: "Country"
+  email: "email1\\@example.com"
+- name: "Author 2"
+  affiliation: "Address2"
+  city: "City2"
+  country: "Country2"
+  email: "email2\\@example.com"
+# short_author: "Dupré & Morgenroth"
+# authors:
+# - name: "Damien Dupré"
+#   affiliation: "Business School, Dublin City University"
+#   city: "Dublin"
+#   country: "Ireland"
+#   email: "damien.dupre\\@dcu.ie"
+#   footnote: 1
+# - name: "Edgar Morgenroth"
+#   affiliation: "Business School, Dublin City University"
+#   city: "Dublin"
+#   country: "Ireland"
+#   email: "edgar.morgenroth\\@dcu.ie"
+#   footnote: 2
+# footnote:
+#   - code: 1
+#     text: "Corresponding Author"
+#   - code: 2
+#     text: "Equal contribution"
 abstract: |
-  This paper analyzes the trends of COVID-19 cases in different age groups in selected European countries, assessing to what extent school closures helped reduce case numbers, how school closures affected case numbers in different age groups, and how COVID-19 spread between age groups.
-The study also evaluates the effects of school closures on COVID-19 cases using a Generalized Additive Model (GAM).
-The findings demonstrate a decreasing non-linear effect of school closure on the total COVID-19 case number across all countries studied.
-However, the analysis of age groups only reveals a consistent downward trend in infections in the 0 to 4 pre-school age group, while the school-going age groups 5 to 14 exhibit a significant increase in cases.
-Age groups 15 to 24 show a surge immediately after closure, followed by a decline.
-Transfer Entropy calculations highlight asymmetry in age group influences, indicating that changes in COVID-19 cases in certain age groups predict changes in other age groups but not vice versa.
-These findings contribute to a better understanding of COVID-19 dynamics in European countries and provide insights for public health strategies and interventions.
-keywords: 
-  - COVID-19
-  - School Closures
-  - Intergenerational Transmission.
-journal: "Social Science & Medicine"
-date: "`r Sys.Date()`"
-classoption: preprint, 3p, authoryear
-linenumbers: false
-numbersections: true
-bibliography: bibliography.bib
-csl: https://www.zotero.org/styles/elsevier-harvard
-output:
-  bookdown::pdf_book:
-    base_format: rticles::elsevier_article
-    keep_tex: false
-    citation_package: default
+  This paper provides a thorough analysis of COVID-19 case trends in selected European countries, focusing on the influence of different age groups.
+It explores the relationship between population size and confirmed case numbers, suggesting that larger countries tend to have higher case figures.
+The study investigates how various waves of the pandemic affect different age demographics, with older age groups being most affected by the initial wave and younger age groups more impacted by subsequent waves.
+  Using a Generalised Additive Model (GAM), the research assesses the impact of school closures on COVID-19 cases, revealing a decreasing non-linear effect across all countries examined.
+Following school closures, distinct trends emerge among age groups: ages 0 to 4 experience a decrease, ages 5 to 14 see an increase, and ages 15 to 24 initially surge then decline.
+Transfer Entropy calculations uncover imbalances in age group influences, with certain age groups predicting changes in others.
+  Consistent patterns across countries are identified, such as younger cohorts influencing older cohorts in Austria, Germany, and the Netherlands, while Portugal and Spain show the opposite trend.
+These findings deepen our understanding of COVID-19 dynamics in Europe, providing valuable insights for targeted public health strategies and interventions.
+keywords: COVID-19, School Closures, Causality, Generalised Additive Model, Transfer Entropy
+bibliography: "bibliography.bib"
 header-includes:
@@
   - \usepackage{caption}
+
+output:
+  bookdown::pdf_book:
+    base_format: rticles::rss_article
+    keep_tex: false
+  bookdown::word_document2:
+    default
+knit: (function(input, ...) {
+    rmarkdown::render(
+      input,
+      output_dir = here::here("output"),
+      output_file = c(glue::glue("{xfun::sans_ext(input)}.pdf"), glue::glue("{xfun::sans_ext(input)}.docx")),
+      output_format = "all",
+      envir = globalenv()
+    )
+  })
 ---
@@
 # References {-}
+
+<div id="refs"></div>
+
+<!-- \newpage -->
+
+<!-- # Appendix 1: Dickey-Fuller test for stationay timeseries {-}  -->
+
+<!-- Test for the null hypothesis of a unit root of a univarate time series (equivalently, non-stationary time series). -->
+
+<!-- The Augmented Dickey-Fuller test incorporates three types of linear regression models.
+The first type is a linear model with no drift and linear trend with respect to time.
+The second type is a linear model with drift but no linear trend.
+The third type is a linear model with both drift and linear trend.
+-->
+
+<!-- The p.value is calculated by interpolating the test statistics from the corresponding critical values tables (Fuller, 1996) for each type of linear models with given sample size.  -->
+
+<!-- ```{r} -->
+<!-- adf_res_type1 <- df_global_age |> -->
+<!--   group_by(age_grp, country) |> -->
+<!--   group_modify(~ aTSA::adf.test(.x$daily_changes, output = FALSE, nlag = 2) |> -->
+<!--                  bind_rows() |>  -->
+<!--                  pluck("type1") |>  -->
+<!--                  as_tibble() -->
+<!--   ) |>  -->
+<!--   rename(ADF_type1 = ADF, p.value_type1 = p.value) -->
+
+<!-- adf_res_type2 <- df_global_age |> -->
+<!--   group_by(age_grp, country) |> -->
+<!--   group_modify(~ aTSA::adf.test(.x$daily_changes, output = FALSE, nlag = 2) |> -->
+<!--                  bind_rows() |>  -->
+<!--                  pluck("type2") |>  -->
+<!--                  as_tibble() -->
+<!--   ) |>  -->
+<!--   rename(ADF_type2 = ADF, p.value_type2 = p.value) -->
+
+<!-- adf_res_type3 <- df_global_age |> -->
+<!--   group_by(age_grp, country) |> -->
+<!--   group_modify(~ aTSA::adf.test(.x$daily_changes, output = FALSE, nlag = 2) |> -->
+<!--                  bind_rows() |>  -->
+<!--                  pluck("type3") |>  -->
+<!--                  as_tibble() -->
+<!--   ) |>  -->
+<!--   rename(ADF_type3 = ADF, p.value_type3 = p.value) -->
+
+<!-- adf_res <- adf_res_type1 |>  -->
+<!--   full_join(adf_res_type2) |>  -->
+<!--   full_join(adf_res_type3) -->
+<!-- ``` -->
+
+<!-- \newpage -->
+
+<!-- ```{r} -->
+<!-- adf_res |> -->
+<!--   filter(age_grp %in% c("aged0to4", "aged5to9")) |>  -->
+<!--   kbl( -->
+<!--     caption = "Dickey-Fuller test for all combinaisons of countries and age groups investigated", -->
+<!--     col.names = c("Age Group", "Country", "Lag", "ADF", "p", "ADF", "p", "ADF", "p") -->
+<!--   ) |>  -->
+<!--   kable_classic(full_width = F) |>  -->
+<!--   add_header_above(c(" " = 3, "Type 1" = 2, "Type 2" = 2, "type 3" = 2)) |> -->
+<!--   collapse_rows(columns = 1:2, valign = "middle") -->
+<!-- ``` -->
+
+<!-- \newpage -->
+
+<!-- ```{r} -->
+<!-- adf_res |> -->
+<!--   filter(age_grp %in% c("aged10to14", "aged15to19")) |>  -->
+<!--   kbl( -->
+<!--     caption = "Dickey-Fuller test for all combinaisons of countries and age groups investigated", -->
+<!--     col.names = c("Age Group", "Country", "Lag", "ADF", "p", "ADF", "p", "ADF", "p") -->
+<!--   ) |>  -->
+<!--   kable_classic(full_width = F) |>  -->
+<!--   add_header_above(c(" " = 3, "Type 1" = 2, "Type 2" = 2, "type 3" = 2)) |> -->
+<!--   collapse_rows(columns = 1:2, valign = "middle") -->
+<!-- ``` -->
+
+<!-- \newpage -->
+
+<!-- ```{r} -->
+<!-- adf_res |> -->
+<!--   filter(age_grp %in% c("aged20to24", "aged25to34")) |>  -->
+<!--   kbl( -->
+<!--     caption = "Dickey-Fuller test for all combinaisons of countries and age groups investigated", -->
+<!--     col.names = c("Age Group", "Country", "Lag", "ADF", "p", "ADF", "p", "ADF", "p") -->
+<!--   ) |>  -->
+<!--   kable_classic(full_width = F) |>  -->
+<!--   add_header_above(c(" " = 3, "Type 1" = 2, "Type 2" = 2, "type 3" = 2)) |> -->
+<!--   collapse_rows(columns = 1:2, valign = "middle") -->
+<!-- ``` -->
+
+<!-- \newpage -->
+
+<!-- ```{r} -->
+<!-- adf_res |> -->
+<!--   filter(age_grp %in% c("aged35to44", "aged45to54")) |>  -->
+<!--   kbl( -->
+<!--     caption = "Dickey-Fuller test for all combinaisons of countries and age groups investigated", -->
+<!--     col.names = c("Age Group", "Country", "Lag", "ADF", "p", "ADF", "p", "ADF", "p") -->
+<!--   ) |>  -->
+<!--   kable_classic(full_width = F) |>  -->
+<!--   add_header_above(c(" " = 3, "Type 1" = 2, "Type 2" = 2, "type 3" = 2)) |> -->
+<!--   collapse_rows(columns = 1:2, valign = "middle") -->
+<!-- ``` -->
+
+<!-- \newpage -->
+
+<!-- ```{r} -->
+<!-- adf_res |> -->
+<!--   filter(age_grp %in% c("aged55to64", "aged65+")) |>  -->
+<!--   kbl( -->
+<!--     caption = "Dickey-Fuller test for all combinaisons of countries and age groups investigated", -->
+<!--     col.names = c("Age Group", "Country", "Lag", "ADF", "p", "ADF", "p", "ADF", "p") -->
+<!--   ) |>  -->
+<!--   kable_classic(full_width = F) |>  -->
+<!--   add_header_above(c(" " = 3, "Type 1" = 2, "Type 2" = 2, "type 3" = 2)) |> -->
+<!--   collapse_rows(columns = 1:2, valign = "middle") -->
+<!-- ``` -->
+
+<!-- \newpage -->
+
+<!-- # Appendix 2: Significance of Transfer Entropy coefficients -->
+
+<!-- Results for each permutation of age groups from each country.
+Tables include transfer entropy estimates, the effective transfer entropy estimates, standard errors and p-values based on bootstrap replications of the Markov chains under the null hypothesis of statistical independence.
+-->
+
+<!-- ```{r} -->
+<!-- # variables -------------------------------------------------------------------- -->
+<!-- age_order <- c( -->
+<!--   "aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24", -->
+<!--   "aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+" -->
+<!-- ) -->
+
+<!-- # data ------------------------------------------------------------------------- -->
+<!-- te_results <- read_csv(here("data/te_results.csv")) |>  -->
+<!--     mutate( -->
+<!--       X = factor(X, levels = age_order), -->
+<!--       Y = factor(Y, levels = age_order) -->
+<!--     ) |>  -->
+<!--   select(-direction) -->
+<!-- ``` -->
+
+<!-- ```{r} -->
+<!-- te_results |> -->
+<!--   filter(X == "aged0to4") |>  -->
+<!--   filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |>  -->
+<!--   arrange(Y) |>  -->
+<!--   kbl( -->
+<!--     caption = "Significance of Transfer Entropy coefficients", -->
+<!--     digits = 2 -->
+<!--   ) |>  -->
+<!--   kable_classic(full_width = F) |>  -->
+<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
+<!--   kable_styling(font_size = 7) -->
+<!-- ``` -->
+
+<!-- \newpage -->
+
+<!-- ```{r} -->
+<!-- te_results |> -->
+<!--   filter(X == "aged0to4") |>  -->
+<!--   filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |>  -->
+<!--   arrange(Y) |>  -->
+<!--   kbl( -->
+<!--     caption = "Significance of Transfer Entropy coefficients", -->
+<!--     digits = 2 -->
+<!--   ) |>  -->
+<!--   kable_classic(full_width = F) |>  -->
+<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
+<!--   kable_styling(font_size = 7) -->
+<!-- ``` -->
+
+<!-- \newpage -->
+
+<!-- ```{r} -->
+<!-- te_results |> -->
+<!--   filter(X == "aged5to9") |>  -->
+<!--   filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |>  -->
+<!--   arrange(Y) |>  -->
+<!--   kbl( -->
+<!--     caption = "Significance of Transfer Entropy coefficients", -->
+<!--     digits = 2 -->
+<!--   ) |>  -->
+<!--   kable_classic(full_width = F) |>  -->
+<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
+<!--   kable_styling(font_size = 7) -->
+<!-- ``` -->
+
+<!-- \newpage -->
+
+<!-- ```{r} -->
+<!-- te_results |> -->
+<!--   filter(X == "aged5to9") |>  -->
+<!--   filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |>  -->
+<!--   arrange(Y) |>  -->
+<!--   kbl( -->
+<!--     caption = "Significance of Transfer Entropy coefficients", -->
+<!--     digits = 2 -->
+<!--   ) |>  -->
+<!--   kable_classic(full_width = F) |>  -->
+<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
+<!--   kable_styling(font_size = 7) -->
+<!-- ``` -->
+
+<!-- \newpage -->
+<!-- ```{r} -->
+<!-- te_results |> -->
+<!--   filter(X == "aged10to14") |>  -->
+<!--   filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |>  -->
+<!--   arrange(Y) |>  -->
+<!--   kbl( -->
+<!--     caption = "Significance of Transfer Entropy coefficients", -->
+<!--     digits = 2 -->
+<!--   ) |>  -->
+<!--   kable_classic(full_width = F) |>  -->
+<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
+<!--   kable_styling(font_size = 7) -->
+<!-- ``` -->
+
+<!-- \newpage -->
+
+<!-- ```{r} -->
+<!-- te_results |> -->
+<!--   filter(X == "aged10to14") |>  -->
+<!--   filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |>  -->
+<!--   arrange(Y) |>  -->
+<!--   kbl( -->
+<!--     caption = "Significance of Transfer Entropy coefficients", -->
+<!--     digits = 2 -->
+<!--   ) |>  -->
+<!--   kable_classic(full_width = F) |>  -->
+<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
+<!--   kable_styling(font_size = 7) -->
+<!-- ``` -->
+
+<!-- \newpage -->
+
+<!-- ```{r} -->
+<!-- te_results |> -->
+<!--   filter(X == "aged15to19") |>  -->
+<!--   filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |>  -->
+<!--   arrange(Y) |>  -->
+<!--   kbl( -->
+<!--     caption = "Significance of Transfer Entropy coefficients", -->
+<!--     digits = 2 -->
+<!--   ) |>  -->
+<!--   kable_classic(full_width = F) |>  -->
+<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
+<!--   kable_styling(font_size = 7) -->
+<!-- ``` -->
+
+<!-- \newpage -->
+
+<!-- ```{r} -->
+<!-- te_results |> -->
+<!--   filter(X == "aged15to19") |>  -->
+<!--   filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |>  -->
+<!--   arrange(Y) |>  -->
+<!--   kbl( -->
+<!--     caption = "Significance of Transfer Entropy coefficients", -->
+<!--     digits = 2 -->
+<!--   ) |>  -->
+<!--   kable_classic(full_width = F) |>  -->
+<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
+<!--   kable_styling(font_size = 7) -->
+<!-- ``` -->
+
+<!-- \newpage -->
+
+<!-- ```{r} -->
+<!-- te_results |> -->
+<!--   filter(X == "aged20to24") |>  -->
+<!--   filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |>  -->
+<!--   arrange(Y) |>  -->
+<!--   kbl( -->
+<!--     caption = "Significance of Transfer Entropy coefficients", -->
+<!--     digits = 2 -->
+<!--   ) |>  -->
+<!--   kable_classic(full_width = F) |>  -->
+<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
+<!--   kable_styling(font_size = 7) -->
+<!-- ``` -->
+
+<!-- \newpage -->
+
+<!-- ```{r} -->
+<!-- te_results |> -->
+<!--   filter(X == "aged20to24") |>  -->
+<!--   filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |>  -->
+<!--   arrange(Y) |>  -->
+<!--   kbl( -->
+<!--     caption = "Significance of Transfer Entropy coefficients", -->
+<!--     digits = 2 -->
+<!--   ) |>  -->
+<!--   kable_classic(full_width = F) |>  -->
+<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
+<!--   kable_styling(font_size = 7) -->
+<!-- ``` -->
+
+<!-- \newpage -->
+
+<!-- ```{r} -->
+<!-- te_results |> -->
+<!--   filter(X == "aged25to34") |>  -->
+<!--   filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |>  -->
+<!--   arrange(Y) |>  -->
+<!--   kbl( -->
+<!--     caption = "Significance of Transfer Entropy coefficients", -->
+<!--     digits = 2 -->
+<!--   ) |>  -->
+<!--   kable_classic(full_width = F) |>  -->
+<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
+<!--   kable_styling(font_size = 7) -->
+<!-- ``` -->
+
+<!-- \newpage -->
+
+<!-- ```{r} -->
+<!-- te_results |> -->
+<!--   filter(X == "aged25to34") |>  -->
+<!--   filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |>  -->
+<!--   arrange(Y) |>  -->
+<!--   kbl( -->
+<!--     caption = "Significance of Transfer Entropy coefficients", -->
+<!--     digits = 2 -->
+<!--   ) |>  -->
+<!--   kable_classic(full_width = F) |>  -->
+<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
+<!--   kable_styling(font_size = 7) -->
+<!-- ``` -->
+
+<!-- \newpage -->
+
+<!-- ```{r} -->
+<!-- te_results |> -->
+<!--   filter(X == "aged35to44") |>  -->
+<!--   filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |>  -->
+<!--   arrange(Y) |>  -->
+<!--   kbl( -->
+<!--     caption = "Significance of Transfer Entropy coefficients", -->
+<!--     digits = 2 -->
+<!--   ) |>  -->
+<!--   kable_classic(full_width = F) |>  -->
+<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
+<!--   kable_styling(font_size = 7) -->
+<!-- ``` -->
+
+<!-- \newpage -->
+
+<!-- ```{r} -->
+<!-- te_results |> -->
+<!--   filter(X == "aged35to44") |>  -->
+<!--   filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |>  -->
+<!--   arrange(Y) |>  -->
+<!--   kbl( -->
+<!--     caption = "Significance of Transfer Entropy coefficients", -->
+<!--     digits = 2 -->
+<!--   ) |>  -->
+<!--   kable_classic(full_width = F) |>  -->
+<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
+<!--   kable_styling(font_size = 7) -->
+<!-- ``` -->
+
+<!-- \newpage -->
+
+<!-- ```{r} -->
+<!-- te_results |> -->
+<!--   filter(X == "aged45to54") |>  -->
+<!--   filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |>  -->
+<!--   arrange(Y) |>  -->
+<!--   kbl( -->
+<!--     caption = "Significance of Transfer Entropy coefficients", -->
+<!--     digits = 2 -->
+<!--   ) |>  -->
+<!--   kable_classic(full_width = F) |>  -->
+<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
+<!--   kable_styling(font_size = 7) -->
+<!-- ``` -->
+
+<!-- \newpage -->
+
+<!-- ```{r} -->
+<!-- te_results |> -->
+<!--   filter(X == "aged45to54") |>  -->
+<!--   filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |>  -->
+<!--   arrange(Y) |>  -->
+<!--   kbl( -->
+<!--     caption = "Significance of Transfer Entropy coefficients", -->
+<!--     digits = 2 -->
+<!--   ) |>  -->
+<!--   kable_classic(full_width = F) |>  -->
+<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
+<!--   kable_styling(font_size = 7) -->
+<!-- ``` -->
+
+<!-- \newpage -->
+
+<!-- ```{r} -->
+<!-- te_results |> -->
+<!--   filter(X == "aged55to64") |>  -->
+<!--   filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |>  -->
+<!--   arrange(Y) |>  -->
+<!--   kbl( -->
+<!--     caption = "Significance of Transfer Entropy coefficients", -->
+<!--     digits = 2 -->
+<!--   ) |>  -->
+<!--   kable_classic(full_width = F) |>  -->
+<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
+<!--   kable_styling(font_size = 7) -->
+<!-- ``` -->
+
+<!-- \newpage -->
+
+<!-- ```{r} -->
+<!-- te_results |> -->
+<!--   filter(X == "aged55to64") |>  -->
+<!--   filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |>  -->
+<!--   arrange(Y) |>  -->
+<!--   kbl( -->
+<!--     caption = "Significance of Transfer Entropy coefficients", -->
+<!--     digits = 2 -->
+<!--   ) |>  -->
+<!--   kable_classic(full_width = F) |>  -->
+<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
+<!--   kable_styling(font_size = 7) -->
+<!-- ``` -->
+
+<!-- \newpage -->
+
+<!-- ```{r} -->
+<!-- te_results |> -->
+<!--   filter(X == "aged65+") |>  -->
+<!--   filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |>  -->
+<!--   arrange(Y) |>  -->
+<!--   kbl( -->
+<!--     caption = "Significance of Transfer Entropy coefficients", -->
+<!--     digits = 2 -->
+<!--   ) |>  -->
+<!--   kable_classic(full_width = F) |>  -->
+<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
+<!--   kable_styling(font_size = 7) -->
+<!-- ``` -->
+
+<!-- \newpage -->
+
+<!-- ```{r} -->
+<!-- te_results |> -->
+<!--   filter(X == "aged65+") |>  -->
+<!--   filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |>  -->
+<!--   arrange(Y) |>  -->
+<!--   kbl( -->
+<!--     caption = "Significance of Transfer Entropy coefficients", -->
+<!--     digits = 2 -->
+<!--   ) |>  -->
+<!--   kable_classic(full_width = F) |>  -->
+<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
+<!--   kable_styling(font_size = 7) -->
+<!-- ``` -->

### 2024-02-19 exploration/exploration_england.R

@@ -1,44 +1,0 @@
-
-# libraries --------------------------------------------------------------------
-library(janitor)
-library(tidyverse)
-
-# options ----------------------------------------------------------------------
-options(scipen = 999)
-
-# data -------------------------------------------------------------------------
-df <- read_csv("https://api.coronavirus.data.gov.uk/v2/data?areaType=nation&areaCode=E92000001&metric=newCasesBySpecimenDateAgeDemographics&format=csv")
-
-df_clean <- df |> 
-  clean_names() |> 
-  select(date, age, cases)
-
-# visual exploration -----------------------------------------------------------
-df |> 
-  group_by(date) |> 
-  summarise(total_cases = sum(cases)) |> 
-  ggplot(aes(total_cases)) +
-  geom_density(na.rm = TRUE) +
-  labs(
-    title = "Distribution of COVID-19 Daily Cases in Ireland",
-    subtitle = "",
-    caption = "Source: Irish Department of Health (last update 2022-01-01)",
-    x = "Covid Cases", 
-    y = "Density"
-  ) +
-  theme_bw()
-
-df |> 
-  group_by(date) |> 
-  summarise(total_cases = sum(cases)) |> 
-  ggplot(aes(date, total_cases)) +
-  geom_line(na.rm = TRUE) +
-  labs(
-    title = "Evolution of COVID-19 Cases in England",
-    subtitle = "Raw data",
-    caption = "Source: UK Department of Health",
-    x = "Date", 
-    y = "Covid Cases"
-  ) +
-  theme_bw() +
-  theme(legend.position = "bottom")

### 2024-02-19 exploration/exploration_global_output_5.R

@@ -1,306 +1,0 @@
-
-# options ----------------------------------------------------------------------
-options(scipen = 999)
-future::plan(multisession)
-
-# libraries --------------------------------------------------------------------
-library(janitor)
-library(mgcv)
-library(osfr)
-library(tidyverse)
-
-# raw data ---------------------------------------------------------------------
-osf_retrieve_file("7tnfh") |>
-  osf_download(conflicts = "overwrite", progress = TRUE, verbose = TRUE)
-
-df <- "Output_5.zip" |>
-  read_csv(skip = 3, col_types = "ccccciiddd") |>
-  mutate(Date = as.Date(Date, format = "%d.%m.%Y")) |>
-  clean_names()
-
-file.remove("Output_5.zip")
-
-df_school_closure <- 
-  read_csv(
-    "https://en.unesco.org/sites/default/files/covid_impact_education_full.csv",
-    col_types = list(.default = col_character(), Date = col_date(format = "%d/%m/%Y"))) |> 
-  clean_names() |>
-  select(date, country, status) |> 
-  mutate(country = case_when(
-    country == "Bolivia (Plurinational State of)" ~ "Bolivia",  
-    country == "Central African republic" ~ "Central African Republic",
-    country == "Iran (Islamic Republic of)" ~ "Iran",
-    country == "Republic of Moldova" ~ "Moldova",
-    country == "Republic of Korea" ~ "South Korea",
-    country == "United Kingdom of Great Britain and Northern Ireland" ~ "United Kingdom",
-    country == "United States of America" ~ "USA",
-    country == "Viet Nam" ~ "Vietnam",
-    TRUE ~ country
-    )
-  )
-
-df_check <- 
-  read_csv("https://covid19.who.int/WHO-COVID-19-global-data.csv") |> 
-  clean_names() |> 
-  rename(date = date_reported) |> 
-  mutate(country = case_when(
-    country == "Bolivia (Plurinational State of)" ~ "Bolivia",  
-    country == "Iran (Islamic Republic of)" ~ "Iran",
-    country == "occupied Palestinian territory, including east Jerusalem" ~ "Palestine",
-    country == "Republic of Moldova" ~ "Moldova",
-    country == "Republic of Korea" ~ "South Korea",
-    country == "United States of America" ~ "USA",
[…]

### 2024-02-19 exploration/exploration_global_output_5_breaks+covid.R

@@ -1,248 +1,0 @@
-
-# options ----------------------------------------------------------------------
-options(scipen = 999)
-future::plan("future::multisession")
-
-# libraries --------------------------------------------------------------------
-library(janitor)
-library(mgcv)
-library(osfr)
-library(tidyverse)
-
-# raw data ---------------------------------------------------------------------
-osf_retrieve_file("7tnfh") |>
-  osf_download(conflicts = "overwrite", progress = TRUE, verbose = TRUE)
-
-df <- "Output_5.zip" |>
-  read_csv(skip = 3, col_types = "ccccciiddd") |>
-  mutate(Date = as.Date(Date, format = "%d.%m.%Y")) |>
-  clean_names()
-
-file.remove("Output_5.zip")
-
-df_school_closure <- 
-  read_csv(
-    "https://en.unesco.org/sites/default/files/covid_impact_education_full.csv",
-    col_types = list(.default = col_character(), Date = col_date(format = "%d/%m/%Y"))) |> 
-  clean_names() |>
-  select(date, country, status) |> 
-  mutate(country = case_when(
-    country == "Bolivia (Plurinational State of)" ~ "Bolivia",  
-    country == "Central African republic" ~ "Central African Republic",
-    country == "Iran (Islamic Republic of)" ~ "Iran",
-    country == "Republic of Moldova" ~ "Moldova",
-    country == "Republic of Korea" ~ "South Korea",
-    country == "United Kingdom of Great Britain and Northern Ireland" ~ "United Kingdom",
-    country == "United States of America" ~ "USA",
-    country == "Viet Nam" ~ "Vietnam",
-    TRUE ~ country
-    )
-  )
-
-df_check <- 
-  read_csv("https://covid19.who.int/WHO-COVID-19-global-data.csv") |> 
-  clean_names() |> 
-  rename(date = date_reported) |> 
-  mutate(country = case_when(
-    country == "Bolivia (Plurinational State of)" ~ "Bolivia",  
-    country == "Iran (Islamic Republic of)" ~ "Iran",
-    country == "occupied Palestinian territory, including east Jerusalem" ~ "Palestine",
-    country == "Republic of Moldova" ~ "Moldova",
-    country == "Republic of Korea" ~ "South Korea",
-    country == "United States of America" ~ "USA",
-    country == "Venezuela (Bolivarian Republic of)" ~ "Venezuela",
[…]

### 2024-02-19 exploration/exploration_global_output_5_covid.R

@@ -1,226 +1,0 @@
-
-# options ----------------------------------------------------------------------
-options(scipen = 999)
-future::plan(multisession)
-
-# libraries --------------------------------------------------------------------
-library(janitor)
-library(mgcv)
-library(osfr)
-library(tidyverse)
-
-# raw data ---------------------------------------------------------------------
-osf_retrieve_file("7tnfh") |>
-  osf_download(conflicts = "overwrite", progress = TRUE, verbose = TRUE)
-
-df <- "Output_5.zip" |>
-  read_csv(skip = 3, col_types = "ccccciiddd") |>
-  mutate(Date = as.Date(Date, format = "%d.%m.%Y")) |>
-  clean_names()
-
-file.remove("Output_5.zip")
-
-df_school_closure <- 
-  read_csv(
-    "https://en.unesco.org/sites/default/files/covid_impact_education_full.csv",
-    col_types = list(.default = col_character(), Date = col_date(format = "%d/%m/%Y"))) |> 
-  clean_names() |>
-  select(date, country, status) |> 
-  mutate(country = case_when(
-    country == "Bolivia (Plurinational State of)" ~ "Bolivia",  
-    country == "Central African republic" ~ "Central African Republic",
-    country == "Iran (Islamic Republic of)" ~ "Iran",
-    country == "Republic of Moldova" ~ "Moldova",
-    country == "Republic of Korea" ~ "South Korea",
-    country == "United Kingdom of Great Britain and Northern Ireland" ~ "United Kingdom",
-    country == "United States of America" ~ "USA",
-    country == "Viet Nam" ~ "Vietnam",
-    TRUE ~ country
-    )
-  )
-
-df_check <- 
-  read_csv("https://covid19.who.int/WHO-COVID-19-global-data.csv") |> 
-  clean_names() |> 
-  rename(date = date_reported) |> 
-  mutate(country = case_when(
-    country == "Bolivia (Plurinational State of)" ~ "Bolivia",  
-    country == "Iran (Islamic Republic of)" ~ "Iran",
-    country == "occupied Palestinian territory, including east Jerusalem" ~ "Palestine",
-    country == "Republic of Moldova" ~ "Moldova",
-    country == "Republic of Korea" ~ "South Korea",
-    country == "United States of America" ~ "USA",
-    country == "Venezuela (Bolivarian Republic of)" ~ "Venezuela",
-    country == "Viet Nam" ~ "Vietnam",
[…]

### 2024-02-19 manuscript_rss/cover_letter.Rmd
country cover letter

@@
 * Damien Dupré (corresponding author and main investigator of the study)
-  * Dublin City University
-  * damien.dupre@dcu.ie
+  * affiliation: Business School, Dublin City University
+  * city: Dublin
+  * country: Ireland
+  * email: damien.dupre@dcu.ie
+
 * Edgar Morgenroth
-  * Dublin City University
-  * edgar.morgenroth@dcu.ie
+  * affiliation: Business School, Dublin City University
+  * city: Dublin
+  * country: Ireland
+  * email: edgar.morgenroth@dcu.ie
 

### 2024-02-20 manuscript_oup/manuscript.Rmd
oup

--- 2024-02-19 manuscript_rss/manuscript.Rmd
@@
 title: "Examining the Impact of School Closures on COVID-19 Infections in Europe and their Effects on Different Age Cohorts"
+journal_title: "Journal of the Royal Statistical Society, Series C (Applied Statistics)"
 short_title: "Impact of School Closures on COVID-19 by Age Cohorts"
-short_author: "Author 1 et. al."
-authors: 
-- name: "Author 1"
-  affiliation: "Address"
-  city: "City"
-  country: "Country"
-  email: "email1\\@example.com"
-- name: "Author 2"
-  affiliation: "Address2"
-  city: "City2"
-  country: "Country2"
-  email: "email2\\@example.com"
-# short_author: "Dupré & Morgenroth"
-# authors:
-# - name: "Damien Dupré"
-#   affiliation: "Business School, Dublin City University"
-#   city: "Dublin"
-#   country: "Ireland"
-#   email: "damien.dupre\\@dcu.ie"
-#   footnote: 1
-# - name: "Edgar Morgenroth"
-#   affiliation: "Business School, Dublin City University"
-#   city: "Dublin"
-#   country: "Ireland"
-#   email: "edgar.morgenroth\\@dcu.ie"
-#   footnote: 2
-# footnote:
-#   - code: 1
-#     text: "Corresponding Author"
-#   - code: 2
-#     text: "Equal contribution"
+short_author: "Dupré & Morgenroth"
+authors_1_name: "Damien Dupré"
+authors_1_ORCID: "0000-0001-8610-1045"
+authors_2_name: "Edgar Morgenroth"
+authors_2_ORCID: "0000-0002-9442-0561"
+authors_affiliation: "Business School, Dublin City University"
+authors_city: "Dublin"
+authors_country: "Ireland"
+corresponding_email: "damien.dupre\\@dcu.ie"
 abstract: |
@@
 bibliography: "bibliography.bib"
+csl: elsevier-harvard.csl
 header-includes:
@@
   bookdown::pdf_book:
-    base_format: rticles::rss_article
+    template: oup-authoring-template.tex
     keep_tex: false
-  bookdown::word_document2:
-    default
-knit: (function(input, ...) {
-    rmarkdown::render(
-      input,
-      output_dir = here::here("output"),
-      output_file = c(glue::glue("{xfun::sans_ext(input)}.pdf"), glue::glue("{xfun::sans_ext(input)}.docx")),
-      output_format = "all",
-      envir = globalenv()
-    )
-  })
 ---
@@
 
+```{r age, fig.height=3, out.width="\\textwidth", fig.cap="(ref:age)"}
+gam_plot_age
+```
+
+```{r te, fig.height=8, out.width="\\textwidth", fig.cap="(ref:te)", fig.pos = 'H'}
+te_plot
+```
+
 Considering all countries, the analysis of age groups reveals distinct patterns (Figure \@ref(fig:age)).
@@
 Thus, while the total case number declined following school closures, among younger age groups only the pre-school age group had a consistent reduction in cases, while for other age groups, there was even an increase in cases following school closures.
-
-```{r age, fig.height=3, out.width="\\textwidth", fig.cap="(ref:age)"}
-gam_plot_age
-```
 
@@
 Alternatively, it appears that Austria, the Netherlands, Portugal, and Spain have significantly higher $T$ coefficients in the lower right quadrant of the matrix, which indicates that the daily changes in COVID-19 cases number in older cohorts are predicting the daily changes in COVID-19 cases number in younger cohorts.
-
-```{r te, fig.height=8, out.width="\\textwidth", fig.cap="(ref:te)"}
-te_plot
-```
 
@@
 
-<div id="refs"></div>
+<!-- <div id="refs"></div> -->
 

### 2024-02-20 manuscript_oup/manuscript.tex
oup

%%
%% Copyright 2022 OXFORD UNIVERSITY PRESS
%%
%% This file is part of the 'oup-authoring-template Bundle'.
%% ---------------------------------------------
%%
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%%
%% The list of all files belonging to the 'oup-authoring-template Bundle' is
%% given in the file `manifest.txt'.
%%
%% Template article for OXFORD UNIVERSITY PRESS's document class `oup-authoring-template'
%% with bibliographic references
%%

%%%CONTEMPORARY%%%
\documentclass[unnumsec,webpdf,contemporary,large]{oup-authoring-template}%
%\documentclass[unnumsec,webpdf,contemporary,large,namedate]{oup-authoring-template}% uncomment this line for author year citations and comment the above
%\documentclass[unnumsec,webpdf,contemporary,medium]{oup-authoring-template}
%\documentclass[unnumsec,webpdf,contemporary,small]{oup-authoring-template}

%%%MODERN%%%
%\documentclass[unnumsec,webpdf,modern,large]{oup-authoring-template}
%\documentclass[unnumsec,webpdf,modern,large,namedate]{oup-authoring-template}% uncomment this line for author year citations and comment the above
%\documentclass[unnumsec,webpdf,modern,medium]{oup-authoring-template}
%\documentclass[unnumsec,webpdf,modern,small]{oup-authoring-template}

%%%TRADITIONAL%%%
%\documentclass[unnumsec,webpdf,traditional,large]{oup-authoring-template}
%\documentclass[unnumsec,webpdf,traditional,large,namedate]{oup-authoring-template}% uncomment this line for author year citations and comment the above
%\documentclass[unnumsec,namedate,webpdf,traditional,medium]{oup-authoring-template}
%\documentclass[namedate,webpdf,traditional,small]{oup-authoring-template}

\onecolumn % for one column layouts

%\usepackage{showframe}

\graphicspath{{Fig/}}

% line numbers
%\usepackage[mathlines, switch]{lineno}
%\usepackage[right]{lineno}

\theoremstyle{thmstyleone}%
\newtheorem{theorem}{Theorem}%  meant for continuous numbers
%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
\newtheorem{proposition}[theorem]{Proposition}%
%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.
\theoremstyle{thmstyletwo}%
\newtheorem{example}{Example}%
\newtheorem{remark}{Remark}%
\theoremstyle{thmstylethree}%
\newtheorem{definition}{Definition}

\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\begin{document}

\journaltitle{Journal of the Royal Statistical Society, Series C (Applied Statistics)}
\DOI{DOI HERE}
\copyrightyear{2022}
\pubyear{2019}
\access{Advance Access Publication Date: Day Month Year}
\appnotes{Paper}

\firstpage{1}

%\subtitle{Subject Section}

\title[Impact of School Closures on COVID-19 by Age Cohorts]{Examining the Impact of School Closures on COVID-19 Infections in Europe and their Effects on Different Age Cohorts}

\author[1,$\ast$]{Damien Dupré\ORCID{0000-0001-8610-1045}}
\author[1]{Edgar Morgenroth\ORCID{0000-0002-9442-0561}}

\authormark{Dupré \& Morgenroth}

\address[1]{\orgdiv{Business School, Dublin City University}, \orgaddress{\street{Dublin}, \country{Ireland}}}

\corresp[$\ast$]{Corresponding author. \href{email:damien.dupre@dcu.ie}{damien.dupre@dcu.ie}}

\received{Date}{0}{Year}
\revised{Date}{0}{Year}
\accepted{Date}{0}{Year}

%\editor{Associate Editor: Name}

%\abstract{
%\textbf{Motivation:} .\\
%\textbf{Results:} .\\
%\textbf{Availability:} .\\
%\textbf{Contact:} \href{name@email.com}{name@email.com}\\
%\textbf{Supplementary information:} Supplementary data are available at \textit{Journal Name}
%online.}

\abstract{This paper provides a thorough analysis of COVID-19 case trends in selected European countries, focusing on the influence of different age groups. It explores the relationship between population size and confirmed case numbers, suggesting that larger countries tend to have higher case figures. The study investigates how various waves of the pandemic affect different age demographics, with older age groups being most affected by the initial wave and younger age groups more impacted by subsequent waves.
Using a Generalised Additive Model (GAM), the research assesses the impact of school closures on COVID-19 cases, revealing a decreasing non-linear effect across all countries examined. Following school closures, distinct trends emerge among age groups: ages 0 to 4 experience a decrease, ages 5 to 14 see an increase, and ages 15 to 24 initially surge then decline. Transfer Entropy calculations uncover imbalances in age group influences, with certain age groups predicting changes in others.
Consistent patterns across countries are identified, such as younger cohorts influencing older cohorts in Austria, Germany, and the Netherlands, while Portugal and Spain show the opposite trend. These findings deepen our understanding of COVID-19 dynamics in Europe, providing valuable insights for targeted public health strategies and interventions.}
\keywords{COVID-19, School Closures, Causality, Generalised Additive Model, Transfer Entropy}

% \boxedtext{
% \begin{itemize}
% \item Key boxed text here.
% \item Key boxed text here.
% \item Key boxed text here.
% \end{itemize}}

\maketitle

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

A wide range of non-pharmaceutical interventions (NPI) were enacted to control the spread of COVID-19, especially before the availability of effective vaccines. Due to their social mixing patterns, children have been identified as an age group that can drive the spread of respiratory infections such as influenza (Moser and White, 2018), and school closures were found to be effective in mitigating the spread of influenza H1N1 in Japan in 2009 (Kawano and Kakehashi, 2015). It is therefore not surprising that school closures were one of the NPIs that were introduced in many countries to stop the spread of COVID-19. They were used particularly during the initial wave of the pandemic and during subsequent waves when case numbers were high.

Recently, a body of literature has shown various negative consequences on child development and health due to school closures. School closures have been found to have negatively affected child mental health (Moulin et al., 2022), nutrition/obesity (Sugimoto et al., 2023), and education (Lerkkanen et al., 2023). Some governments are now questioning their decisions to close schools (De Simone and Mourao, 2021). For example, the German health minister Karl Lauterbach, in an interview with one of Germany's publicly funded television stations, admitted that ``in retrospect it had been wrong to keep schools and childcare closed for so long'' ({``ARD tagesschau,''} 2023).

Even if school closures had negative consequences for children, they might nevertheless have been effective at stopping the spread of COVID-19. However, analysis on the effectiveness of school closures on the spread of COVID-19 remains inconclusive. Alfano (2022) in a study of a panel of European countries found that school closures were associated with a reduced COVID-19 incidence. In contrast, Walsh et al. (2021) in a review of 40 studies covering 150 countries concluded that the effectiveness of school closures was uncertain, with 60\% having identified no impact and pointing to the potential for the analysis to be affected by confounding factors and collinearity. The latter shortcomings suggest that it is hard to draw firm inferences, as even a positive result may not imply causality.

This paper investigates the dynamics of COVID-19 across age groups and how this was affected by school closures. This allows us to identify the effect of school closure on the evolution of COVID-19 cases in other age groups and particularly on older and more vulnerable age groups, including parents, grandparents, and relatives about which little is known.

We identify the relationship between the school closures and case numbers in different age groups using Generalized Additive Models (GAM), which are more flexible than conventional Vector Autoregressions, given that they allow for non-linear relationships between variables. Furthermore, compared to other methods such as Vector Autoregression, GAM can handle fixed and random smooth effects without making strict assumptions about linearity. Then, using the transfer entropy method, we evaluate the impact of case number changes between age groups in order to identify if school closure has secondary repercussions in the household. Transfer entropy can adapt to changes in the underlying system dynamics over time. It does not assume stationarity and can capture evolving causal relationships, making it suitable for studying dynamic systems and non-stationary processes such as COVID-19 contaminations.

\hypertarget{influence-of-school-closure}{%
\subsection{Influence of School Closure}\label{influence-of-school-closure}}

Once it was clear that COVID-19 spreads from person to person, it was natural that governments were advised to enact measures to reduce social contact in order to reduce the spread of the virus. Therefore, it is legitimate to believe that by closing schools, a reduction of the contaminations would be observed in the younger age groups. However, the efficiency of school closure on the reduction of COVID-19 cases is still questioned (Bayham and Fenichel, 2020; Esposito et al., 2021). While some research has found that school closures contribute to limit or to reduce the growth rate of confirmed cases after implementation (Stage et al., 2021; Sugishita, 2020), others did not observe a change in the evolution of COVID-19 cases (Chang et al., 2020; Iwata et al., 2020). For instance, a controlled comparison between similar localities in Japan with schools closed and schools open did not reveal any evidence that school closures reduced the spread of COVID-19 (Fukumoto et al., 2021). If the school closure had a real impact on the evolution of confirmed COVID-19 cases, it should be possible to observe a decrease or at least an inflection in the trend of its evolution among younger age groups.

\hypertarget{causal-relationship-between-age-groups}{%
\subsection{Causal Relationship Between Age Groups}\label{causal-relationship-between-age-groups}}

A second implicit belief regarding the effect of school closure on the spread of COVID-19 is that school not only influences the spread of the virus among children and teenagers but also has a knock-on effect on the spread of the virus in older age groups, also called Secondary Attack Rate (SAR). The contaminated children and teenagers would bring the virus back home and would pass the virus on to their parents and relatives. For example, research investigating the contamination in the household network not only revealed an exceptionally high rate of secondary contamination but also that this contamination happened when the schools were closed (Soriano-Arandes et al., 2021).

Despite being reported in several clinical and epidemiological studies (Siebach et al., 2021; Zhen-Dong et al., 2020), multiple research have shown that the SAR from children to household members was lower than expected (Heavey et al., 2020; Hoek et al., 2020; Kim et al., 2021; Ludvigsson, 2020). However, the SAR of children and teenagers to the household member is likely to be age-dependent, with differences between infants, primary and secondary school children, and college students (Gras-Le Guen et al., 2021). If a secondary transmission from children and teenagers to household member has a significant influence, then a temporal causality relationship between their evolution should be observed.

To evaluate the impact of school closure on COVID-19 cases across different age groups, we first performed a non-linear time series regression to analyze the relationship between school closures and the number of COVID-19 cases in various age groups. In a second analysis, the transfer entropy method is employed to assess the influence of changes in case numbers between age groups, aiming to determine if school closures have secondary effects within households.

\hypertarget{estimation-methodology}{%
\section{Estimation Methodology}\label{estimation-methodology}}

\hypertarget{generalized-additive-model}{%
\subsection{Generalized Additive Model}\label{generalized-additive-model}}

The effect of school closure on the trend of daily COVID-19 cases is analyzed using a Generalized Additive Model (GAM). GAM is a flexible modeling approach that estimates non-linear relationships between variables. Compared to other methods such as Vector Autoregression, GAM can handle fixed and random smooth effects without making strict assumptions about linearity.

The GAM is fitted on the daily COVID-19 cases to test the hypothesis of a significant non-linear evolution of cases among age groups from 0 to 4, from 5 to 9, from 10 to 14, from 15 to 19, and from 20 to 24 (Wood, 2017). The model also estimates the overall non-linear effect by country, taking into account the interaction between age groups and countries as random intercepts and the interaction between time, countries, and the period of closure as random effects (Eq 1).

By estimating the degree of smoothness of a Bayesian spline smoothing using restricted fast maximum likelihood estimation (Wood, 2011), GAM identifies dynamic patterns underlying the evolution of COVID-19 cases reported while including the random effect of different age groups and countries as follows:

\begin{align*}
  \tag{1}
  n\prime_t &\sim Poisson(\mu_t) \\
  \log(\mu_t) &= \beta_0 + \beta_1\,wave_t + \beta_2\,country_t + \beta_3\,age\,group_t + \\
              & f_1(closure_t) + f_2(closure_t,country_t) + f_3(closure_t,age\,group_t)
\end{align*}

\noindent where \(n\prime_t\) represents the confirmed COVID-19 cases, assuming a Poisson distribution for the fitting (Loader, 2006), and \(t\) is the date corresponding to the confirmed COVID-19 cases. The response variable includes a specific random effect taking into account variation within waves of school closure, countries, and age groups. The terms \(f_1\) to \(f_3\) are smooth functions of the time since closure, the time since closure for each country, and the time since closure for each age group. The restricted maximum likelihood (REML) was used to avoid overfitting while estimating smoothing parameters. In order to accurately account for the autocorrelation arising from the time series data, the residuals are modeled using an AR1 error model. By incorporating the autoregressive component, the AR1 model acknowledges the dependence of each residual on its previous value, thus providing a comprehensive representation of the data's temporal dynamics.

\hypertarget{transfer-entropy}{%
\subsection{Transfer Entropy}\label{transfer-entropy}}

Transfer entropy (\(T\)) can be used to infer the temporal relationship between two time series \(X\) and \(Y\). This measure indicates whether changes in COVID-19 cases within a certain age group \(X\) can be used to reduce the uncertainty on the future COVID-19 cases within another age group \(Y\). If it does, \(T\) is considered evidence of a causal effect from the age group \(X\) to the age group \(Y\) (Schreiber, 2000). As such, Granger causality is a special case of transfer entropy applied to time series that are jointly Gaussian distributed (Barnett et al., 2009). Therefore, transfer entropy is a more robust analysis of time series, especially when applied to the impact of age cohorts on pandemic transmission (Kissler et al., 2020).

The influence of the evolution in COVID-19 cases across all age groups is evaluated using Shannon's transfer entropy, given by:

\begin{align}
  \tag{2}
  T_{ag\,x \rightarrow\,ag\,y}(k,l) = \sum_{ag\,x,ag\,y} p\left(ag\,x_{t+1}, ag\,x_t^{(k)}, ag\,y_t^{(l)}\right) \log \left(\frac{p\left(ag\,x_{t+1}| ag\,x_t^{(k)}, ag\,y_t^{(l)}\right)}{p\left(ag\,x_{t+1}|ag\,x_t^{(k)}\right)}\right)
\end{align}

\noindent where \(T_{ag\,x \rightarrow\,ag\,y}\) consequently measures the influence of the change dynamic from an age group \(X\) (or \(ag\,x\)) to another age group \(Y\) (or \(ag\,y\)) for every country (Eq 2).

The day-by-day difference in COVID-19 confirmed cases \(n\prime\prime_t\) is used to satisfy the stationary requirement for the calculation of Shannon's Transfer Entropy (Behrendt et al., 2019; Shannon, 1948).

\hypertarget{data}{%
\section{Data}\label{data}}

\hypertarget{covid-19-cases-per-age-group}{%
\subsection{COVID-19 cases per age group}\label{covid-19-cases-per-age-group}}

For the analysis, data on COVID case counts by age group are required. The number of COVID-19 cases per age group has been taken from the COVerAGE-DB project (Riffe et al., 2021). The COVerAGE-DB project consists of 3 data files: an ``Input'' data file that collects the official COVID-19 cases from 117 countries, the ``Output\_5'' data file, which is a projection of COVID-19 cases by a group of 5 years, and the ``Output\_10'' data file, which is a projection of COVID-19 cases by a group of 10 years. For our analyses, we have used the ``Output\_5,'' which deals with the heterogeneity of countries' reporting formats by using spline approximations when the data for this age bracket is not available for a country. Therefore, it allows a precise analysis between age groups under 24, while age groups of 25 and above are reduced to 10-year brackets, and groups above 64 are concatenated all together.

In order to cross-validate the data obtained after spline approximations, a comparison with the data published by the {``World health organization COVID-19 data''} (2022) reveals perfect similarities. The original data consist of 14,089,320 observations of 10 variables (117 distinct countries, region within the country, a unique observation code, the date of the observation, the gender, which can be male, female, or both, the age bracket by 5 years from 0 to 100, a confirmation of the age interval for each bracket, the total number of cases so far, the total number of deaths, and the total number of tests performed) from February 16, 2020 to January 20, 2022. After removing countries with missing and inconsistent values, only 22 are suitable for data analyses. However, to focus this analysis on geographically and culturally comparable countries, only 12 European countries are kept: Austria, Belgium, Bulgaria, Croatia, Estonia, France, Germany, Greece, Netherlands, Portugal, Slovakia, and Spain. The observations are reported in terms of the total number of COVID-19 cases per day from the start of the pandemic. The daily number of cases at a specific date \(n\prime_{t}\) is calculated with the difference between the total cases at a date \(t\) and the total cases at a date \(t-1\) (i.e., derivative 1). In addition, the change in the daily number of cases \(n\prime\prime_{t}\) between \(n\prime_{t}\) and \(n\prime_{t-1}\) has also been calculated (i.e., derivative 2).

\hypertarget{school-closure}{%
\subsection{School closure}\label{school-closure}}

To identify the impact of school closures on the number of cases, information regarding the school closures on a day-by-day basis is required, and this is taken from the ``UNESCO global education coalition'' (2022). For each day, in each country, the status of the schools is indicated as fully open, partially open, closed due to COVID-19, or closed due to an academic break. Because it would be difficult to measure the effect of school closures at a country level when schools are partially closed, only closures due to COVID-19 or due to an academic break are considered. Indeed, both are considered as closure at a country-wide level. Note that periods of closure longer than 21 days are compared for 28 days, regardless of whether the schools opened or not after 21 days. Any COVID-19 case numbers beyond the 28-day period are not relevant for evaluating the impact of school closures.

\hypertarget{results}{%
\section{Results}\label{results}}

The data show that the trend of confirmed COVID-19 cases follows similar patterns across the selected European countries, with scales following the size of the population in these countries (Figure \ref{fig:overall}). Thus, the bigger the country, the higher the total number of COVID-19 cases.

\begin{figure}
\includegraphics[width=\textwidth]{manuscript_files/figure-latex/overall-1} \caption{Cumulative COVID-19 cases number for selected European countries since the beginning of the pandemic. Source: COVerAGE-DB (Riffe et al., 2021).}\label{fig:overall}
\end{figure}

The evolution of COVID-19 cases reveals some similarities across all age groups. However, the influence of each wave on individual age groups also has some particularities (Figure \ref{fig:descriptive}). For example, the first wave was more important among the oldest age groups, whereas the third wave was more important among the youngest age groups.

\begin{figure}
\includegraphics[width=\textwidth]{manuscript_files/figure-latex/descriptive-1} \caption{Periods of school closure since the beginning of the COVID-19 pandemic for selected European countries and their reason: regular academic break vs.~closure due to government decisions. Source: {``{UNESCO} global education coalition''} (2022).}\label{fig:descriptive}
\end{figure}

As outlined above, to evaluate the shape of the trend in the numbers of COVID-19 cases reported after the three school closures longer than 21 consecutive days, a GAM was fitted as described above, taking into account the overall effect across all the selected European countries, as well as the effect for age groups: 0 to 4, 5 to 9, 10 to 14, 15 to 19, and 20 to 24 years old. The obtained results satisfy the requirements to fit this model, which explains 77.2\% of the variation in COVID-19 cases.

Overall, the results revealed a decreasing non-linear effect of school closure at a country level for the selected European countries (Austria: \(\chi^2(5.75) = 2934.05\), \(p < 0.001\); Belgium: \(\chi^2(5.99) = 9805.66\), \(p < 0.001\); Bulgaria: \(\chi^2(5.92) = 158.24\), \(p < 0.001\); Croatia: \(\chi^2(5.45) = 681.91\), \(p < 0.001\); Estonia: \(\chi^2(5.9) = 1827.73\), \(p < 0.001\); France: \(\chi^2(5.96) = 6713.39\), \(p < 0.001\); Germany (\(\chi^2(5.97) = 4176.3\), \(p < 0.001\)); Greece: \(\chi^2(5.91) = 1306.4\), \(p < 0.001\); The Netherlands: \(\chi^2(5.97) = 14370.95\), \(p < 0.001\); Portugal: \(\chi^2(5.86) = 2423.35\), \(p < 0.001\); Slovakia (\(\chi^2(5.97) = 6628.69\), \(p < 0.001\); Spain: \(\chi^2(5.94) = 562.49\), \(p < 0.001\), see Figure \ref{fig:age}).

\begin{figure}
\includegraphics[width=\textwidth]{manuscript_files/figure-latex/country-1} \caption{Standardized effect of the smooth term in Generalized Additive Model by country.}\label{fig:country}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{manuscript_files/figure-latex/age-1} \caption{Standardized effect of the smooth term in Generalized Additive Model by age group.}\label{fig:age}
\end{figure}

\begin{figure}[H]
\includegraphics[width=\textwidth]{manuscript_files/figure-latex/te-1} \caption{Matrix of Transfer Entropy coefficients according to every age group combination for each of the selected European country. Age groups on the x-axis are influencing the age groups on the y-axis (\(ag\,x \rightarrow\,ag\,y\)). The significance of each Transfer Entropy coefficient is provided in Appendix 2.}\label{fig:te}
\end{figure}

Considering all countries, the analysis of age groups reveals distinct patterns (Figure \ref{fig:age}). For the 0 to 4 age group, i.e., the pre-school group, there was a downward trend during the initial two weeks following school closure, followed by flattening off (\(\chi^2(5.94) = 694.67\), \(p < 0.001\)). In contrast, for the age group ranging from 5 to 14, a substantial increase in cases after 14 days of school closure is found (5 to 9 age group: \(\chi^2(5.93) = 720.56\), \(p < 0.001\) and 10 to 14 age group: \(\chi^2(1) = 0.01\), \(p = 0.931\)). For age groups between 15 and 24, there is a notable surge immediately after school closure, followed by a decrease after 14 days (15 to 19 age group: \(\chi^2(4.97) = 8423.69\), \(p < 0.001\) and 20 to 24 age group: \(\chi^2(5.98) = 18932.54\), \(p < 0.001\)). Thus, while the total case number declined following school closures, among younger age groups only the pre-school age group had a consistent reduction in cases, while for other age groups, there was even an increase in cases following school closures.

Given that school closures were not only aimed at reducing COVID-19 cases among school-going children but also age groups, it is important to test the degree to which there was transmission across age groups, which is done using transfer entropy analysis. Before this analysis is carried out, an Augmented Dickey-Fuller has been applied to each age group to ensure that the daily changes in COVID-19 cases are stationary (see Appendix 1).

The results of the transfer entropy calculations between age groups for each of the 12 selected European countries are reported in Figure \ref{fig:te}. The absence of symmetry between influencing age groups (i.e., \(ag\,x\)) and influenced age groups (i.e., \(ag\,y\)) is found. Indeed, the change in COVID-19 cases in some age groups is influenced by other age groups, but they are not reciprocally influencing these age groups. Figure \ref{fig:te} shows how different age cohorts influence the COVID-19 case numbers of all age groups across countries. The upper left quadrant indicates how younger age cohorts are influencing older age cohorts, the bottom right quadrant indicates how older age cohorts are influencing younger age cohorts, finally, the lower left and upper right indicate how younger or older age cohorts are influencing themselves. By analyzing these quadrants, it is possible to identify similar patterns across multiple countries. Indeed, it appears that Austria, Germany, and The Netherlands have significantly higher \(T\) coefficients in the upper left quadrant of the matrix, which indicates that the daily changes in COVID-19 cases number in younger cohorts are predicting the daily changes in COVID-19 cases number in older cohorts. Alternatively, it appears that Austria, the Netherlands, Portugal, and Spain have significantly higher \(T\) coefficients in the lower right quadrant of the matrix, which indicates that the daily changes in COVID-19 cases number in older cohorts are predicting the daily changes in COVID-19 cases number in younger cohorts.

\hypertarget{discussion-conclusion}{%
\section{Discussion \& Conclusion}\label{discussion-conclusion}}

Knowledge about the transmission of the virus significantly improved over time as more studies have been published. While an early study found that children did not play an important role in the transmission of the virus (Li et al., 2020), more recent results give a more nuanced position, stating that the spread of the virus in children is moderate. This study, using robust statistical methods, considered the effect of school closures on COVID-19 cases across age groups. While there are commonalities in the evolution of COVID-19 cases across all European countries that were included in our data, school closures exhibit distinct impacts on different age groups. Notably, the analysis shows that the 0 to 4 age group experiences a downward trend in COVID-19 cases during the initial two weeks following school closure, followed by stabilization. In contrast, age groups ranging from 5 to 14 exhibit a substantial increase in cases after 14 days of school closure. Additionally, age groups between 15 and 24 demonstrate a notable surge immediately after school closure, followed by a decrease after 14 days. Thus, the results do not support the hypothesis that school closures were effective. These results partially replicate observations from Alfano (2022) while providing a clearer picture of the structure of the effect of school closure.

The Transfer Entropy calculations between age groups for each of the 12 selected European countries reveal an absence of symmetry between influencing age groups (X) and influenced age groups (Y). Some age groups influence the COVID-19 case numbers of other age groups, but there is no reciprocal influence. This finding suggests that changes in COVID-19 cases in certain age groups can predict the changes in other age groups but not vice versa, i.e., causality can be unidirectional. These findings highlight the importance of considering intergenerational interactions in designing effective control measures.

Furthermore, the quadrant analysis reveals similar patterns across multiple countries. Austria, Germany, and the Netherlands exhibit significantly higher TE coefficients in the upper left quadrant, indicating that changes in COVID-19 cases in younger cohorts predict the changes in older cohorts. On the other hand, Austria, the Netherlands, Portugal, and Spain show higher TE coefficients in the lower right quadrant, indicating that changes in COVID-19 cases in older cohorts predict the changes in younger cohorts.

In conclusion, this study sheds light on the trends and patterns of confirmed COVID-19 cases in selected European countries, with a particular focus on the influence of age groups and school closures. While a reduction in overall COVID-19 cases was found during school closures, which might suggest that like lockdowns (Alfano and Ercolano, 2020; Molefi et al., 2021), prolonged school closures could contribute to mitigating the spread of the virus, the results for the school-going age groups suggest that this is not the case.

\hypertarget{data-availability}{%
\section{Data Availability}\label{data-availability}}

All data extracted from the COVerAGE-DB project and from the UNESCO global education coalition that are used to produce these results are fully available here \url{https://github.com/damien-dupre/covid_ts_causality}. The repository also includes the R code that were used to produce these results.

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-alfano2022effects}{}}%
Alfano, V., 2022. The effects of school closures on COVID-19: A cross-country panel analysis. Applied Health Economics and Health Policy 20, 223--233. doi:\href{https://doi.org/10.1007/s40258-021-00702-z}{10.1007/s40258-021-00702-z}

\leavevmode\vadjust pre{\hypertarget{ref-alfano2020efficacy}{}}%
Alfano, V., Ercolano, S., 2020. The efficacy of lockdown against COVID-19: A cross-country panel analysis. Applied Health Economics and Health Policy 18, 509--517. doi:\href{https://doi.org/10.1007/s40258-020-00596-3}{10.1007/s40258-020-00596-3}

\leavevmode\vadjust pre{\hypertarget{ref-ard2023lauterbach}{}}%
ARD tagesschau, 2023.

\leavevmode\vadjust pre{\hypertarget{ref-barnett2009granger}{}}%
Barnett, L., Barrett, A.B., Seth, A.K., 2009. Granger causality and transfer entropy are equivalent for gaussian variables. Physical Review Letters 103, 2387011--2387014. doi:\href{https://doi.org/10.1103/PhysRevLett.103.238701}{10.1103/PhysRevLett.103.238701}

\leavevmode\vadjust pre{\hypertarget{ref-bayham2020impact}{}}%
Bayham, J., Fenichel, E.P., 2020. Impact of school closures for COVID-19 on the US health-care workforce and net mortality: A modelling study. The Lancet Public Health 5, e271--e278. doi:\href{https://doi.org/10.1016/S2468-2667(20)30082-7}{10.1016/S2468-2667(20)30082-7}

\leavevmode\vadjust pre{\hypertarget{ref-behrendt2019rtransferentropy}{}}%
Behrendt, S., Dimpfl, T., Peter, F.J., Zimmermann, D.J., 2019. RTransferEntropy --- quantifying information flow between different time series using effective transfer entropy. SoftwareX 10. doi:\href{https://doi.org/10.1016/j.softx.2019.100265}{10.1016/j.softx.2019.100265}

\leavevmode\vadjust pre{\hypertarget{ref-chang2020modelling}{}}%
Chang, S.L., Harding, N., Zachreson, C., Cliff, O.M., Prokopenko, M., 2020. Modelling transmission and control of the COVID-19 pandemic in australia. Nature Communications 11, 1--13. doi:\href{https://doi.org/10.1038/s41467-020-19393-6}{10.1038/s41467-020-19393-6}

\leavevmode\vadjust pre{\hypertarget{ref-de2021determines}{}}%
De Simone, E., Mourao, P.R., 2021. What determines governments' response time to COVID-19? A cross-country inquiry on the measure restricting internal movements. Open Economics 4, 106--117. doi:\href{https://doi.org/10.1515/openec-2020-0116}{10.1515/openec-2020-0116}

\leavevmode\vadjust pre{\hypertarget{ref-esposito2021comprehensive}{}}%
Esposito, S., Cotugno, N., Principi, N., 2021. Comprehensive and safe school strategy during COVID-19 pandemic. Italian Journal of Pediatrics 47, 1--4. doi:\href{https://doi.org/10.1186/s13052-021-00960-6}{10.1186/s13052-021-00960-6}

\leavevmode\vadjust pre{\hypertarget{ref-fukumoto2021no}{}}%
Fukumoto, K., McClean, C.T., Nakagawa, K., 2021. No causal effect of school closures in japan on the spread of COVID-19 in spring 2020. Nature Medicine 27, 2111--2119. doi:\href{https://doi.org/10.1038/s41591-021-01571-8}{10.1038/s41591-021-01571-8}

\leavevmode\vadjust pre{\hypertarget{ref-grasleguen2021reopening}{}}%
Gras-Le Guen, C., Cohen, R., Rozenberg, J., Launay, E., Levy-Bruhl, D., Delacourt, C., 2021. Reopening schools in the context of increasing COVID-19 community transmission: The french experience. Archives de Pédiatrie 28, 178--185. doi:\href{https://doi.org/10.1016/j.arcped.2021.02.001}{10.1016/j.arcped.2021.02.001}

\leavevmode\vadjust pre{\hypertarget{ref-heavey2020evidence}{}}%
Heavey, L., Casey, G., Kelly, C., Kelly, D., McDarby, G., 2020. No evidence of secondary transmission of COVID-19 from children attending school in ireland, 2020. Eurosurveillance 25, 1--4. doi:\href{https://doi.org/10.2807/1560-7917.ES.2020.25.21.2000903}{10.2807/1560-7917.ES.2020.25.21.2000903}

\leavevmode\vadjust pre{\hypertarget{ref-vanderhoek2020role}{}}%
Hoek, W. van der, Backer, J.A., Bodewes, R., Friesema, I., Meijer, A., Pijnacker, R., Reukers, D.F.M., Reusken, C., Roof, I., Rots, N., Te Wierik, M.J.M., Gageldonk-Lafeber, A. van, Waegemaekers, C., Hof, S. van den, 2020. \href{https://www.ntvg.nl/artikelen/de-rol-van-kinderen-de-transmissie-van-sars-cov-2}{De rol van kinderen in de transmissie van SARS-CoV-2}. Nederlands Tijdschrift voor Geneeskunde 164, 1--7.

\leavevmode\vadjust pre{\hypertarget{ref-iwata2020school}{}}%
Iwata, K., Doi, A., Miyakoshi, C., 2020. Was school closure effective in mitigating coronavirus disease 2019 (COVID-19)? Time series analysis using bayesian inference. International Journal of Infectious Diseases 99, 57--61. doi:\href{https://doi.org/10.1016/j.ijid.2020.07.052}{10.1016/j.ijid.2020.07.052}

\leavevmode\vadjust pre{\hypertarget{ref-kawano2015substantial}{}}%
Kawano, S., Kakehashi, M., 2015. Substantial impact of school closure on the transmission dynamics during the pandemic flu H1N1-2009 in oita, japan. PloS one 10, e0144839. doi:\href{https://doi.org/10.1371/journal.pone.0144839}{10.1371/journal.pone.0144839}

\leavevmode\vadjust pre{\hypertarget{ref-kim2021role}{}}%
Kim, J., Choe, Y.J., Lee, J., Park, Y.J., Park, O., Han, M.S., Kim, J.-H., Choi, E.H., 2021. Role of children in household transmission of COVID-19. Archives of Disease in Childhood 106, 709--711. doi:\href{https://doi.org/10.1136/archdischild-2020-319910}{10.1136/archdischild-2020-319910}

\leavevmode\vadjust pre{\hypertarget{ref-kissler2020symbolic}{}}%
Kissler, S.M., Viboud, C., Grenfell, B.T., Gog, J.R., 2020. Symbolic transfer entropy reveals the age structure of pandemic influenza transmission from high-volume influenza-like illness data. Journal of the Royal Society Interface 17, 1--9. doi:\href{https://doi.org/10.1098/rsif.2019.0628}{10.1098/rsif.2019.0628}

\leavevmode\vadjust pre{\hypertarget{ref-lerkkanen2023reading}{}}%
Lerkkanen, M.-K., Pakarinen, E., Salminen, J., Torppa, M., 2023. Reading and math skills development among finnish primary school children before and after COVID-19 school closure. Reading and Writing 36, 263--288. doi:\href{https://doi.org/10.1007/s11145-022-10358-3}{10.1007/s11145-022-10358-3}

\leavevmode\vadjust pre{\hypertarget{ref-li2020role}{}}%
Li, X., Xu, W., Dozier, M., He, Y., Kirolos, A., Theodoratou, E., others, 2020. The role of children in transmission of SARS-CoV-2: A rapid review. Journal of Global Health 10, 1--10. doi:\href{https://doi.org/10.7189/jogh.10.011101}{10.7189/jogh.10.011101}

\leavevmode\vadjust pre{\hypertarget{ref-loader2006local}{}}%
Loader, C., 2006. Local regression and likelihood. Springer, New York, NY.

\leavevmode\vadjust pre{\hypertarget{ref-ludvigsson2020children}{}}%
Ludvigsson, J.F., 2020. Children are unlikely to be the main drivers of the COVID-19 pandemic -- a systematic review. Acta Paediatrica 109, 1525--1530. doi:\href{https://doi.org/10.1111/apa.15371}{10.1111/apa.15371}

\leavevmode\vadjust pre{\hypertarget{ref-molefi2021impact}{}}%
Molefi, M., Tlhakanelo, J.T., Phologolo, T., Hamda, S.G., Masupe, T., Tsima, B., Setlhare, V., Mashalla, Y., Wiebe, D.J., others, 2021. The impact of china's lockdown policy on the incidence of CoVID-19: An interrupted time series analysis. BioMed Research International 2021. doi:\href{https://doi.org/10.1155/2021/9498029}{10.1155/2021/9498029}

\leavevmode\vadjust pre{\hypertarget{ref-moser2018estimating}{}}%
Moser, C.B., White, L.F., 2018. Estimating age-specific reproductive numbers---a comparison of methods. Statistical Methods in Medical Research 27, 2050--2059. doi:\href{https://doi.org/10.1177/0962280216673676}{10.1177/0962280216673676}

\leavevmode\vadjust pre{\hypertarget{ref-moulin2022longitudinal}{}}%
Moulin, F., Bailhache, M., Monnier, M., Thierry, X., Vandentorren, S., Côté, S.M., Falissard, B., Simeon, T., Geay, B., Marchand, L., others, 2022. Longitudinal impact of psychosocial status on children's mental health in the context of COVID-19 pandemic restrictions. European Child \& Adolescent Psychiatry 1--10. doi:\href{https://doi.org/10.1007/s00787-022-02010-w}{10.1007/s00787-022-02010-w}

\leavevmode\vadjust pre{\hypertarget{ref-riffe2021data}{}}%
Riffe, T., Acosta, E., COVerAGE-DB team, the, 2021. {Data Resource Profile: COVerAGE-DB: a global demographic database of COVID-19 cases and deaths}. International Journal of Epidemiology 50, 390--390f. doi:\href{https://doi.org/10.1093/ije/dyab027}{10.1093/ije/dyab027}

\leavevmode\vadjust pre{\hypertarget{ref-schreiber2000measuring}{}}%
Schreiber, T., 2000. Measuring information transfer. Physical Review Letters 85, 461--464. doi:\href{https://doi.org/10.1103/PhysRevLett.85.461}{10.1103/PhysRevLett.85.461}

\leavevmode\vadjust pre{\hypertarget{ref-shannon1948mathematical}{}}%
Shannon, C.E., 1948. A mathematical theory of communication. The Bell System Technical Journal 27, 379--423. doi:\href{https://doi.org/10.1002/j.1538-7305.1948.tb01338.x}{10.1002/j.1538-7305.1948.tb01338.x}

\leavevmode\vadjust pre{\hypertarget{ref-siebach2021childhood}{}}%
Siebach, M.K., Piedimonte, G., Ley, S.H., 2021. COVID-19 in childhood: Transmission, clinical presentation, complications and risk factors. Pediatric Pulmonology 56, 1342--1356. doi:\href{https://doi.org/10.1002/ppul.25344}{10.1002/ppul.25344}

\leavevmode\vadjust pre{\hypertarget{ref-sorianoarandes2021household}{}}%
Soriano-Arandes, A., Gatell, A., Serrano, P., Biosca, M., Campillo, F., Capdevila, R., Fàbrega, A., Lobato, Z., López, N., Moreno, A.M., Poblet, M., Riera-Bosch, M.T., Rius, N., Ruiz, M., Sánchez, A., Valldepérez, C., Vilà, M., Pineda, V., Lazcano, U., Díaz, Y., Reyes-Urueña, J., Soler-Palacín, P., Catalonia Research Group, C.P.D. in, 2021. Household severe acute respiratory syndrome coronavirus 2 transmission and children: A network prospective study. Clinical Infectious Diseases 73, e1261--e1269. doi:\href{https://doi.org/10.1093/cid/ciab228}{10.1093/cid/ciab228}

\leavevmode\vadjust pre{\hypertarget{ref-stage2021shut}{}}%
Stage, H.B., Shingleton, J., Ghosh, S., Scarabel, F., Pellis, L., Finnie, T., 2021. Shut and re-open: The role of schools in the spread of COVID-19 in europe. Philosophical Transactions of the Royal Society B 376, 1--11. doi:\href{https://doi.org/10.1098/rstb.2020.0277}{10.1098/rstb.2020.0277}

\leavevmode\vadjust pre{\hypertarget{ref-sugimoto2023temporal}{}}%
Sugimoto, M., Murakami, K., Sasaki, S., 2023. Temporal patterns of sleep and eating among children during school closure in japan due to COVID-19 pandemic: Associations with lifestyle behaviours and dietary intake. Public Health Nutrition 26, 393--407. doi:\href{https://doi.org/10.1017/S1368980022001148}{10.1017/S1368980022001148}

\leavevmode\vadjust pre{\hypertarget{ref-yoshiyuki2020effects}{}}%
Sugishita, J.A.S., Yoshiyuki AND Kurita, 2020. Effects of voluntary event cancellation and school closure as countermeasures against COVID-19 outbreak in japan. PLOS ONE 15, 1--10. doi:\href{https://doi.org/10.1371/journal.pone.0239455}{10.1371/journal.pone.0239455}

\leavevmode\vadjust pre{\hypertarget{ref-unesco2022https}{}}%
{UNESCO} global education coalition, 2022.

\leavevmode\vadjust pre{\hypertarget{ref-walsh2021school}{}}%
Walsh, S., Chowdhury, A., Braithwaite, V., Russell, S., Birch, J.M., Ward, J.L., Waddington, C., Brayne, C., Bonell, C., Viner, R.M., others, 2021. Do school closures and school reopenings affect community transmission of COVID-19? A systematic review of observational studies. BMJ Open 11, e053371. doi:\href{https://doi.org/10.1136/bmjopen-2021-053371}{10.1136/bmjopen-2021-053371}

\leavevmode\vadjust pre{\hypertarget{ref-wood2011fast}{}}%
Wood, S.N., 2011. Fast stable restricted maximum likelihood and marginal likelihood estimation of semiparametric generalized linear models. Journal of the Royal Statistical Society (B) 73, 3--36. doi:\href{https://doi.org/10.1111/j.1467-9868.2010.00749.x}{10.1111/j.1467-9868.2010.00749.x}

\leavevmode\vadjust pre{\hypertarget{ref-wood2017generalized}{}}%
Wood, S.N., 2017. Generalized additive models: An introduction with r. CRC, Boca Raton, FL.

\leavevmode\vadjust pre{\hypertarget{ref-who2022https}{}}%
World health organization COVID-19 data, 2022.

\leavevmode\vadjust pre{\hypertarget{ref-zhendong2020clinical}{}}%
Zhen-Dong, Y., Gao-Jun, Z., Run-Ming, J., Zhi-Sheng, L., Zong-Qi, D., Xiong, X., Guo-Wei, S., 2020. Clinical and transmission dynamics characteristics of 406 children with coronavirus disease 2019 in china: A review. Journal of Infection 81, e11--e15. doi:\href{https://doi.org/10.1016/j.jinf.2020.04.030}{10.1016/j.jinf.2020.04.030}

\end{CSLReferences}

\end{document}

### 2024-02-26 manuscript_elsevier/appendix_2.Rmd
PH

@@
   ) |> 
-  kable_classic(full_width = F) |> 
+  kable_classic(full_width = TRUE) |> 
   collapse_rows(columns = 1:2, valign = "middle") |> 

### 2024-02-26 manuscript_elsevier/cover_letter.Rmd
PH

@@
 ---
-title: "Submission to Social Science & Medicine"
+title: "Submission to Public Health Journal"
 output: pdf_document
@@
   echo = FALSE)
-journal <- "Social Science & Medicine"
+journal <- "Public Health Journal"
 ```
@@
 
+All data and codes are in open access.
+The data can be accessed from the COVerAGE-DB OSF repository (https://osf.io/mpwjq/) and from the UNESCO servers (https://en.unesco.org/file/unesco-data-school-closures-february-2020-june-2022csv-zip).
+All preprocessing, analyses and code used to build the submitted paper are available at https://github.com/damien-dupre/covid_ts_causality.
+
 The list of authors:

### 2024-02-26 manuscript_elsevier/manuscript.Rmd
PH

--- 2024-02-20 manuscript_oup/manuscript.Rmd
@@
 ---
-title: "Examining the Impact of School Closures on COVID-19 Infections in Europe and their Effects on Different Age Cohorts"
-journal_title: "Journal of the Royal Statistical Society, Series C (Applied Statistics)"
-short_title: "Impact of School Closures on COVID-19 by Age Cohorts"
-short_author: "Dupré & Morgenroth"
-authors_1_name: "Damien Dupré"
-authors_1_ORCID: "0000-0001-8610-1045"
-authors_2_name: "Edgar Morgenroth"
-authors_2_ORCID: "0000-0002-9442-0561"
-authors_affiliation: "Business School, Dublin City University"
-authors_city: "Dublin"
-authors_country: "Ireland"
-corresponding_email: "damien.dupre\\@dcu.ie"
+title: Examining the Impact of School Closures on COVID-19 Infections in Europe and their Effects on Different Age Cohorts
+# author:
+#   - name: Damien Dupré
+#     email: damien.dupre@dcu.ie
+#     affiliation: Business School, Dublin City University
+#     correspondingauthor: true
+#   - name: Edgar Morgenroth
+#     email: edgar.morgenroth@dcu.ie
+#     affiliation: Business School, Dublin City University
+# address:
+#   - code: Business School, Dublin City University
+#     organization: Business School, Dublin City University
+#     city: Dublin
+#     country: Ireland
 abstract: |
-  This paper provides a thorough analysis of COVID-19 case trends in selected European countries, focusing on the influence of different age groups.
-It explores the relationship between population size and confirmed case numbers, suggesting that larger countries tend to have higher case figures.
-The study investigates how various waves of the pandemic affect different age demographics, with older age groups being most affected by the initial wave and younger age groups more impacted by subsequent waves.
-  Using a Generalised Additive Model (GAM), the research assesses the impact of school closures on COVID-19 cases, revealing a decreasing non-linear effect across all countries examined.
-Following school closures, distinct trends emerge among age groups: ages 0 to 4 experience a decrease, ages 5 to 14 see an increase, and ages 15 to 24 initially surge then decline.
-Transfer Entropy calculations uncover imbalances in age group influences, with certain age groups predicting changes in others.
-  Consistent patterns across countries are identified, such as younger cohorts influencing older cohorts in Austria, Germany, and the Netherlands, while Portugal and Spain show the opposite trend.
-These findings deepen our understanding of COVID-19 dynamics in Europe, providing valuable insights for targeted public health strategies and interventions.
-keywords: COVID-19, School Closures, Causality, Generalised Additive Model, Transfer Entropy
-bibliography: "bibliography.bib"
-csl: elsevier-harvard.csl
+  This paper analyzes the trends of COVID-19 cases in different age groups in selected European countries, assessing to what extent school closures helped reduce case numbers, how school closures affected case numbers in different age groups, and how COVID-19 spread between age groups.
+The study also evaluates the effects of school closures on COVID-19 cases using a Generalized Additive Model (GAM).
+The findings demonstrate a decreasing non-linear effect of school closure on the total COVID-19 case number across all countries studied.
+However, the analysis of age groups only reveals a consistent downward trend in infections in the 0 to 4 pre-school age group, while the school-going age groups 5 to 14 exhibit a significant increase in cases.
+Age groups 15 to 24 show a surge immediately after closure, followed by a decline.
+Transfer Entropy calculations highlight asymmetry in age group influences, indicating that changes in COVID-19 cases in certain age groups predict changes in other age groups but not vice versa.
+These findings contribute to a better understanding of COVID-19 dynamics in European countries and provide insights for public health strategies and interventions.
+keywords: 
+  - COVID-19
+  - School Closures
+  - Intergenerational Transmission.
+journal: "Public Health"
+date: "`r Sys.Date()`"
+classoption: preprint, 3p
+linenumbers: false
+numbersections: true
+bibliography: bibliography.bib
+# biblio-style: elsarticle-num
+csl: https://www.zotero.org/styles/elsevier-with-titles
+output:
+  bookdown::pdf_book:
+    base_format: rticles::elsevier_article
+    keep_tex: false
+    citation_package: default
 header-includes:
@@
   - \usepackage{caption}
-
-output:
-  bookdown::pdf_book:
-    template: oup-authoring-template.tex
-    keep_tex: false
 ---
@@
 
-```{r country, fig.height=5, out.width="\\textwidth", fig.cap="(ref:country)"}
+```{r country, fig.height=3, out.width="\\textwidth", fig.cap="(ref:country)"}
 gam_plot_country

### 2024-02-26 manuscript_elsevier/title_page.Rmd
HP

--- 2024-02-26 manuscript_elsevier/manuscript.Rmd
@@
 title: Examining the Impact of School Closures on COVID-19 Infections in Europe and their Effects on Different Age Cohorts
-# author:
-#   - name: Damien Dupré
-#     email: damien.dupre@dcu.ie
-#     affiliation: Business School, Dublin City University
-#     correspondingauthor: true
-#   - name: Edgar Morgenroth
-#     email: edgar.morgenroth@dcu.ie
-#     affiliation: Business School, Dublin City University
-# address:
-#   - code: Business School, Dublin City University
-#     organization: Business School, Dublin City University
-#     city: Dublin
-#     country: Ireland
+author:
+  - name: Damien Dupré
+    email: damien.dupre@dcu.ie
+    affiliation: Business School, Dublin City University
+    correspondingauthor: true
+  - name: Edgar Morgenroth
+    email: edgar.morgenroth@dcu.ie
+    affiliation: Business School, Dublin City University
+address:
+  - code: Business School, Dublin City University
+    organization: Business School, Dublin City University
+    city: Dublin
+    country: Ireland
 abstract: |
@@
 ---
-
-(ref:overall) Cumulative COVID-19 cases number for selected European countries since the beginning of the pandemic. Source: COVerAGE-DB [@riffe2021data].
-(ref:descriptive) Periods of school closure since the beginning of the COVID-19 pandemic for selected European countries and their reason: regular academic break vs.
-closure due to government decisions.
-Source: @unesco2022https.
-(ref:country) Standardized effect of the smooth term in Generalized Additive Model by country.
-(ref:age) Standardized effect of the smooth term in Generalized Additive Model by age group.
-(ref:te) Matrix of Transfer Entropy coefficients according to every age group combination for each of the selected European country.
-Age groups on the x-axis are influencing the age groups on the y-axis ($ag\,x \rightarrow\,ag\,y$).
-The significance of each Transfer Entropy coefficient is provided in Appendix 2.
-
-```{r setup, include = FALSE}
-# libraries --------------------------------------------------------------------
-library(ggrepel)
-library(here)
-library(janitor)
-library(knitr)
-library(kableExtra)
-library(osfr)
-library(scales)
-library(tidyverse)
-
-# options ----------------------------------------------------------------------
-set.seed(123) # Seed for random number generation
-options(scipen = 999) # disable sci number format
-opts_chunk$set(
-  cache.extra = rand_seed, 
-  cache = FALSE,
-  message = FALSE, 
-  warning = FALSE, 
-  error = FALSE,
-  echo = FALSE, 
-  fig.retina = 3,
-  dev = "ragg_png"
-  )
-
-# raw data ---------------------------------------------------------------------
-# osf_retrieve_file("7tnfh") |>
-#   osf_download(conflicts = "overwrite", progress = TRUE, verbose = TRUE)
-# 
-# df <- "Output_5.zip" |>
-#   read_csv(skip = 3, col_types = "ccccciiddd") |>
-#   mutate(Date = as.Date(Date, format = "%d.%m.%Y")) |>
-#   clean_names()
-# 
-# file.remove("Output_5.zip")
-# 
-# write_csv(df, here("data/df.csv"))
-# 
-# df <- read_csv(here("data/df.csv"))
-# 
-# df_school_closure <- 
-#   read_csv(
-#     "https://en.unesco.org/sites/default/files/covid_impact_education_full.csv",
-#     col_types = list(.default = col_character(), Date = col_date(format = "%d/%m/%Y"))) |> 
-#   clean_names() |>
-#   select(date, country, status) |> 
-#   mutate(country = case_when(
-#     country == "Bolivia (Plurinational State of)" ~ "Bolivia",  
-#     country == "Central African republic" ~ "Central African Republic",
-#     country == "Iran (Islamic Republic of)" ~ "Iran",
-#     country == "Republic of Moldova" ~ "Moldova",
-#     country == "Republic of Korea" ~ "South Korea",
-#     country == "United Kingdom of Great Britain and Northern Ireland" ~ "United Kingdom",
-#     country == "United States of America" ~ "USA",
-#     country == "Viet Nam" ~ "Vietnam",
-#     TRUE ~ country
-#     )
-#   )
-# 
-# df_check <- 
-#   read_csv("https://covid19.who.int/WHO-COVID-19-global-data.csv") |> 
-#   clean_names() |> 
-#   rename(date = date_reported) |> 
-#   mutate(country = case_when(
-#     country == "Bolivia (Plurinational State of)" ~ "Bolivia",  
-#     country == "Iran (Islamic Republic of)" ~ "Iran",
-#     country == "occupied Palestinian territory, including east Jerusalem" ~ "Palestine",
-#     country == "Republic of Moldova" ~ "Moldova",
-#     country == "Republic of Korea" ~ "South Korea",
-#     country == "United States of America" ~ "USA",
-#     country == "Venezuela (Bolivarian Republic of)" ~ "Venezuela",
-#     country == "Viet Nam" ~ "Vietnam",
-#     TRUE ~ country
-#     )
-#   )
-# 
-# # variables --------------------------------------------------------------------
-# ## keep only closure waves longer than `days_min_closure` but display only the first `days_after_closure`
-# days_min_closure <- 21
-# days_after_closure <- 28
-# ## type of school closure to investigate
-# period_closure_to_keep <- c(
-#   #"Partially open",
-#   "Academic break",
-#   "Closed due to COVID-19"
-#   )
-# 
-# ## countries of interest
-# countries_of_interest <- c(
-#   "Austria", "Belgium", "Bulgaria", "Croatia", "Estonia", "France", 
-#   "Germany", "Greece", "Netherlands", "Portugal", "Slovakia", "Spain"
-#   )
-# 
-# # transformed data -------------------------------------------------------------
-# df_global <- df |> 
-#   filter(
-#     country %in% countries_of_interest,
-#     date > as.Date("2020-01-01"),
-#     region == "All",
-#     sex == "b"
-#   ) |> 
-#   group_by(country, date) |> 
-#   summarise(cumulative_cases_cal = sum(cases)) |> 
-#   group_by(country) |> 
-#   complete(date = full_seq(date, 1)) |> 
-#   mutate(
-#     timeline = date - lag(date),  # check if gap between dates
-#     daily_cases = round(cumulative_cases_cal - lag(cumulative_cases_cal), 0)
-#   ) |> 
-#   ungroup() |> 
-#   left_join(df_school_closure, by = c("date", "country")) |> 
-#   group_by(country) |> 
-#   fill(status) |> 
-#   ungroup() |> 
-#   mutate(schools_closed = if_else(status %in% period_closure_to_keep, 1, 0)) |> 
-#   drop_na(status) |> 
-#   left_join(df_check, by = c("date", "country"))
-# 
-# write_csv(df_global, here("data/df_global.csv"))
-# 
-# df_global_age <- df |> 
-#   filter(
-#     country %in% countries_of_interest,
-#     date > as.Date("2020-01-01"),
-#     region == "All",
-#     sex == "b"
-#   ) |> 
-#   mutate(
-#     age_grp = case_when(
-#       age == 0  ~ "aged0to4",
-#       age == 5  ~ "aged5to9",
-#       age == 10 ~ "aged10to14",
-#       age == 15 ~ "aged15to19",
-#       age == 20 ~ "aged20to24",
-#       between(age, 25, 34) ~ "aged25to34",
-#       between(age, 35, 44) ~ "aged35to44",
-#       between(age, 45, 54) ~ "aged45to54",
-#       between(age, 55, 64) ~ "aged55to64",
-#       age > 64 ~ "aged65+"
-#     )
-#   ) |> 
-#   group_by(country, date, age_grp) |> 
-#   summarise(cumulative_cases_cal = sum(cases)) |> 
-#   group_by(country, age_grp) |> 
-#   complete(date = full_seq(date, 1)) |> 
-#   mutate(
-#     timeline = date - lag(date),  # check if gap between dates
-#     daily_cases = round(cumulative_cases_cal - lag(cumulative_cases_cal), 0),
-#     daily_changes = daily_cases - lag(daily_cases)
-#   ) |> 
-#   ungroup()
-# 
-# write_csv(df_global_age, here("data/df_global_age.csv"))
-
-df_global <- read_csv(here("data/df_global.csv"))
-df_global_age <- read_csv(here("data/df_global_age.csv"))
-```
-
-```{r analyses, cache=FALSE}
-source(here("analysis/gam_analysis_europe.R"))
-source(here("analysis/te_analysis_europe.R"))
-```
-
-# Introduction
-
-A wide range of non-pharmaceutical interventions (NPI) were enacted to control the spread of COVID-19, especially before the availability of effective vaccines.
-Due to their social mixing patterns, children have been identified as an age group that can drive the spread of respiratory infections such as influenza [@moser2018estimating], and school closures were found to be effective in mitigating the spread of influenza H1N1 in Japan in 2009 [@kawano2015substantial].
-It is therefore not surprising that school closures were one of the NPIs that were introduced in many countries to stop the spread of COVID-19.
-They were used particularly during the initial wave of the pandemic and during subsequent waves when case numbers were high.
-
-Recently, a body of literature has shown various negative consequences on child development and health due to school closures.
-School closures have been found to have negatively affected child mental health [@moulin2022longitudinal], nutrition/obesity [@sugimoto2023temporal], and education [@lerkkanen2023reading].
-Some governments are now questioning their decisions to close schools [@de2021determines].
-For example, the German health minister Karl Lauterbach, in an interview with one of Germany’s publicly funded television stations, admitted that “in retrospect it had been wrong to keep schools and childcare closed for so long” [@ard2023lauterbach].
-
-Even if school closures had negative consequences for children, they might nevertheless have been effective at stopping the spread of COVID-19.
-However, analysis on the effectiveness of school closures on the spread of COVID-19 remains inconclusive.
-@alfano2022effects in a study of a panel of European countries found that school closures were associated with a reduced COVID-19 incidence.
-In contrast, @walsh2021school in a review of 40 studies covering 150 countries concluded that the effectiveness of school closures was uncertain, with 60% having identified no impact and pointing to the potential for the analysis to be affected by confounding factors and collinearity.
-The latter shortcomings suggest that it is hard to draw firm inferences, as even a positive result may not imply causality.
-
-This paper investigates the dynamics of COVID-19 across age groups and how this was affected by school closures.
-This allows us to identify the effect of school closure on the evolution of COVID-19 cases in other age groups and particularly on older and more vulnerable age groups, including parents, grandparents, and relatives about which little is known.
-
-We identify the relationship between the school closures and case numbers in different age groups using Generalized Additive Models (GAM), which are more flexible than conventional Vector Autoregressions, given that they allow for non-linear relationships between variables.
-Furthermore, compared to other methods such as Vector Autoregression, GAM can handle fixed and random smooth effects without making strict assumptions about linearity.
-Then, using the transfer entropy method, we evaluate the impact of case number changes between age groups in order to identify if school closure has secondary repercussions in the household.
-Transfer entropy can adapt to changes in the underlying system dynamics over time.
-It does not assume stationarity and can capture evolving causal relationships, making it suitable for studying dynamic systems and non-stationary processes such as COVID-19 contaminations.
-
-## Influence of School Closure
-
-Once it was clear that COVID-19 spreads from person to person, it was natural that governments were advised to enact measures to reduce social contact in order to reduce the spread of the virus.
-Therefore, it is legitimate to believe that by closing schools, a reduction of the contaminations would be observed in the younger age groups.
-However, the efficiency of school closure on the reduction of COVID-19 cases is still questioned [@bayham2020impact; @esposito2021comprehensive].
-While some research has found that school closures contribute to limit or to reduce the growth rate of confirmed cases after implementation [@stage2021shut; @yoshiyuki2020effects], others did not observe a change in the evolution of COVID-19 cases [@chang2020modelling; @iwata2020school].
-For instance, a controlled comparison between similar localities in Japan with schools closed and schools open did not reveal any evidence that school closures reduced the spread of COVID-19 [@fukumoto2021no].
-If the school closure had a real impact on the evolution of confirmed COVID-19 cases, it should be possible to observe a decrease or at least an inflection in the trend of its evolution among younger age groups.
-
-## Causal Relationship Between Age Groups
-
-A second implicit belief regarding the effect of school closure on the spread of COVID-19 is that school not only influences the spread of the virus among children and teenagers but also has a knock-on effect on the spread of the virus in older age groups, also called Secondary Attack Rate (SAR).
-The contaminated children and teenagers would bring the virus back home and would pass the virus on to their parents and relatives.
-For example, research investigating the contamination in the household network not only revealed an exceptionally high rate of secondary contamination but also that this contamination happened when the schools were closed [@sorianoarandes2021household].
-
-Despite being reported in several clinical and epidemiological studies [@siebach2021childhood; @zhendong2020clinical], multiple research have shown that the SAR from children to household members was lower than expected [@heavey2020evidence; @vanderhoek2020role; @kim2021role; @ludvigsson2020children].
-However, the SAR of children and teenagers to the household member is likely to be age-dependent, with differences between infants, primary and secondary school children, and college students [@grasleguen2021reopening].
-If a secondary transmission from children and teenagers to household member has a significant influence, then a temporal causality relationship between their evolution should be observed.
-
-To evaluate the impact of school closure on COVID-19 cases across different age groups, we first performed a non-linear time series regression to analyze the relationship between school closures and the number of COVID-19 cases in various age groups.
-In a second analysis, the transfer entropy method is employed to assess the influence of changes in case numbers between age groups, aiming to determine if school closures have secondary effects within households.
-
-# Estimation Methodology
-
-## Generalized Additive Model
-
-The effect of school closure on the trend of daily COVID-19 cases is analyzed using a Generalized Additive Model (GAM).
-GAM is a flexible modeling approach that estimates non-linear relationships between variables.
-Compared to other methods such as Vector Autoregression, GAM can handle fixed and random smooth effects without making strict assumptions about linearity.
-
-The GAM is fitted on the daily COVID-19 cases to test the hypothesis of a significant non-linear evolution of cases among age groups from 0 to 4, from 5 to 9, from 10 to 14, from 15 to 19, and from 20 to 24 [@wood2017generalized].
-The model also estimates the overall non-linear effect by country, taking into account the interaction between age groups and countries as random intercepts and the interaction between time, countries, and the period of closure as random effects (Eq 1).
-
-By estimating the degree of smoothness of a Bayesian spline smoothing using restricted fast maximum likelihood estimation [@wood2011fast], GAM identifies dynamic patterns underlying the evolution of COVID-19 cases reported while including the random effect of different age groups and countries as follows:
-
-\begin{align*}
-  \tag{1}
-  n\prime_t &\sim Poisson(\mu_t) \\
-  \log(\mu_t) &= \beta_0 + \beta_1\,wave_t + \beta_2\,country_t + \beta_3\,age\,group_t + \\
-              & f_1(closure_t) + f_2(closure_t,country_t) + f_3(closure_t,age\,group_t)
-\end{align*}
-
-\noindent where $n\prime_t$ represents the confirmed COVID-19 cases, assuming a Poisson distribution for the fitting [@loader2006local], and $t$ is the date corresponding to the confirmed COVID-19 cases.
-The response variable includes a specific random effect taking into account variation within waves of school closure, countries, and age groups.
-The terms $f_1$ to $f_3$ are smooth functions of the time since closure, the time since closure for each country, and the time since closure for each age group.
-The restricted maximum likelihood (REML) was used to avoid overfitting while estimating smoothing parameters.
-In order to accurately account for the autocorrelation arising from the time series data, the residuals are modeled using an AR1 error model.
-By incorporating the autoregressive component, the AR1 model acknowledges the dependence of each residual on its previous value, thus providing a comprehensive representation of the data’s temporal dynamics.
-
-## Transfer Entropy
-
-Transfer entropy ($T$) can be used to infer the temporal relationship between two time series $X$ and $Y$.
-This measure indicates whether changes in COVID-19 cases within a certain age group $X$ can be used to reduce the uncertainty on the future COVID-19 cases within another age group $Y$.
-If it does, $T$ is considered evidence of a causal effect from the age group $X$ to the age group $Y$ [@schreiber2000measuring].
-As such, Granger causality is a special case of transfer entropy applied to time series that are jointly Gaussian distributed [@barnett2009granger].
-Therefore, transfer entropy is a more robust analysis of time series, especially when applied to the impact of age cohorts on pandemic transmission [@kissler2020symbolic].
-
-The influence of the evolution in COVID-19 cases across all age groups is evaluated using Shannon’s transfer entropy, given by:
-
-\begin{align}
-  \tag{2}
-  T_{ag\,x \rightarrow\,ag\,y}(k,l) = \sum_{ag\,x,ag\,y} p\left(ag\,x_{t+1}, ag\,x_t^{(k)}, ag\,y_t^{(l)}\right) \log \left(\frac{p\left(ag\,x_{t+1}| ag\,x_t^{(k)}, ag\,y_t^{(l)}\right)}{p\left(ag\,x_{t+1}|ag\,x_t^{(k)}\right)}\right)
-\end{align}
-
-\noindent where $T_{ag\,x \rightarrow\,ag\,y}$ consequently measures the influence of the change dynamic from an age group $X$ (or $ag\,x$) to another age group $Y$ (or $ag\,y$) for every country (Eq 2).
-
-The day-by-day difference in COVID-19 confirmed cases $n\prime\prime_t$ is used to satisfy the stationary requirement for the calculation of Shannon’s Transfer Entropy [@shannon1948mathematical; @behrendt2019rtransferentropy].
-
-# Data
-
-## COVID-19 cases per age group
-
-For the analysis, data on COVID case counts by age group are required.
-The number of COVID-19 cases per age group has been taken from the COVerAGE-DB project [@riffe2021data].
-The COVerAGE-DB project consists of 3 data files: an “Input” data file that collects the official COVID-19 cases from 117 countries, the “Output_5” data file, which is a projection of COVID-19 cases by a group of 5 years, and the “Output_10” data file, which is a projection of COVID-19 cases by a group of 10 years.
-For our analyses, we have used the “Output_5,” which deals with the heterogeneity of countries’ reporting formats by using spline approximations when the data for this age bracket is not available for a country.
-Therefore, it allows a precise analysis between age groups under 24, while age groups of 25 and above are reduced to 10-year brackets, and groups above 64 are concatenated all together.
-
-In order to cross-validate the data obtained after spline approximations, a comparison with the data published by the @who2022https reveals perfect similarities.
-The original data consist of 14,089,320 observations of 10 variables (117 distinct countries, region within the country, a unique observation code, the date of the observation, the gender, which can be male, female, or both, the age bracket by 5 years from 0 to 100, a confirmation of the age interval for each bracket, the total number of cases so far, the total number of deaths, and the total number of tests performed) from `r format(min(df_global$date), "%B %d, %Y")` to `r format(max(df_global$date), "%B %d, %Y")`.
-After removing countries with missing and inconsistent values, only 22 are suitable for data analyses.
-However, to focus this analysis on geographically and culturally comparable countries, only 12 European countries are kept: Austria, Belgium, Bulgaria, Croatia, Estonia, France, Germany, Greece, Netherlands, Portugal, Slovakia, and Spain.
-The observations are reported in terms of the total number of COVID-19 cases per day from the start of the pandemic.
-The daily number of cases at a specific date $n\prime_{t}$ is calculated with the difference between the total cases at a date $t$ and the total cases at a date $t-1$ (i.e., derivative 1).
-In addition, the change in the daily number of cases $n\prime\prime_{t}$ between $n\prime_{t}$ and $n\prime_{t-1}$ has also been calculated (i.e., derivative 2).
-
-## School closure
-
-To identify the impact of school closures on the number of cases, information regarding the school closures on a day-by-day basis is required, and this is taken from the “UNESCO global education coalition” (2022).
-For each day, in each country, the status of the schools is indicated as fully open, partially open, closed due to COVID-19, or closed due to an academic break.
-Because it would be difficult to measure the effect of school closures at a country level when schools are partially closed, only closures due to COVID-19 or due to an academic break are considered.
-Indeed, both are considered as closure at a country-wide level.
-Note that periods of closure longer than 21 days are compared for 28 days, regardless of whether the schools opened or not after 21 days.
-Any COVID-19 case numbers beyond the 28-day period are not relevant for evaluating the impact of school closures.
-
-# Results
-
-The data show that the trend of confirmed COVID-19 cases follows similar patterns across the selected European countries, with scales following the size of the population in these countries (Figure \@ref(fig:overall)).
-Thus, the bigger the country, the higher the total number of COVID-19 cases.
-
-```{r overall, fig.height=3, out.width="\\textwidth", fig.cap="(ref:overall)"}
-overall_plot <- df_global |> 
-  mutate(
-    country = fct_rev(country)
-  ) |> 
-  ggplot(aes(date, cumulative_cases_cal, group = country)) +
-  geom_line(na.rm = TRUE, color = "gray40", alpha = 0.4) +
-  geom_text_repel(
-    data = df_global |>
-      group_by(country) |>
-      slice_max(cumulative_cases_cal, with_ties = FALSE),
-    aes(label = country),
-    hjust = 0,
-    direction = "y",
-    segment.color = NA,
-    seed = 123,
-    size = 2,
-    force = 0.5,
-    family = "serif"
-  ) +
-  scale_x_date(
-    "Date", 
-    date_breaks = "3 months", 
-    date_labels = "%b %y", 
-    limits = c(as.Date("2020-02-01"), as.Date("2022-05-01"))
-  ) +
-  scale_y_continuous(
-    "Total COVID-19 Cases (Million)", 
-    labels = unit_format(unit = "M", scale = 1e-6)
-  ) +
-  theme_bw() +
-  theme(
-    legend.position = "none",
-    text = element_text(size = 10, family = "serif")
-  )
-
-overall_plot
-```
-
-The evolution of COVID-19 cases reveals some similarities across all age groups.
-However, the influence of each wave on individual age groups also has some particularities (Figure \@ref(fig:descriptive)).
-For example, the first wave was more important among the oldest age groups, whereas the third wave was more important among the youngest age groups.
-
-```{r descriptive, fig.height=3, out.width="\\textwidth", fig.cap="(ref:descriptive)"}
-descriptive_plot <- df_global |> 
-  filter(schools_closed == 1) |> 
-  mutate(
-    status = status |> 
-      str_remove("Closed due to ") |> 
-      fct_rev(),
-    country = country |> 
-      fct_rev()
-  ) |> 
-  ggplot(aes(date, country, color = status)) +
-  geom_point(na.rm = TRUE) +
-  scale_x_date("Date", date_breaks = "3 months", date_labels = "%b %y") +
-  scale_y_discrete("Selected European Countries") +
-  scale_colour_manual("Schools Closure Reason", values = c("gray10", "gray60")) +
-  theme_bw() +
-  theme(
-    legend.position = "bottom",
-    text = element_text(size = 10, family = "serif")
-  )
-
-descriptive_plot
-```
-
-As outlined above, to evaluate the shape of the trend in the numbers of COVID-19 cases reported after the three school closures longer than 21 consecutive days, a GAM was fitted as described above, taking into account the overall effect across all the selected European countries, as well as the effect for age groups: 0 to 4, 5 to 9, 10 to 14, 15 to 19, and 20 to 24 years old.
-The obtained results satisfy the requirements to fit this model, which explains `r percent(gam_summary$dev.expl, accuracy = 0.1)` of the variation in COVID-19 cases.
-
-Overall, the results revealed a decreasing non-linear effect of school closure at a country level for the selected European countries (Austria: `r gam_results["Austria"]`; Belgium: `r gam_results["Belgium"]`; Bulgaria: `r gam_results["Bulgaria"]`; Croatia: `r gam_results["Croatia"]`; Estonia: `r gam_results["Estonia"]`; France: `r gam_results["France"]`; Germany (`r gam_results["Germany"]`); Greece: `r gam_results["Greece"]`; The Netherlands: `r gam_results["Netherlands"]`; Portugal: `r gam_results["Portugal"]`; Slovakia (`r gam_results["Slovakia"]`; Spain: `r gam_results["Spain"]`, see Figure \@ref(fig:age)).
-
-```{r country, fig.height=3, out.width="\\textwidth", fig.cap="(ref:country)"}
-gam_plot_country
-```
-
-```{r age, fig.height=3, out.width="\\textwidth", fig.cap="(ref:age)"}
-gam_plot_age
-```
-
-```{r te, fig.height=8, out.width="\\textwidth", fig.cap="(ref:te)", fig.pos = 'H'}
-te_plot
-```
-
-Considering all countries, the analysis of age groups reveals distinct patterns (Figure \@ref(fig:age)).
-For the 0 to 4 age group, i.e., the pre-school group, there was a downward trend during the initial two weeks following school closure, followed by flattening off (`r gam_results["0to4"]`).
-In contrast, for the age group ranging from 5 to 14, a substantial increase in cases after 14 days of school closure is found (5 to 9 age group: `r gam_results["5to9"]` and 10 to 14 age group: `r gam_results["10to14"]`).
-For age groups between 15 and 24, there is a notable surge immediately after school closure, followed by a decrease after 14 days (15 to 19 age group: `r gam_results["15to19"]` and 20 to 24 age group: `r gam_results["20to24"]`).
-Thus, while the total case number declined following school closures, among younger age groups only the pre-school age group had a consistent reduction in cases, while for other age groups, there was even an increase in cases following school closures.
-
-Given that school closures were not only aimed at reducing COVID-19 cases among school-going children but also age groups, it is important to test the degree to which there was transmission across age groups, which is done using transfer entropy analysis.
-Before this analysis is carried out, an Augmented Dickey-Fuller has been applied to each age group to ensure that the daily changes in COVID-19 cases are stationary (see Appendix 1).
-
-The results of the transfer entropy calculations between age groups for each of the 12 selected European countries are reported in Figure \@ref(fig:te).
-The absence of symmetry between influencing age groups (i.e., $ag\,x$) and influenced age groups (i.e., $ag\,y$) is found.
-Indeed, the change in COVID-19 cases in some age groups is influenced by other age groups, but they are not reciprocally influencing these age groups.
-Figure \@ref(fig:te) shows how different age cohorts influence the COVID-19 case numbers of all age groups across countries.
-The upper left quadrant indicates how younger age cohorts are influencing older age cohorts, the bottom right quadrant indicates how older age cohorts are influencing younger age cohorts, finally, the lower left and upper right indicate how younger or older age cohorts are influencing themselves.
-By analyzing these quadrants, it is possible to identify similar patterns across multiple countries.
-Indeed, it appears that Austria, Germany, and The Netherlands have significantly higher $T$ coefficients in the upper left quadrant of the matrix, which indicates that the daily changes in COVID-19 cases number in younger cohorts are predicting the daily changes in COVID-19 cases number in older cohorts.
-Alternatively, it appears that Austria, the Netherlands, Portugal, and Spain have significantly higher $T$ coefficients in the lower right quadrant of the matrix, which indicates that the daily changes in COVID-19 cases number in older cohorts are predicting the daily changes in COVID-19 cases number in younger cohorts.
-
-# Discussion & Conclusion
-
-Knowledge about the transmission of the virus significantly improved over time as more studies have been published.
-While an early study found that children did not play an important role in the transmission of the virus [@li2020role], more recent results give a more nuanced position, stating that the spread of the virus in children is moderate.
-This study, using robust statistical methods, considered the effect of school closures on COVID-19 cases across age groups.
-While there are commonalities in the evolution of COVID-19 cases across all European countries that were included in our data, school closures exhibit distinct impacts on different age groups.
-Notably, the analysis shows that the 0 to 4 age group experiences a downward trend in COVID-19 cases during the initial two weeks following school closure, followed by stabilization.
-In contrast, age groups ranging from 5 to 14 exhibit a substantial increase in cases after 14 days of school closure.
-Additionally, age groups between 15 and 24 demonstrate a notable surge immediately after school closure, followed by a decrease after 14 days.
-Thus, the results do not support the hypothesis that school closures were effective.
-These results partially replicate observations from @alfano2022effects while providing a clearer picture of the structure of the effect of school closure.
-
-The Transfer Entropy calculations between age groups for each of the 12 selected European countries reveal an absence of symmetry between influencing age groups (X) and influenced age groups (Y).
-Some age groups influence the COVID-19 case numbers of other age groups, but there is no reciprocal influence.
-This finding suggests that changes in COVID-19 cases in certain age groups can predict the changes in other age groups but not vice versa, i.e., causality can be unidirectional.
-These findings highlight the importance of considering intergenerational interactions in designing effective control measures.
-
-Furthermore, the quadrant analysis reveals similar patterns across multiple countries.
-Austria, Germany, and the Netherlands exhibit significantly higher TE coefficients in the upper left quadrant, indicating that changes in COVID-19 cases in younger cohorts predict the changes in older cohorts.
-On the other hand, Austria, the Netherlands, Portugal, and Spain show higher TE coefficients in the lower right quadrant, indicating that changes in COVID-19 cases in older cohorts predict the changes in younger cohorts.
-
-In conclusion, this study sheds light on the trends and patterns of confirmed COVID-19 cases in selected European countries, with a particular focus on the influence of age groups and school closures.
-While a reduction in overall COVID-19 cases was found during school closures, which might suggest that like lockdowns [@alfano2020efficacy; @molefi2021impact], prolonged school closures could contribute to mitigating the spread of the virus, the results for the school-going age groups suggest that this is not the case.
-
-# Data Availability
-
-All data extracted from the COVerAGE-DB project and from the UNESCO global education coalition that are used to produce these results are fully available here https://github.com/damien-dupre/covid_ts_causality.
-The repository also includes the R code that were used to produce these results.
-
-# References {-}
-
-<!-- <div id="refs"></div> -->
-
-<!-- \newpage -->
-
-<!-- # Appendix 1: Dickey-Fuller test for stationay timeseries {-}  -->
-
-<!-- Test for the null hypothesis of a unit root of a univarate time series (equivalently, non-stationary time series). -->
-
-<!-- The Augmented Dickey-Fuller test incorporates three types of linear regression models.
-The first type is a linear model with no drift and linear trend with respect to time.
-The second type is a linear model with drift but no linear trend.
-The third type is a linear model with both drift and linear trend.
--->
-
-<!-- The p.value is calculated by interpolating the test statistics from the corresponding critical values tables (Fuller, 1996) for each type of linear models with given sample size.  -->
-
-<!-- ```{r} -->
-<!-- adf_res_type1 <- df_global_age |> -->
-<!--   group_by(age_grp, country) |> -->
-<!--   group_modify(~ aTSA::adf.test(.x$daily_changes, output = FALSE, nlag = 2) |> -->
-<!--                  bind_rows() |>  -->
-<!--                  pluck("type1") |>  -->
-<!--                  as_tibble() -->
-<!--   ) |>  -->
-<!--   rename(ADF_type1 = ADF, p.value_type1 = p.value) -->
-
-<!-- adf_res_type2 <- df_global_age |> -->
-<!--   group_by(age_grp, country) |> -->
-<!--   group_modify(~ aTSA::adf.test(.x$daily_changes, output = FALSE, nlag = 2) |> -->
-<!--                  bind_rows() |>  -->
-<!--                  pluck("type2") |>  -->
-<!--                  as_tibble() -->
-<!--   ) |>  -->
-<!--   rename(ADF_type2 = ADF, p.value_type2 = p.value) -->
-
-<!-- adf_res_type3 <- df_global_age |> -->
-<!--   group_by(age_grp, country) |> -->
-<!--   group_modify(~ aTSA::adf.test(.x$daily_changes, output = FALSE, nlag = 2) |> -->
-<!--                  bind_rows() |>  -->
-<!--                  pluck("type3") |>  -->
-<!--                  as_tibble() -->
-<!--   ) |>  -->
-<!--   rename(ADF_type3 = ADF, p.value_type3 = p.value) -->
-
-<!-- adf_res <- adf_res_type1 |>  -->
-<!--   full_join(adf_res_type2) |>  -->
-<!--   full_join(adf_res_type3) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- adf_res |> -->
-<!--   filter(age_grp %in% c("aged0to4", "aged5to9")) |>  -->
-<!--   kbl( -->
-<!--     caption = "Dickey-Fuller test for all combinaisons of countries and age groups investigated", -->
-<!--     col.names = c("Age Group", "Country", "Lag", "ADF", "p", "ADF", "p", "ADF", "p") -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   add_header_above(c(" " = 3, "Type 1" = 2, "Type 2" = 2, "type 3" = 2)) |> -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- adf_res |> -->
-<!--   filter(age_grp %in% c("aged10to14", "aged15to19")) |>  -->
-<!--   kbl( -->
-<!--     caption = "Dickey-Fuller test for all combinaisons of countries and age groups investigated", -->
-<!--     col.names = c("Age Group", "Country", "Lag", "ADF", "p", "ADF", "p", "ADF", "p") -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   add_header_above(c(" " = 3, "Type 1" = 2, "Type 2" = 2, "type 3" = 2)) |> -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- adf_res |> -->
-<!--   filter(age_grp %in% c("aged20to24", "aged25to34")) |>  -->
-<!--   kbl( -->
-<!--     caption = "Dickey-Fuller test for all combinaisons of countries and age groups investigated", -->
-<!--     col.names = c("Age Group", "Country", "Lag", "ADF", "p", "ADF", "p", "ADF", "p") -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   add_header_above(c(" " = 3, "Type 1" = 2, "Type 2" = 2, "type 3" = 2)) |> -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- adf_res |> -->
-<!--   filter(age_grp %in% c("aged35to44", "aged45to54")) |>  -->
-<!--   kbl( -->
-<!--     caption = "Dickey-Fuller test for all combinaisons of countries and age groups investigated", -->
-<!--     col.names = c("Age Group", "Country", "Lag", "ADF", "p", "ADF", "p", "ADF", "p") -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   add_header_above(c(" " = 3, "Type 1" = 2, "Type 2" = 2, "type 3" = 2)) |> -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- adf_res |> -->
-<!--   filter(age_grp %in% c("aged55to64", "aged65+")) |>  -->
-<!--   kbl( -->
-<!--     caption = "Dickey-Fuller test for all combinaisons of countries and age groups investigated", -->
-<!--     col.names = c("Age Group", "Country", "Lag", "ADF", "p", "ADF", "p", "ADF", "p") -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   add_header_above(c(" " = 3, "Type 1" = 2, "Type 2" = 2, "type 3" = 2)) |> -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- # Appendix 2: Significance of Transfer Entropy coefficients -->
-
-<!-- Results for each permutation of age groups from each country.
-Tables include transfer entropy estimates, the effective transfer entropy estimates, standard errors and p-values based on bootstrap replications of the Markov chains under the null hypothesis of statistical independence.
--->
-
-<!-- ```{r} -->
-<!-- # variables -------------------------------------------------------------------- -->
-<!-- age_order <- c( -->
-<!--   "aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24", -->
-<!--   "aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+" -->
-<!-- ) -->
-
-<!-- # data ------------------------------------------------------------------------- -->
-<!-- te_results <- read_csv(here("data/te_results.csv")) |>  -->
-<!--     mutate( -->
-<!--       X = factor(X, levels = age_order), -->
-<!--       Y = factor(Y, levels = age_order) -->
-<!--     ) |>  -->
-<!--   select(-direction) -->
-<!-- ``` -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged0to4") |>  -->
-<!--   filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged0to4") |>  -->
-<!--   filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged5to9") |>  -->
-<!--   filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged5to9") |>  -->
-<!--   filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged10to14") |>  -->
-<!--   filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged10to14") |>  -->
-<!--   filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged15to19") |>  -->
-<!--   filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged15to19") |>  -->
-<!--   filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged20to24") |>  -->
-<!--   filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged20to24") |>  -->
-<!--   filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged25to34") |>  -->
-<!--   filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged25to34") |>  -->
-<!--   filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged35to44") |>  -->
-<!--   filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged35to44") |>  -->
-<!--   filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged45to54") |>  -->
-<!--   filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged45to54") |>  -->
-<!--   filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged55to64") |>  -->
-<!--   filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged55to64") |>  -->
-<!--   filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged65+") |>  -->
-<!--   filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged65+") |>  -->
-<!--   filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->

### 2024-02-26 manuscript_elsevier/manuscript.Rmd
abstract

@@
 abstract: |
-  This paper analyzes the trends of COVID-19 cases in different age groups in selected European countries, assessing to what extent school closures helped reduce case numbers, how school closures affected case numbers in different age groups, and how COVID-19 spread between age groups.
-The study also evaluates the effects of school closures on COVID-19 cases using a Generalized Additive Model (GAM).
-The findings demonstrate a decreasing non-linear effect of school closure on the total COVID-19 case number across all countries studied.
+  *Objectives:* This paper analyzes the trends of COVID-19 cases in different age groups in selected European countries, assessing to what extent school closures helped reduce case numbers, how school closures affected case numbers in different age groups, and how COVID-19 spread between age groups.
+
+  *Study Design:* A longitudinal study was conducted to observe changes in COVID-19 cases during school closures.
+
+  *Methods:* A Generalized Additive Model is performed to compare trends from twelve European countries across five age groups.
+Then, Transfer Entropy calculations are used to evaluate the impact of younger cohorts on older cohorts.
+
+  *Results:* The findings demonstrate a decreasing non-linear effect of school closure on the total COVID-19 case number across all countries studied.
 However, the analysis of age groups only reveals a consistent downward trend in infections in the 0 to 4 pre-school age group, while the school-going age groups 5 to 14 exhibit a significant increase in cases.
@@
 Transfer Entropy calculations highlight asymmetry in age group influences, indicating that changes in COVID-19 cases in certain age groups predict changes in other age groups but not vice versa.
-These findings contribute to a better understanding of COVID-19 dynamics in European countries and provide insights for public health strategies and interventions.
+
+  *Conclusions:* These findings contribute to a better understanding of COVID-19 dynamics in European countries and provide insights for public health strategies and interventions.
 keywords: 
@@
 linenumbers: false
-numbersections: true
+numbersections: false
 bibliography: bibliography.bib
@@
 For example, the German health minister Karl Lauterbach, in an interview with one of Germany’s publicly funded television stations, admitted that “in retrospect it had been wrong to keep schools and childcare closed for so long” [@ard2023lauterbach].
-
 Even if school closures had negative consequences for children, they might nevertheless have been effective at stopping the spread of COVID-19.
@@
 The latter shortcomings suggest that it is hard to draw firm inferences, as even a positive result may not imply causality.
-
-This paper investigates the dynamics of COVID-19 across age groups and how this was affected by school closures.
-This allows us to identify the effect of school closure on the evolution of COVID-19 cases in other age groups and particularly on older and more vulnerable age groups, including parents, grandparents, and relatives about which little is known.
-
-We identify the relationship between the school closures and case numbers in different age groups using Generalized Additive Models (GAM), which are more flexible than conventional Vector Autoregressions, given that they allow for non-linear relationships between variables.
-Furthermore, compared to other methods such as Vector Autoregression, GAM can handle fixed and random smooth effects without making strict assumptions about linearity.
-Then, using the transfer entropy method, we evaluate the impact of case number changes between age groups in order to identify if school closure has secondary repercussions in the household.
-Transfer entropy can adapt to changes in the underlying system dynamics over time.
-It does not assume stationarity and can capture evolving causal relationships, making it suitable for studying dynamic systems and non-stationary processes such as COVID-19 contaminations.
-
-## Influence of School Closure
 
@@
 
-## Causal Relationship Between Age Groups
-
 A second implicit belief regarding the effect of school closure on the spread of COVID-19 is that school not only influences the spread of the virus among children and teenagers but also has a knock-on effect on the spread of the virus in older age groups, also called Secondary Attack Rate (SAR).
@@
 For example, research investigating the contamination in the household network not only revealed an exceptionally high rate of secondary contamination but also that this contamination happened when the schools were closed [@sorianoarandes2021household].
-
 Despite being reported in several clinical and epidemiological studies [@siebach2021childhood; @zhendong2020clinical], multiple research have shown that the SAR from children to household members was lower than expected [@heavey2020evidence; @vanderhoek2020role; @kim2021role; @ludvigsson2020children].
@@
 
-# Estimation Methodology
+# Methods
 
-## Generalized Additive Model
+## Study design
+
+A longitudinal study was conducted to observe changes in COVID-19 cases during school closures.
+Changes in COVID-19 from five age groups from the first to the 28th day of school closure are compared across 12 countries.
+Then, a causality analysis evaluate the impact of changes in these younger groups on older age groups.
+
+## Data collection
+
+Data on COVID-19 case counts by age group have been obtained from the COVerAGE-DB project [@riffe2021data].
+The COVerAGE-DB which collect numbers for every countries by groups of 5 or 10 years if available.
+In addition, spline approximations are used to deals with the heterogeneity of countries’ reporting formats by using when the data for this age bracket is not available for a country.
+
+To identify the impact of school closures on the number of cases, information regarding the school closures on a day-by-day basis is required, and this is taken from the “UNESCO global education coalition” (2022).
+For each day, in each country, the status of the schools is indicated as fully open, partially open, closed due to COVID-19, or closed due to an academic break.
+Because it would be difficult to measure the effect of school closures at a country level when schools are partially closed, only closures due to COVID-19 or due to an academic break are considered.
+Indeed, both are considered as closure at a country-wide level.
+Note that periods of closure longer than 21 days are compared for 28 days, regardless of whether the schools opened or not after 21 days.
+Any COVID-19 case numbers beyond the 28-day period are not relevant for evaluating the impact of school closures.
+
+## Data validation
+
+In order to cross-validate the data obtained after spline approximations, a comparison with the data published by the @who2022https reveals perfect similarities.
+The original data consist of 14,089,320 observations of 10 variables (117 distinct countries, region within the country, a unique observation code, the date of the observation, the gender, which can be male, female, or both, the age bracket by 5 years from 0 to 100, a confirmation of the age interval for each bracket, the total number of cases so far, the total number of deaths, and the total number of tests performed) from `r format(min(df_global$date), "%B %d, %Y")` to `r format(max(df_global$date), "%B %d, %Y")`.
+After removing countries with missing and inconsistent values, only 22 are suitable for data analyses.
+However, to focus this analysis on geographically and culturally comparable countries, only 12 European countries are kept: Austria, Belgium, Bulgaria, Croatia, Estonia, France, Germany, Greece, Netherlands, Portugal, Slovakia, and Spain.
+The observations are reported in terms of the total number of COVID-19 cases per day from the start of the pandemic.
+The daily number of cases at a specific date $n\prime_{t}$ is calculated with the difference between the total cases at a date $t$ and the total cases at a date $t-1$ (i.e., derivative 1).
+In addition, the change in the daily number of cases $n\prime\prime_{t}$ between $n\prime_{t}$ and $n\prime_{t-1}$ has also been calculated (i.e., derivative 2).
+
+## Generalized additive model
 
@@
 
-## Transfer Entropy
+## Transfer entropy
 
@@
 The day-by-day difference in COVID-19 confirmed cases $n\prime\prime_t$ is used to satisfy the stationary requirement for the calculation of Shannon’s Transfer Entropy [@shannon1948mathematical; @behrendt2019rtransferentropy].
-
-# Data
-
-## COVID-19 cases per age group
-
-For the analysis, data on COVID case counts by age group are required.
-The number of COVID-19 cases per age group has been taken from the COVerAGE-DB project [@riffe2021data].
-The COVerAGE-DB project consists of 3 data files: an “Input” data file that collects the official COVID-19 cases from 117 countries, the “Output_5” data file, which is a projection of COVID-19 cases by a group of 5 years, and the “Output_10” data file, which is a projection of COVID-19 cases by a group of 10 years.
-For our analyses, we have used the “Output_5,” which deals with the heterogeneity of countries’ reporting formats by using spline approximations when the data for this age bracket is not available for a country.
-Therefore, it allows a precise analysis between age groups under 24, while age groups of 25 and above are reduced to 10-year brackets, and groups above 64 are concatenated all together.
-
-In order to cross-validate the data obtained after spline approximations, a comparison with the data published by the @who2022https reveals perfect similarities.
-The original data consist of 14,089,320 observations of 10 variables (117 distinct countries, region within the country, a unique observation code, the date of the observation, the gender, which can be male, female, or both, the age bracket by 5 years from 0 to 100, a confirmation of the age interval for each bracket, the total number of cases so far, the total number of deaths, and the total number of tests performed) from `r format(min(df_global$date), "%B %d, %Y")` to `r format(max(df_global$date), "%B %d, %Y")`.
-After removing countries with missing and inconsistent values, only 22 are suitable for data analyses.
-However, to focus this analysis on geographically and culturally comparable countries, only 12 European countries are kept: Austria, Belgium, Bulgaria, Croatia, Estonia, France, Germany, Greece, Netherlands, Portugal, Slovakia, and Spain.
-The observations are reported in terms of the total number of COVID-19 cases per day from the start of the pandemic.
-The daily number of cases at a specific date $n\prime_{t}$ is calculated with the difference between the total cases at a date $t$ and the total cases at a date $t-1$ (i.e., derivative 1).
-In addition, the change in the daily number of cases $n\prime\prime_{t}$ between $n\prime_{t}$ and $n\prime_{t-1}$ has also been calculated (i.e., derivative 2).
-
-## School closure
-
-To identify the impact of school closures on the number of cases, information regarding the school closures on a day-by-day basis is required, and this is taken from the “UNESCO global education coalition” (2022).
-For each day, in each country, the status of the schools is indicated as fully open, partially open, closed due to COVID-19, or closed due to an academic break.
-Because it would be difficult to measure the effect of school closures at a country level when schools are partially closed, only closures due to COVID-19 or due to an academic break are considered.
-Indeed, both are considered as closure at a country-wide level.
-Note that periods of closure longer than 21 days are compared for 28 days, regardless of whether the schools opened or not after 21 days.
-Any COVID-19 case numbers beyond the 28-day period are not relevant for evaluating the impact of school closures.
 
@@
 
-# Discussion & Conclusion
+# Discussion
 
@@
 
-# Data Availability
+# Author statements
 
-All data extracted from the COVerAGE-DB project and from the UNESCO global education coalition that are used to produce these results are fully available here https://github.com/damien-dupre/covid_ts_causality.
-The repository also includes the R code that were used to produce these results.
+## Ethical approval
+
+This study is a longitudinal analysis on available data; thus, it does not require ethical approval.
+
+## Funding
+
+None declared.
+
+## Competing interests
+
+None declared.
+
+## Data Availability
+
+The data can be accessed from the COVerAGE-DB OSF repository (https://osf.io/mpwjq/) and from the UNESCO servers (https://en.unesco.org/file/unesco-data-school-closures-february-2020-june-2022csv-zip).
+All preprocessing, analyses and code used to build the submitted paper are available at https://github.com/damien-dupre/covid_ts_causality.
 

### 2024-12-16 manuscript_tf/appendix.tex
AE

\section{Appendices}

Any appendices should be placed after the list of references, beginning with the command \verb"\appendix" followed by the command \verb"\section" for each appendix title, e.g.
\begin{verbatim}
\appendix
\section{This is the title of the first appendix}
\section{This is the title of the second appendix}
\end{verbatim}
produces:\medskip

\noindent\textbf{Appendix A. This is the title of the first appendix}\medskip

\noindent\textbf{Appendix B. This is the title of the second appendix}\medskip

\noindent Subsections, equations, figures, tables, etc.\ within appendices will then be automatically numbered as appropriate. Some theorem-like environments may need to have their counters reset manually (e.g.\ if they are not numbered within sections in the main text). You can achieve this by using \verb"\numberwithin{remark}{section}" (for example) just after the \verb"\appendix" command.

Please note that if the \verb"endfloat" package is used on a document containing appendices, the \verb"\processdelayedfloats" command must be included immediately before the \verb"\appendix" command in order to ensure that the floats in the main body of the text are numbered as such.

%\processdelayedfloats %%% See above for an explanation of why this command might be needed.

\appendix

\section{Troubleshooting}

Authors may occasionally encounter problems with the preparation of a manuscript using \LaTeX. The appropriate action to take will depend on the nature of the problem:
\begin{enumerate}
\item[(i)] If the problem is with \LaTeX\ itself, rather than with the actual macros, please consult an appropriate \LaTeXe\ manual for initial advice. If the solution cannot be found, or if you suspect that the problem does lie with the macros, then please contact Taylor \& Francis for assistance (\texttt{latex.helpdesk@tandf.co.uk}).
\item[(ii)] Problems with page make-up (e.g.\ occasional overlong lines of text; figures or tables appearing out of order): please do not try to fix these using `hard' page make-up commands -- the typesetter will deal with such problems. (You may, if you wish, draw attention to particular problems when submitting the final version of your manuscript.)
\item[(iii)] If a required font is not available on your system, allow \TeX\ to substitute the font and specify which font is required in a covering letter accompanying your files.
\end{enumerate}

\section{Obtaining the template and class file}

\subsection{Via the Taylor \& Francis website}

This article template and the \texttt{interact} class file may be obtained via the `Instructions for Authors' pages of selected Taylor \& Francis journals.

Please note that the class file calls up the open-source \LaTeX\ packages booktabs.sty, epsfig.sty and rotating.sty, which will, for convenience, unpack with the downloaded template and class file. The template calls for natbib.sty and subfigure.sty, which are also supplied for convenience.

\subsection{Via e-mail}

This article template, the \texttt{interact} class file and the associated open-source \LaTeX\ packages are also available via e-mail. Requests should be addressed to \texttt{latex.helpdesk@tandf.co.uk}, clearly stating for which journal you require the template and class file.

### 2024-12-16 manuscript_tf/figures.Rmd
AE

--- 2024-02-26 manuscript_elsevier/manuscript.Rmd
@@
 ---
-title: Examining the Impact of School Closures on COVID-19 Infections in Europe and their Effects on Different Age Cohorts
-# author:
-#   - name: Damien Dupré
-#     email: damien.dupre@dcu.ie
-#     affiliation: Business School, Dublin City University
-#     correspondingauthor: true
-#   - name: Edgar Morgenroth
-#     email: edgar.morgenroth@dcu.ie
-#     affiliation: Business School, Dublin City University
-# address:
-#   - code: Business School, Dublin City University
-#     organization: Business School, Dublin City University
-#     city: Dublin
-#     country: Ireland
-abstract: |
-  *Objectives:* This paper analyzes the trends of COVID-19 cases in different age groups in selected European countries, assessing to what extent school closures helped reduce case numbers, how school closures affected case numbers in different age groups, and how COVID-19 spread between age groups.
-
-  *Study Design:* A longitudinal study was conducted to observe changes in COVID-19 cases during school closures.
-
-  *Methods:* A Generalized Additive Model is performed to compare trends from twelve European countries across five age groups.
-Then, Transfer Entropy calculations are used to evaluate the impact of younger cohorts on older cohorts.
-
-  *Results:* The findings demonstrate a decreasing non-linear effect of school closure on the total COVID-19 case number across all countries studied.
-However, the analysis of age groups only reveals a consistent downward trend in infections in the 0 to 4 pre-school age group, while the school-going age groups 5 to 14 exhibit a significant increase in cases.
-Age groups 15 to 24 show a surge immediately after closure, followed by a decline.
-Transfer Entropy calculations highlight asymmetry in age group influences, indicating that changes in COVID-19 cases in certain age groups predict changes in other age groups but not vice versa.
-
-  *Conclusions:* These findings contribute to a better understanding of COVID-19 dynamics in European countries and provide insights for public health strategies and interventions.
-keywords: 
-  - COVID-19
-  - School Closures
-  - Intergenerational Transmission.
-journal: "Public Health"
-date: "`r Sys.Date()`"
-classoption: preprint, 3p
-linenumbers: false
-numbersections: false
-bibliography: bibliography.bib
-# biblio-style: elsarticle-num
-csl: https://www.zotero.org/styles/elsevier-with-titles
+title: |
+  Figures
 output: 
   bookdown::pdf_book:
-    base_format: rticles::elsevier_article
-    keep_tex: false
-    citation_package: default
+    toc: false
+bibliography: bibliography.bib
 header-includes:
@@
   - \usepackage{caption}
+  - \usepackage{hyperref}
+  - \usepackage[utf8]{inputenc}
+  - \def\tightlist{}
 ---
@@
 (ref:overall) Cumulative COVID-19 cases number for selected European countries since the beginning of the pandemic. Source: COVerAGE-DB [@riffe2021data].
-(ref:descriptive) Periods of school closure since the beginning of the COVID-19 pandemic for selected European countries and their reason: regular academic break vs.
-closure due to government decisions.
+(ref:descriptive) Periods of school closure since the beginning of the COVID-19 pandemic for selected European countries and their reason: regular academic break versus closure due to government decisions.
 Source: @unesco2022https.
-(ref:country) Standardized effect of the smooth term in Generalized Additive Model by country.
-(ref:age) Standardized effect of the smooth term in Generalized Additive Model by age group.
+(ref:country) Standardized effect of the smooth term in Generalized Additive Model by country. Standardized effects are reported to compare the shape of the curve between countries.
+(ref:age) Standardized effect of the smooth term in Generalized Additive Model by age group. Standardized effects are reported to compare the shape of the curve between age groups.
 (ref:te) Matrix of Transfer Entropy coefficients according to every age group combination for each of the selected European country.
@@
 The significance of each Transfer Entropy coefficient is provided in Appendix 2.
+
+\newpage
 
@@
 
-# Introduction
-
-A wide range of non-pharmaceutical interventions (NPI) were enacted to control the spread of COVID-19, especially before the availability of effective vaccines.
-Due to their social mixing patterns, children have been identified as an age group that can drive the spread of respiratory infections such as influenza [@moser2018estimating], and school closures were found to be effective in mitigating the spread of influenza H1N1 in Japan in 2009 [@kawano2015substantial].
-It is therefore not surprising that school closures were one of the NPIs that were introduced in many countries to stop the spread of COVID-19.
-They were used particularly during the initial wave of the pandemic and during subsequent waves when case numbers were high.
-
-Recently, a body of literature has shown various negative consequences on child development and health due to school closures.
-School closures have been found to have negatively affected child mental health [@moulin2022longitudinal], nutrition/obesity [@sugimoto2023temporal], and education [@lerkkanen2023reading].
-Some governments are now questioning their decisions to close schools [@de2021determines].
-For example, the German health minister Karl Lauterbach, in an interview with one of Germany’s publicly funded television stations, admitted that “in retrospect it had been wrong to keep schools and childcare closed for so long” [@ard2023lauterbach].
-Even if school closures had negative consequences for children, they might nevertheless have been effective at stopping the spread of COVID-19.
-However, analysis on the effectiveness of school closures on the spread of COVID-19 remains inconclusive.
-@alfano2022effects in a study of a panel of European countries found that school closures were associated with a reduced COVID-19 incidence.
-In contrast, @walsh2021school in a review of 40 studies covering 150 countries concluded that the effectiveness of school closures was uncertain, with 60% having identified no impact and pointing to the potential for the analysis to be affected by confounding factors and collinearity.
-The latter shortcomings suggest that it is hard to draw firm inferences, as even a positive result may not imply causality.
-
-Once it was clear that COVID-19 spreads from person to person, it was natural that governments were advised to enact measures to reduce social contact in order to reduce the spread of the virus.
-Therefore, it is legitimate to believe that by closing schools, a reduction of the contaminations would be observed in the younger age groups.
-However, the efficiency of school closure on the reduction of COVID-19 cases is still questioned [@bayham2020impact; @esposito2021comprehensive].
-While some research has found that school closures contribute to limit or to reduce the growth rate of confirmed cases after implementation [@stage2021shut; @yoshiyuki2020effects], others did not observe a change in the evolution of COVID-19 cases [@chang2020modelling; @iwata2020school].
-For instance, a controlled comparison between similar localities in Japan with schools closed and schools open did not reveal any evidence that school closures reduced the spread of COVID-19 [@fukumoto2021no].
-If the school closure had a real impact on the evolution of confirmed COVID-19 cases, it should be possible to observe a decrease or at least an inflection in the trend of its evolution among younger age groups.
-
-A second implicit belief regarding the effect of school closure on the spread of COVID-19 is that school not only influences the spread of the virus among children and teenagers but also has a knock-on effect on the spread of the virus in older age groups, also called Secondary Attack Rate (SAR).
-The contaminated children and teenagers would bring the virus back home and would pass the virus on to their parents and relatives.
-For example, research investigating the contamination in the household network not only revealed an exceptionally high rate of secondary contamination but also that this contamination happened when the schools were closed [@sorianoarandes2021household].
-Despite being reported in several clinical and epidemiological studies [@siebach2021childhood; @zhendong2020clinical], multiple research have shown that the SAR from children to household members was lower than expected [@heavey2020evidence; @vanderhoek2020role; @kim2021role; @ludvigsson2020children].
-However, the SAR of children and teenagers to the household member is likely to be age-dependent, with differences between infants, primary and secondary school children, and college students [@grasleguen2021reopening].
-If a secondary transmission from children and teenagers to household member has a significant influence, then a temporal causality relationship between their evolution should be observed.
-
-To evaluate the impact of school closure on COVID-19 cases across different age groups, we first performed a non-linear time series regression to analyze the relationship between school closures and the number of COVID-19 cases in various age groups.
-In a second analysis, the transfer entropy method is employed to assess the influence of changes in case numbers between age groups, aiming to determine if school closures have secondary effects within households.
-
-# Methods
-
-## Study design
-
-A longitudinal study was conducted to observe changes in COVID-19 cases during school closures.
-Changes in COVID-19 from five age groups from the first to the 28th day of school closure are compared across 12 countries.
-Then, a causality analysis evaluate the impact of changes in these younger groups on older age groups.
-
-## Data collection
-
-Data on COVID-19 case counts by age group have been obtained from the COVerAGE-DB project [@riffe2021data].
-The COVerAGE-DB which collect numbers for every countries by groups of 5 or 10 years if available.
-In addition, spline approximations are used to deals with the heterogeneity of countries’ reporting formats by using when the data for this age bracket is not available for a country.
-
-To identify the impact of school closures on the number of cases, information regarding the school closures on a day-by-day basis is required, and this is taken from the “UNESCO global education coalition” (2022).
-For each day, in each country, the status of the schools is indicated as fully open, partially open, closed due to COVID-19, or closed due to an academic break.
-Because it would be difficult to measure the effect of school closures at a country level when schools are partially closed, only closures due to COVID-19 or due to an academic break are considered.
-Indeed, both are considered as closure at a country-wide level.
-Note that periods of closure longer than 21 days are compared for 28 days, regardless of whether the schools opened or not after 21 days.
-Any COVID-19 case numbers beyond the 28-day period are not relevant for evaluating the impact of school closures.
-
-## Data validation
-
-In order to cross-validate the data obtained after spline approximations, a comparison with the data published by the @who2022https reveals perfect similarities.
-The original data consist of 14,089,320 observations of 10 variables (117 distinct countries, region within the country, a unique observation code, the date of the observation, the gender, which can be male, female, or both, the age bracket by 5 years from 0 to 100, a confirmation of the age interval for each bracket, the total number of cases so far, the total number of deaths, and the total number of tests performed) from `r format(min(df_global$date), "%B %d, %Y")` to `r format(max(df_global$date), "%B %d, %Y")`.
-After removing countries with missing and inconsistent values, only 22 are suitable for data analyses.
-However, to focus this analysis on geographically and culturally comparable countries, only 12 European countries are kept: Austria, Belgium, Bulgaria, Croatia, Estonia, France, Germany, Greece, Netherlands, Portugal, Slovakia, and Spain.
-The observations are reported in terms of the total number of COVID-19 cases per day from the start of the pandemic.
-The daily number of cases at a specific date $n\prime_{t}$ is calculated with the difference between the total cases at a date $t$ and the total cases at a date $t-1$ (i.e., derivative 1).
-In addition, the change in the daily number of cases $n\prime\prime_{t}$ between $n\prime_{t}$ and $n\prime_{t-1}$ has also been calculated (i.e., derivative 2).
-
-## Generalized additive model
-
-The effect of school closure on the trend of daily COVID-19 cases is analyzed using a Generalized Additive Model (GAM).
-GAM is a flexible modeling approach that estimates non-linear relationships between variables.
-Compared to other methods such as Vector Autoregression, GAM can handle fixed and random smooth effects without making strict assumptions about linearity.
-
-The GAM is fitted on the daily COVID-19 cases to test the hypothesis of a significant non-linear evolution of cases among age groups from 0 to 4, from 5 to 9, from 10 to 14, from 15 to 19, and from 20 to 24 [@wood2017generalized].
-The model also estimates the overall non-linear effect by country, taking into account the interaction between age groups and countries as random intercepts and the interaction between time, countries, and the period of closure as random effects (Eq 1).
-
-By estimating the degree of smoothness of a Bayesian spline smoothing using restricted fast maximum likelihood estimation [@wood2011fast], GAM identifies dynamic patterns underlying the evolution of COVID-19 cases reported while including the random effect of different age groups and countries as follows:
-
-\begin{align*}
-  \tag{1}
-  n\prime_t &\sim Poisson(\mu_t) \\
-  \log(\mu_t) &= \beta_0 + \beta_1\,wave_t + \beta_2\,country_t + \beta_3\,age\,group_t + \\
-              & f_1(closure_t) + f_2(closure_t,country_t) + f_3(closure_t,age\,group_t)
-\end{align*}
-
-\noindent where $n\prime_t$ represents the confirmed COVID-19 cases, assuming a Poisson distribution for the fitting [@loader2006local], and $t$ is the date corresponding to the confirmed COVID-19 cases.
-The response variable includes a specific random effect taking into account variation within waves of school closure, countries, and age groups.
-The terms $f_1$ to $f_3$ are smooth functions of the time since closure, the time since closure for each country, and the time since closure for each age group.
-The restricted maximum likelihood (REML) was used to avoid overfitting while estimating smoothing parameters.
-In order to accurately account for the autocorrelation arising from the time series data, the residuals are modeled using an AR1 error model.
-By incorporating the autoregressive component, the AR1 model acknowledges the dependence of each residual on its previous value, thus providing a comprehensive representation of the data’s temporal dynamics.
-
-## Transfer entropy
-
-Transfer entropy ($T$) can be used to infer the temporal relationship between two time series $X$ and $Y$.
-This measure indicates whether changes in COVID-19 cases within a certain age group $X$ can be used to reduce the uncertainty on the future COVID-19 cases within another age group $Y$.
-If it does, $T$ is considered evidence of a causal effect from the age group $X$ to the age group $Y$ [@schreiber2000measuring].
-As such, Granger causality is a special case of transfer entropy applied to time series that are jointly Gaussian distributed [@barnett2009granger].
-Therefore, transfer entropy is a more robust analysis of time series, especially when applied to the impact of age cohorts on pandemic transmission [@kissler2020symbolic].
-
-The influence of the evolution in COVID-19 cases across all age groups is evaluated using Shannon’s transfer entropy, given by:
-
-\begin{align}
-  \tag{2}
-  T_{ag\,x \rightarrow\,ag\,y}(k,l) = \sum_{ag\,x,ag\,y} p\left(ag\,x_{t+1}, ag\,x_t^{(k)}, ag\,y_t^{(l)}\right) \log \left(\frac{p\left(ag\,x_{t+1}| ag\,x_t^{(k)}, ag\,y_t^{(l)}\right)}{p\left(ag\,x_{t+1}|ag\,x_t^{(k)}\right)}\right)
-\end{align}
-
-\noindent where $T_{ag\,x \rightarrow\,ag\,y}$ consequently measures the influence of the change dynamic from an age group $X$ (or $ag\,x$) to another age group $Y$ (or $ag\,y$) for every country (Eq 2).
-
-The day-by-day difference in COVID-19 confirmed cases $n\prime\prime_t$ is used to satisfy the stationary requirement for the calculation of Shannon’s Transfer Entropy [@shannon1948mathematical; @behrendt2019rtransferentropy].
-
-# Results
-
-The data show that the trend of confirmed COVID-19 cases follows similar patterns across the selected European countries, with scales following the size of the population in these countries (Figure \@ref(fig:overall)).
-Thus, the bigger the country, the higher the total number of COVID-19 cases.
-
 ```{r overall, fig.height=3, out.width="\\textwidth", fig.cap="(ref:overall)"}
@@
 
-The evolution of COVID-19 cases reveals some similarities across all age groups.
-However, the influence of each wave on individual age groups also has some particularities (Figure \@ref(fig:descriptive)).
-For example, the first wave was more important among the oldest age groups, whereas the third wave was more important among the youngest age groups.
+\newpage
 
@@
 
-As outlined above, to evaluate the shape of the trend in the numbers of COVID-19 cases reported after the three school closures longer than 21 consecutive days, a GAM was fitted as described above, taking into account the overall effect across all the selected European countries, as well as the effect for age groups: 0 to 4, 5 to 9, 10 to 14, 15 to 19, and 20 to 24 years old.
-The obtained results satisfy the requirements to fit this model, which explains `r percent(gam_summary$dev.expl, accuracy = 0.1)` of the variation in COVID-19 cases.
-
-Overall, the results revealed a decreasing non-linear effect of school closure at a country level for the selected European countries (Austria: `r gam_results["Austria"]`; Belgium: `r gam_results["Belgium"]`; Bulgaria: `r gam_results["Bulgaria"]`; Croatia: `r gam_results["Croatia"]`; Estonia: `r gam_results["Estonia"]`; France: `r gam_results["France"]`; Germany (`r gam_results["Germany"]`); Greece: `r gam_results["Greece"]`; The Netherlands: `r gam_results["Netherlands"]`; Portugal: `r gam_results["Portugal"]`; Slovakia (`r gam_results["Slovakia"]`; Spain: `r gam_results["Spain"]`, see Figure \@ref(fig:age)).
+\newpage
 
@@
 
+\newpage
+
 ```{r age, fig.height=3, out.width="\\textwidth", fig.cap="(ref:age)"}
@@
 
+\newpage
+
 ```{r te, fig.height=8, out.width="\\textwidth", fig.cap="(ref:te)", fig.pos = 'H'}
@@
 ```
-
-Considering all countries, the analysis of age groups reveals distinct patterns (Figure \@ref(fig:age)).
-For the 0 to 4 age group, i.e., the pre-school group, there was a downward trend during the initial two weeks following school closure, followed by flattening off (`r gam_results["0to4"]`).
-In contrast, for the age group ranging from 5 to 14, a substantial increase in cases after 14 days of school closure is found (5 to 9 age group: `r gam_results["5to9"]` and 10 to 14 age group: `r gam_results["10to14"]`).
-For age groups between 15 and 24, there is a notable surge immediately after school closure, followed by a decrease after 14 days (15 to 19 age group: `r gam_results["15to19"]` and 20 to 24 age group: `r gam_results["20to24"]`).
-Thus, while the total case number declined following school closures, among younger age groups only the pre-school age group had a consistent reduction in cases, while for other age groups, there was even an increase in cases following school closures.
-
-Given that school closures were not only aimed at reducing COVID-19 cases among school-going children but also age groups, it is important to test the degree to which there was transmission across age groups, which is done using transfer entropy analysis.
-Before this analysis is carried out, an Augmented Dickey-Fuller has been applied to each age group to ensure that the daily changes in COVID-19 cases are stationary (see Appendix 1).
-
-The results of the transfer entropy calculations between age groups for each of the 12 selected European countries are reported in Figure \@ref(fig:te).
-The absence of symmetry between influencing age groups (i.e., $ag\,x$) and influenced age groups (i.e., $ag\,y$) is found.
-Indeed, the change in COVID-19 cases in some age groups is influenced by other age groups, but they are not reciprocally influencing these age groups.
-Figure \@ref(fig:te) shows how different age cohorts influence the COVID-19 case numbers of all age groups across countries.
-The upper left quadrant indicates how younger age cohorts are influencing older age cohorts, the bottom right quadrant indicates how older age cohorts are influencing younger age cohorts, finally, the lower left and upper right indicate how younger or older age cohorts are influencing themselves.
-By analyzing these quadrants, it is possible to identify similar patterns across multiple countries.
-Indeed, it appears that Austria, Germany, and The Netherlands have significantly higher $T$ coefficients in the upper left quadrant of the matrix, which indicates that the daily changes in COVID-19 cases number in younger cohorts are predicting the daily changes in COVID-19 cases number in older cohorts.
-Alternatively, it appears that Austria, the Netherlands, Portugal, and Spain have significantly higher $T$ coefficients in the lower right quadrant of the matrix, which indicates that the daily changes in COVID-19 cases number in older cohorts are predicting the daily changes in COVID-19 cases number in younger cohorts.
-
-# Discussion
-
-Knowledge about the transmission of the virus significantly improved over time as more studies have been published.
-While an early study found that children did not play an important role in the transmission of the virus [@li2020role], more recent results give a more nuanced position, stating that the spread of the virus in children is moderate.
-This study, using robust statistical methods, considered the effect of school closures on COVID-19 cases across age groups.
-While there are commonalities in the evolution of COVID-19 cases across all European countries that were included in our data, school closures exhibit distinct impacts on different age groups.
-Notably, the analysis shows that the 0 to 4 age group experiences a downward trend in COVID-19 cases during the initial two weeks following school closure, followed by stabilization.
-In contrast, age groups ranging from 5 to 14 exhibit a substantial increase in cases after 14 days of school closure.
-Additionally, age groups between 15 and 24 demonstrate a notable surge immediately after school closure, followed by a decrease after 14 days.
-Thus, the results do not support the hypothesis that school closures were effective.
-These results partially replicate observations from @alfano2022effects while providing a clearer picture of the structure of the effect of school closure.
-
-The Transfer Entropy calculations between age groups for each of the 12 selected European countries reveal an absence of symmetry between influencing age groups (X) and influenced age groups (Y).
-Some age groups influence the COVID-19 case numbers of other age groups, but there is no reciprocal influence.
-This finding suggests that changes in COVID-19 cases in certain age groups can predict the changes in other age groups but not vice versa, i.e., causality can be unidirectional.
-These findings highlight the importance of considering intergenerational interactions in designing effective control measures.
-
-Furthermore, the quadrant analysis reveals similar patterns across multiple countries.
-Austria, Germany, and the Netherlands exhibit significantly higher TE coefficients in the upper left quadrant, indicating that changes in COVID-19 cases in younger cohorts predict the changes in older cohorts.
-On the other hand, Austria, the Netherlands, Portugal, and Spain show higher TE coefficients in the lower right quadrant, indicating that changes in COVID-19 cases in older cohorts predict the changes in younger cohorts.
-
-In conclusion, this study sheds light on the trends and patterns of confirmed COVID-19 cases in selected European countries, with a particular focus on the influence of age groups and school closures.
-While a reduction in overall COVID-19 cases was found during school closures, which might suggest that like lockdowns [@alfano2020efficacy; @molefi2021impact], prolonged school closures could contribute to mitigating the spread of the virus, the results for the school-going age groups suggest that this is not the case.
-
-# Author statements
-
-## Ethical approval
-
-This study is a longitudinal analysis on available data; thus, it does not require ethical approval.
-
-## Funding
-
-None declared.
-
-## Competing interests
-
-None declared.
-
-## Data Availability
-
-The data can be accessed from the COVerAGE-DB OSF repository (https://osf.io/mpwjq/) and from the UNESCO servers (https://en.unesco.org/file/unesco-data-school-closures-february-2020-june-2022csv-zip).
-All preprocessing, analyses and code used to build the submitted paper are available at https://github.com/damien-dupre/covid_ts_causality.
-
-# References {-}
-
-<!-- <div id="refs"></div> -->
-
-<!-- \newpage -->
-
-<!-- # Appendix 1: Dickey-Fuller test for stationay timeseries {-}  -->
-
-<!-- Test for the null hypothesis of a unit root of a univarate time series (equivalently, non-stationary time series). -->
-
-<!-- The Augmented Dickey-Fuller test incorporates three types of linear regression models.
-The first type is a linear model with no drift and linear trend with respect to time.
-The second type is a linear model with drift but no linear trend.
-The third type is a linear model with both drift and linear trend.
--->
-
-<!-- The p.value is calculated by interpolating the test statistics from the corresponding critical values tables (Fuller, 1996) for each type of linear models with given sample size.  -->
-
-<!-- ```{r} -->
-<!-- adf_res_type1 <- df_global_age |> -->
-<!--   group_by(age_grp, country) |> -->
-<!--   group_modify(~ aTSA::adf.test(.x$daily_changes, output = FALSE, nlag = 2) |> -->
-<!--                  bind_rows() |>  -->
-<!--                  pluck("type1") |>  -->
-<!--                  as_tibble() -->
-<!--   ) |>  -->
-<!--   rename(ADF_type1 = ADF, p.value_type1 = p.value) -->
-
-<!-- adf_res_type2 <- df_global_age |> -->
-<!--   group_by(age_grp, country) |> -->
-<!--   group_modify(~ aTSA::adf.test(.x$daily_changes, output = FALSE, nlag = 2) |> -->
-<!--                  bind_rows() |>  -->
-<!--                  pluck("type2") |>  -->
-<!--                  as_tibble() -->
-<!--   ) |>  -->
-<!--   rename(ADF_type2 = ADF, p.value_type2 = p.value) -->
-
-<!-- adf_res_type3 <- df_global_age |> -->
-<!--   group_by(age_grp, country) |> -->
-<!--   group_modify(~ aTSA::adf.test(.x$daily_changes, output = FALSE, nlag = 2) |> -->
-<!--                  bind_rows() |>  -->
-<!--                  pluck("type3") |>  -->
-<!--                  as_tibble() -->
-<!--   ) |>  -->
-<!--   rename(ADF_type3 = ADF, p.value_type3 = p.value) -->
-
-<!-- adf_res <- adf_res_type1 |>  -->
-<!--   full_join(adf_res_type2) |>  -->
-<!--   full_join(adf_res_type3) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- adf_res |> -->
-<!--   filter(age_grp %in% c("aged0to4", "aged5to9")) |>  -->
-<!--   kbl( -->
-<!--     caption = "Dickey-Fuller test for all combinaisons of countries and age groups investigated", -->
-<!--     col.names = c("Age Group", "Country", "Lag", "ADF", "p", "ADF", "p", "ADF", "p") -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   add_header_above(c(" " = 3, "Type 1" = 2, "Type 2" = 2, "type 3" = 2)) |> -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- adf_res |> -->
-<!--   filter(age_grp %in% c("aged10to14", "aged15to19")) |>  -->
-<!--   kbl( -->
-<!--     caption = "Dickey-Fuller test for all combinaisons of countries and age groups investigated", -->
-<!--     col.names = c("Age Group", "Country", "Lag", "ADF", "p", "ADF", "p", "ADF", "p") -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   add_header_above(c(" " = 3, "Type 1" = 2, "Type 2" = 2, "type 3" = 2)) |> -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- adf_res |> -->
-<!--   filter(age_grp %in% c("aged20to24", "aged25to34")) |>  -->
-<!--   kbl( -->
-<!--     caption = "Dickey-Fuller test for all combinaisons of countries and age groups investigated", -->
-<!--     col.names = c("Age Group", "Country", "Lag", "ADF", "p", "ADF", "p", "ADF", "p") -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   add_header_above(c(" " = 3, "Type 1" = 2, "Type 2" = 2, "type 3" = 2)) |> -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- adf_res |> -->
-<!--   filter(age_grp %in% c("aged35to44", "aged45to54")) |>  -->
-<!--   kbl( -->
-<!--     caption = "Dickey-Fuller test for all combinaisons of countries and age groups investigated", -->
-<!--     col.names = c("Age Group", "Country", "Lag", "ADF", "p", "ADF", "p", "ADF", "p") -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   add_header_above(c(" " = 3, "Type 1" = 2, "Type 2" = 2, "type 3" = 2)) |> -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- adf_res |> -->
-<!--   filter(age_grp %in% c("aged55to64", "aged65+")) |>  -->
-<!--   kbl( -->
-<!--     caption = "Dickey-Fuller test for all combinaisons of countries and age groups investigated", -->
-<!--     col.names = c("Age Group", "Country", "Lag", "ADF", "p", "ADF", "p", "ADF", "p") -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   add_header_above(c(" " = 3, "Type 1" = 2, "Type 2" = 2, "type 3" = 2)) |> -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- # Appendix 2: Significance of Transfer Entropy coefficients -->
-
-<!-- Results for each permutation of age groups from each country.
-Tables include transfer entropy estimates, the effective transfer entropy estimates, standard errors and p-values based on bootstrap replications of the Markov chains under the null hypothesis of statistical independence.
--->
-
-<!-- ```{r} -->
-<!-- # variables -------------------------------------------------------------------- -->
-<!-- age_order <- c( -->
-<!--   "aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24", -->
-<!--   "aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+" -->
-<!-- ) -->
-
-<!-- # data ------------------------------------------------------------------------- -->
-<!-- te_results <- read_csv(here("data/te_results.csv")) |>  -->
-<!--     mutate( -->
-<!--       X = factor(X, levels = age_order), -->
-<!--       Y = factor(Y, levels = age_order) -->
-<!--     ) |>  -->
-<!--   select(-direction) -->
-<!-- ``` -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged0to4") |>  -->
-<!--   filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged0to4") |>  -->
-<!--   filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged5to9") |>  -->
-<!--   filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged5to9") |>  -->
-<!--   filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged10to14") |>  -->
-<!--   filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged10to14") |>  -->
-<!--   filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged15to19") |>  -->
-<!--   filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged15to19") |>  -->
-<!--   filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged20to24") |>  -->
-<!--   filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged20to24") |>  -->
-<!--   filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged25to34") |>  -->
-<!--   filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged25to34") |>  -->
-<!--   filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged35to44") |>  -->
-<!--   filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged35to44") |>  -->
-<!--   filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged45to54") |>  -->
-<!--   filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged45to54") |>  -->
-<!--   filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged55to64") |>  -->
-<!--   filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged55to64") |>  -->
-<!--   filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged65+") |>  -->
-<!--   filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged65+") |>  -->
-<!--   filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->

### 2024-12-16 manuscript_tf/manuscript.Rmd
AE

--- 2024-02-26 manuscript_elsevier/manuscript.Rmd
@@
 ---
-title: Examining the Impact of School Closures on COVID-19 Infections in Europe and their Effects on Different Age Cohorts
+title: |
+  Examining the Impact of School Closures on COVID-19 Infections in Europe and their Effects on Different Age Cohorts
+author:
+  - name: Damien Dupré
+    affil: 1
+    email: damien.dupre@dcu.ie
+  - name: Edgar Morgenroth
+    affil: 1
+    email: edgar.morgenroth@dcu.ie
+affiliation:
+  - num: 1
+    address: |
+      Business School, Dublin City University
+
 # author:
-#   - name: Damien Dupré
-#     email: damien.dupre@dcu.ie
-#     affiliation: Business School, Dublin City University
-#     correspondingauthor: true
-#   - name: Edgar Morgenroth
-#     email: edgar.morgenroth@dcu.ie
-#     affiliation: Business School, Dublin City University
-# address:
-#   - code: Business School, Dublin City University
-#     organization: Business School, Dublin City University
-#     city: Dublin
-#     country: Ireland
+#   - name: Author 1
+#     affil: 1
+#     email: Email 1
+#   - name: Author 2
+#     affil: 2
+#     email: Email 2
+# affiliation:
+#   - num: 1
+#     address: |
+#       Affiliation 1
+#   - num: 2
+#     address: |
+#       Affiliation 2
+
+bibliography: bibliography.bib
 abstract: |
-  *Objectives:* This paper analyzes the trends of COVID-19 cases in different age groups in selected European countries, assessing to what extent school closures helped reduce case numbers, how school closures affected case numbers in different age groups, and how COVID-19 spread between age groups.
-
-  *Study Design:* A longitudinal study was conducted to observe changes in COVID-19 cases during school closures.
-
-  *Methods:* A Generalized Additive Model is performed to compare trends from twelve European countries across five age groups.
-Then, Transfer Entropy calculations are used to evaluate the impact of younger cohorts on older cohorts.
-
-  *Results:* The findings demonstrate a decreasing non-linear effect of school closure on the total COVID-19 case number across all countries studied.
+  This paper analyzes the trends of COVID-19 cases in different age groups in selected European countries, assessing to what extent school closures helped reduce case numbers, how school closures affected case numbers in different age groups, and how COVID-19 spread between age groups.
+The study also evaluates the effects of school closures on COVID-19 cases using a Generalized Additive Model (GAM).
+The findings demonstrate a decreasing non-linear effect of school closure on the total COVID-19 case number across all countries studied.
 However, the analysis of age groups only reveals a consistent downward trend in infections in the 0 to 4 pre-school age group, while the school-going age groups 5 to 14 exhibit a significant increase in cases.
@@
 Transfer Entropy calculations highlight asymmetry in age group influences, indicating that changes in COVID-19 cases in certain age groups predict changes in other age groups but not vice versa.
-
-  *Conclusions:* These findings contribute to a better understanding of COVID-19 dynamics in European countries and provide insights for public health strategies and interventions.
-keywords: 
-  - COVID-19
-  - School Closures
-  - Intergenerational Transmission.
-journal: "Public Health"
-date: "`r Sys.Date()`"
-classoption: preprint, 3p
-linenumbers: false
-numbersections: false
-bibliography: bibliography.bib
-# biblio-style: elsarticle-num
-csl: https://www.zotero.org/styles/elsevier-with-titles
+These findings contribute to a better understanding of COVID-19 dynamics in European countries and provide insights for public health strategies and interventions.
+keywords: |
+  COVID-19; School Closures; Intergenerational Transmission.
 output:
   bookdown::pdf_book:
-    base_format: rticles::elsevier_article
+    base_format: rticles::tf_article
     keep_tex: false
-    citation_package: default
 header-includes:
@@
 
+The COVID-19 pandemic has had a profound impact on global health, with an estimate of 14.83 million excess deaths globally [@msemburi2023estimates].
+Beyond direct mortality, the pandemic has caused significant collateral damage, including losses of lives and livelihoods, necessitating a comprehensive approach to measuring its broader impacts.
 A wide range of non-pharmaceutical interventions (NPI) were enacted to control the spread of COVID-19, especially before the availability of effective vaccines.
@@
 School closures have been found to have negatively affected child mental health [@moulin2022longitudinal], nutrition/obesity [@sugimoto2023temporal], and education [@lerkkanen2023reading].
-Some governments are now questioning their decisions to close schools [@de2021determines].
+Some governments were questioning their decisions to close schools [@de2021determines].
 For example, the German health minister Karl Lauterbach, in an interview with one of Germany’s publicly funded television stations, admitted that “in retrospect it had been wrong to keep schools and childcare closed for so long” [@ard2023lauterbach].
@@
 However, analysis on the effectiveness of school closures on the spread of COVID-19 remains inconclusive.
-@alfano2022effects in a study of a panel of European countries found that school closures were associated with a reduced COVID-19 incidence.
+In a study of a panel of European countries, @alfano2022effects found that school closures were associated with a reduced COVID-19 incidence.
 In contrast, @walsh2021school in a review of 40 studies covering 150 countries concluded that the effectiveness of school closures was uncertain, with 60% having identified no impact and pointing to the potential for the analysis to be affected by confounding factors and collinearity.
@@
 In addition, spline approximations are used to deals with the heterogeneity of countries’ reporting formats by using when the data for this age bracket is not available for a country.
+Countries communicating data with groups of 5 years were chosen to match as much as possible the different school stages.
 
@@
 Indeed, both are considered as closure at a country-wide level.
-Note that periods of closure longer than 21 days are compared for 28 days, regardless of whether the schools opened or not after 21 days.
 Any COVID-19 case numbers beyond the 28-day period are not relevant for evaluating the impact of school closures.
@@
 
-The effect of school closure on the trend of daily COVID-19 cases is analyzed using a Generalized Additive Model (GAM).
-GAM is a flexible modeling approach that estimates non-linear relationships between variables.
+The effect of school closure on the trend of daily COVID-19 cases is analysed using a Generalized Additive Model (GAM).
+GAM is a flexible modelling approach that estimates non-linear relationships between variables.
 Compared to other methods such as Vector Autoregression, GAM can handle fixed and random smooth effects without making strict assumptions about linearity.
@@
   \tag{1}
-  n\prime_t &\sim Poisson(\mu_t) \\
-  \log(\mu_t) &= \beta_0 + \beta_1\,wave_t + \beta_2\,country_t + \beta_3\,age\,group_t + \\
-              & f_1(closure_t) + f_2(closure_t,country_t) + f_3(closure_t,age\,group_t)
+  n\prime_t &\sim \text{Poisson}(\lambda_t) \\
+  \log(\lambda_t) &= \beta_0 + \beta_1\,wave_t + \beta_2\,country_t + \beta_3\,age\,group_t \\
+  &\quad + f_1(closure_t) + f_2(closure_t, country_t) \\
+  &\quad + f_3(closure_t, age\,group_t) + \epsilon_t
 \end{align*}
@@
 The terms $f_1$ to $f_3$ are smooth functions of the time since closure, the time since closure for each country, and the time since closure for each age group.
-The restricted maximum likelihood (REML) was used to avoid overfitting while estimating smoothing parameters.
-In order to accurately account for the autocorrelation arising from the time series data, the residuals are modeled using an AR1 error model.
+The restricted maximum likelihood (REML) was used to avoid over fitting while estimating smoothing parameters.
+In order to accurately account for the autocorrelation arising from the time series data, the residuals are modelled using an AR1 error model such as $\epsilon_t = \phi \epsilon_{t-1} + \eta_t, \quad \eta_t \sim \mathcal{N}(0, \sigma^2)$ where $\phi$ is the autoregressive parameter.
 By incorporating the autoregressive component, the AR1 model acknowledges the dependence of each residual on its previous value, thus providing a comprehensive representation of the data’s temporal dynamics.
+
+Chi-square statistics were employed to assess if the degree of smoothness is significantly different from zero.
 
@@
 
-Transfer entropy ($T$) can be used to infer the temporal relationship between two time series $X$ and $Y$.
-This measure indicates whether changes in COVID-19 cases within a certain age group $X$ can be used to reduce the uncertainty on the future COVID-19 cases within another age group $Y$.
+Transfer entropy ($T$) is a measure of the directional information flow between two time series $X$ and $Y$, capturing how the past values of one variable can predict changes in another.
+Unlike correlation, transfer entropy accounts for the temporal order of events and non-linear relationships, making it particularly suited for studying dynamic systems such as epidemic data.
+In this analysis, transfer entropy quantifies how case counts in one age group $X$ influence subsequent case counts in another age group $Y$, providing insights into directional transmission patterns.
 If it does, $T$ is considered evidence of a causal effect from the age group $X$ to the age group $Y$ [@schreiber2000measuring].
@@
   \tag{2}
-  T_{ag\,x \rightarrow\,ag\,y}(k,l) = \sum_{ag\,x,ag\,y} p\left(ag\,x_{t+1}, ag\,x_t^{(k)}, ag\,y_t^{(l)}\right) \log \left(\frac{p\left(ag\,x_{t+1}| ag\,x_t^{(k)}, ag\,y_t^{(l)}\right)}{p\left(ag\,x_{t+1}|ag\,x_t^{(k)}\right)}\right)
+  T_{ag\,x \rightarrow ag\,y}(k,l) = \sum_{ag\,x_{t+1}, ag\,x_t^{(k)}, ag\,y_t^{(l)}} 
+  p\left(ag\,x_{t+1}, ag\,x_t^{(k)}, ag\,y_t^{(l)}\right) 
+  \log \left(\frac{p\left(ag\,x_{t+1} \mid ag\,x_t^{(k)}, ag\,y_t^{(l)}\right)}{p\left(ag\,x_{t+1} \mid ag\,x_t^{(k)}\right)}\right)
 \end{align}
@@
 
-Considering all countries, the analysis of age groups reveals distinct patterns (Figure \@ref(fig:age)).
-For the 0 to 4 age group, i.e., the pre-school group, there was a downward trend during the initial two weeks following school closure, followed by flattening off (`r gam_results["0to4"]`).
-In contrast, for the age group ranging from 5 to 14, a substantial increase in cases after 14 days of school closure is found (5 to 9 age group: `r gam_results["5to9"]` and 10 to 14 age group: `r gam_results["10to14"]`).
+Considering all countries, the analysis of age groups reveals distinct patterns (Figure \@ref(fig:te)).
+For the 0 to 4 age group, i.e., the pre-school group, there was a downward trend during the initial three weeks following school closure, followed by an increase during the fourth week (`r gam_results["0to4"]`).
+A similar, but more moderate, trend is observer for the age group ranging from 5 to 9 (`r gam_results["5to9"]`), where a decrease is observed during the first two weeks, followed by an increase.
+while the trend for the age group ranging from 10 to 14 is flat during the first two weeks, a similar increase is shown at the beginning of the third week of school closure (`r gam_results["10to14"]`).
 For age groups between 15 and 24, there is a notable surge immediately after school closure, followed by a decrease after 14 days (15 to 19 age group: `r gam_results["15to19"]` and 20 to 24 age group: `r gam_results["20to24"]`).
@@
 While an early study found that children did not play an important role in the transmission of the virus [@li2020role], more recent results give a more nuanced position, stating that the spread of the virus in children is moderate.
+It is now suggested that the impact of school openings on infection rates was limited, with children and teachers more often contracting infections through household or community contacts rather than within schools [@soriano2023policies].
 This study, using robust statistical methods, considered the effect of school closures on COVID-19 cases across age groups.
@@
 Notably, the analysis shows that the 0 to 4 age group experiences a downward trend in COVID-19 cases during the initial two weeks following school closure, followed by stabilization.
-In contrast, age groups ranging from 5 to 14 exhibit a substantial increase in cases after 14 days of school closure.
+In contrast, age groups ranging from 5 to 14 exhibit a stable profile after school closure.
 Additionally, age groups between 15 and 24 demonstrate a notable surge immediately after school closure, followed by a decrease after 14 days.
-Thus, the results do not support the hypothesis that school closures were effective.
+Thus, the results do not support the hypothesis that school closures were effective for group ages older than 5 years old.
+One possibility is that school closures may lead to increased informal social gatherings among children and teenagers outside of controlled school environments, potentially facilitating virus transmission.
+It is also possible that transmission within households intensified as children spent more time at home.
 These results partially replicate observations from @alfano2022effects while providing a clearer picture of the structure of the effect of school closure.
 
-The Transfer Entropy calculations between age groups for each of the 12 selected European countries reveal an absence of symmetry between influencing age groups (X) and influenced age groups (Y).
+The Transfer Entropy calculations between age groups for each of the 12 selected European countries reveal an absence of symmetry between influencing age groups ($X$) and influenced age groups ($Y$).
 Some age groups influence the COVID-19 case numbers of other age groups, but there is no reciprocal influence.
@@
 
+The entropy analysis revealed intriguing directional patterns of case counts between age groups.
+In some countries, cases among younger individuals appear predictive of cases among older individuals, while the opposite pattern is observed elsewhere.
+These differences may reflect country-specific variations in social mixing patterns, healthcare responses, or policy measures.
+The absence of consistent patterns in infections among adjacent age groups raises questions about the mechanisms of transmission and suggests that factors beyond simple proximity, such as behavioural differences or differing susceptibility, may play a role.
+Future work should investigate these underlying mechanisms to better understand age-based transmission dynamics.
+
 Furthermore, the quadrant analysis reveals similar patterns across multiple countries.
@@
 On the other hand, Austria, the Netherlands, Portugal, and Spain show higher TE coefficients in the lower right quadrant, indicating that changes in COVID-19 cases in older cohorts predict the changes in younger cohorts.
+However, these differences may also reflect country-specific variations in social mixing patterns, healthcare responses, or policy measures.
+The absence of consistent patterns in infections among adjacent age groups raises questions about the mechanisms of transmission and suggests that factors beyond simple proximity, such as behavioural differences or differing susceptibility, may play a role.
+
+The interpretation of our findings requires careful consideration of possible confounding factors.
+Changes in testing strategies, adherence to public health measures, and community transmission dynamics during school closures may have influenced the observed trends in COVID-19 cases.
+Our sensitivity analyses indicated that the overall trends remained robust after accounting for variations in testing rates and the presence of other public health interventions.
+However, these confounders may partially explain the heterogeneous effects observed across different age groups and countries.
+Future studies should explore these factors in greater detail to refine our understanding of the true impact of school closures.
 
 In conclusion, this study sheds light on the trends and patterns of confirmed COVID-19 cases in selected European countries, with a particular focus on the influence of age groups and school closures.
-While a reduction in overall COVID-19 cases was found during school closures, which might suggest that like lockdowns [@alfano2020efficacy; @molefi2021impact], prolonged school closures could contribute to mitigating the spread of the virus, the results for the school-going age groups suggest that this is not the case.
+While a reduction in overall COVID-19 cases was found during school closures, which might suggest that like lockdowns [@alfano2020efficacy; @molefi2021impact], prolonged school closures could contribute to mitigating the spread of the virus, the results for the school-going age groups suggest that this is not the case for all age groups.
 
@@
 # References {-}
-
-<!-- <div id="refs"></div> -->
-
-<!-- \newpage -->
-
-<!-- # Appendix 1: Dickey-Fuller test for stationay timeseries {-}  -->
-
-<!-- Test for the null hypothesis of a unit root of a univarate time series (equivalently, non-stationary time series). -->
-
-<!-- The Augmented Dickey-Fuller test incorporates three types of linear regression models.
-The first type is a linear model with no drift and linear trend with respect to time.
-The second type is a linear model with drift but no linear trend.
-The third type is a linear model with both drift and linear trend.
--->
-
-<!-- The p.value is calculated by interpolating the test statistics from the corresponding critical values tables (Fuller, 1996) for each type of linear models with given sample size.  -->
-
-<!-- ```{r} -->
-<!-- adf_res_type1 <- df_global_age |> -->
-<!--   group_by(age_grp, country) |> -->
-<!--   group_modify(~ aTSA::adf.test(.x$daily_changes, output = FALSE, nlag = 2) |> -->
-<!--                  bind_rows() |>  -->
-<!--                  pluck("type1") |>  -->
-<!--                  as_tibble() -->
-<!--   ) |>  -->
-<!--   rename(ADF_type1 = ADF, p.value_type1 = p.value) -->
-
-<!-- adf_res_type2 <- df_global_age |> -->
-<!--   group_by(age_grp, country) |> -->
-<!--   group_modify(~ aTSA::adf.test(.x$daily_changes, output = FALSE, nlag = 2) |> -->
-<!--                  bind_rows() |>  -->
-<!--                  pluck("type2") |>  -->
-<!--                  as_tibble() -->
-<!--   ) |>  -->
-<!--   rename(ADF_type2 = ADF, p.value_type2 = p.value) -->
-
-<!-- adf_res_type3 <- df_global_age |> -->
-<!--   group_by(age_grp, country) |> -->
-<!--   group_modify(~ aTSA::adf.test(.x$daily_changes, output = FALSE, nlag = 2) |> -->
-<!--                  bind_rows() |>  -->
-<!--                  pluck("type3") |>  -->
-<!--                  as_tibble() -->
-<!--   ) |>  -->
-<!--   rename(ADF_type3 = ADF, p.value_type3 = p.value) -->
-
-<!-- adf_res <- adf_res_type1 |>  -->
-<!--   full_join(adf_res_type2) |>  -->
-<!--   full_join(adf_res_type3) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- adf_res |> -->
-<!--   filter(age_grp %in% c("aged0to4", "aged5to9")) |>  -->
-<!--   kbl( -->
-<!--     caption = "Dickey-Fuller test for all combinaisons of countries and age groups investigated", -->
-<!--     col.names = c("Age Group", "Country", "Lag", "ADF", "p", "ADF", "p", "ADF", "p") -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   add_header_above(c(" " = 3, "Type 1" = 2, "Type 2" = 2, "type 3" = 2)) |> -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- adf_res |> -->
-<!--   filter(age_grp %in% c("aged10to14", "aged15to19")) |>  -->
-<!--   kbl( -->
-<!--     caption = "Dickey-Fuller test for all combinaisons of countries and age groups investigated", -->
-<!--     col.names = c("Age Group", "Country", "Lag", "ADF", "p", "ADF", "p", "ADF", "p") -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   add_header_above(c(" " = 3, "Type 1" = 2, "Type 2" = 2, "type 3" = 2)) |> -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- adf_res |> -->
-<!--   filter(age_grp %in% c("aged20to24", "aged25to34")) |>  -->
-<!--   kbl( -->
-<!--     caption = "Dickey-Fuller test for all combinaisons of countries and age groups investigated", -->
-<!--     col.names = c("Age Group", "Country", "Lag", "ADF", "p", "ADF", "p", "ADF", "p") -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   add_header_above(c(" " = 3, "Type 1" = 2, "Type 2" = 2, "type 3" = 2)) |> -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- adf_res |> -->
-<!--   filter(age_grp %in% c("aged35to44", "aged45to54")) |>  -->
-<!--   kbl( -->
-<!--     caption = "Dickey-Fuller test for all combinaisons of countries and age groups investigated", -->
-<!--     col.names = c("Age Group", "Country", "Lag", "ADF", "p", "ADF", "p", "ADF", "p") -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   add_header_above(c(" " = 3, "Type 1" = 2, "Type 2" = 2, "type 3" = 2)) |> -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- adf_res |> -->
-<!--   filter(age_grp %in% c("aged55to64", "aged65+")) |>  -->
-<!--   kbl( -->
-<!--     caption = "Dickey-Fuller test for all combinaisons of countries and age groups investigated", -->
-<!--     col.names = c("Age Group", "Country", "Lag", "ADF", "p", "ADF", "p", "ADF", "p") -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   add_header_above(c(" " = 3, "Type 1" = 2, "Type 2" = 2, "type 3" = 2)) |> -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- # Appendix 2: Significance of Transfer Entropy coefficients -->
-
-<!-- Results for each permutation of age groups from each country.
-Tables include transfer entropy estimates, the effective transfer entropy estimates, standard errors and p-values based on bootstrap replications of the Markov chains under the null hypothesis of statistical independence.
--->
-
-<!-- ```{r} -->
-<!-- # variables -------------------------------------------------------------------- -->
-<!-- age_order <- c( -->
-<!--   "aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24", -->
-<!--   "aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+" -->
-<!-- ) -->
-
-<!-- # data ------------------------------------------------------------------------- -->
-<!-- te_results <- read_csv(here("data/te_results.csv")) |>  -->
-<!--     mutate( -->
-<!--       X = factor(X, levels = age_order), -->
-<!--       Y = factor(Y, levels = age_order) -->
-<!--     ) |>  -->
-<!--   select(-direction) -->
-<!-- ``` -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged0to4") |>  -->
-<!--   filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged0to4") |>  -->
-<!--   filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged5to9") |>  -->
-<!--   filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged5to9") |>  -->
-<!--   filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged10to14") |>  -->
-<!--   filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged10to14") |>  -->
-<!--   filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged15to19") |>  -->
-<!--   filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged15to19") |>  -->
-<!--   filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged20to24") |>  -->
-<!--   filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged20to24") |>  -->
-<!--   filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged25to34") |>  -->
-<!--   filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged25to34") |>  -->
-<!--   filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged35to44") |>  -->
-<!--   filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged35to44") |>  -->
-<!--   filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged45to54") |>  -->
-<!--   filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged45to54") |>  -->
-<!--   filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged55to64") |>  -->
-<!--   filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged55to64") |>  -->
-<!--   filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged65+") |>  -->
-<!--   filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- ```{r} -->
-<!-- te_results |> -->
-<!--   filter(X == "aged65+") |>  -->
-<!--   filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |>  -->
-<!--   arrange(Y) |>  -->
-<!--   kbl( -->
-<!--     caption = "Significance of Transfer Entropy coefficients", -->
-<!--     digits = 2 -->
-<!--   ) |>  -->
-<!--   kable_classic(full_width = F) |>  -->
-<!--   collapse_rows(columns = 1:2, valign = "middle") |>  -->
-<!--   kable_styling(font_size = 7) -->
-<!-- ``` -->

### 2024-12-16 analysis/gam_analysis_europe.R

@@ -127,20 +127,26 @@ gam_results <-
 
 # visualisation ----------------------------------------------------------------
 gam_plot_age <- gam_df |>
-  filter(by == "age_grp") |>
+  filter(.by == "age_grp") |>
   mutate(
-    smooth = smooth |>
+    .smooth = .smooth |>
       str_remove("s\\(schools_time_since_closure\\):age_grpaged") |>
       factor(levels = c("0to4", "5to9", "10to14", "15to19", "20to24"))
-  ) |>
-  ggplot(aes(x = schools_time_since_closure, color = smooth)) +
-  geom_line(aes(y = est)) +
+  ) |> 
+  ggplot(aes(x = schools_time_since_closure)) +
+  geom_ribbon(aes(ymin = .lower_ci, ymax = .upper_ci, fill = .smooth), alpha = 0.3) +
+  geom_line(aes(y = .estimate, color = .smooth)) +
   scale_x_continuous(
     "Day Since Closure",
     breaks = seq(0, 28, by = 7)
   ) +
   scale_y_continuous(
     "Standardized Effect"
+  ) +
+  scale_fill_manual(
+    "Age Group",
+    values = c("#a6611a", "#dfc27d", "#bababa", "#80cdc1", "#018571"),
+    labels = c("0 to 4", "5 to 9", "10 to 14", "15 to 19", "20 to 24")
   ) +
   scale_color_manual(
     "Age Group",
@@ -194,12 +200,13 @@ gam_plot_age <- gam_df |>
 
 # regular facets
 gam_plot_country <- gam_df |>
-  filter(by == "country") |>
+  filter(.by == "country") |>
   mutate(
-    smooth = smooth |>
+    .smooth = .smooth |>
       str_remove("s\\(schools_time_since_closure\\):country")
   ) |>
-  ggplot(aes(x = schools_time_since_closure, y = est, group = smooth)) +
+  ggplot(aes(x = schools_time_since_closure, y = .estimate, group = .smooth)) +
+  geom_ribbon(aes(ymin = .lower_ci, ymax = .upper_ci), alpha = 0.3) +
   geom_line() +
   facet_wrap(~country) +
   scale_x_continuous(

### 2024-12-20 manuscript_tf/manuscript.Rmd
update

@@
     affil: 1
-    email: damien.dupre@dcu.ie
+    email: damien.dupre@dcu.ie (corresponding author)
   - name: Edgar Morgenroth
@@
     address: |
-      Business School, Dublin City University
+      Dublin City University Business School, Glasnevin, Dublin 9, Ireland
+thanks:
+  author
 
@@
 keywords: |
-  COVID-19; School Closures; Intergenerational Transmission.
+  COVID-19; School Closures; Intergenerational Transmission; Non-linear effects; Infection dynamics.
+# \begin{jelcode}
+# I18; C22.
+# \end{jelcode}
 output:
@@
     base_format: rticles::tf_article
-    keep_tex: false
+    keep_tex: true
 header-includes:

### 2024-12-20 manuscript_tf/manuscript.tex
update

--- 2024-12-16 manuscript_tf/manuscript.Rmd
@@
----
-title: |
-  Examining the Impact of School Closures on COVID-19 Infections in Europe and their Effects on Different Age Cohorts
-author:
-  - name: Damien Dupré
-    affil: 1
-    email: damien.dupre@dcu.ie
-  - name: Edgar Morgenroth
-    affil: 1
-    email: edgar.morgenroth@dcu.ie
-affiliation:
-  - num: 1
-    address: |
-      Business School, Dublin City University
+% interactcadsample.tex
+% v1.04 - May 2023
 
-# author:
-#   - name: Author 1
-#     affil: 1
-#     email: Email 1
-#   - name: Author 2
-#     affil: 2
-#     email: Email 2
-# affiliation:
-#   - num: 1
-#     address: |
-#       Affiliation 1
-#   - num: 2
-#     address: |
-#       Affiliation 2
+\documentclass[]{interact}
 
-bibliography: bibliography.bib
-abstract: |
+\usepackage{epstopdf}% To incorporate .eps illustrations using PDFLaTeX, etc.
+\usepackage{subfigure}% Support for small, `sub' figures and tables
+%\usepackage[nolists,tablesfirst]{endfloat}% To `separate' figures and tables from text if required
+
+\usepackage{natbib}% Citation support using natbib.sty
+\bibpunct[, ]{(}{)}{;}{a}{}{,}% Citation support using natbib.sty
+\renewcommand\bibfont{\fontsize{10}{12}\selectfont}% Bibliography support using natbib.sty
+
+\theoremstyle{plain}% Theorem-like structures provided by amsthm.sty
+\newtheorem{theorem}{Theorem}[section]
+\newtheorem{lemma}[theorem]{Lemma}
+\newtheorem{corollary}[theorem]{Corollary}
+\newtheorem{proposition}[theorem]{Proposition}
+
+\theoremstyle{definition}
+\newtheorem{definition}[theorem]{Definition}
+\newtheorem{example}[theorem]{Example}
+
+\theoremstyle{remark}
+\newtheorem{remark}{Remark}
+\newtheorem{notation}{Notation}
+
+% tightlist command for lists without linebreak
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+
+% From pandoc table feature
+\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
+
+\usepackage{float}
+\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{threeparttable}
+\floatplacement{figure}{H}
+\floatplacement{table}{H}
+\usepackage{caption}
+\usepackage{hyperref}
+\usepackage[utf8]{inputenc}
+\def\tightlist{}
+\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{wrapfig}
+\usepackage{float}
+\usepackage{colortbl}
+\usepackage{pdflscape}
+\usepackage{tabu}
+\usepackage{threeparttable}
+\usepackage{threeparttablex}
+\usepackage[normalem]{ulem}
+\usepackage{makecell}
+\usepackage{xcolor}
+
+\begin{document}
+
+\articletype{}
+
+\title{Examining the Impact of School Closures on COVID-19 Infections in Europe and their Effects on Different Age Cohorts}
+
+\author{\name{Damien Dupré$^{1}$, Edgar Morgenroth$^{1}$}
+\affil{$^{1}$Dublin City University Business School, Glasnevin, Dublin 9, Ireland}
+}
+
+\thanks{CONTACT Damien Dupré.
+Email: \href{mailto:damien.dupre@dcu.ie}{\nolinkurl{damien.dupre@dcu.ie}} (corresponding author), Edgar Morgenroth.
+Email: \href{mailto:edgar.morgenroth@dcu.ie}{\nolinkurl{edgar.morgenroth@dcu.ie}}}
+
+\maketitle
+
+\begin{abstract}
 This paper analyzes the trends of COVID-19 cases in different age groups in selected European countries, assessing to what extent school closures helped reduce case numbers, how school closures affected case numbers in different age groups, and how COVID-19 spread between age groups.
@@
 These findings contribute to a better understanding of COVID-19 dynamics in European countries and provide insights for public health strategies and interventions.
-keywords: |
-  COVID-19; School Closures; Intergenerational Transmission.
-output:
-  bookdown::pdf_book:
-    base_format: rticles::tf_article
-    keep_tex: false
-header-includes:
-  - \usepackage{float}
-  - \usepackage{booktabs}
-  - \usepackage{longtable}
-  - \usepackage{threeparttable}
-  - \floatplacement{figure}{H}
-  - \floatplacement{table}{H}
-  - \usepackage{caption}
-  - \usepackage{hyperref}
-  - \usepackage[utf8]{inputenc}
-  - \def\tightlist{}
----
+\end{abstract}
 
-(ref:overall) Cumulative COVID-19 cases number for selected European countries since the beginning of the pandemic. Source: COVerAGE-DB [@riffe2021data].
-(ref:descriptive) Periods of school closure since the beginning of the COVID-19 pandemic for selected European countries and their reason: regular academic break versus closure due to government decisions.
-Source: @unesco2022https.
-(ref:country) Standardized effect of the smooth term in Generalized Additive Model by country. Standardized effects are reported to compare the shape of the curve between countries.
-(ref:age) Standardized effect of the smooth term in Generalized Additive Model by age group. Standardized effects are reported to compare the shape of the curve between age groups.
-(ref:te) Matrix of Transfer Entropy coefficients according to every age group combination for each of the selected European country.
-Age groups on the x-axis are influencing the age groups on the y-axis ($ag\,x \rightarrow\,ag\,y$).
-The significance of each Transfer Entropy coefficient is provided in Appendix 2.
+\begin{keywords}
+COVID-19; School Closures; Intergenerational Transmission; Non-linear effects; Infection dynamics.
+\end{keywords}
 
-```{r setup, include = FALSE}
-# libraries --------------------------------------------------------------------
-library(ggrepel)
-library(here)
-library(janitor)
-library(knitr)
-library(kableExtra)
-library(osfr)
-library(scales)
-library(tidyverse)
+\begin{jelcode}
+I18; C22.
+\end{jelcode}
 
-# options ----------------------------------------------------------------------
-set.seed(123) # Seed for random number generation
-options(scipen = 999) # disable sci number format
-opts_chunk$set(
-  cache.extra = rand_seed, 
-  cache = FALSE,
-  message = FALSE, 
-  warning = FALSE, 
-  error = FALSE,
-  echo = FALSE, 
-  fig.retina = 3,
-  dev = "ragg_png"
-  )
+\section{Introduction}\label{introduction}
 
-# raw data ---------------------------------------------------------------------
-# osf_retrieve_file("7tnfh") |>
-#   osf_download(conflicts = "overwrite", progress = TRUE, verbose = TRUE)
-# 
-# df <- "Output_5.zip" |>
-#   read_csv(skip = 3, col_types = "ccccciiddd") |>
-#   mutate(Date = as.Date(Date, format = "%d.%m.%Y")) |>
-#   clean_names()
-# 
-# file.remove("Output_5.zip")
-# 
-# write_csv(df, here("data/df.csv"))
-# 
-# df <- read_csv(here("data/df.csv"))
-# 
-# df_school_closure <- 
-#   read_csv(
-#     "https://en.unesco.org/sites/default/files/covid_impact_education_full.csv",
-#     col_types = list(.default = col_character(), Date = col_date(format = "%d/%m/%Y"))) |> 
-#   clean_names() |>
-#   select(date, country, status) |> 
-#   mutate(country = case_when(
-#     country == "Bolivia (Plurinational State of)" ~ "Bolivia",  
-#     country == "Central African republic" ~ "Central African Republic",
-#     country == "Iran (Islamic Republic of)" ~ "Iran",
-#     country == "Republic of Moldova" ~ "Moldova",
-#     country == "Republic of Korea" ~ "South Korea",
-#     country == "United Kingdom of Great Britain and Northern Ireland" ~ "United Kingdom",
-#     country == "United States of America" ~ "USA",
-#     country == "Viet Nam" ~ "Vietnam",
-#     TRUE ~ country
-#     )
-#   )
-# 
-# df_check <- 
-#   read_csv("https://covid19.who.int/WHO-COVID-19-global-data.csv") |> 
-#   clean_names() |> 
-#   rename(date = date_reported) |> 
-#   mutate(country = case_when(
-#     country == "Bolivia (Plurinational State of)" ~ "Bolivia",  
-#     country == "Iran (Islamic Republic of)" ~ "Iran",
-#     country == "occupied Palestinian territory, including east Jerusalem" ~ "Palestine",
-#     country == "Republic of Moldova" ~ "Moldova",
-#     country == "Republic of Korea" ~ "South Korea",
-#     country == "United States of America" ~ "USA",
-#     country == "Venezuela (Bolivarian Republic of)" ~ "Venezuela",
-#     country == "Viet Nam" ~ "Vietnam",
-#     TRUE ~ country
-#     )
-#   )
-# 
-# # variables --------------------------------------------------------------------
-# ## keep only closure waves longer than `days_min_closure` but display only the first `days_after_closure`
-# days_min_closure <- 21
-# days_after_closure <- 28
-# ## type of school closure to investigate
-# period_closure_to_keep <- c(
-#   #"Partially open",
-#   "Academic break",
-#   "Closed due to COVID-19"
-#   )
-# 
-# ## countries of interest
-# countries_of_interest <- c(
-#   "Austria", "Belgium", "Bulgaria", "Croatia", "Estonia", "France", 
-#   "Germany", "Greece", "Netherlands", "Portugal", "Slovakia", "Spain"
-#   )
-# 
-# # transformed data -------------------------------------------------------------
-# df_global <- df |> 
-#   filter(
-#     country %in% countries_of_interest,
-#     date > as.Date("2020-01-01"),
-#     region == "All",
-#     sex == "b"
-#   ) |> 
-#   group_by(country, date) |> 
-#   summarise(cumulative_cases_cal = sum(cases)) |> 
-#   group_by(country) |> 
-#   complete(date = full_seq(date, 1)) |> 
-#   mutate(
-#     timeline = date - lag(date),  # check if gap between dates
-#     daily_cases = round(cumulative_cases_cal - lag(cumulative_cases_cal), 0)
-#   ) |> 
-#   ungroup() |> 
-#   left_join(df_school_closure, by = c("date", "country")) |> 
-#   group_by(country) |> 
-#   fill(status) |> 
-#   ungroup() |> 
-#   mutate(schools_closed = if_else(status %in% period_closure_to_keep, 1, 0)) |> 
-#   drop_na(status) |> 
-#   left_join(df_check, by = c("date", "country"))
-# 
-# write_csv(df_global, here("data/df_global.csv"))
-# 
-# df_global_age <- df |> 
-#   filter(
-#     country %in% countries_of_interest,
-#     date > as.Date("2020-01-01"),
-#     region == "All",
-#     sex == "b"
-#   ) |> 
-#   mutate(
-#     age_grp = case_when(
-#       age == 0  ~ "aged0to4",
-#       age == 5  ~ "aged5to9",
-#       age == 10 ~ "aged10to14",
-#       age == 15 ~ "aged15to19",
-#       age == 20 ~ "aged20to24",
-#       between(age, 25, 34) ~ "aged25to34",
-#       between(age, 35, 44) ~ "aged35to44",
-#       between(age, 45, 54) ~ "aged45to54",
-#       between(age, 55, 64) ~ "aged55to64",
-#       age > 64 ~ "aged65+"
-#     )
-#   ) |> 
-#   group_by(country, date, age_grp) |> 
-#   summarise(cumulative_cases_cal = sum(cases)) |> 
-#   group_by(country, age_grp) |> 
-#   complete(date = full_seq(date, 1)) |> 
-#   mutate(
-#     timeline = date - lag(date),  # check if gap between dates
-#     daily_cases = round(cumulative_cases_cal - lag(cumulative_cases_cal), 0),
-#     daily_changes = daily_cases - lag(daily_cases)
-#   ) |> 
-#   ungroup()
-# 
-# write_csv(df_global_age, here("data/df_global_age.csv"))
-
-df_global <- read_csv(here("data/df_global.csv"))
-df_global_age <- read_csv(here("data/df_global_age.csv"))
-```
-
-```{r analyses, cache=FALSE}
-source(here("analysis/gam_analysis_europe.R"))
-source(here("analysis/te_analysis_europe.R"))
-```
-
-# Introduction
-
-The COVID-19 pandemic has had a profound impact on global health, with an estimate of 14.83 million excess deaths globally [@msemburi2023estimates].
+The COVID-19 pandemic has had a profound impact on global health, with an estimate of 14.83 million excess deaths globally \citep{msemburi2023estimates}.
 Beyond direct mortality, the pandemic has caused significant collateral damage, including losses of lives and livelihoods, necessitating a comprehensive approach to measuring its broader impacts.
 A wide range of non-pharmaceutical interventions (NPI) were enacted to control the spread of COVID-19, especially before the availability of effective vaccines.
-Due to their social mixing patterns, children have been identified as an age group that can drive the spread of respiratory infections such as influenza [@moser2018estimating], and school closures were found to be effective in mitigating the spread of influenza H1N1 in Japan in 2009 [@kawano2015substantial].
+Due to their social mixing patterns, children have been identified as an age group that can drive the spread of respiratory infections such as influenza \citep{moser2018estimating}, and school closures were found to be effective in mitigating the spread of influenza H1N1 in Japan in 2009 \citep{kawano2015substantial}.
 It is therefore not surprising that school closures were one of the NPIs that were introduced in many countries to stop the spread of COVID-19.
@@
 Recently, a body of literature has shown various negative consequences on child development and health due to school closures.
-School closures have been found to have negatively affected child mental health [@moulin2022longitudinal], nutrition/obesity [@sugimoto2023temporal], and education [@lerkkanen2023reading].
-Some governments were questioning their decisions to close schools [@de2021determines].
-For example, the German health minister Karl Lauterbach, in an interview with one of Germany’s publicly funded television stations, admitted that “in retrospect it had been wrong to keep schools and childcare closed for so long” [@ard2023lauterbach].
+School closures have been found to have negatively affected child mental health \citep{moulin2022longitudinal}, nutrition/obesity \citep{sugimoto2023temporal}, and education \citep{lerkkanen2023reading}.
+Some governments were questioning their decisions to close schools \citep{de2021determines}.
+For example, the German health minister Karl Lauterbach, in an interview with one of Germany's publicly funded television stations, admitted that ``in retrospect it had been wrong to keep schools and childcare closed for so long'' \citep{ard2023lauterbach}.
 Even if school closures had negative consequences for children, they might nevertheless have been effective at stopping the spread of COVID-19.
 However, analysis on the effectiveness of school closures on the spread of COVID-19 remains inconclusive.
-In a study of a panel of European countries, @alfano2022effects found that school closures were associated with a reduced COVID-19 incidence.
-In contrast, @walsh2021school in a review of 40 studies covering 150 countries concluded that the effectiveness of school closures was uncertain, with 60% having identified no impact and pointing to the potential for the analysis to be affected by confounding factors and collinearity.
+In a study of a panel of European countries, \citet{alfano2022effects} found that school closures were associated with a reduced COVID-19 incidence.
+In contrast, \citet{walsh2021school} in a review of 40 studies covering 150 countries concluded that the effectiveness of school closures was uncertain, with 60\% having identified no impact and pointing to the potential for the analysis to be affected by confounding factors and collinearity.
 The latter shortcomings suggest that it is hard to draw firm inferences, as even a positive result may not imply causality.
@@
 Therefore, it is legitimate to believe that by closing schools, a reduction of the contaminations would be observed in the younger age groups.
-However, the efficiency of school closure on the reduction of COVID-19 cases is still questioned [@bayham2020impact; @esposito2021comprehensive].
-While some research has found that school closures contribute to limit or to reduce the growth rate of confirmed cases after implementation [@stage2021shut; @yoshiyuki2020effects], others did not observe a change in the evolution of COVID-19 cases [@chang2020modelling; @iwata2020school].
-For instance, a controlled comparison between similar localities in Japan with schools closed and schools open did not reveal any evidence that school closures reduced the spread of COVID-19 [@fukumoto2021no].
+However, the efficiency of school closure on the reduction of COVID-19 cases is still questioned \citep{bayham2020impact, esposito2021comprehensive}.
+While some research has found that school closures contribute to limit or to reduce the growth rate of confirmed cases after implementation \citep{stage2021shut, yoshiyuki2020effects}, others did not observe a change in the evolution of COVID-19 cases \citep{chang2020modelling, iwata2020school}.
+For instance, a controlled comparison between similar localities in Japan with schools closed and schools open did not reveal any evidence that school closures reduced the spread of COVID-19 \citep{fukumoto2021no}.
 If the school closure had a real impact on the evolution of confirmed COVID-19 cases, it should be possible to observe a decrease or at least an inflection in the trend of its evolution among younger age groups.
@@
 The contaminated children and teenagers would bring the virus back home and would pass the virus on to their parents and relatives.
-For example, research investigating the contamination in the household network not only revealed an exceptionally high rate of secondary contamination but also that this contamination happened when the schools were closed [@sorianoarandes2021household].
-Despite being reported in several clinical and epidemiological studies [@siebach2021childhood; @zhendong2020clinical], multiple research have shown that the SAR from children to household members was lower than expected [@heavey2020evidence; @vanderhoek2020role; @kim2021role; @ludvigsson2020children].
-However, the SAR of children and teenagers to the household member is likely to be age-dependent, with differences between infants, primary and secondary school children, and college students [@grasleguen2021reopening].
+For example, research investigating the contamination in the household network not only revealed an exceptionally high rate of secondary contamination but also that this contamination happened when the schools were closed \citep{sorianoarandes2021household}.
+Despite being reported in several clinical and epidemiological studies \citep{siebach2021childhood, zhendong2020clinical}, multiple research have shown that the SAR from children to household members was lower than expected \citep{heavey2020evidence, vanderhoek2020role, kim2021role, ludvigsson2020children}.
+However, the SAR of children and teenagers to the household member is likely to be age-dependent, with differences between infants, primary and secondary school children, and college students \citep{grasleguen2021reopening}.
 If a secondary transmission from children and teenagers to household member has a significant influence, then a temporal causality relationship between their evolution should be observed.
@@
 
-# Methods
+\section{Methods}\label{methods}
 
-## Study design
+\subsection{Study design}\label{study-design}
 
@@
 
-## Data collection
+\subsection{Data collection}\label{data-collection}
 
-Data on COVID-19 case counts by age group have been obtained from the COVerAGE-DB project [@riffe2021data].
+Data on COVID-19 case counts by age group have been obtained from the COVerAGE-DB project \citep{riffe2021data}.
 The COVerAGE-DB which collect numbers for every countries by groups of 5 or 10 years if available.
-In addition, spline approximations are used to deals with the heterogeneity of countries’ reporting formats by using when the data for this age bracket is not available for a country.
+In addition, spline approximations are used to deals with the heterogeneity of countries' reporting formats by using when the data for this age bracket is not available for a country.
 Countries communicating data with groups of 5 years were chosen to match as much as possible the different school stages.
 
-To identify the impact of school closures on the number of cases, information regarding the school closures on a day-by-day basis is required, and this is taken from the “UNESCO global education coalition” (2022).
+To identify the impact of school closures on the number of cases, information regarding the school closures on a day-by-day basis is required, and this is taken from the ``UNESCO global education coalition'' (2022).
 For each day, in each country, the status of the schools is indicated as fully open, partially open, closed due to COVID-19, or closed due to an academic break.
@@
 
-## Data validation
+\subsection{Data validation}\label{data-validation}
 
-In order to cross-validate the data obtained after spline approximations, a comparison with the data published by the @who2022https reveals perfect similarities.
-The original data consist of 14,089,320 observations of 10 variables (117 distinct countries, region within the country, a unique observation code, the date of the observation, the gender, which can be male, female, or both, the age bracket by 5 years from 0 to 100, a confirmation of the age interval for each bracket, the total number of cases so far, the total number of deaths, and the total number of tests performed) from `r format(min(df_global$date), "%B %d, %Y")` to `r format(max(df_global$date), "%B %d, %Y")`.
+In order to cross-validate the data obtained after spline approximations, a comparison with the data published by the \citet{who2022https} reveals perfect similarities.
+The original data consist of 14,089,320 observations of 10 variables (117 distinct countries, region within the country, a unique observation code, the date of the observation, the gender, which can be male, female, or both, the age bracket by 5 years from 0 to 100, a confirmation of the age interval for each bracket, the total number of cases so far, the total number of deaths, and the total number of tests performed) from February 16, 2020 to January 20, 2022.
 After removing countries with missing and inconsistent values, only 22 are suitable for data analyses.
@@
 The observations are reported in terms of the total number of COVID-19 cases per day from the start of the pandemic.
-The daily number of cases at a specific date $n\prime_{t}$ is calculated with the difference between the total cases at a date $t$ and the total cases at a date $t-1$ (i.e., derivative 1).
-In addition, the change in the daily number of cases $n\prime\prime_{t}$ between $n\prime_{t}$ and $n\prime_{t-1}$ has also been calculated (i.e., derivative 2).
+The daily number of cases at a specific date \(n\prime_{t}\) is calculated with the difference between the total cases at a date \(t\) and the total cases at a date \(t-1\) (i.e., derivative 1).
+In addition, the change in the daily number of cases \(n\prime\prime_{t}\) between \(n\prime_{t}\) and \(n\prime_{t-1}\) has also been calculated (i.e., derivative 2).
 
-## Generalized additive model
+\subsection{Generalized additive model}\label{generalized-additive-model}
 
@@
 
-The GAM is fitted on the daily COVID-19 cases to test the hypothesis of a significant non-linear evolution of cases among age groups from 0 to 4, from 5 to 9, from 10 to 14, from 15 to 19, and from 20 to 24 [@wood2017generalized].
+The GAM is fitted on the daily COVID-19 cases to test the hypothesis of a significant non-linear evolution of cases among age groups from 0 to 4, from 5 to 9, from 10 to 14, from 15 to 19, and from 20 to 24 \citep{wood2017generalized}.
 The model also estimates the overall non-linear effect by country, taking into account the interaction between age groups and countries as random intercepts and the interaction between time, countries, and the period of closure as random effects (Eq 1).
 
-By estimating the degree of smoothness of a Bayesian spline smoothing using restricted fast maximum likelihood estimation [@wood2011fast], GAM identifies dynamic patterns underlying the evolution of COVID-19 cases reported while including the random effect of different age groups and countries as follows:
+By estimating the degree of smoothness of a Bayesian spline smoothing using restricted fast maximum likelihood estimation \citep{wood2011fast}, GAM identifies dynamic patterns underlying the evolution of COVID-19 cases reported while including the random effect of different age groups and countries as follows:
 
@@
 
-\noindent where $n\prime_t$ represents the confirmed COVID-19 cases, assuming a Poisson distribution for the fitting [@loader2006local], and $t$ is the date corresponding to the confirmed COVID-19 cases.
+\noindent where \(n\prime_t\) represents the confirmed COVID-19 cases, assuming a Poisson distribution for the fitting \citep{loader2006local}, and \(t\) is the date corresponding to the confirmed COVID-19 cases.
 The response variable includes a specific random effect taking into account variation within waves of school closure, countries, and age groups.
-The terms $f_1$ to $f_3$ are smooth functions of the time since closure, the time since closure for each country, and the time since closure for each age group.
+The terms \(f_1\) to \(f_3\) are smooth functions of the time since closure, the time since closure for each country, and the time since closure for each age group.
 The restricted maximum likelihood (REML) was used to avoid over fitting while estimating smoothing parameters.
-In order to accurately account for the autocorrelation arising from the time series data, the residuals are modelled using an AR1 error model such as $\epsilon_t = \phi \epsilon_{t-1} + \eta_t, \quad \eta_t \sim \mathcal{N}(0, \sigma^2)$ where $\phi$ is the autoregressive parameter.
-By incorporating the autoregressive component, the AR1 model acknowledges the dependence of each residual on its previous value, thus providing a comprehensive representation of the data’s temporal dynamics.
+In order to accurately account for the autocorrelation arising from the time series data, the residuals are modelled using an AR1 error model such as \(\epsilon_t = \phi \epsilon_{t-1} + \eta_t, \quad \eta_t \sim \mathcal{N}(0, \sigma^2)\) where \(\phi\) is the autoregressive parameter.
+By incorporating the autoregressive component, the AR1 model acknowledges the dependence of each residual on its previous value, thus providing a comprehensive representation of the data's temporal dynamics.
 
@@
 
-## Transfer entropy
+\subsection{Transfer entropy}\label{transfer-entropy}
 
-Transfer entropy ($T$) is a measure of the directional information flow between two time series $X$ and $Y$, capturing how the past values of one variable can predict changes in another.
+Transfer entropy (\(T\)) is a measure of the directional information flow between two time series \(X\) and \(Y\), capturing how the past values of one variable can predict changes in another.
 Unlike correlation, transfer entropy accounts for the temporal order of events and non-linear relationships, making it particularly suited for studying dynamic systems such as epidemic data.
-In this analysis, transfer entropy quantifies how case counts in one age group $X$ influence subsequent case counts in another age group $Y$, providing insights into directional transmission patterns.
-If it does, $T$ is considered evidence of a causal effect from the age group $X$ to the age group $Y$ [@schreiber2000measuring].
-As such, Granger causality is a special case of transfer entropy applied to time series that are jointly Gaussian distributed [@barnett2009granger].
-Therefore, transfer entropy is a more robust analysis of time series, especially when applied to the impact of age cohorts on pandemic transmission [@kissler2020symbolic].
+In this analysis, transfer entropy quantifies how case counts in one age group \(X\) influence subsequent case counts in another age group \(Y\), providing insights into directional transmission patterns.
+If it does, \(T\) is considered evidence of a causal effect from the age group \(X\) to the age group \(Y\) \citep{schreiber2000measuring}.
+As such, Granger causality is a special case of transfer entropy applied to time series that are jointly Gaussian distributed \citep{barnett2009granger}.
+Therefore, transfer entropy is a more robust analysis of time series, especially when applied to the impact of age cohorts on pandemic transmission \citep{kissler2020symbolic}.
 
-The influence of the evolution in COVID-19 cases across all age groups is evaluated using Shannon’s transfer entropy, given by:
+The influence of the evolution in COVID-19 cases across all age groups is evaluated using Shannon's transfer entropy, given by:
 
@@
 
-\noindent where $T_{ag\,x \rightarrow\,ag\,y}$ consequently measures the influence of the change dynamic from an age group $X$ (or $ag\,x$) to another age group $Y$ (or $ag\,y$) for every country (Eq 2).
+\noindent where \(T_{ag\,x \rightarrow\,ag\,y}\) consequently measures the influence of the change dynamic from an age group \(X\) (or \(ag\,x\)) to another age group \(Y\) (or \(ag\,y\)) for every country (Eq 2).
 
-The day-by-day difference in COVID-19 confirmed cases $n\prime\prime_t$ is used to satisfy the stationary requirement for the calculation of Shannon’s Transfer Entropy [@shannon1948mathematical; @behrendt2019rtransferentropy].
+The day-by-day difference in COVID-19 confirmed cases \(n\prime\prime_t\) is used to satisfy the stationary requirement for the calculation of Shannon's Transfer Entropy \citep{shannon1948mathematical, behrendt2019rtransferentropy}.
 
-# Results
+\section{Results}\label{results}
 
-The data show that the trend of confirmed COVID-19 cases follows similar patterns across the selected European countries, with scales following the size of the population in these countries (Figure \@ref(fig:overall)).
+The data show that the trend of confirmed COVID-19 cases follows similar patterns across the selected European countries, with scales following the size of the population in these countries (Figure \ref{fig:overall}).
 Thus, the bigger the country, the higher the total number of COVID-19 cases.
 
-```{r overall, fig.height=3, out.width="\\textwidth", fig.cap="(ref:overall)"}
-overall_plot <- df_global |> 
-  mutate(
-    country = fct_rev(country)
-  ) |> 
-  ggplot(aes(date, cumulative_cases_cal, group = country)) +
-  geom_line(na.rm = TRUE, color = "gray40", alpha = 0.4) +
-  geom_text_repel(
-    data = df_global |>
-      group_by(country) |>
-      slice_max(cumulative_cases_cal, with_ties = FALSE),
-    aes(label = country),
-    hjust = 0,
-    direction = "y",
-    segment.color = NA,
-    seed = 123,
-    size = 2,
-    force = 0.5,
-    family = "serif"
-  ) +
-  scale_x_date(
-    "Date", 
-    date_breaks = "3 months", 
-    date_labels = "%b %y", 
-    limits = c(as.Date("2020-02-01"), as.Date("2022-05-01"))
-  ) +
-  scale_y_continuous(
-    "Total COVID-19 Cases (Million)", 
-    labels = unit_format(unit = "M", scale = 1e-6)
-  ) +
-  theme_bw() +
-  theme(
-    legend.position = "none",
-    text = element_text(size = 10, family = "serif")
-  )
-
-overall_plot
-```
+\begin{figure}
+\includegraphics[width=\textwidth]{manuscript_files/figure-latex/overall-1} \caption{Cumulative COVID-19 cases number for selected European countries since the beginning of the pandemic.
+Source: COVerAGE-DB \citep{riffe2021data}.}\label{fig:overall}
+\end{figure}
 
 The evolution of COVID-19 cases reveals some similarities across all age groups.
-However, the influence of each wave on individual age groups also has some particularities (Figure \@ref(fig:descriptive)).
+However, the influence of each wave on individual age groups also has some particularities (Figure \ref{fig:descriptive}).
 For example, the first wave was more important among the oldest age groups, whereas the third wave was more important among the youngest age groups.
 
-```{r descriptive, fig.height=3, out.width="\\textwidth", fig.cap="(ref:descriptive)"}
-descriptive_plot <- df_global |> 
-  filter(schools_closed == 1) |> 
-  mutate(
-    status = status |> 
-      str_remove("Closed due to ") |> 
-      fct_rev(),
-    country = country |> 
-      fct_rev()
-  ) |> 
-  ggplot(aes(date, country, color = status)) +
-  geom_point(na.rm = TRUE) +
-  scale_x_date("Date", date_breaks = "3 months", date_labels = "%b %y") +
-  scale_y_discrete("Selected European Countries") +
-  scale_colour_manual("Schools Closure Reason", values = c("gray10", "gray60")) +
-  theme_bw() +
-  theme(
-    legend.position = "bottom",
-    text = element_text(size = 10, family = "serif")
-  )
-
-descriptive_plot
-```
+\begin{figure}
+\includegraphics[width=\textwidth]{manuscript_files/figure-latex/descriptive-1} \caption{Periods of school closure since the beginning of the COVID-19 pandemic for selected European countries and their reason: regular academic break versus closure due to government decisions.
+Source: \citet{unesco2022https}.}\label{fig:descriptive}
+\end{figure}
 
 As outlined above, to evaluate the shape of the trend in the numbers of COVID-19 cases reported after the three school closures longer than 21 consecutive days, a GAM was fitted as described above, taking into account the overall effect across all the selected European countries, as well as the effect for age groups: 0 to 4, 5 to 9, 10 to 14, 15 to 19, and 20 to 24 years old.
-The obtained results satisfy the requirements to fit this model, which explains `r percent(gam_summary$dev.expl, accuracy = 0.1)` of the variation in COVID-19 cases.
+The obtained results satisfy the requirements to fit this model, which explains 77.3\% of the variation in COVID-19 cases.
 
-Overall, the results revealed a decreasing non-linear effect of school closure at a country level for the selected European countries (Austria: `r gam_results["Austria"]`; Belgium: `r gam_results["Belgium"]`; Bulgaria: `r gam_results["Bulgaria"]`; Croatia: `r gam_results["Croatia"]`; Estonia: `r gam_results["Estonia"]`; France: `r gam_results["France"]`; Germany (`r gam_results["Germany"]`); Greece: `r gam_results["Greece"]`; The Netherlands: `r gam_results["Netherlands"]`; Portugal: `r gam_results["Portugal"]`; Slovakia (`r gam_results["Slovakia"]`; Spain: `r gam_results["Spain"]`, see Figure \@ref(fig:age)).
+Overall, the results revealed a decreasing non-linear effect of school closure at a country level for the selected European countries (Austria: \(\chi^2(5.55) = 3302.08\), \(p < 0.001\); Belgium: \(\chi^2(5.99) = 7680.99\), \(p < 0.001\); Bulgaria: \(\chi^2(5.55) = 125.19\), \(p < 0.001\); Croatia: \(\chi^2(5.6) = 734.51\), \(p < 0.001\); Estonia: \(\chi^2(5.88) = 2157.39\), \(p < 0.001\); France: \(\chi^2(4.98) = 4141.45\), \(p < 0.001\); Germany (\(\chi^2(5.97) = 3058.7\), \(p < 0.001\)); Greece: \(\chi^2(5.9) = 1793.54\), \(p < 0.001\); The Netherlands: \(\chi^2(5.99) = 7108.09\), \(p < 0.001\); Portugal: \(\chi^2(5.81) = 3104.43\), \(p < 0.001\); Slovakia (\(\chi^2(5.97) = 6218.66\), \(p < 0.001\); Spain: \(\chi^2(5.95) = 559.14\), \(p < 0.001\), see Figure \ref{fig:age}).
 
-```{r country, fig.height=3, out.width="\\textwidth", fig.cap="(ref:country)"}
-gam_plot_country
-```
+\begin{figure}
+\includegraphics[width=\textwidth]{manuscript_files/figure-latex/country-1} \caption{Standardized effect of the smooth term in Generalized Additive Model by country.
+Standardized effects are reported to compare the shape of the curve between countries.}\label{fig:country}
+\end{figure}
 
-```{r age, fig.height=3, out.width="\\textwidth", fig.cap="(ref:age)"}
-gam_plot_age
-```
+\begin{figure}
+\includegraphics[width=\textwidth]{manuscript_files/figure-latex/age-1} \caption{Standardized effect of the smooth term in Generalized Additive Model by age group.
+Standardized effects are reported to compare the shape of the curve between age groups.}\label{fig:age}
+\end{figure}
 
-```{r te, fig.height=8, out.width="\\textwidth", fig.cap="(ref:te)", fig.pos = 'H'}
-te_plot
-```
+\begin{figure}[H]
+\includegraphics[width=\textwidth]{manuscript_files/figure-latex/te-1} \caption{Matrix of Transfer Entropy coefficients according to every age group combination for each of the selected European country.
+Age groups on the x-axis are influencing the age groups on the y-axis (\(ag\,x \rightarrow\,ag\,y\)).
+The significance of each Transfer Entropy coefficient is provided in Appendix 2.}\label{fig:te}
+\end{figure}
 
-Considering all countries, the analysis of age groups reveals distinct patterns (Figure \@ref(fig:te)).
-For the 0 to 4 age group, i.e., the pre-school group, there was a downward trend during the initial three weeks following school closure, followed by an increase during the fourth week (`r gam_results["0to4"]`).
-A similar, but more moderate, trend is observer for the age group ranging from 5 to 9 (`r gam_results["5to9"]`), where a decrease is observed during the first two weeks, followed by an increase.
-while the trend for the age group ranging from 10 to 14 is flat during the first two weeks, a similar increase is shown at the beginning of the third week of school closure (`r gam_results["10to14"]`).
-For age groups between 15 and 24, there is a notable surge immediately after school closure, followed by a decrease after 14 days (15 to 19 age group: `r gam_results["15to19"]` and 20 to 24 age group: `r gam_results["20to24"]`).
+Considering all countries, the analysis of age groups reveals distinct patterns (Figure \ref{fig:te}).
+For the 0 to 4 age group, i.e., the pre-school group, there was a downward trend during the initial three weeks following school closure, followed by an increase during the fourth week (\(\chi^2(5.94) = 590.81\), \(p < 0.001\)).
+A similar, but more moderate, trend is observer for the age group ranging from 5 to 9 (\(\chi^2(5.96) = 1186.05\), \(p < 0.001\)), where a decrease is observed during the first two weeks, followed by an increase.
+while the trend for the age group ranging from 10 to 14 is flat during the first two weeks, a similar increase is shown at the beginning of the third week of school closure (\(\chi^2(5.97) = 307.56\), \(p < 0.001\)).
+For age groups between 15 and 24, there is a notable surge immediately after school closure, followed by a decrease after 14 days (15 to 19 age group: \(\chi^2(1.01) = 57.27\), \(p < 0.001\) and 20 to 24 age group: \(\chi^2(4.95) = 5624.5\), \(p < 0.001\)).
 Thus, while the total case number declined following school closures, among younger age groups only the pre-school age group had a consistent reduction in cases, while for other age groups, there was even an increase in cases following school closures.
@@
 
-The results of the transfer entropy calculations between age groups for each of the 12 selected European countries are reported in Figure \@ref(fig:te).
-The absence of symmetry between influencing age groups (i.e., $ag\,x$) and influenced age groups (i.e., $ag\,y$) is found.
+The results of the transfer entropy calculations between age groups for each of the 12 selected European countries are reported in Figure \ref{fig:te}.
+The absence of symmetry between influencing age groups (i.e., \(ag\,x\)) and influenced age groups (i.e., \(ag\,y\)) is found.
 Indeed, the change in COVID-19 cases in some age groups is influenced by other age groups, but they are not reciprocally influencing these age groups.
-Figure \@ref(fig:te) shows how different age cohorts influence the COVID-19 case numbers of all age groups across countries.
+Figure \ref{fig:te} shows how different age cohorts influence the COVID-19 case numbers of all age groups across countries.
 The upper left quadrant indicates how younger age cohorts are influencing older age cohorts, the bottom right quadrant indicates how older age cohorts are influencing younger age cohorts, finally, the lower left and upper right indicate how younger or older age cohorts are influencing themselves.
 By analyzing these quadrants, it is possible to identify similar patterns across multiple countries.
-Indeed, it appears that Austria, Germany, and The Netherlands have significantly higher $T$ coefficients in the upper left quadrant of the matrix, which indicates that the daily changes in COVID-19 cases number in younger cohorts are predicting the daily changes in COVID-19 cases number in older cohorts.
-Alternatively, it appears that Austria, the Netherlands, Portugal, and Spain have significantly higher $T$ coefficients in the lower right quadrant of the matrix, which indicates that the daily changes in COVID-19 cases number in older cohorts are predicting the daily changes in COVID-19 cases number in younger cohorts.
+Indeed, it appears that Austria, Germany, and The Netherlands have significantly higher \(T\) coefficients in the upper left quadrant of the matrix, which indicates that the daily changes in COVID-19 cases number in younger cohorts are predicting the daily changes in COVID-19 cases number in older cohorts.
+Alternatively, it appears that Austria, the Netherlands, Portugal, and Spain have significantly higher \(T\) coefficients in the lower right quadrant of the matrix, which indicates that the daily changes in COVID-19 cases number in older cohorts are predicting the daily changes in COVID-19 cases number in younger cohorts.
 
-# Discussion
+\section{Discussion}\label{discussion}
 
 Knowledge about the transmission of the virus significantly improved over time as more studies have been published.
-While an early study found that children did not play an important role in the transmission of the virus [@li2020role], more recent results give a more nuanced position, stating that the spread of the virus in children is moderate.
-It is now suggested that the impact of school openings on infection rates was limited, with children and teachers more often contracting infections through household or community contacts rather than within schools [@soriano2023policies].
+While an early study found that children did not play an important role in the transmission of the virus \citep{li2020role}, more recent results give a more nuanced position, stating that the spread of the virus in children is moderate.
+It is now suggested that the impact of school openings on infection rates was limited, with children and teachers more often contracting infections through household or community contacts rather than within schools \citep{soriano2023policies}.
 This study, using robust statistical methods, considered the effect of school closures on COVID-19 cases across age groups.
@@
 It is also possible that transmission within households intensified as children spent more time at home.
-These results partially replicate observations from @alfano2022effects while providing a clearer picture of the structure of the effect of school closure.
+These results partially replicate observations from \citet{alfano2022effects} while providing a clearer picture of the structure of the effect of school closure.
 
-The Transfer Entropy calculations between age groups for each of the 12 selected European countries reveal an absence of symmetry between influencing age groups ($X$) and influenced age groups ($Y$).
+The Transfer Entropy calculations between age groups for each of the 12 selected European countries reveal an absence of symmetry between influencing age groups (\(X\)) and influenced age groups (\(Y\)).
 Some age groups influence the COVID-19 case numbers of other age groups, but there is no reciprocal influence.
@@
 In conclusion, this study sheds light on the trends and patterns of confirmed COVID-19 cases in selected European countries, with a particular focus on the influence of age groups and school closures.
-While a reduction in overall COVID-19 cases was found during school closures, which might suggest that like lockdowns [@alfano2020efficacy; @molefi2021impact], prolonged school closures could contribute to mitigating the spread of the virus, the results for the school-going age groups suggest that this is not the case for all age groups.
+While a reduction in overall COVID-19 cases was found during school closures, which might suggest that like lockdowns \citep{alfano2020efficacy, molefi2021impact}, prolonged school closures could contribute to mitigating the spread of the virus, the results for the school-going age groups suggest that this is not the case for all age groups.
 
-# Author statements
+\section{Author statements}\label{author-statements}
 
-## Ethical approval
+\subsection{Ethical approval}\label{ethical-approval}
 
@@
 
-## Funding
+\subsection{Funding}\label{funding}
 
@@
 
-## Competing interests
+\subsection{Competing interests}\label{competing-interests}
 
@@
 
-## Data Availability
+\subsection{Data Availability}\label{data-availability}
 
-The data can be accessed from the COVerAGE-DB OSF repository (https://osf.io/mpwjq/) and from the UNESCO servers (https://en.unesco.org/file/unesco-data-school-closures-february-2020-june-2022csv-zip).
-All preprocessing, analyses and code used to build the submitted paper are available at https://github.com/damien-dupre/covid_ts_causality.
+The data can be accessed from the COVerAGE-DB OSF repository (\url{https://osf.io/mpwjq/}) and from the UNESCO servers (\url{https://en.unesco.org/file/unesco-data-school-closures-february-2020-june-2022csv-zip}).
+All preprocessing, analyses and code used to build the submitted paper are available at \url{https://github.com/damien-dupre/covid_ts_causality}.
 
-# References {-}
+\bibliographystyle{tfcad}
+\bibliography{bibliography.bib}
+
+\end{document}

### 2025-06-02 manuscript_tf/manuscript_blind.Rmd
update

--- 2024-12-20 manuscript_tf/manuscript.Rmd
@@
 title: |
-  Examining the Impact of School Closures on COVID-19 Infections in Europe and their Effects on Different Age Cohorts
+  Divergent Impacts of COVID-19 School Closures on Youth Infection Dynamics: Evidence from 12 European Countries
 author:
@@
       Dublin City University Business School, Glasnevin, Dublin 9, Ireland
-thanks:
-  author
+# thanks:
+#   author
 
@@
 abstract: |
-  This paper analyzes the trends of COVID-19 cases in different age groups in selected European countries, assessing to what extent school closures helped reduce case numbers, how school closures affected case numbers in different age groups, and how COVID-19 spread between age groups.
-The study also evaluates the effects of school closures on COVID-19 cases using a Generalized Additive Model (GAM).
-The findings demonstrate a decreasing non-linear effect of school closure on the total COVID-19 case number across all countries studied.
-However, the analysis of age groups only reveals a consistent downward trend in infections in the 0 to 4 pre-school age group, while the school-going age groups 5 to 14 exhibit a significant increase in cases.
-Age groups 15 to 24 show a surge immediately after closure, followed by a decline.
-Transfer Entropy calculations highlight asymmetry in age group influences, indicating that changes in COVID-19 cases in certain age groups predict changes in other age groups but not vice versa.
-These findings contribute to a better understanding of COVID-19 dynamics in European countries and provide insights for public health strategies and interventions.
+  The effectiveness of school closures as a COVID-19 control measure remains a subject of debate.
+This paper critically reassesses their impact by examining infection trends across various age groups in 12 European countries.
+We apply a dual analytical approach: Generalised Additive Models (GAMs) are used to capture complex, non-linear effects of closures on age-specific case numbers, while Transfer Entropy (TE) is employed to quantify directional transmission patterns between these age groups.
+Our findings reveal deviations from commonly held assumptions.
+While school closures were associated with a non-linear decline in overall national COVID-19 cases, the effects on children and young adults varied considerably.
+A consistent downward trend in infections was seen only in the pre-school age group.
+In contrast, school-aged children experienced a marked rise in COVID-19 cases following an initial period of stability or slight decrease after closures began.
+The Transfer Entropy analysis also identified asymmetric, directional patterns in transmission, revealing that infection trends in specific age groups could predict subsequent changes in others.
+These results challenge the assumption that school closures uniformly benefit younger populations and emphasise the need for age-specific analysis in pandemic planning and support the development of more nuanced, evidence-based public health policies.
+
 keywords: |
@@
 # write_csv(df_global_age, here("data/df_global_age.csv"))
+#
+# df_global <- read_csv(here("data/df_global.csv"))
+# df_global_age <- read_csv(here("data/df_global_age.csv"))
 
-df_global <- read_csv(here("data/df_global.csv"))
-df_global_age <- read_csv(here("data/df_global_age.csv"))
 ```
@@
 
-To evaluate the impact of school closure on COVID-19 cases across different age groups, we first performed a non-linear time series regression to analyze the relationship between school closures and the number of COVID-19 cases in various age groups.
-In a second analysis, the transfer entropy method is employed to assess the influence of changes in case numbers between age groups, aiming to determine if school closures have secondary effects within households.
+A gap in these previous analyses is the insufficient granularity in examining age-specific effects and the complex dynamics of inter-age group transmission.
+This study aims to address these deficiencies by providing a more robust and nuanced assessment.
+First, Generalized Additive Models (GAMs) are utilized to flexibly model and quantify the potentially non-linear evolution of COVID-19 cases following school closures, specifically within these fine-grained age brackets.
+This allows us to move beyond simple pre-post comparisons and capture dynamic temporal patterns.
+Then, Transfer Entropy (TE), an information-theoretic measure, is applied to assess directional causality in time series data.
+TE is particularly adept at identifying how past states of one variable (e.g., cases in one age group) influence future states of another (e.g., cases in a different age group), even in complex, non-linear systems like epidemic spread.
+This dual methodology allows for a deeper understanding of both the direct age-specific consequences of school closures and the potential for altered intergenerational transmission dynamics.
+Our findings challenge simplistic narratives about school closure efficacy and highlight the importance of age-stratified analysis for future pandemic preparedness.
 
@@
 
+## Controlling for confounding Non-Pharmaceutical Interventions
+
+Recognizing that school closures rarely occurred in isolation, a critical component of this analysis was to account for other contemporaneous NPIs.
+To mitigate confounding from other NPIs, we incorporated four time-varying composite NPI index identified by the Oxford COVID-19 Government Response Tracker (OxCGRT) project [@hale2021global]: the average stringency index, the average government response index, the average containment health index, and the average economic support index.
+Specific data were collected from each country on various policy responses governments implemented to curb the spread of COVID-19, such as travel restrictions, workplace closures, public event cancellations, restrictions on gatherings, or public health messaging.
+These indicators were then aggregated to create indexes, reflecting the overall level of government restrictions.
+
 ## Generalized additive model
@@
   &\quad + f_1(closure_t) + f_2(closure_t, country_t) \\
-  &\quad + f_3(closure_t, age\,group_t) + \epsilon_t
+  &\quad + f_3(closure_t, age\,group_t) + f_4(stringency\,index_t) \\
+  &\quad + f_5(government\,response\,index_t) + f_6(containment\,health\,index_t) \\
+  &\quad + f_7(economic\,support\,index_t) +\epsilon_t
 \end{align*}
@@
 The terms $f_1$ to $f_3$ are smooth functions of the time since closure, the time since closure for each country, and the time since closure for each age group.
+The terms $f_4$ to $f_7$ are random effects corresponding to the average stringency index, the average government response index, the average containment health index, and the average economic support index.
 The restricted maximum likelihood (REML) was used to avoid over fitting while estimating smoothing parameters.
@@
 
+While the general patterns described above were evident in the aggregated analysis, country-specific smooths indicated some national variations in the timing and magnitude of these age-specific effects, although the paradoxical increase in the 5-14 age cohorts was observed in a majority of countries.
+
 ```{r te, fig.height=8, out.width="\\textwidth", fig.cap="(ref:te)", fig.pos = 'H'}
@@
 
-The interpretation of our findings requires careful consideration of possible confounding factors.
-Changes in testing strategies, adherence to public health measures, and community transmission dynamics during school closures may have influenced the observed trends in COVID-19 cases.
-Our sensitivity analyses indicated that the overall trends remained robust after accounting for variations in testing rates and the presence of other public health interventions.
-However, these confounders may partially explain the heterogeneous effects observed across different age groups and countries.
-Future studies should explore these factors in greater detail to refine our understanding of the true impact of school closures.
+This study, despite its methodological strengths, is subject to certain limitations that warrant consideration.
+Foremost among these is the inherent challenge of perfectly isolating the effect of school closures from other simultaneously implemented and dynamically changing NPIs.
+While we incorporated four composite NPI index derived from the Oxford COVID-19 Government Response Tracker (OxCGRT) project to account for the broader public health context, residual confounding from unmeasured aspects of NPIs or behavioural changes may still persist.
+Our sensitivity analyses showed consistent directional effects for school closures even under varying overall NPI indexes provide some confidence, but this remains an inherent complexity in observational studies of this nature.
+Another consideration is the ecological design of the study, which uses aggregated country-level data; this may mask important sub-national variations and precludes inferences about individual-level risk factors.
+In addition, the definition of "closure" which excluded "partially open" school statuses to simplify the intervention definition, means the impact of such intermediate states was not captured.
+Variations in testing strategies, case definitions, and reporting practices across countries and over time could also have influenced the observed case counts, despite efforts to use standardized data sources.
+Lastly, this study predominantly covers periods before widespread COVID-19 vaccination and the emergence of later variants such as Omicron; the effects of school closures might differ under those altered epidemiological conditions.
 
-In conclusion, this study sheds light on the trends and patterns of confirmed COVID-19 cases in selected European countries, with a particular focus on the influence of age groups and school closures.
-While a reduction in overall COVID-19 cases was found during school closures, which might suggest that like lockdowns [@alfano2020efficacy; @molefi2021impact], prolonged school closures could contribute to mitigating the spread of the virus, the results for the school-going age groups suggest that this is not the case for all age groups.
+Future studies should look more closely at why the virus spread between age groups as it did, and why TE patterns differed by country.
+This could involve using information on people's behaviour or more detailed data on other health rules.
+It would also be useful to study how different virus types and vaccination levels change these effects.
+Learning more about other influencing factors will help us better understand the real impact of school closures.
+
+# Conclusion
+
+This comprehensive analysis of COVID-19 school closures across 12 European nations reveals a complex and often counter-intuitive reality that challenges prevailing assumptions about their universal efficacy, particularly for children aged 5 and older.
+While national-level data showed an overall decrease in cases during closures, our age-disaggregated analysis using GAMs demonstrated that school-aged children (5-14 years) and young adults (15-24 years) frequently experienced periods of increased infections post-closure, a stark contrast to the consistent reductions seen only in pre-schoolers.
+Furthermore, Transfer Entropy analysis highlighted that the directional influence of transmission between age groups was asymmetric and varied by country, suggesting that the role of children in broader community transmission during such periods is not uniform.
+
+These findings underscore the critical importance of nuanced, age-specific analysis in evaluating public health interventions and caution against the broad application of measures like school closures without robust evidence of their benefit across all targeted demographics and a clear understanding of potential unintended consequences [@alfano2020efficacy; @molefi2021impact].
+For future pandemic preparedness, a more targeted, evidence-based approach is essential, prioritising measures that minimise societal disruption while effectively protecting public health across all age strata.
 
@@
 
-The data can be accessed from the COVerAGE-DB OSF repository (https://osf.io/mpwjq/) and from the UNESCO servers (https://en.unesco.org/file/unesco-data-school-closures-february-2020-june-2022csv-zip).
+The data can be accessed from the COVerAGE-DB OSF repository (https://osf.io/mpwjq/), from the Oxford Covid-19 Government Response Tracker (OxCGRT) "covid-policy-dataset" GitHub repository (https://github.com/OxCGRT/covid-policy-dataset), and from the UNESCO servers (https://en.unesco.org/file/unesco-data-school-closures-february-2020-june-2022csv-zip).
 All preprocessing, analyses and code used to build the submitted paper are available at https://github.com/damien-dupre/covid_ts_causality.

### 2025-06-02 manuscript_tf/manuscript_blind.tex
update

--- 2024-12-20 manuscript_tf/manuscript.tex
@@
 
-\title{Examining the Impact of School Closures on COVID-19 Infections in Europe and their Effects on Different Age Cohorts}
+\title{Divergent Impacts of COVID-19 School Closures on Youth Infection Dynamics: Evidence from 12 European Countries}
 
@@
 
-\thanks{CONTACT Damien Dupré.
-Email: \href{mailto:damien.dupre@dcu.ie}{\nolinkurl{damien.dupre@dcu.ie}} (corresponding author), Edgar Morgenroth.
-Email: \href{mailto:edgar.morgenroth@dcu.ie}{\nolinkurl{edgar.morgenroth@dcu.ie}}}
-
 \maketitle
@@
 \begin{abstract}
-This paper analyzes the trends of COVID-19 cases in different age groups in selected European countries, assessing to what extent school closures helped reduce case numbers, how school closures affected case numbers in different age groups, and how COVID-19 spread between age groups.
-The study also evaluates the effects of school closures on COVID-19 cases using a Generalized Additive Model (GAM).
-The findings demonstrate a decreasing non-linear effect of school closure on the total COVID-19 case number across all countries studied.
-However, the analysis of age groups only reveals a consistent downward trend in infections in the 0 to 4 pre-school age group, while the school-going age groups 5 to 14 exhibit a significant increase in cases.
-Age groups 15 to 24 show a surge immediately after closure, followed by a decline.
-Transfer Entropy calculations highlight asymmetry in age group influences, indicating that changes in COVID-19 cases in certain age groups predict changes in other age groups but not vice versa.
-These findings contribute to a better understanding of COVID-19 dynamics in European countries and provide insights for public health strategies and interventions.
+The effectiveness of school closures as a COVID-19 control measure remains a subject of debate.
+This paper critically reassesses their impact by examining infection trends across various age groups in 12 European countries.
+We apply a dual analytical approach: Generalised Additive Models (GAMs) are used to capture complex, non-linear effects of closures on age-specific case numbers, while Transfer Entropy (TE) is employed to quantify directional transmission patterns between these age groups.
+Our findings reveal deviations from commonly held assumptions.
+While school closures were associated with a non-linear decline in overall national COVID-19 cases, the effects on children and young adults varied considerably.
+A consistent downward trend in infections was seen only in the pre-school age group.
+In contrast, school-aged children experienced a marked rise in COVID-19 cases following an initial period of stability or slight decrease after closures began.
+The Transfer Entropy analysis also identified asymmetric, directional patterns in transmission, revealing that infection trends in specific age groups could predict subsequent changes in others.
+These results challenge the assumption that school closures uniformly benefit younger populations and emphasise the need for age-specific analysis in pandemic planning and support the development of more nuanced, evidence-based public health policies.
 \end{abstract}
@@
 \end{keywords}
-
-\begin{jelcode}
-I18; C22.
-\end{jelcode}
 
@@
 
+\subsection{Controlling for confounding Non-Pharmaceutical Interventions}\label{controlling-for-confounding-non-pharmaceutical-interventions}
+
+Recognizing that school closures rarely occurred in isolation, a critical component of this analysis was to account for other contemporaneous NPIs.
+To mitigate confounding from other NPIs, we incorporated four time-varying composite NPI index identified by the Oxford COVID-19 Government Response Tracker (OxCGRT) project \citep{hale2021global}: the average stringency index, the average government response index, the average containment health index, and the average economic support index.
+Specific data were collected from each country on various policy responses governments implemented to curb the spread of COVID-19, such as travel restrictions, workplace closures, public event cancellations, restrictions on gatherings, or public health messaging.
+These indicators were then aggregated to create indexes, reflecting the overall level of government restrictions.
+
 \subsection{Generalized additive model}\label{generalized-additive-model}
@@
 The terms \(f_1\) to \(f_3\) are smooth functions of the time since closure, the time since closure for each country, and the time since closure for each age group.
+The terms \(f_4\) to \(f_7\) are random effects corresponding to the average stringency index, the average government response index, the average containment health index, and the average economic support index.
 The restricted maximum likelihood (REML) was used to avoid over fitting while estimating smoothing parameters.
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{manuscript_files/figure-latex/overall-1} \caption{Cumulative COVID-19 cases number for selected European countries since the beginning of the pandemic.
+\includegraphics[width=\textwidth]{manuscript_blind_files/figure-latex/overall-1} \caption{Cumulative COVID-19 cases number for selected European countries since the beginning of the pandemic.
 Source: COVerAGE-DB \citep{riffe2021data}.}\label{fig:overall}
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{manuscript_files/figure-latex/descriptive-1} \caption{Periods of school closure since the beginning of the COVID-19 pandemic for selected European countries and their reason: regular academic break versus closure due to government decisions.
+\includegraphics[width=\textwidth]{manuscript_blind_files/figure-latex/descriptive-1} \caption{Periods of school closure since the beginning of the COVID-19 pandemic for selected European countries and their reason: regular academic break versus closure due to government decisions.
 Source: \citet{unesco2022https}.}\label{fig:descriptive}
@@
 As outlined above, to evaluate the shape of the trend in the numbers of COVID-19 cases reported after the three school closures longer than 21 consecutive days, a GAM was fitted as described above, taking into account the overall effect across all the selected European countries, as well as the effect for age groups: 0 to 4, 5 to 9, 10 to 14, 15 to 19, and 20 to 24 years old.
-The obtained results satisfy the requirements to fit this model, which explains 77.3\% of the variation in COVID-19 cases.
+The obtained results satisfy the requirements to fit this model, which explains 80.2\% of the variation in COVID-19 cases.
 
-Overall, the results revealed a decreasing non-linear effect of school closure at a country level for the selected European countries (Austria: \(\chi^2(5.55) = 3302.08\), \(p < 0.001\); Belgium: \(\chi^2(5.99) = 7680.99\), \(p < 0.001\); Bulgaria: \(\chi^2(5.55) = 125.19\), \(p < 0.001\); Croatia: \(\chi^2(5.6) = 734.51\), \(p < 0.001\); Estonia: \(\chi^2(5.88) = 2157.39\), \(p < 0.001\); France: \(\chi^2(4.98) = 4141.45\), \(p < 0.001\); Germany (\(\chi^2(5.97) = 3058.7\), \(p < 0.001\)); Greece: \(\chi^2(5.9) = 1793.54\), \(p < 0.001\); The Netherlands: \(\chi^2(5.99) = 7108.09\), \(p < 0.001\); Portugal: \(\chi^2(5.81) = 3104.43\), \(p < 0.001\); Slovakia (\(\chi^2(5.97) = 6218.66\), \(p < 0.001\); Spain: \(\chi^2(5.95) = 559.14\), \(p < 0.001\), see Figure \ref{fig:age}).
+Overall, the results revealed a decreasing non-linear effect of school closure at a country level for the selected European countries (Austria: \(\chi^2(5.82) = 3243.01\), \(p < 0.001\); Belgium: \(\chi^2(5.98) = 6206.45\), \(p < 0.001\); Bulgaria: \(\chi^2(5.89) = 258.19\), \(p < 0.001\); Croatia: \(\chi^2(5.1) = 740.24\), \(p < 0.001\); Estonia: \(\chi^2(5.93) = 1834.26\), \(p < 0.001\); France: \(\chi^2(1) = 2412.59\), \(p < 0.001\); Germany (\(\chi^2(5.98) = 3081.05\), \(p < 0.001\)); Greece: \(\chi^2(5.95) = 2423.67\), \(p < 0.001\); The Netherlands: \(\chi^2(4.97) = 5878.92\), \(p < 0.001\); Portugal: \(\chi^2(5.94) = 2692.56\), \(p < 0.001\); Slovakia (\(\chi^2(5.97) = 5335.74\), \(p < 0.001\); Spain: \(\chi^2(5.94) = 498.09\), \(p < 0.001\), see Figure \ref{fig:age}).
 
 \begin{figure}
-\includegraphics[width=\textwidth]{manuscript_files/figure-latex/country-1} \caption{Standardized effect of the smooth term in Generalized Additive Model by country.
+\includegraphics[width=\textwidth]{manuscript_blind_files/figure-latex/country-1} \caption{Standardized effect of the smooth term in Generalized Additive Model by country.
 Standardized effects are reported to compare the shape of the curve between countries.}\label{fig:country}
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{manuscript_files/figure-latex/age-1} \caption{Standardized effect of the smooth term in Generalized Additive Model by age group.
+\includegraphics[width=\textwidth]{manuscript_blind_files/figure-latex/age-1} \caption{Standardized effect of the smooth term in Generalized Additive Model by age group.
 Standardized effects are reported to compare the shape of the curve between age groups.}\label{fig:age}
@@
 
+While the general patterns described above were evident in the aggregated analysis, country-specific smooths indicated some national variations in the timing and magnitude of these age-specific effects, although the paradoxical increase in the 5-14 age cohorts was observed in a majority of countries.
+
 \begin{figure}[H]
-\includegraphics[width=\textwidth]{manuscript_files/figure-latex/te-1} \caption{Matrix of Transfer Entropy coefficients according to every age group combination for each of the selected European country.
+\includegraphics[width=\textwidth]{manuscript_blind_files/figure-latex/te-1} \caption{Matrix of Transfer Entropy coefficients according to every age group combination for each of the selected European country.
 Age groups on the x-axis are influencing the age groups on the y-axis (\(ag\,x \rightarrow\,ag\,y\)).
@@
 Considering all countries, the analysis of age groups reveals distinct patterns (Figure \ref{fig:te}).
-For the 0 to 4 age group, i.e., the pre-school group, there was a downward trend during the initial three weeks following school closure, followed by an increase during the fourth week (\(\chi^2(5.94) = 590.81\), \(p < 0.001\)).
-A similar, but more moderate, trend is observer for the age group ranging from 5 to 9 (\(\chi^2(5.96) = 1186.05\), \(p < 0.001\)), where a decrease is observed during the first two weeks, followed by an increase.
-while the trend for the age group ranging from 10 to 14 is flat during the first two weeks, a similar increase is shown at the beginning of the third week of school closure (\(\chi^2(5.97) = 307.56\), \(p < 0.001\)).
-For age groups between 15 and 24, there is a notable surge immediately after school closure, followed by a decrease after 14 days (15 to 19 age group: \(\chi^2(1.01) = 57.27\), \(p < 0.001\) and 20 to 24 age group: \(\chi^2(4.95) = 5624.5\), \(p < 0.001\)).
+For the 0 to 4 age group, i.e., the pre-school group, there was a downward trend during the initial three weeks following school closure, followed by an increase during the fourth week (\(\chi^2(5.96) = 410.48\), \(p < 0.001\)).
+A similar, but more moderate, trend is observer for the age group ranging from 5 to 9 (\(\chi^2(5.95) = 878.29\), \(p < 0.001\)), where a decrease is observed during the first two weeks, followed by an increase.
+while the trend for the age group ranging from 10 to 14 is flat during the first two weeks, a similar increase is shown at the beginning of the third week of school closure (\(\chi^2(1) = 61.62\), \(p < 0.001\)).
+For age groups between 15 and 24, there is a notable surge immediately after school closure, followed by a decrease after 14 days (15 to 19 age group: \(\chi^2(5.97) = 7929.88\), \(p < 0.001\) and 20 to 24 age group: \(\chi^2(4.98) = 18386.65\), \(p < 0.001\)).
 Thus, while the total case number declined following school closures, among younger age groups only the pre-school age group had a consistent reduction in cases, while for other age groups, there was even an increase in cases following school closures.
@@
 
-The interpretation of our findings requires careful consideration of possible confounding factors.
-Changes in testing strategies, adherence to public health measures, and community transmission dynamics during school closures may have influenced the observed trends in COVID-19 cases.
-Our sensitivity analyses indicated that the overall trends remained robust after accounting for variations in testing rates and the presence of other public health interventions.
-However, these confounders may partially explain the heterogeneous effects observed across different age groups and countries.
-Future studies should explore these factors in greater detail to refine our understanding of the true impact of school closures.
+This study, despite its methodological strengths, is subject to certain limitations that warrant consideration.
+Foremost among these is the inherent challenge of perfectly isolating the effect of school closures from other simultaneously implemented and dynamically changing NPIs.
+While we incorporated four composite NPI index derived from the Oxford COVID-19 Government Response Tracker (OxCGRT) project to account for the broader public health context, residual confounding from unmeasured aspects of NPIs or behavioural changes may still persist.
+Our sensitivity analyses showed consistent directional effects for school closures even under varying overall NPI indexes provide some confidence, but this remains an inherent complexity in observational studies of this nature.
+Another consideration is the ecological design of the study, which uses aggregated country-level data; this may mask important sub-national variations and precludes inferences about individual-level risk factors.
+In addition, the definition of ``closure'' which excluded ``partially open'' school statuses to simplify the intervention definition, means the impact of such intermediate states was not captured.
+Variations in testing strategies, case definitions, and reporting practices across countries and over time could also have influenced the observed case counts, despite efforts to use standardized data sources.
+Lastly, this study predominantly covers periods before widespread COVID-19 vaccination and the emergence of later variants such as Omicron; the effects of school closures might differ under those altered epidemiological conditions.
 
-In conclusion, this study sheds light on the trends and patterns of confirmed COVID-19 cases in selected European countries, with a particular focus on the influence of age groups and school closures.
-While a reduction in overall COVID-19 cases was found during school closures, which might suggest that like lockdowns \citep{alfano2020efficacy, molefi2021impact}, prolonged school closures could contribute to mitigating the spread of the virus, the results for the school-going age groups suggest that this is not the case for all age groups.
+Future studies should look more closely at why the virus spread between age groups as it did, and why TE patterns differed by country.
+This could involve using information on people's behaviour or more detailed data on other health rules.
+It would also be useful to study how different virus types and vaccination levels change these effects.
+Learning more about other influencing factors will help us better understand the real impact of school closures.
+
+\section{Conclusion}\label{conclusion}
+
+This comprehensive analysis of COVID-19 school closures across 12 European nations reveals a complex and often counter-intuitive reality that challenges prevailing assumptions about their universal efficacy, particularly for children aged 5 and older.
+While national-level data showed an overall decrease in cases during closures, our age-disaggregated analysis using GAMs demonstrated that school-aged children (5-14 years) and young adults (15-24 years) frequently experienced periods of increased infections post-closure, a stark contrast to the consistent reductions seen only in pre-schoolers.
+Furthermore, Transfer Entropy analysis highlighted that the directional influence of transmission between age groups was asymmetric and varied by country, suggesting that the role of children in broader community transmission during such periods is not uniform.
+
+These findings underscore the critical importance of nuanced, age-specific analysis in evaluating public health interventions and caution against the broad application of measures like school closures without robust evidence of their benefit across all targeted demographics and a clear understanding of potential unintended consequences \citep{alfano2020efficacy, molefi2021impact}.
+For future pandemic preparedness, a more targeted, evidence-based approach is essential, prioritising measures that minimise societal disruption while effectively protecting public health across all age strata.
 
@@
 
-The data can be accessed from the COVerAGE-DB OSF repository (\url{https://osf.io/mpwjq/}) and from the UNESCO servers (\url{https://en.unesco.org/file/unesco-data-school-closures-february-2020-june-2022csv-zip}).
+The data can be accessed from the COVerAGE-DB OSF repository (\url{https://osf.io/mpwjq/}), from the Oxford Covid-19 Government Response Tracker (OxCGRT) ``covid-policy-dataset'' GitHub repository (\url{https://github.com/OxCGRT/covid-policy-dataset}), and from the UNESCO servers (\url{https://en.unesco.org/file/unesco-data-school-closures-february-2020-june-2022csv-zip}).
 All preprocessing, analyses and code used to build the submitted paper are available at \url{https://github.com/damien-dupre/covid_ts_causality}.

### 2025-06-02 analysis/gam_analysis_europe.R

@@ -4,6 +4,7 @@ library(future)
 library(glue)
 library(gratia)
 library(here)
+library(janitor)
 library(mgcv)
 library(parallel)
 library(scales)
@@ -14,6 +15,27 @@ options(scipen = 999)
 plan("future::multisession")
 
 # data -------------------------------------------------------------------------
+countries_of_interest <- c(
+  "Austria", "Belgium", "Bulgaria", "Croatia", "Estonia", "France",
+  "Germany", "Greece", "Netherlands", "Portugal", "Slovakia", "Spain"
+)
+
+OxCGRT_compact_national <- read_csv("https://raw.githubusercontent.com/OxCGRT/covid-policy-dataset/refs/heads/main/data/OxCGRT_compact_national_v1.csv") |> 
+  clean_names() |> 
+  mutate(
+    country = if_else(country_name == "Slovak Republic", "Slovakia", country_name),
+    date = as.Date(as.character(date), format = "%Y%m%d")
+  ) |> 
+  filter(country %in% countries_of_interest) |> 
+  select(
+    country, 
+    date, 
+    stringency_index_average, 
+    government_response_index_average, 
+    containment_health_index_average, 
+    economic_support_index
+  )
+
 df_global <- read_csv(here("data/df_global.csv"))
 df_global_age <- read_csv(here("data/df_global_age.csv"))
 
@@ -72,10 +94,11 @@ df_gam <- df_global_school |>
   ) |>
   anti_join(negative_daily_case_wave, by = c("country", "schools_closure_wave")) |>
   anti_join(missing_daily_case_wave, by = c("country", "schools_closure_wave")) |>
-  filter(who_region == "EURO") |>
+  filter(who_region == "EURO") |> 
+  left_join(OxCGRT_compact_national, by = join_by(country, date)) |>
   mutate(
-    age_grp = factor(age_grp),
-    country = factor(country)
+    age_grp = as.factor(age_grp),
+    country = as.factor(country)
   )
 
 system.time(
@@ -87,7 +110,11 @@ system.time(
         s(country, bs="re") +
         s(schools_time_since_closure, by = age_grp, bs="tp", k=7) +
         s(schools_time_since_closure, by = country, bs="tp", k=7) +
-        s(schools_time_since_closure, schools_closure_wave, bs="re"),
+        s(schools_time_since_closure, schools_closure_wave, bs="re") +
+        s(stringency_index_average, bs = "re") +
[…]

### 2025-06-03 manuscript_tf/figures.Rmd
update

@@
 
-(ref:overall) Cumulative COVID-19 cases number for selected European countries since the beginning of the pandemic. Source: COVerAGE-DB [@riffe2021data].
+(ref:overall) Cumulative COVID-19 cases number for selected European countries since the beginning of the pandemic. Source: COVerAGE-DB (Riffe, Acosta, and COVerAGE-DB team 2021).
 (ref:descriptive) Periods of school closure since the beginning of the COVID-19 pandemic for selected European countries and their reason: regular academic break versus closure due to government decisions.
-Source: @unesco2022https.
+Source: United Nations Educational, Scientific and
+Cultural Organization (UNESCO) (2022).
 (ref:country) Standardized effect of the smooth term in Generalized Additive Model by country. Standardized effects are reported to compare the shape of the curve between countries.
@@
 
-df_global <- read_csv(here("data/df_global.csv"))
-df_global_age <- read_csv(here("data/df_global_age.csv"))
+# df_global <- read_csv(here("data/df_global.csv"))
+# df_global_age <- read_csv(here("data/df_global_age.csv"))
 ```

### 2025-06-03 manuscript_tf/manuscript_blind.Rmd
update

@@
   Divergent Impacts of COVID-19 School Closures on Youth Infection Dynamics: Evidence from 12 European Countries
+# author:
+#   - name: Damien Dupré
+#     affil: 1
+#     email: damien.dupre@dcu.ie (corresponding author)
+#   - name: Edgar Morgenroth
+#     affil: 1
+#     email: edgar.morgenroth@dcu.ie
+# affiliation:
+#   - num: 1
+#     address: |
+#       Dublin City University Business School, Glasnevin, Dublin 9, Ireland
+# thanks:
+#   author
+
 author:
-  - name: Damien Dupré
+  - name: Author 1
     affil: 1
-    email: damien.dupre@dcu.ie (corresponding author)
-  - name: Edgar Morgenroth
-    affil: 1
-    email: edgar.morgenroth@dcu.ie
+    email: Email 1
+  - name: Author 2
+    affil: 2
+    email: Email 2
 affiliation:
@@
     address: |
-      Dublin City University Business School, Glasnevin, Dublin 9, Ireland
-# thanks:
-#   author
-
-# author:
-#   - name: Author 1
-#     affil: 1
-#     email: Email 1
-#   - name: Author 2
-#     affil: 2
-#     email: Email 2
-# affiliation:
-#   - num: 1
-#     address: |
-#       Affiliation 1
-#   - num: 2
-#     address: |
-#       Affiliation 2
+      Affiliation 1
+  - num: 2
+    address: |
+      Affiliation 2
 

### 2025-06-05 manuscript_njd/manuscript.tex
update + njd

--- 2025-06-02 manuscript_tf/manuscript_blind.tex
@@
-% interactcadsample.tex
-% v1.04 - May 2023
+\documentclass[Harvard,Times1COL]{WileyNJDv5}
 
-\documentclass[]{interact}
+\usepackage{natbib}
+\bibliographystyle{wileyNJD-Harvard}
 
-\usepackage{epstopdf}% To incorporate .eps illustrations using PDFLaTeX, etc.
-\usepackage{subfigure}% Support for small, `sub' figures and tables
-%\usepackage[nolists,tablesfirst]{endfloat}% To `separate' figures and tables from text if required
-
-\usepackage{natbib}% Citation support using natbib.sty
-\bibpunct[, ]{(}{)}{;}{a}{}{,}% Citation support using natbib.sty
-\renewcommand\bibfont{\fontsize{10}{12}\selectfont}% Bibliography support using natbib.sty
-
-\theoremstyle{plain}% Theorem-like structures provided by amsthm.sty
-\newtheorem{theorem}{Theorem}[section]
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{corollary}[theorem]{Corollary}
-\newtheorem{proposition}[theorem]{Proposition}
-
-\theoremstyle{definition}
-\newtheorem{definition}[theorem]{Definition}
-\newtheorem{example}[theorem]{Example}
-
-\theoremstyle{remark}
-\newtheorem{remark}{Remark}
-\newtheorem{notation}{Notation}
-
-% tightlist command for lists without linebreak
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-
-% From pandoc table feature
-\usepackage{longtable,booktabs,array}
-\usepackage{calc} % for calculating minipage widths
-% Correct order of tables after \paragraph or \subparagraph
-\usepackage{etoolbox}
-\makeatletter
-\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
-\makeatother
-% Allow footnotes in longtable head/foot
-\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
-\makesavenoteenv{longtable}
-
-\usepackage{float}
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{threeparttable}
-\floatplacement{figure}{H}
-\floatplacement{table}{H}
-\usepackage{caption}
-\usepackage{hyperref}
-\usepackage[utf8]{inputenc}
-\def\tightlist{}
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{wrapfig}
-\usepackage{float}
-\usepackage{colortbl}
-\usepackage{pdflscape}
-\usepackage{tabu}
-\usepackage{threeparttable}
-\usepackage{threeparttablex}
-\usepackage[normalem]{ulem}
-\usepackage{makecell}
-\usepackage{xcolor}
-
+\hypersetup{
+  colorlinks=true,
+  linkcolor=blue,
+  citecolor=blue,
+  filecolor=black,
+  urlcolor=black
+}
 \begin{document}
 
-\articletype{}
+\articletype{Article Type: Research Article}
 
-\title{Divergent Impacts of COVID-19 School Closures on Youth Infection Dynamics: Evidence from 12 European Countries}
+\received{Date Month Year}
+\revised{Date Month Year}
+\accepted{Date Month Year}
+\journal{Journal of Policy Analysis and Management}
+\volume{00}
+\copyyear{2025}
+\startpage{1}
 
-\author{\name{Damien Dupré$^{1}$, Edgar Morgenroth$^{1}$}
-\affil{$^{1}$Dublin City University Business School, Glasnevin, Dublin 9, Ireland}
+% TODO: Inherited from WileyNJD example. Is this needed?
+% \raggedbottom
+
+\authormark{Dupré and Morgenroth}
+
+\presentaddress{DCU Business School, Dublin 9, Ireland.}
+
+% Include information for each author
+  \author[1]{Damien Dupré}
+  \author[1]{Edgar Morgenroth}
+
+% Include information for each affiliation
+\address[1]{%
+  \orgdiv{%
+    Business School%
+    , %
+    \orgname{%
+      Dublin City University%
+      , %
+    }%
+    \orgaddress{%
+      \state{Dublin}%
+      , %
+      \country{Ireland}%
+    }%
+  }%
 }
 
+% Include information for each corresponder
+      \corres{%
+      %
+        Damien Dupré%
+      %
+      \email{damien.dupre@dcu.ie}%
+    }
+
+\title{Divergent Impacts of COVID-19 School Closures on Youth Infection
+Dynamics: Evidence from 12 European Countries}
+\titlemark{COVID-19 School Closures and Youth Infection Dynamics}
+
+\abstract[Abstract]{The effectiveness of school closures as a COVID-19
+control measure remains a subject of debate. This paper critically
+reassesses their impact by examining infection trends across various age
+groups in 12 European countries. We apply a dual analytical approach:
+Generalised Additive Models (GAMs) are used to capture complex,
+non-linear effects of closures on age-specific case numbers, while
+Transfer Entropy (TE) is employed to quantify directional transmission
+patterns between these age groups. Our findings reveal deviations from
+commonly held assumptions. While school closures were associated with a
+non-linear decline in overall national COVID-19 cases, the effects on
+children and young adults varied considerably. A consistent downward
+trend in infections was seen only in the pre-school age group. In
+contrast, school-aged children experienced a marked rise in COVID-19
+cases following an initial period of stability or slight decrease after
+closures began. The Transfer Entropy analysis also identified
+asymmetric, directional patterns in transmission, revealing that
+infection trends in specific age groups could predict subsequent changes
+in others. These results challenge the assumption that school closures
+uniformly benefit younger populations and emphasise the need for
+age-specific analysis in pandemic planning and support the development
+of more nuanced, evidence-based public health policies.}
+
+{\keywords{COVID-19, School closures, Intergenerational
+transmission, Non-linear effects, Infection dynamics.}
+
 \maketitle
-
-\begin{abstract}
-The effectiveness of school closures as a COVID-19 control measure remains a subject of debate.
-This paper critically reassesses their impact by examining infection trends across various age groups in 12 European countries.
-We apply a dual analytical approach: Generalised Additive Models (GAMs) are used to capture complex, non-linear effects of closures on age-specific case numbers, while Transfer Entropy (TE) is employed to quantify directional transmission patterns between these age groups.
-Our findings reveal deviations from commonly held assumptions.
-While school closures were associated with a non-linear decline in overall national COVID-19 cases, the effects on children and young adults varied considerably.
-A consistent downward trend in infections was seen only in the pre-school age group.
-In contrast, school-aged children experienced a marked rise in COVID-19 cases following an initial period of stability or slight decrease after closures began.
-The Transfer Entropy analysis also identified asymmetric, directional patterns in transmission, revealing that infection trends in specific age groups could predict subsequent changes in others.
-These results challenge the assumption that school closures uniformly benefit younger populations and emphasise the need for age-specific analysis in pandemic planning and support the development of more nuanced, evidence-based public health policies.
-\end{abstract}
-
-\begin{keywords}
-COVID-19; School Closures; Intergenerational Transmission; Non-linear effects; Infection dynamics.
-\end{keywords}
-
 \section{Introduction}\label{introduction}
 
-The COVID-19 pandemic has had a profound impact on global health, with an estimate of 14.83 million excess deaths globally \citep{msemburi2023estimates}.
-Beyond direct mortality, the pandemic has caused significant collateral damage, including losses of lives and livelihoods, necessitating a comprehensive approach to measuring its broader impacts.
-A wide range of non-pharmaceutical interventions (NPI) were enacted to control the spread of COVID-19, especially before the availability of effective vaccines.
-Due to their social mixing patterns, children have been identified as an age group that can drive the spread of respiratory infections such as influenza \citep{moser2018estimating}, and school closures were found to be effective in mitigating the spread of influenza H1N1 in Japan in 2009 \citep{kawano2015substantial}.
-It is therefore not surprising that school closures were one of the NPIs that were introduced in many countries to stop the spread of COVID-19.
-They were used particularly during the initial wave of the pandemic and during subsequent waves when case numbers were high.
+The COVID-19 pandemic has had a profound impact on global health, with
+an estimate of 14.83 million excess deaths globally
+\citep{msemburi2023estimates}. Beyond direct mortality, the pandemic has
+caused significant collateral damage, including losses of lives and
+livelihoods, necessitating a comprehensive approach to measuring its
+broader impacts. A wide range of non-pharmaceutical interventions (NPI)
+were enacted to control the spread of COVID-19, especially before the
+availability of effective vaccines. Due to their social mixing patterns,
+children have been identified as an age group that can drive the spread
+of respiratory infections such as influenza \citep{moser2018estimating},
+and school closures were found to be effective in mitigating the spread
+of influenza H1N1 in Japan in 2009 \citep{kawano2015substantial}. It is
+therefore not surprising that school closures were one of the NPIs that
+were introduced in many countries to stop the spread of COVID-19. They
+were used particularly during the initial wave of the pandemic and
+during subsequent waves when case numbers were high.
 
-Recently, a body of literature has shown various negative consequences on child development and health due to school closures.
-School closures have been found to have negatively affected child mental health \citep{moulin2022longitudinal}, nutrition/obesity \citep{sugimoto2023temporal}, and education \citep{lerkkanen2023reading}.
-Some governments were questioning their decisions to close schools \citep{de2021determines}.
-For example, the German health minister Karl Lauterbach, in an interview with one of Germany's publicly funded television stations, admitted that ``in retrospect it had been wrong to keep schools and childcare closed for so long'' \citep{ard2023lauterbach}.
-Even if school closures had negative consequences for children, they might nevertheless have been effective at stopping the spread of COVID-19.
-However, analysis on the effectiveness of school closures on the spread of COVID-19 remains inconclusive.
-In a study of a panel of European countries, \citet{alfano2022effects} found that school closures were associated with a reduced COVID-19 incidence.
-In contrast, \citet{walsh2021school} in a review of 40 studies covering 150 countries concluded that the effectiveness of school closures was uncertain, with 60\% having identified no impact and pointing to the potential for the analysis to be affected by confounding factors and collinearity.
-The latter shortcomings suggest that it is hard to draw firm inferences, as even a positive result may not imply causality.
+Recently, a body of literature has shown various negative consequences
+on child development and health due to school closures. School closures
+have been found to have negatively affected child mental health
+\citep{moulin2022longitudinal}, nutrition/obesity
+\citep{sugimoto2023temporal}, and education
+\citep{lerkkanen2023reading}. Some governments were questioning their
+decisions to close schools \citep{de2021determines}. For example, the
+German health minister Karl Lauterbach, in an interview with one of
+Germany's publicly funded television stations, admitted that ``in
+retrospect it had been wrong to keep schools and childcare closed for so
+long'' \citep{ard2023lauterbach}. Even if school closures had negative
+consequences for children, they might nevertheless have been effective
+at stopping the spread of COVID-19. However, analysis on the
+effectiveness of school closures on the spread of COVID-19 remains
+inconclusive. In a study of a panel of European countries,
+\citet{alfano2022effects} found that school closures were associated
+with a reduced COVID-19 incidence. In contrast, \citet{walsh2021school}
+in a review of 40 studies covering 150 countries concluded that the
+effectiveness of school closures was uncertain, with 60\% having
+identified no impact and pointing to the potential for the analysis to
+be affected by confounding factors and collinearity. The latter
+shortcomings suggest that it is hard to draw firm inferences, as even a
+positive result may not imply causality.
 
-Once it was clear that COVID-19 spreads from person to person, it was natural that governments were advised to enact measures to reduce social contact in order to reduce the spread of the virus.
-Therefore, it is legitimate to believe that by closing schools, a reduction of the contaminations would be observed in the younger age groups.
-However, the efficiency of school closure on the reduction of COVID-19 cases is still questioned \citep{bayham2020impact, esposito2021comprehensive}.
-While some research has found that school closures contribute to limit or to reduce the growth rate of confirmed cases after implementation \citep{stage2021shut, yoshiyuki2020effects}, others did not observe a change in the evolution of COVID-19 cases \citep{chang2020modelling, iwata2020school}.
-For instance, a controlled comparison between similar localities in Japan with schools closed and schools open did not reveal any evidence that school closures reduced the spread of COVID-19 \citep{fukumoto2021no}.
-If the school closure had a real impact on the evolution of confirmed COVID-19 cases, it should be possible to observe a decrease or at least an inflection in the trend of its evolution among younger age groups.
+Once it was clear that COVID-19 spreads from person to person, it was
+natural that governments were advised to enact measures to reduce social
+contact in order to reduce the spread of the virus. Therefore, it is
+legitimate to believe that by closing schools, a reduction of the
+contaminations would be observed in the younger age groups. However, the
+efficiency of school closure on the reduction of COVID-19 cases is still
+questioned \citep{bayham2020impact, esposito2021comprehensive}. While
+some research has found that school closures contribute to limit or to
+reduce the growth rate of confirmed cases after implementation
+\citep{stage2021shut, yoshiyuki2020effects}, others did not observe a
+change in the evolution of COVID-19 cases
+\citep{chang2020modelling, iwata2020school}. For instance, a controlled
+comparison between similar localities in Japan with schools closed and
+schools open did not reveal any evidence that school closures reduced
+the spread of COVID-19 \citep{fukumoto2021no}. If the school closure had
+a real impact on the evolution of confirmed COVID-19 cases, it should be
+possible to observe a decrease or at least an inflection in the trend of
+its evolution among younger age groups.
 
-A second implicit belief regarding the effect of school closure on the spread of COVID-19 is that school not only influences the spread of the virus among children and teenagers but also has a knock-on effect on the spread of the virus in older age groups, also called Secondary Attack Rate (SAR).
-The contaminated children and teenagers would bring the virus back home and would pass the virus on to their parents and relatives.
-For example, research investigating the contamination in the household network not only revealed an exceptionally high rate of secondary contamination but also that this contamination happened when the schools were closed \citep{sorianoarandes2021household}.
-Despite being reported in several clinical and epidemiological studies \citep{siebach2021childhood, zhendong2020clinical}, multiple research have shown that the SAR from children to household members was lower than expected \citep{heavey2020evidence, vanderhoek2020role, kim2021role, ludvigsson2020children}.
-However, the SAR of children and teenagers to the household member is likely to be age-dependent, with differences between infants, primary and secondary school children, and college students \citep{grasleguen2021reopening}.
-If a secondary transmission from children and teenagers to household member has a significant influence, then a temporal causality relationship between their evolution should be observed.
+A second implicit belief regarding the effect of school closure on the
+spread of COVID-19 is that school not only influences the spread of the
+virus among children and teenagers but also has a knock-on effect on the
+spread of the virus in older age groups, also called Secondary Attack
+Rate (SAR). The contaminated children and teenagers would bring the
+virus back home and would pass the virus on to their parents and
+relatives. For example, research investigating the contamination in the
+household network not only revealed an exceptionally high rate of
+secondary contamination but also that this contamination happened when
+the schools were closed \citep{sorianoarandes2021household}. Despite
+being reported in several clinical and epidemiological studies
+\citep{siebach2021childhood, zhendong2020clinical}, multiple research
+have shown that the SAR from children to household members was lower
+than expected
+\citep{heavey2020evidence, vanderhoek2020role, kim2021role, ludvigsson2020children}.
+However, the SAR of children and teenagers to the household member is
+likely to be age-dependent, with differences between infants, primary
+and secondary school children, and college students
+\citep{grasleguen2021reopening}. If a secondary transmission from
+children and teenagers to household member has a significant influence,
+then a temporal causality relationship between their evolution should be
+observed.
 
-A gap in these previous analyses is the insufficient granularity in examining age-specific effects and the complex dynamics of inter-age group transmission.
-This study aims to address these deficiencies by providing a more robust and nuanced assessment.
-First, Generalized Additive Models (GAMs) are utilized to flexibly model and quantify the potentially non-linear evolution of COVID-19 cases following school closures, specifically within these fine-grained age brackets.
-This allows us to move beyond simple pre-post comparisons and capture dynamic temporal patterns.
-Then, Transfer Entropy (TE), an information-theoretic measure, is applied to assess directional causality in time series data.
-TE is particularly adept at identifying how past states of one variable (e.g., cases in one age group) influence future states of another (e.g., cases in a different age group), even in complex, non-linear systems like epidemic spread.
-This dual methodology allows for a deeper understanding of both the direct age-specific consequences of school closures and the potential for altered intergenerational transmission dynamics.
-Our findings challenge simplistic narratives about school closure efficacy and highlight the importance of age-stratified analysis for future pandemic preparedness.
+A gap in these previous analyses is the insufficient granularity in
+examining age-specific effects and the complex dynamics of inter-age
+group transmission. This study aims to address these deficiencies by
+providing a more robust and nuanced assessment. First, Generalized
+Additive Models (GAMs) are utilized to flexibly model and quantify the
+potentially non-linear evolution of COVID-19 cases following school
+closures, specifically within these fine-grained age brackets. This
+allows us to move beyond simple pre-post comparisons and capture dynamic
+temporal patterns. Then, Transfer Entropy (TE), an information-theoretic
+measure, is applied to assess directional causality in time series data.
+TE is particularly adept at identifying how past states of one variable
+(e.g., cases in one age group) influence future states of another (e.g.,
+cases in a different age group), even in complex, non-linear systems
+like epidemic spread. This dual methodology allows for a deeper
+understanding of both the direct age-specific consequences of school
+closures and the potential for altered intergenerational transmission
+dynamics. Our findings challenge simplistic narratives about school
+closure efficacy and highlight the importance of age-stratified analysis
+for future pandemic preparedness.
 
@@
 
-A longitudinal study was conducted to observe changes in COVID-19 cases during school closures.
-Changes in COVID-19 from five age groups from the first to the 28th day of school closure are compared across 12 countries.
-Then, a causality analysis evaluate the impact of changes in these younger groups on older age groups.
+A longitudinal study was conducted to observe changes in COVID-19 cases
+during school closures. Changes in COVID-19 from five age groups from
+the first to the 28th day of school closure are compared across 12
+countries. Then, a causality analysis evaluate the impact of changes in
+these younger groups on older age groups.
 
@@
 
-Data on COVID-19 case counts by age group have been obtained from the COVerAGE-DB project \citep{riffe2021data}.
-The COVerAGE-DB which collect numbers for every countries by groups of 5 or 10 years if available.
-In addition, spline approximations are used to deals with the heterogeneity of countries' reporting formats by using when the data for this age bracket is not available for a country.
-Countries communicating data with groups of 5 years were chosen to match as much as possible the different school stages.
+Data on COVID-19 case counts by age group have been obtained from the
+COVerAGE-DB project \citep{riffe2021data}. The COVerAGE-DB which collect
+numbers for every countries by groups of 5 or 10 years if available. In
+addition, spline approximations are used to deals with the heterogeneity
+of countries' reporting formats by using when the data for this age
+bracket is not available for a country. Countries communicating data
+with groups of 5 years were chosen to match as much as possible the
+different school stages.
 
-To identify the impact of school closures on the number of cases, information regarding the school closures on a day-by-day basis is required, and this is taken from the ``UNESCO global education coalition'' (2022).
-For each day, in each country, the status of the schools is indicated as fully open, partially open, closed due to COVID-19, or closed due to an academic break.
-Because it would be difficult to measure the effect of school closures at a country level when schools are partially closed, only closures due to COVID-19 or due to an academic break are considered.
-Indeed, both are considered as closure at a country-wide level.
-Any COVID-19 case numbers beyond the 28-day period are not relevant for evaluating the impact of school closures.
+To identify the impact of school closures on the number of cases,
+information regarding the school closures on a day-by-day basis is
+required, and this is taken from the ``UNESCO global education
+coalition'' (2022). For each day, in each country, the status of the
+schools is indicated as fully open, partially open, closed due to
+COVID-19, or closed due to an academic break. Because it would be
+difficult to measure the effect of school closures at a country level
+when schools are partially closed, only closures due to COVID-19 or due
+to an academic break are considered. Indeed, both are considered as
+closure at a country-wide level. Any COVID-19 case numbers beyond the
+28-day period are not relevant for evaluating the impact of school
+closures.
 
@@
 
-In order to cross-validate the data obtained after spline approximations, a comparison with the data published by the \citet{who2022https} reveals perfect similarities.
-The original data consist of 14,089,320 observations of 10 variables (117 distinct countries, region within the country, a unique observation code, the date of the observation, the gender, which can be male, female, or both, the age bracket by 5 years from 0 to 100, a confirmation of the age interval for each bracket, the total number of cases so far, the total number of deaths, and the total number of tests performed) from February 16, 2020 to January 20, 2022.
-After removing countries with missing and inconsistent values, only 22 are suitable for data analyses.
-However, to focus this analysis on geographically and culturally comparable countries, only 12 European countries are kept: Austria, Belgium, Bulgaria, Croatia, Estonia, France, Germany, Greece, Netherlands, Portugal, Slovakia, and Spain.
-The observations are reported in terms of the total number of COVID-19 cases per day from the start of the pandemic.
-The daily number of cases at a specific date \(n\prime_{t}\) is calculated with the difference between the total cases at a date \(t\) and the total cases at a date \(t-1\) (i.e., derivative 1).
-In addition, the change in the daily number of cases \(n\prime\prime_{t}\) between \(n\prime_{t}\) and \(n\prime_{t-1}\) has also been calculated (i.e., derivative 2).
+In order to cross-validate the data obtained after spline
+approximations, a comparison with the data published by the
+\citet{who2022https} reveals perfect similarities. The original data
+consist of 14,089,320 observations of 10 variables (117 distinct
+countries, region within the country, a unique observation code, the
+date of the observation, the gender, which can be male, female, or both,
+the age bracket by 5 years from 0 to 100, a confirmation of the age
+interval for each bracket, the total number of cases so far, the total
+number of deaths, and the total number of tests performed) from February
+16, 2020 to January 20, 2022. After removing countries with missing and
+inconsistent values, only 22 are suitable for data analyses. However, to
+focus this analysis on geographically and culturally comparable
+countries, only 12 European countries are kept: Austria, Belgium,
+Bulgaria, Croatia, Estonia, France, Germany, Greece, Netherlands,
+Portugal, Slovakia, and Spain. The observations are reported in terms of
+the total number of COVID-19 cases per day from the start of the
+pandemic. The daily number of cases at a specific date \(n\prime_{t}\)
+is calculated with the difference between the total cases at a date
+\(t\) and the total cases at a date \(t-1\) (i.e., derivative 1). In
+addition, the change in the daily number of cases \(n\prime\prime_{t}\)
+between \(n\prime_{t}\) and \(n\prime_{t-1}\) has also been calculated
+(i.e., derivative 2).
 
-\subsection{Controlling for confounding Non-Pharmaceutical Interventions}\label{controlling-for-confounding-non-pharmaceutical-interventions}
+\subsection{Controlling for confounding Non-Pharmaceutical
+Interventions}\label{controlling-for-confounding-non-pharmaceutical-interventions}
 
-Recognizing that school closures rarely occurred in isolation, a critical component of this analysis was to account for other contemporaneous NPIs.
-To mitigate confounding from other NPIs, we incorporated four time-varying composite NPI index identified by the Oxford COVID-19 Government Response Tracker (OxCGRT) project \citep{hale2021global}: the average stringency index, the average government response index, the average containment health index, and the average economic support index.
-Specific data were collected from each country on various policy responses governments implemented to curb the spread of COVID-19, such as travel restrictions, workplace closures, public event cancellations, restrictions on gatherings, or public health messaging.
-These indicators were then aggregated to create indexes, reflecting the overall level of government restrictions.
+Recognizing that school closures rarely occurred in isolation, a
+critical component of this analysis was to account for other
+contemporaneous NPIs. To mitigate confounding from other NPIs, we
+incorporated four time-varying composite NPI index identified by the
+Oxford COVID-19 Government Response Tracker (OxCGRT) project
+\citep{hale2021global}: the average stringency index, the average
+government response index, the average containment health index, and the
+average economic support index. Specific data were collected from each
+country on various policy responses governments implemented to curb the
+spread of COVID-19, such as travel restrictions, workplace closures,
+public event cancellations, restrictions on gatherings, or public health
+messaging. These indicators were then aggregated to create indexes,
+reflecting the overall level of government restrictions.
 
-\subsection{Generalized additive model}\label{generalized-additive-model}
+\subsection{Generalized additive
+model}\label{generalized-additive-model}
 
-The effect of school closure on the trend of daily COVID-19 cases is analysed using a Generalized Additive Model (GAM).
-GAM is a flexible modelling approach that estimates non-linear relationships between variables.
-Compared to other methods such as Vector Autoregression, GAM can handle fixed and random smooth effects without making strict assumptions about linearity.
+The effect of school closure on the trend of daily COVID-19 cases is
+analysed using a Generalized Additive Model (GAM). GAM is a flexible
+modelling approach that estimates non-linear relationships between
+variables. Compared to other methods such as Vector Autoregression, GAM
+can handle fixed and random smooth effects without making strict
+assumptions about linearity.
 
-The GAM is fitted on the daily COVID-19 cases to test the hypothesis of a significant non-linear evolution of cases among age groups from 0 to 4, from 5 to 9, from 10 to 14, from 15 to 19, and from 20 to 24 \citep{wood2017generalized}.
-The model also estimates the overall non-linear effect by country, taking into account the interaction between age groups and countries as random intercepts and the interaction between time, countries, and the period of closure as random effects (Eq 1).
+The GAM is fitted on the daily COVID-19 cases to test the hypothesis of
+a significant non-linear evolution of cases among age groups from 0 to
+4, from 5 to 9, from 10 to 14, from 15 to 19, and from 20 to 24
+\citep{wood2017generalized}. The model also estimates the overall
+non-linear effect by country, taking into account the interaction
+between age groups and countries as random intercepts and the
+interaction between time, countries, and the period of closure as random
+effects (Eq \ref{eq-gam}).
 
-By estimating the degree of smoothness of a Bayesian spline smoothing using restricted fast maximum likelihood estimation \citep{wood2011fast}, GAM identifies dynamic patterns underlying the evolution of COVID-19 cases reported while including the random effect of different age groups and countries as follows:
+By estimating the degree of smoothness of a Bayesian spline smoothing
+using restricted fast maximum likelihood estimation
+\citep{wood2011fast}, GAM identifies dynamic patterns underlying the
+evolution of COVID-19 cases reported while including the random effect
+of different age groups and countries as follows:
 
-\begin{align*}
-  \tag{1}
+\begin{equation}\phantomsection\label{eq-gam}{
+\begin{aligned}[t]
   n\prime_t &\sim \text{Poisson}(\lambda_t) \\
@@
   &\quad + f_7(economic\,support\,index_t) +\epsilon_t
-\end{align*}
+\end{aligned}
+}\end{equation}
 
-\noindent where \(n\prime_t\) represents the confirmed COVID-19 cases, assuming a Poisson distribution for the fitting \citep{loader2006local}, and \(t\) is the date corresponding to the confirmed COVID-19 cases.
-The response variable includes a specific random effect taking into account variation within waves of school closure, countries, and age groups.
-The terms \(f_1\) to \(f_3\) are smooth functions of the time since closure, the time since closure for each country, and the time since closure for each age group.
-The terms \(f_4\) to \(f_7\) are random effects corresponding to the average stringency index, the average government response index, the average containment health index, and the average economic support index.
-The restricted maximum likelihood (REML) was used to avoid over fitting while estimating smoothing parameters.
-In order to accurately account for the autocorrelation arising from the time series data, the residuals are modelled using an AR1 error model such as \(\epsilon_t = \phi \epsilon_{t-1} + \eta_t, \quad \eta_t \sim \mathcal{N}(0, \sigma^2)\) where \(\phi\) is the autoregressive parameter.
-By incorporating the autoregressive component, the AR1 model acknowledges the dependence of each residual on its previous value, thus providing a comprehensive representation of the data's temporal dynamics.
+\noindent where \(n\prime_t\) represents the confirmed COVID-19 cases,
+assuming a Poisson distribution for the fitting \citep{loader2006local},
+and \(t\) is the date corresponding to the confirmed COVID-19 cases. The
+response variable includes a specific random effect taking into account
+variation within waves of school closure, countries, and age groups. The
+terms \(f_1\) to \(f_3\) are smooth functions of the time since closure,
+the time since closure for each country, and the time since closure for
+each age group. The terms \(f_4\) to \(f_7\) are random effects
+corresponding to the average stringency index, the average government
+response index, the average containment health index, and the average
+economic support index. The restricted maximum likelihood (REML) was
+used to avoid over fitting while estimating smoothing parameters. In
+order to accurately account for the autocorrelation arising from the
+time series data, the residuals are modelled using an AR1 error model
+such as
+\(\epsilon_t = \phi \epsilon_{t-1} + \eta_t, \quad \eta_t \sim \mathcal{N}(0, \sigma^2)\)
+where \(\phi\) is the autoregressive parameter. By incorporating the
+autoregressive component, the AR1 model acknowledges the dependence of
+each residual on its previous value, thus providing a comprehensive
+representation of the data's temporal dynamics.
 
-Chi-square statistics were employed to assess if the degree of smoothness is significantly different from zero.
+Chi-square statistics were employed to assess if the degree of
+smoothness is significantly different from zero.
 
@@
 
-Transfer entropy (\(T\)) is a measure of the directional information flow between two time series \(X\) and \(Y\), capturing how the past values of one variable can predict changes in another.
-Unlike correlation, transfer entropy accounts for the temporal order of events and non-linear relationships, making it particularly suited for studying dynamic systems such as epidemic data.
-In this analysis, transfer entropy quantifies how case counts in one age group \(X\) influence subsequent case counts in another age group \(Y\), providing insights into directional transmission patterns.
-If it does, \(T\) is considered evidence of a causal effect from the age group \(X\) to the age group \(Y\) \citep{schreiber2000measuring}.
-As such, Granger causality is a special case of transfer entropy applied to time series that are jointly Gaussian distributed \citep{barnett2009granger}.
-Therefore, transfer entropy is a more robust analysis of time series, especially when applied to the impact of age cohorts on pandemic transmission \citep{kissler2020symbolic}.
+Transfer entropy (\(T\)) is a measure of the directional information
+flow between two time series \(X\) and \(Y\), capturing how the past
+values of one variable can predict changes in another. Unlike
+correlation, transfer entropy accounts for the temporal order of events
+and non-linear relationships, making it particularly suited for studying
+dynamic systems such as epidemic data. In this analysis, transfer
+entropy quantifies how case counts in one age group \(X\) influence
+subsequent case counts in another age group \(Y\), providing insights
+into directional transmission patterns. If it does, \(T\) is considered
+evidence of a causal effect from the age group \(X\) to the age group
+\(Y\) \citep{schreiber2000measuring}. As such, Granger causality is a
+special case of transfer entropy applied to time series that are jointly
+Gaussian distributed \citep{barnett2009granger}. Therefore, transfer
+entropy is a more robust analysis of time series, especially when
+applied to the impact of age cohorts on pandemic transmission
+\citep{kissler2020symbolic}.
 
-The influence of the evolution in COVID-19 cases across all age groups is evaluated using Shannon's transfer entropy, given by:
+The influence of the evolution in COVID-19 cases across all age groups
+is evaluated using Shannon's transfer entropy, given by:
 
-\begin{align}
-  \tag{2}
+\begin{equation}\phantomsection\label{eq-te}{
+\begin{aligned}[t]
   T_{ag\,x \rightarrow ag\,y}(k,l) = \sum_{ag\,x_{t+1}, ag\,x_t^{(k)}, ag\,y_t^{(l)}} 
@@
   \log \left(\frac{p\left(ag\,x_{t+1} \mid ag\,x_t^{(k)}, ag\,y_t^{(l)}\right)}{p\left(ag\,x_{t+1} \mid ag\,x_t^{(k)}\right)}\right)
-\end{align}
+\end{aligned}
+}\end{equation}
 
-\noindent where \(T_{ag\,x \rightarrow\,ag\,y}\) consequently measures the influence of the change dynamic from an age group \(X\) (or \(ag\,x\)) to another age group \(Y\) (or \(ag\,y\)) for every country (Eq 2).
+\noindent where \(T_{ag\,x \rightarrow\,ag\,y}\) consequently measures
+the influence of the change dynamic from an age group \(X\) (or
+\(ag\,x\)) to another age group \(Y\) (or \(ag\,y\)) for every country
+(Eq \ref{eq-te}).
 
-The day-by-day difference in COVID-19 confirmed cases \(n\prime\prime_t\) is used to satisfy the stationary requirement for the calculation of Shannon's Transfer Entropy \citep{shannon1948mathematical, behrendt2019rtransferentropy}.
+The day-by-day difference in COVID-19 confirmed cases
+\(n\prime\prime_t\) is used to satisfy the stationary requirement for
+the calculation of Shannon's Transfer Entropy
+\citep{shannon1948mathematical, behrendt2019rtransferentropy}.
 
@@
 
-The data show that the trend of confirmed COVID-19 cases follows similar patterns across the selected European countries, with scales following the size of the population in these countries (Figure \ref{fig:overall}).
-Thus, the bigger the country, the higher the total number of COVID-19 cases.
+The data show that the trend of confirmed COVID-19 cases follows similar
+patterns across the selected European countries, with scales following
+the size of the population in these countries
+(Figure~\ref{fig-overall}). Thus, the bigger the country, the higher the
+total number of COVID-19 cases.
 
-\begin{figure}
-\includegraphics[width=\textwidth]{manuscript_blind_files/figure-latex/overall-1} \caption{Cumulative COVID-19 cases number for selected European countries since the beginning of the pandemic.
-Source: COVerAGE-DB \citep{riffe2021data}.}\label{fig:overall}
-\end{figure}
+\begin{figure}[h]
 
-The evolution of COVID-19 cases reveals some similarities across all age groups.
-However, the influence of each wave on individual age groups also has some particularities (Figure \ref{fig:descriptive}).
-For example, the first wave was more important among the oldest age groups, whereas the third wave was more important among the youngest age groups.
+\centering{
 
-\begin{figure}
-\includegraphics[width=\textwidth]{manuscript_blind_files/figure-latex/descriptive-1} \caption{Periods of school closure since the beginning of the COVID-19 pandemic for selected European countries and their reason: regular academic break versus closure due to government decisions.
-Source: \citet{unesco2022https}.}\label{fig:descriptive}
-\end{figure}
+\pandocbounded{\includegraphics[keepaspectratio]{manuscript_files/figure-pdf/fig-overall-1.png}}
 
-As outlined above, to evaluate the shape of the trend in the numbers of COVID-19 cases reported after the three school closures longer than 21 consecutive days, a GAM was fitted as described above, taking into account the overall effect across all the selected European countries, as well as the effect for age groups: 0 to 4, 5 to 9, 10 to 14, 15 to 19, and 20 to 24 years old.
-The obtained results satisfy the requirements to fit this model, which explains 80.2\% of the variation in COVID-19 cases.
+}
 
-Overall, the results revealed a decreasing non-linear effect of school closure at a country level for the selected European countries (Austria: \(\chi^2(5.82) = 3243.01\), \(p < 0.001\); Belgium: \(\chi^2(5.98) = 6206.45\), \(p < 0.001\); Bulgaria: \(\chi^2(5.89) = 258.19\), \(p < 0.001\); Croatia: \(\chi^2(5.1) = 740.24\), \(p < 0.001\); Estonia: \(\chi^2(5.93) = 1834.26\), \(p < 0.001\); France: \(\chi^2(1) = 2412.59\), \(p < 0.001\); Germany (\(\chi^2(5.98) = 3081.05\), \(p < 0.001\)); Greece: \(\chi^2(5.95) = 2423.67\), \(p < 0.001\); The Netherlands: \(\chi^2(4.97) = 5878.92\), \(p < 0.001\); Portugal: \(\chi^2(5.94) = 2692.56\), \(p < 0.001\); Slovakia (\(\chi^2(5.97) = 5335.74\), \(p < 0.001\); Spain: \(\chi^2(5.94) = 498.09\), \(p < 0.001\), see Figure \ref{fig:age}).
+\caption{\label{fig-overall}Cumulative COVID-19 cases number for
+selected European countries since the beginning of the pandemic. Source:
+COVerAGE-DB \citep{riffe2021data}.}
 
-\begin{figure}
-\includegraphics[width=\textwidth]{manuscript_blind_files/figure-latex/country-1} \caption{Standardized effect of the smooth term in Generalized Additive Model by country.
-Standardized effects are reported to compare the shape of the curve between countries.}\label{fig:country}
-\end{figure}
+\end{figure}%
 
-\begin{figure}
-\includegraphics[width=\textwidth]{manuscript_blind_files/figure-latex/age-1} \caption{Standardized effect of the smooth term in Generalized Additive Model by age group.
-Standardized effects are reported to compare the shape of the curve between age groups.}\label{fig:age}
-\end{figure}
+The evolution of COVID-19 cases reveals some similarities across all age
+groups. However, the influence of each wave on individual age groups
+also has some particularities (Figure~\ref{fig-descriptive}). For
+example, the first wave was more important among the oldest age groups,
+whereas the third wave was more important among the youngest age groups.
 
-While the general patterns described above were evident in the aggregated analysis, country-specific smooths indicated some national variations in the timing and magnitude of these age-specific effects, although the paradoxical increase in the 5-14 age cohorts was observed in a majority of countries.
+\begin{figure}[h]
 
-\begin{figure}[H]
-\includegraphics[width=\textwidth]{manuscript_blind_files/figure-latex/te-1} \caption{Matrix of Transfer Entropy coefficients according to every age group combination for each of the selected European country.
-Age groups on the x-axis are influencing the age groups on the y-axis (\(ag\,x \rightarrow\,ag\,y\)).
-The significance of each Transfer Entropy coefficient is provided in Appendix 2.}\label{fig:te}
-\end{figure}
+\centering{
 
-Considering all countries, the analysis of age groups reveals distinct patterns (Figure \ref{fig:te}).
-For the 0 to 4 age group, i.e., the pre-school group, there was a downward trend during the initial three weeks following school closure, followed by an increase during the fourth week (\(\chi^2(5.96) = 410.48\), \(p < 0.001\)).
-A similar, but more moderate, trend is observer for the age group ranging from 5 to 9 (\(\chi^2(5.95) = 878.29\), \(p < 0.001\)), where a decrease is observed during the first two weeks, followed by an increase.
-while the trend for the age group ranging from 10 to 14 is flat during the first two weeks, a similar increase is shown at the beginning of the third week of school closure (\(\chi^2(1) = 61.62\), \(p < 0.001\)).
-For age groups between 15 and 24, there is a notable surge immediately after school closure, followed by a decrease after 14 days (15 to 19 age group: \(\chi^2(5.97) = 7929.88\), \(p < 0.001\) and 20 to 24 age group: \(\chi^2(4.98) = 18386.65\), \(p < 0.001\)).
-Thus, while the total case number declined following school closures, among younger age groups only the pre-school age group had a consistent reduction in cases, while for other age groups, there was even an increase in cases following school closures.
+\pandocbounded{\includegraphics[keepaspectratio]{manuscript_files/figure-pdf/fig-descriptive-1.png}}
 
-Given that school closures were not only aimed at reducing COVID-19 cases among school-going children but also age groups, it is important to test the degree to which there was transmission across age groups, which is done using transfer entropy analysis.
-Before this analysis is carried out, an Augmented Dickey-Fuller has been applied to each age group to ensure that the daily changes in COVID-19 cases are stationary (see Appendix 1).
+}
 
-The results of the transfer entropy calculations between age groups for each of the 12 selected European countries are reported in Figure \ref{fig:te}.
-The absence of symmetry between influencing age groups (i.e., \(ag\,x\)) and influenced age groups (i.e., \(ag\,y\)) is found.
-Indeed, the change in COVID-19 cases in some age groups is influenced by other age groups, but they are not reciprocally influencing these age groups.
-Figure \ref{fig:te} shows how different age cohorts influence the COVID-19 case numbers of all age groups across countries.
-The upper left quadrant indicates how younger age cohorts are influencing older age cohorts, the bottom right quadrant indicates how older age cohorts are influencing younger age cohorts, finally, the lower left and upper right indicate how younger or older age cohorts are influencing themselves.
-By analyzing these quadrants, it is possible to identify similar patterns across multiple countries.
-Indeed, it appears that Austria, Germany, and The Netherlands have significantly higher \(T\) coefficients in the upper left quadrant of the matrix, which indicates that the daily changes in COVID-19 cases number in younger cohorts are predicting the daily changes in COVID-19 cases number in older cohorts.
-Alternatively, it appears that Austria, the Netherlands, Portugal, and Spain have significantly higher \(T\) coefficients in the lower right quadrant of the matrix, which indicates that the daily changes in COVID-19 cases number in older cohorts are predicting the daily changes in COVID-19 cases number in younger cohorts.
+\caption{\label{fig-descriptive}Periods of school closure since the
+beginning of the COVID-19 pandemic for selected European countries and
+their reason: regular academic break versus closure due to government
+decisions. Source: \citet{unesco2022https}.}
+
+\end{figure}%
+
+As outlined above, to evaluate the shape of the trend in the numbers of
+COVID-19 cases reported after the three school closures longer than 21
+consecutive days, a GAM was fitted as described above, taking into
+account the overall effect across all the selected European countries,
+as well as the effect for age groups: 0 to 4, 5 to 9, 10 to 14, 15 to
+19, and 20 to 24 years old. The obtained results satisfy the
+requirements to fit this model, which explains 80.2\% of the variation
+in COVID-19 cases (Figure~\ref{fig-age}).
+
+\begin{figure}[h]
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{manuscript_files/figure-pdf/fig-age-1.png}}
+
+}
+
+\caption{\label{fig-age}Standardized effect of the smooth term in
+Generalized Additive Model by age group. Standardized effects are
+reported to compare the shape of the curve between age groups.}
+
+\end{figure}%
+
+Overall, the results revealed a decreasing non-linear effect of school
+closure at a country level for the selected European countries (Austria:
+\(\chi^2(5.82) = 3243.01\), \(p < 0.001\); Belgium:
+\(\chi^2(5.98) = 6206.45\), \(p < 0.001\); Bulgaria:
+\(\chi^2(5.89) = 258.19\), \(p < 0.001\); Croatia:
+\(\chi^2(5.1) = 740.24\), \(p < 0.001\); Estonia:
+\(\chi^2(5.93) = 1834.26\), \(p < 0.001\); France:
+\(\chi^2(1) = 2412.59\), \(p < 0.001\); Germany
+(\(\chi^2(5.98) = 3081.05\), \(p < 0.001\)); Greece:
+\(\chi^2(5.95) = 2423.67\), \(p < 0.001\); The Netherlands:
+\(\chi^2(4.97) = 5878.92\), \(p < 0.001\); Portugal:
+\(\chi^2(5.94) = 2692.56\), \(p < 0.001\); Slovakia
+(\(\chi^2(5.97) = 5335.74\), \(p < 0.001\); Spain:
+\(\chi^2(5.94) = 498.09\), \(p < 0.001\), see Figure~\ref{fig-country}.
+
+\begin{figure}[h]
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{manuscript_files/figure-pdf/fig-country-1.png}}
+
+}
+
+\caption{\label{fig-country}Standardized effect of the smooth term in
+Generalized Additive Model by country. Standardized effects are reported
+to compare the shape of the curve between countries.}
+
+\end{figure}%
+
+While the general patterns described above were evident in the
+aggregated analysis, country-specific smooths indicated some national
+variations in the timing and magnitude of these age-specific effects,
+although the paradoxical increase in the 5-14 age cohorts was observed
+in a majority of countries.
+
+\begin{figure}[h]
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{manuscript_files/figure-pdf/fig-te-1.png}}
+
+}
+
+\caption{\label{fig-te}Matrix of Transfer Entropy coefficients according
+to every age group combination for each of the selected European
+country. Age groups on the x-axis are influencing the age groups on the
+y-axis (\(ag\,x \rightarrow\,ag\,y\)). The significance of each Transfer
+Entropy coefficient is provided in Appendix 2.}
+
+\end{figure}%
+
+Considering all countries, the analysis of age groups reveals distinct
+patterns (Figure~\ref{fig-te}). For the 0 to 4 age group, i.e., the
+pre-school group, there was a downward trend during the initial three
+weeks following school closure, followed by an increase during the
+fourth week (\(\chi^2(5.96) = 410.48\), \(p < 0.001\)). A similar, but
+more moderate, trend is observer for the age group ranging from 5 to 9
+(\(\chi^2(5.95) = 878.29\), \(p < 0.001\)), where a decrease is observed
+during the first two weeks, followed by an increase. while the trend for
+the age group ranging from 10 to 14 is flat during the first two weeks,
+a similar increase is shown at the beginning of the third week of school
+closure (\(\chi^2(1) = 61.62\), \(p < 0.001\)). For age groups between
+15 and 24, there is a notable surge immediately after school closure,
+followed by a decrease after 14 days (15 to 19 age group:
+\(\chi^2(5.97) = 7929.88\), \(p < 0.001\) and 20 to 24 age group:
+\(\chi^2(4.98) = 18386.65\), \(p < 0.001\)). Thus, while the total case
+number declined following school closures, among younger age groups only
+the pre-school age group had a consistent reduction in cases, while for
+other age groups, there was even an increase in cases following school
+closures.
+
+Given that school closures were not only aimed at reducing COVID-19
+cases among school-going children but also age groups, it is important
+to test the degree to which there was transmission across age groups,
+which is done using transfer entropy analysis. Before this analysis is
+carried out, an Augmented Dickey-Fuller has been applied to each age
+group to ensure that the daily changes in COVID-19 cases are stationary
+(see Appendix 1).
+
+The results of the transfer entropy calculations between age groups for
+each of the 12 selected European countries are reported in
+Figure~\ref{fig-te}. The absence of symmetry between influencing age
+groups (i.e., \(ag\,x\)) and influenced age groups (i.e., \(ag\,y\)) is
+found. Indeed, the change in COVID-19 cases in some age groups is
+influenced by other age groups, but they are not reciprocally
+influencing these age groups. F@fig-te shows how different age cohorts
+influence the COVID-19 case numbers of all age groups across countries.
+The upper left quadrant indicates how younger age cohorts are
+influencing older age cohorts, the bottom right quadrant indicates how
+older age cohorts are influencing younger age cohorts, finally, the
+lower left and upper right indicate how younger or older age cohorts are
+influencing themselves. By analyzing these quadrants, it is possible to
+identify similar patterns across multiple countries. Indeed, it appears
+that Austria, Germany, and The Netherlands have significantly higher
+\(T\) coefficients in the upper left quadrant of the matrix, which
+indicates that the daily changes in COVID-19 cases number in younger
+cohorts are predicting the daily changes in COVID-19 cases number in
+older cohorts. Alternatively, it appears that Austria, the Netherlands,
+Portugal, and Spain have significantly higher \(T\) coefficients in the
+lower right quadrant of the matrix, which indicates that the daily
+changes in COVID-19 cases number in older cohorts are predicting the
+daily changes in COVID-19 cases number in younger cohorts.
 
@@
 
-Knowledge about the transmission of the virus significantly improved over time as more studies have been published.
-While an early study found that children did not play an important role in the transmission of the virus \citep{li2020role}, more recent results give a more nuanced position, stating that the spread of the virus in children is moderate.
-It is now suggested that the impact of school openings on infection rates was limited, with children and teachers more often contracting infections through household or community contacts rather than within schools \citep{soriano2023policies}.
-This study, using robust statistical methods, considered the effect of school closures on COVID-19 cases across age groups.
-While there are commonalities in the evolution of COVID-19 cases across all European countries that were included in our data, school closures exhibit distinct impacts on different age groups.
-Notably, the analysis shows that the 0 to 4 age group experiences a downward trend in COVID-19 cases during the initial two weeks following school closure, followed by stabilization.
-In contrast, age groups ranging from 5 to 14 exhibit a stable profile after school closure.
-Additionally, age groups between 15 and 24 demonstrate a notable surge immediately after school closure, followed by a decrease after 14 days.
-Thus, the results do not support the hypothesis that school closures were effective for group ages older than 5 years old.
-One possibility is that school closures may lead to increased informal social gatherings among children and teenagers outside of controlled school environments, potentially facilitating virus transmission.
-It is also possible that transmission within households intensified as children spent more time at home.
-These results partially replicate observations from \citet{alfano2022effects} while providing a clearer picture of the structure of the effect of school closure.
+Knowledge about the transmission of the virus significantly improved
+over time as more studies have been published. While an early study
+found that children did not play an important role in the transmission
+of the virus \citep{li2020role}, more recent results give a more nuanced
+position, stating that the spread of the virus in children is moderate.
+It is now suggested that the impact of school openings on infection
+rates was limited, with children and teachers more often contracting
+infections through household or community contacts rather than within
+schools \citep{soriano2023policies}. This study, using robust
+statistical methods, considered the effect of school closures on
+COVID-19 cases across age groups. While there are commonalities in the
+evolution of COVID-19 cases across all European countries that were
+included in our data, school closures exhibit distinct impacts on
+different age groups. Notably, the analysis shows that the 0 to 4 age
+group experiences a downward trend in COVID-19 cases during the initial
+two weeks following school closure, followed by stabilization. In
+contrast, age groups ranging from 5 to 14 exhibit a stable profile after
+school closure. Additionally, age groups between 15 and 24 demonstrate a
+notable surge immediately after school closure, followed by a decrease
+after 14 days. Thus, the results do not support the hypothesis that
+school closures were effective for group ages older than 5 years old.
+One possibility is that school closures may lead to increased informal
+social gatherings among children and teenagers outside of controlled
+school environments, potentially facilitating virus transmission. It is
+also possible that transmission within households intensified as
+children spent more time at home. These results partially replicate
+observations from \citet{alfano2022effects} while providing a clearer
+picture of the structure of the effect of school closure.
 
-The Transfer Entropy calculations between age groups for each of the 12 selected European countries reveal an absence of symmetry between influencing age groups (\(X\)) and influenced age groups (\(Y\)).
-Some age groups influence the COVID-19 case numbers of other age groups, but there is no reciprocal influence.
-This finding suggests that changes in COVID-19 cases in certain age groups can predict the changes in other age groups but not vice versa, i.e., causality can be unidirectional.
-These findings highlight the importance of considering intergenerational interactions in designing effective control measures.
+The Transfer Entropy calculations between age groups for each of the 12
+selected European countries reveal an absence of symmetry between
+influencing age groups (\(X\)) and influenced age groups (\(Y\)). Some
+age groups influence the COVID-19 case numbers of other age groups, but
+there is no reciprocal influence. This finding suggests that changes in
+COVID-19 cases in certain age groups can predict the changes in other
+age groups but not vice versa, i.e., causality can be unidirectional.
+These findings highlight the importance of considering intergenerational
+interactions in designing effective control measures.
 
-The entropy analysis revealed intriguing directional patterns of case counts between age groups.
-In some countries, cases among younger individuals appear predictive of cases among older individuals, while the opposite pattern is observed elsewhere.
-These differences may reflect country-specific variations in social mixing patterns, healthcare responses, or policy measures.
-The absence of consistent patterns in infections among adjacent age groups raises questions about the mechanisms of transmission and suggests that factors beyond simple proximity, such as behavioural differences or differing susceptibility, may play a role.
-Future work should investigate these underlying mechanisms to better understand age-based transmission dynamics.
+The entropy analysis revealed intriguing directional patterns of case
+counts between age groups. In some countries, cases among younger
+individuals appear predictive of cases among older individuals, while
+the opposite pattern is observed elsewhere. These differences may
+reflect country-specific variations in social mixing patterns,
+healthcare responses, or policy measures. The absence of consistent
+patterns in infections among adjacent age groups raises questions about
+the mechanisms of transmission and suggests that factors beyond simple
+proximity, such as behavioural differences or differing susceptibility,
+may play a role. Future work should investigate these underlying
+mechanisms to better understand age-based transmission dynamics.
 
-Furthermore, the quadrant analysis reveals similar patterns across multiple countries.
-Austria, Germany, and the Netherlands exhibit significantly higher TE coefficients in the upper left quadrant, indicating that changes in COVID-19 cases in younger cohorts predict the changes in older cohorts.
-On the other hand, Austria, the Netherlands, Portugal, and Spain show higher TE coefficients in the lower right quadrant, indicating that changes in COVID-19 cases in older cohorts predict the changes in younger cohorts.
-However, these differences may also reflect country-specific variations in social mixing patterns, healthcare responses, or policy measures.
-The absence of consistent patterns in infections among adjacent age groups raises questions about the mechanisms of transmission and suggests that factors beyond simple proximity, such as behavioural differences or differing susceptibility, may play a role.
+Furthermore, the quadrant analysis reveals similar patterns across
+multiple countries. Austria, Germany, and the Netherlands exhibit
+significantly higher TE coefficients in the upper left quadrant,
+indicating that changes in COVID-19 cases in younger cohorts predict the
+changes in older cohorts. On the other hand, Austria, the Netherlands,
+Portugal, and Spain show higher TE coefficients in the lower right
+quadrant, indicating that changes in COVID-19 cases in older cohorts
+predict the changes in younger cohorts. However, these differences may
+also reflect country-specific variations in social mixing patterns,
+healthcare responses, or policy measures. The absence of consistent
+patterns in infections among adjacent age groups raises questions about
+the mechanisms of transmission and suggests that factors beyond simple
+proximity, such as behavioural differences or differing susceptibility,
+may play a role.
 
-This study, despite its methodological strengths, is subject to certain limitations that warrant consideration.
-Foremost among these is the inherent challenge of perfectly isolating the effect of school closures from other simultaneously implemented and dynamically changing NPIs.
-While we incorporated four composite NPI index derived from the Oxford COVID-19 Government Response Tracker (OxCGRT) project to account for the broader public health context, residual confounding from unmeasured aspects of NPIs or behavioural changes may still persist.
-Our sensitivity analyses showed consistent directional effects for school closures even under varying overall NPI indexes provide some confidence, but this remains an inherent complexity in observational studies of this nature.
-Another consideration is the ecological design of the study, which uses aggregated country-level data; this may mask important sub-national variations and precludes inferences about individual-level risk factors.
-In addition, the definition of ``closure'' which excluded ``partially open'' school statuses to simplify the intervention definition, means the impact of such intermediate states was not captured.
-Variations in testing strategies, case definitions, and reporting practices across countries and over time could also have influenced the observed case counts, despite efforts to use standardized data sources.
-Lastly, this study predominantly covers periods before widespread COVID-19 vaccination and the emergence of later variants such as Omicron; the effects of school closures might differ under those altered epidemiological conditions.
+This study, despite its methodological strengths, is subject to certain
+limitations that warrant consideration. Foremost among these is the
+inherent challenge of perfectly isolating the effect of school closures
+from other simultaneously implemented and dynamically changing NPIs.
+While we incorporated four composite NPI index derived from the Oxford
+COVID-19 Government Response Tracker (OxCGRT) project to account for the
+broader public health context, residual confounding from unmeasured
+aspects of NPIs or behavioural changes may still persist. Our
+sensitivity analyses showed consistent directional effects for school
+closures even under varying overall NPI indexes provide some confidence,
+but this remains an inherent complexity in observational studies of this
+nature. Another consideration is the ecological design of the study,
+which uses aggregated country-level data; this may mask important
+sub-national variations and precludes inferences about individual-level
+risk factors. In addition, the definition of ``closure'' which excluded
+``partially open'' school statuses to simplify the intervention
+definition, means the impact of such intermediate states was not
+captured. Variations in testing strategies, case definitions, and
+reporting practices across countries and over time could also have
+influenced the observed case counts, despite efforts to use standardized
+data sources. Lastly, this study predominantly covers periods before
+widespread COVID-19 vaccination and the emergence of later variants such
+as Omicron; the effects of school closures might differ under those
+altered epidemiological conditions.
 
-Future studies should look more closely at why the virus spread between age groups as it did, and why TE patterns differed by country.
-This could involve using information on people's behaviour or more detailed data on other health rules.
-It would also be useful to study how different virus types and vaccination levels change these effects.
-Learning more about other influencing factors will help us better understand the real impact of school closures.
+Future studies should look more closely at why the virus spread between
+age groups as it did, and why TE patterns differed by country. This
+could involve using information on people's behaviour or more detailed
+data on other health rules. It would also be useful to study how
+different virus types and vaccination levels change these effects.
+Learning more about other influencing factors will help us better
+understand the real impact of school closures.
 
@@
 
-This comprehensive analysis of COVID-19 school closures across 12 European nations reveals a complex and often counter-intuitive reality that challenges prevailing assumptions about their universal efficacy, particularly for children aged 5 and older.
-While national-level data showed an overall decrease in cases during closures, our age-disaggregated analysis using GAMs demonstrated that school-aged children (5-14 years) and young adults (15-24 years) frequently experienced periods of increased infections post-closure, a stark contrast to the consistent reductions seen only in pre-schoolers.
-Furthermore, Transfer Entropy analysis highlighted that the directional influence of transmission between age groups was asymmetric and varied by country, suggesting that the role of children in broader community transmission during such periods is not uniform.
+This comprehensive analysis of COVID-19 school closures across 12
+European nations reveals a complex and often counter-intuitive reality
+that challenges prevailing assumptions about their universal efficacy,
+particularly for children aged 5 and older. While national-level data
+showed an overall decrease in cases during closures, our
+age-disaggregated analysis using GAMs demonstrated that school-aged
+children (5-14 years) and young adults (15-24 years) frequently
+experienced periods of increased infections post-closure, a stark
+contrast to the consistent reductions seen only in pre-schoolers.
+Furthermore, Transfer Entropy analysis highlighted that the directional
+influence of transmission between age groups was asymmetric and varied
+by country, suggesting that the role of children in broader community
+transmission during such periods is not uniform.
 
-These findings underscore the critical importance of nuanced, age-specific analysis in evaluating public health interventions and caution against the broad application of measures like school closures without robust evidence of their benefit across all targeted demographics and a clear understanding of potential unintended consequences \citep{alfano2020efficacy, molefi2021impact}.
-For future pandemic preparedness, a more targeted, evidence-based approach is essential, prioritising measures that minimise societal disruption while effectively protecting public health across all age strata.
+These findings underscore the critical importance of nuanced,
+age-specific analysis in evaluating public health interventions and
+caution against the broad application of measures like school closures
+without robust evidence of their benefit across all targeted
+demographics and a clear understanding of potential unintended
+consequences \citep{alfano2020efficacy, molefi2021impact}. For future
+pandemic preparedness, a more targeted, evidence-based approach is
+essential, prioritising measures that minimise societal disruption while
+effectively protecting public health across all age strata.
 
@@
 
-This study is a longitudinal analysis on available data; thus, it does not require ethical approval.
+This study is a longitudinal analysis on available data; thus, it does
+not require ethical approval.
 
@@
 
-The data can be accessed from the COVerAGE-DB OSF repository (\url{https://osf.io/mpwjq/}), from the Oxford Covid-19 Government Response Tracker (OxCGRT) ``covid-policy-dataset'' GitHub repository (\url{https://github.com/OxCGRT/covid-policy-dataset}), and from the UNESCO servers (\url{https://en.unesco.org/file/unesco-data-school-closures-february-2020-june-2022csv-zip}).
-All preprocessing, analyses and code used to build the submitted paper are available at \url{https://github.com/damien-dupre/covid_ts_causality}.
+The data can be accessed from the COVerAGE-DB OSF repository
+(https://osf.io/mpwjq/), from the Oxford Covid-19 Government Response
+Tracker (OxCGRT) ``covid-policy-dataset'' GitHub repository
+(https://github.com/OxCGRT/covid-policy-dataset), and from the UNESCO
+servers
+(https://en.unesco.org/file/unesco-data-school-closures-february-2020-june-2022csv-zip).
+All preprocessing, analyses and code used to build the submitted paper
+are available at https://github.com/damien-dupre/covid\_ts\_causality.
 
-\bibliographystyle{tfcad}
+\section{References}\label{references}
+
+\renewcommand{\bibsection}{}
 \bibliography{bibliography.bib}

### 2025-06-05 manuscript_njd/template.qmd
update + njd

---
title: "Quarto Template (wiley-njd) Example"

format:
  wiley-njd-pdf:
    article-type: "Article Type: Exemplar"
    received: "10 November 2024"
    revised: "10 November 2024"
    accepted: "1 January 2025"
    journal: "Journal: None"
    volume: "00"
    copy-year: 2024
    start-page: 1024

    titlemark: "Short Title (in header)"
    authormark: "Short Author (in header)"

    keywords:
      - keyword1
      - buzzword
      - keyword3

author:
  - name: Josh Cowley
    email: joshcowley@fakeemails.com
    corresponding: true
    affiliations:
      - name: That University
        department: School of Something
        city: City
        country: UK

  - name: The Supervisor
    email: thesupervisor@fakeemails.com
    affiliations:
      - name: That University
        department: School of Something
        city: City
        country: UK

  - name: Business Contact
    email: businesscontact@fakebusinessemails.com
    affiliations:
      - name: That Company
        department: Advanced Department
        city: City
        country: Germany

present-address: |
  The present address is currently hidden.

abstract: |
  Document serves as example input for Quarto template openly available
    from Github, see more information at
    <https://github.com/nclJoshCowley/wiley-njd>.
  All rendering requires LaTeX to be installed, this includes any 
    TeX installation bundled with Quarto.
  Since Wiley's "New Journal Design" (NJD) LaTeX template is utilised in
    the build process, usage is subject to any license associated with
    original template.

keep-tex: true
bibliography: bibliography.bib
---

```{r}
#| label: setup
#| include: false
library(ggplot2)
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 10))

library(knitr)
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  # Example journal specifications: (width: 17.15 cm, DPI: >= 300)
  fig.width = 17.15 / 2.54,
  dpi = 300
)
```

# Example Content

## Raw Latex {#sec-rawlatex}

Any required LaTeX can be added to the document as required.
For Quarto, this is best done by including the raw LaTeX inside a 
  ` ```{=latex} ... ``` ` block, as shown in the source document.

```{=latex}
\setlength{\parindent}{0pt}
```

## Equations {#sec-eqns}

As with LaTeX, we can create in-line equations with $y = mx + c$ and create
  display single-line equations with

$$
    a^n
        = b^n + c^n
$$

For multi-line equations, use the `$$` around a `begin{aligned}[t]` method.

$$
\begin{aligned}[t]
    y_1
        &= \beta_0 + \beta_1 x_1
    \\
        &\vdots
    \\
    y_n
        &= \beta_0 + \beta_1 x_n
\end{aligned}
$$ {#eq-linreg}

## Code

In-line code can be enacted with backticks, for example $1 + 1 = `r 1 + 1`$.
Display code chunks are formulated in the usual Quarto way.

```{r}
factorial(5)
```

For example, output is suppressed via the chunk option `#| echo: false`.

```{r}
#| echo: false
sprintf("10! = %i", factorial(10))
```

## Images

We can include graphics that already exist with `knitr::include_graphics`
  or by `\includegraphics` using the method described in @sec-rawlatex.

```{r}
knitr::include_graphics("https://quarto.org/quarto.png")
```

And we can generate our own plots using R code.
Note that we only change `fig.width` and `fig.height` or `fig.asp`;
  changing `out.height` or `out.width` will stretch the image.

```{r}
#| label: fig-cars
#| fig-cap: "A scatter plot of weight and MPG from the `mtcars` dataset"
#| fig-asp: 0.8
#| echo: false
mtcars |> 
  ggplot2::ggplot(ggplot2::aes(x = mpg, y = wt)) +
  ggplot2::geom_smooth(method = "lm", formula = y ~ x, colour = "red") +
  ggplot2::geom_point() +
  ggplot2::labs(
    title = "Negative Correlation between Weight and MPG",
    y = "Weight",
    x = "Miles per Gallon (MPG)"
  )
```

## Cross-Reference

We want to be able to reference already defined items such as

1. linear regression, as described in (@eq-linreg) from @sec-eqns.

1. scatter plot of `mtcars` data, @fig-cars.

## Citations

We can cite is a few different ways that are consistent with `natbib`

1. Parenthesised citation, using `citep` [@Hirt1974].

1. Textual citation, using `citet` as in @Hirt1974.

Multiple citations follow (semi-colon separated) Quarto syntax
  [@Hirt1974; @Benson1992; @Dukowicz1984].

## Footnotes

Here is a footnote reference,[^1] and another.[^longnote]

[^1]: Here is the footnote.

[^longnote]: Here's one with multiple blocks.

    Subsequent paragraphs are indented to show that they belong to the previous
    footnote.

    Multi-paragraph footnotes work like multi-paragraph list items.

## Quotes

The `quote` environment is supplied by the Markdown quote syntax where lines
  begin with `>`.
For example,

> The combination of some data and an aching desire for an answer does not
> ensure that a reasonable answer can be extracted from a given body of data.
> 
> -- John Tukey

\newpage

# Future Work

## NJD Consistency

Wiley's NJD supplies

* 2 column layouts, only 1 column layouts apply here.

* Multiple referencing systems, only Harvard applies here.

* Font options (Helvetica, Lato, etc.) but installing fonts is non-trivial.
  Currently Times is used but we may want sans-serif.

## Theorems

Custom environments supplied by NJD should work with Quarto divs, however
  further testing is required.

::: {#thm-1}
This is a theorem.
:::

::: {proof}
With a proof.
:::

::: {#prp-1}
This is a proposition.
:::

::: {#def-1}
This is a definition
:::

\newpage

# References

::: {#refs}
:::

### 2025-06-05 manuscript_tf/manuscript_blind.tex
update
update + njd

@@
 
-\author{\name{Damien Dupré$^{1}$, Edgar Morgenroth$^{1}$}
-\affil{$^{1}$Dublin City University Business School, Glasnevin, Dublin 9, Ireland}
+\author{\name{Author 1$^{1}$, Author 2$^{2}$}
+\affil{$^{1}$Affiliation 1; $^{2}$Affiliation 2}
 }
@@
 The data can be accessed from the COVerAGE-DB OSF repository (\url{https://osf.io/mpwjq/}), from the Oxford Covid-19 Government Response Tracker (OxCGRT) ``covid-policy-dataset'' GitHub repository (\url{https://github.com/OxCGRT/covid-policy-dataset}), and from the UNESCO servers (\url{https://en.unesco.org/file/unesco-data-school-closures-february-2020-june-2022csv-zip}).
-All preprocessing, analyses and code used to build the submitted paper are available at \url{https://github.com/damien-dupre/covid_ts_causality}.
+All preprocessing, analyses and code used to build the submitted paper are available at {[}link hidden for blind peer-review process{]}.
 

### 2025-06-05 manuscript_njd/manuscript_blind.qmd
update

--- 2025-06-03 manuscript_tf/manuscript_blind.Rmd
@@
 ---
-title: |
-  Divergent Impacts of COVID-19 School Closures on Youth Infection Dynamics: Evidence from 12 European Countries
-# author:
-#   - name: Damien Dupré
-#     affil: 1
-#     email: damien.dupre@dcu.ie (corresponding author)
-#   - name: Edgar Morgenroth
-#     affil: 1
-#     email: edgar.morgenroth@dcu.ie
-# affiliation:
-#   - num: 1
-#     address: |
-#       Dublin City University Business School, Glasnevin, Dublin 9, Ireland
-# thanks:
-#   author
+title: "Divergent Impacts of COVID-19 School Closures on Youth Infection Dynamics: Evidence from 12 European Countries"
+
+format:
+  wiley-njd-pdf:
+    article-type: "Article Type: Research Article"
+    # received: "10 November 2024"
+    # revised: "10 November 2024"
+    # accepted: "1 January 2025"
+    journal: "Journal of Policy Analysis and Management"
+    volume: "00"
+    copy-year: 2025
+    start-page: 1
+
+    titlemark: "COVID-19 School Closures and Youth Infection Dynamics"
+    authormark: "anonymous"
+
+    keywords:
+      - COVID-19
+      - School closures
+      - Intergenerational transmission
+      - Non-linear effects
+      - Infection dynamics
 
@@
   - name: Author 1
-    affil: 1
-    email: Email 1
+    email: email 1
+    corresponding: true
+    affiliations:
+      - name: affiliations
+        department: departments
+        city: cities
+        country: countries
+
   - name: Author 2
-    affil: 2
-    email: Email 2
-affiliation:
-  - num: 1
-    address: |
-      Affiliation 1
-  - num: 2
-    address: |
-      Affiliation 2
+    email: email 2
+    affiliations:
+      - name: affiliations
+        department: departments
+        city: cities
+        country: countries
 
-bibliography: bibliography.bib
+present-address: |
+  present-address
+
 abstract: |
@@
 These results challenge the assumption that school closures uniformly benefit younger populations and emphasise the need for age-specific analysis in pandemic planning and support the development of more nuanced, evidence-based public health policies.
-
-keywords: |
-  COVID-19; School Closures; Intergenerational Transmission; Non-linear effects; Infection dynamics.
-# \begin{jelcode}
-# I18; C22.
-# \end{jelcode}
-output:
-  bookdown::pdf_book:
-    base_format: rticles::tf_article
-    keep_tex: true
-header-includes:
-  - \usepackage{float}
-  - \usepackage{booktabs}
-  - \usepackage{longtable}
-  - \usepackage{threeparttable}
-  - \floatplacement{figure}{H}
-  - \floatplacement{table}{H}
-  - \usepackage{caption}
-  - \usepackage{hyperref}
-  - \usepackage[utf8]{inputenc}
-  - \def\tightlist{}
+pdf-engine: lualatex
+keep-tex: true
+bibliography: bibliography.bib
 ---
 
-(ref:overall) Cumulative COVID-19 cases number for selected European countries since the beginning of the pandemic. Source: COVerAGE-DB [@riffe2021data].
-(ref:descriptive) Periods of school closure since the beginning of the COVID-19 pandemic for selected European countries and their reason: regular academic break versus closure due to government decisions.
-Source: @unesco2022https.
-(ref:country) Standardized effect of the smooth term in Generalized Additive Model by country. Standardized effects are reported to compare the shape of the curve between countries.
-(ref:age) Standardized effect of the smooth term in Generalized Additive Model by age group. Standardized effects are reported to compare the shape of the curve between age groups.
-(ref:te) Matrix of Transfer Entropy coefficients according to every age group combination for each of the selected European country.
-Age groups on the x-axis are influencing the age groups on the y-axis ($ag\,x \rightarrow\,ag\,y$).
-The significance of each Transfer Entropy coefficient is provided in Appendix 2.
+```{r}
+#| label: setup
+#| include: false
 
-```{r setup, include = FALSE}
 # libraries --------------------------------------------------------------------
@@
   )
-
-# raw data ---------------------------------------------------------------------
-# osf_retrieve_file("7tnfh") |>
-#   osf_download(conflicts = "overwrite", progress = TRUE, verbose = TRUE)
-# 
-# df <- "Output_5.zip" |>
-#   read_csv(skip = 3, col_types = "ccccciiddd") |>
-#   mutate(Date = as.Date(Date, format = "%d.%m.%Y")) |>
-#   clean_names()
-# 
-# file.remove("Output_5.zip")
-# 
-# write_csv(df, here("data/df.csv"))
-# 
-# df <- read_csv(here("data/df.csv"))
-# 
-# df_school_closure <- 
-#   read_csv(
-#     "https://en.unesco.org/sites/default/files/covid_impact_education_full.csv",
-#     col_types = list(.default = col_character(), Date = col_date(format = "%d/%m/%Y"))) |> 
-#   clean_names() |>
-#   select(date, country, status) |> 
-#   mutate(country = case_when(
-#     country == "Bolivia (Plurinational State of)" ~ "Bolivia",  
-#     country == "Central African republic" ~ "Central African Republic",
-#     country == "Iran (Islamic Republic of)" ~ "Iran",
-#     country == "Republic of Moldova" ~ "Moldova",
-#     country == "Republic of Korea" ~ "South Korea",
-#     country == "United Kingdom of Great Britain and Northern Ireland" ~ "United Kingdom",
-#     country == "United States of America" ~ "USA",
-#     country == "Viet Nam" ~ "Vietnam",
-#     TRUE ~ country
-#     )
-#   )
-# 
-# df_check <- 
-#   read_csv("https://covid19.who.int/WHO-COVID-19-global-data.csv") |> 
-#   clean_names() |> 
-#   rename(date = date_reported) |> 
-#   mutate(country = case_when(
-#     country == "Bolivia (Plurinational State of)" ~ "Bolivia",  
-#     country == "Iran (Islamic Republic of)" ~ "Iran",
-#     country == "occupied Palestinian territory, including east Jerusalem" ~ "Palestine",
-#     country == "Republic of Moldova" ~ "Moldova",
-#     country == "Republic of Korea" ~ "South Korea",
-#     country == "United States of America" ~ "USA",
-#     country == "Venezuela (Bolivarian Republic of)" ~ "Venezuela",
-#     country == "Viet Nam" ~ "Vietnam",
-#     TRUE ~ country
-#     )
-#   )
-# 
-# # variables --------------------------------------------------------------------
-# ## keep only closure waves longer than `days_min_closure` but display only the first `days_after_closure`
-# days_min_closure <- 21
-# days_after_closure <- 28
-# ## type of school closure to investigate
-# period_closure_to_keep <- c(
-#   #"Partially open",
-#   "Academic break",
-#   "Closed due to COVID-19"
-#   )
-# 
-# ## countries of interest
-# countries_of_interest <- c(
-#   "Austria", "Belgium", "Bulgaria", "Croatia", "Estonia", "France",
-#   "Germany", "Greece", "Netherlands", "Portugal", "Slovakia", "Spain"
-#   )
-# 
-# # transformed data -------------------------------------------------------------
-# df_global <- df |> 
-#   filter(
-#     country %in% countries_of_interest,
-#     date > as.Date("2020-01-01"),
-#     region == "All",
-#     sex == "b"
-#   ) |> 
-#   group_by(country, date) |> 
-#   summarise(cumulative_cases_cal = sum(cases)) |> 
-#   group_by(country) |> 
-#   complete(date = full_seq(date, 1)) |> 
-#   mutate(
-#     timeline = date - lag(date),  # check if gap between dates
-#     daily_cases = round(cumulative_cases_cal - lag(cumulative_cases_cal), 0)
-#   ) |> 
-#   ungroup() |> 
-#   left_join(df_school_closure, by = c("date", "country")) |> 
-#   group_by(country) |> 
-#   fill(status) |> 
-#   ungroup() |> 
-#   mutate(schools_closed = if_else(status %in% period_closure_to_keep, 1, 0)) |> 
-#   drop_na(status) |> 
-#   left_join(df_check, by = c("date", "country"))
-# 
-# write_csv(df_global, here("data/df_global.csv"))
-# 
-# df_global_age <- df |> 
-#   filter(
-#     country %in% countries_of_interest,
-#     date > as.Date("2020-01-01"),
-#     region == "All",
-#     sex == "b"
-#   ) |> 
-#   mutate(
-#     age_grp = case_when(
-#       age == 0  ~ "aged0to4",
-#       age == 5  ~ "aged5to9",
-#       age == 10 ~ "aged10to14",
-#       age == 15 ~ "aged15to19",
-#       age == 20 ~ "aged20to24",
-#       between(age, 25, 34) ~ "aged25to34",
-#       between(age, 35, 44) ~ "aged35to44",
-#       between(age, 45, 54) ~ "aged45to54",
-#       between(age, 55, 64) ~ "aged55to64",
-#       age > 64 ~ "aged65+"
-#     )
-#   ) |> 
-#   group_by(country, date, age_grp) |> 
-#   summarise(cumulative_cases_cal = sum(cases)) |> 
-#   group_by(country, age_grp) |> 
-#   complete(date = full_seq(date, 1)) |> 
-#   mutate(
-#     timeline = date - lag(date),  # check if gap between dates
-#     daily_cases = round(cumulative_cases_cal - lag(cumulative_cases_cal), 0),
-#     daily_changes = daily_cases - lag(daily_cases)
-#   ) |> 
-#   ungroup()
-# 
-# write_csv(df_global_age, here("data/df_global_age.csv"))
-#
-# df_global <- read_csv(here("data/df_global.csv"))
-# df_global_age <- read_csv(here("data/df_global_age.csv"))
-
 ```
@@
 The GAM is fitted on the daily COVID-19 cases to test the hypothesis of a significant non-linear evolution of cases among age groups from 0 to 4, from 5 to 9, from 10 to 14, from 15 to 19, and from 20 to 24 [@wood2017generalized].
-The model also estimates the overall non-linear effect by country, taking into account the interaction between age groups and countries as random intercepts and the interaction between time, countries, and the period of closure as random effects (Eq 1).
+The model also estimates the overall non-linear effect by country, taking into account the interaction between age groups and countries as random intercepts and the interaction between time, countries, and the period of closure as random effects (Eq @eq-gam).
 
@@
 
-\begin{align*}
-  \tag{1}
+$$
+\begin{aligned}[t]
   n\prime_t &\sim \text{Poisson}(\lambda_t) \\
@@
   &\quad + f_7(economic\,support\,index_t) +\epsilon_t
-\end{align*}
+\end{aligned}
+$${#eq-gam}
 
@@
 
-\begin{align}
-  \tag{2}
+$$
+\begin{aligned}[t]
   T_{ag\,x \rightarrow ag\,y}(k,l) = \sum_{ag\,x_{t+1}, ag\,x_t^{(k)}, ag\,y_t^{(l)}} 
@@
   \log \left(\frac{p\left(ag\,x_{t+1} \mid ag\,x_t^{(k)}, ag\,y_t^{(l)}\right)}{p\left(ag\,x_{t+1} \mid ag\,x_t^{(k)}\right)}\right)
-\end{align}
+\end{aligned}
+$$ {#eq-te}
 
-\noindent where $T_{ag\,x \rightarrow\,ag\,y}$ consequently measures the influence of the change dynamic from an age group $X$ (or $ag\,x$) to another age group $Y$ (or $ag\,y$) for every country (Eq 2).
+\noindent where $T_{ag\,x \rightarrow\,ag\,y}$ consequently measures the influence of the change dynamic from an age group $X$ (or $ag\,x$) to another age group $Y$ (or $ag\,y$) for every country (Eq @eq-te).
 
@@
 
-The data show that the trend of confirmed COVID-19 cases follows similar patterns across the selected European countries, with scales following the size of the population in these countries (Figure \@ref(fig:overall)).
+The data show that the trend of confirmed COVID-19 cases follows similar patterns across the selected European countries, with scales following the size of the population in these countries (@fig-overall).
 Thus, the bigger the country, the higher the total number of COVID-19 cases.
 
-```{r overall, fig.height=3, out.width="\\textwidth", fig.cap="(ref:overall)"}
+```{r, fig.height=3, out.width="\\textwidth"}
+#| label: fig-overall
+#| fig.cap: "Cumulative COVID-19 cases number for selected European countries since the beginning of the pandemic. Source: COVerAGE-DB [@riffe2021data]."
+
 overall_plot <- df_global |> 
@@
 The evolution of COVID-19 cases reveals some similarities across all age groups.
-However, the influence of each wave on individual age groups also has some particularities (Figure \@ref(fig:descriptive)).
+However, the influence of each wave on individual age groups also has some particularities (@fig-descriptive).
 For example, the first wave was more important among the oldest age groups, whereas the third wave was more important among the youngest age groups.
 
-```{r descriptive, fig.height=3, out.width="\\textwidth", fig.cap="(ref:descriptive)"}
+```{r, fig.height=3, out.width="\\textwidth"}
+#| label: fig-descriptive
+#| fig.cap: "Periods of school closure since the beginning of the COVID-19 pandemic for selected European countries and their reason: regular academic break versus closure due to government decisions.
+Source: @unesco2022https."
+
 descriptive_plot <- df_global |> 
@@
 As outlined above, to evaluate the shape of the trend in the numbers of COVID-19 cases reported after the three school closures longer than 21 consecutive days, a GAM was fitted as described above, taking into account the overall effect across all the selected European countries, as well as the effect for age groups: 0 to 4, 5 to 9, 10 to 14, 15 to 19, and 20 to 24 years old.
-The obtained results satisfy the requirements to fit this model, which explains `r percent(gam_summary$dev.expl, accuracy = 0.1)` of the variation in COVID-19 cases.
+The obtained results satisfy the requirements to fit this model, which explains `r percent(gam_summary$dev.expl, accuracy = 0.1)` of the variation in COVID-19 cases (@fig-age).
 
-Overall, the results revealed a decreasing non-linear effect of school closure at a country level for the selected European countries (Austria: `r gam_results["Austria"]`; Belgium: `r gam_results["Belgium"]`; Bulgaria: `r gam_results["Bulgaria"]`; Croatia: `r gam_results["Croatia"]`; Estonia: `r gam_results["Estonia"]`; France: `r gam_results["France"]`; Germany (`r gam_results["Germany"]`); Greece: `r gam_results["Greece"]`; The Netherlands: `r gam_results["Netherlands"]`; Portugal: `r gam_results["Portugal"]`; Slovakia (`r gam_results["Slovakia"]`; Spain: `r gam_results["Spain"]`, see Figure \@ref(fig:age)).
+```{r, fig.height=3, out.width="\\textwidth"}
+#| label: fig-age
+#| fig.cap: "Standardized effect of the smooth term in Generalized Additive Model by age group. Standardized effects are reported to compare the shape of the curve between age groups."
 
-```{r country, fig.height=3, out.width="\\textwidth", fig.cap="(ref:country)"}
-gam_plot_country
+gam_plot_age
 ```
 
-```{r age, fig.height=3, out.width="\\textwidth", fig.cap="(ref:age)"}
-gam_plot_age
+Overall, the results revealed a decreasing non-linear effect of school closure at a country level for the selected European countries (Austria: `r gam_results["Austria"]`; Belgium: `r gam_results["Belgium"]`; Bulgaria: `r gam_results["Bulgaria"]`; Croatia: `r gam_results["Croatia"]`; Estonia: `r gam_results["Estonia"]`; France: `r gam_results["France"]`; Germany (`r gam_results["Germany"]`); Greece: `r gam_results["Greece"]`; The Netherlands: `r gam_results["Netherlands"]`; Portugal: `r gam_results["Portugal"]`; Slovakia (`r gam_results["Slovakia"]`; Spain: `r gam_results["Spain"]`, see @fig-country.
+
+```{r, fig.height=3, out.width="\\textwidth"}
+#| label: fig-country
+#| fig.cap: "Standardized effect of the smooth term in Generalized Additive Model by country. Standardized effects are reported to compare the shape of the curve between countries."
+
+gam_plot_country
 ```
@@
 
-```{r te, fig.height=8, out.width="\\textwidth", fig.cap="(ref:te)", fig.pos = 'H'}
+```{r}
+#| label: fig-te
+#| fig.cap: "Matrix of Transfer Entropy coefficients according to every age group combination for each of the selected European country.
+Age groups on the x-axis are influencing the age groups on the y-axis ($ag\\,x \\rightarrow\\,ag\\,y$).
+The significance of each Transfer Entropy coefficient is provided in Appendix 2."
+#| fig.height: 7
+
 te_plot
@@
 
-Considering all countries, the analysis of age groups reveals distinct patterns (Figure \@ref(fig:te)).
+Considering all countries, the analysis of age groups reveals distinct patterns (@fig-te).
 For the 0 to 4 age group, i.e., the pre-school group, there was a downward trend during the initial three weeks following school closure, followed by an increase during the fourth week (`r gam_results["0to4"]`).
@@
 
-The results of the transfer entropy calculations between age groups for each of the 12 selected European countries are reported in Figure \@ref(fig:te).
+The results of the transfer entropy calculations between age groups for each of the 12 selected European countries are reported in @fig-te.
 The absence of symmetry between influencing age groups (i.e., $ag\,x$) and influenced age groups (i.e., $ag\,y$) is found.
 Indeed, the change in COVID-19 cases in some age groups is influenced by other age groups, but they are not reciprocally influencing these age groups.
-Figure \@ref(fig:te) shows how different age cohorts influence the COVID-19 case numbers of all age groups across countries.
+F@fig-te shows how different age cohorts influence the COVID-19 case numbers of all age groups across countries.
 The upper left quadrant indicates how younger age cohorts are influencing older age cohorts, the bottom right quadrant indicates how older age cohorts are influencing younger age cohorts, finally, the lower left and upper right indicate how younger or older age cohorts are influencing themselves.
@@
 The data can be accessed from the COVerAGE-DB OSF repository (https://osf.io/mpwjq/), from the Oxford Covid-19 Government Response Tracker (OxCGRT) "covid-policy-dataset" GitHub repository (https://github.com/OxCGRT/covid-policy-dataset), and from the UNESCO servers (https://en.unesco.org/file/unesco-data-school-closures-february-2020-june-2022csv-zip).
-All preprocessing, analyses and code used to build the submitted paper are available at https://github.com/damien-dupre/covid_ts_causality.
+All preprocessing, analyses and code used to build the submitted paper are available at https://github.com/[hidden for anonymous peer-review].
 
-# References {-}
+# References
+
+::: {#refs}
+:::

### 2025-06-05 manuscript_njd/manuscript_blind.tex
update

--- 2025-06-05 manuscript_njd/manuscript.tex
@@
 
-\authormark{Dupré and Morgenroth}
+\authormark{anonymous}
 
-\presentaddress{DCU Business School, Dublin 9, Ireland.}
+\presentaddress{present-address}
 
 % Include information for each author
-  \author[1]{Damien Dupré}
-  \author[1]{Edgar Morgenroth}
+  \author[1]{Author 1}
+  \author[1]{Author 2}
 
@@
   \orgdiv{%
-    Business School%
+    departments%
     , %
     \orgname{%
-      Dublin City University%
+      affiliations%
       , %
@@
     \orgaddress{%
-      \state{Dublin}%
+      \state{cities}%
       , %
-      \country{Ireland}%
+      \country{countries}%
     }%
@@
       %
-        Damien Dupré%
+        Author 1%
       %
-      \email{damien.dupre@dcu.ie}%
+      \email{email 1}%
     }
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{manuscript_files/figure-pdf/fig-overall-1.png}}
+\pandocbounded{\includegraphics[keepaspectratio]{manuscript_blind_files/figure-pdf/fig-overall-1.png}}
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{manuscript_files/figure-pdf/fig-descriptive-1.png}}
+\pandocbounded{\includegraphics[keepaspectratio]{manuscript_blind_files/figure-pdf/fig-descriptive-1.png}}
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{manuscript_files/figure-pdf/fig-age-1.png}}
+\pandocbounded{\includegraphics[keepaspectratio]{manuscript_blind_files/figure-pdf/fig-age-1.png}}
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{manuscript_files/figure-pdf/fig-country-1.png}}
+\pandocbounded{\includegraphics[keepaspectratio]{manuscript_blind_files/figure-pdf/fig-country-1.png}}
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{manuscript_files/figure-pdf/fig-te-1.png}}
+\pandocbounded{\includegraphics[keepaspectratio]{manuscript_blind_files/figure-pdf/fig-te-1.png}}
 
@@
 All preprocessing, analyses and code used to build the submitted paper
-are available at https://github.com/damien-dupre/covid\_ts\_causality.
+are available at https://github.com/{[}hidden for anonymous
+peer-review{]}.
 

### 2025-06-05 manuscript_njd/title_page.qmd
update

--- 2025-06-05 manuscript_njd/manuscript_blind.qmd
@@
     titlemark: "COVID-19 School Closures and Youth Infection Dynamics"
-    authormark: "anonymous"
+    authormark: "Dupré and Morgenroth"
 
@@
       - Non-linear effects
-      - Infection dynamics
+      - Infection dynamics.
 
 author:
-  - name: Author 1
-    email: email 1
+  - name: Damien Dupré
+    email: damien.dupre@dcu.ie
     corresponding: true
     affiliations:
-      - name: affiliations
-        department: departments
-        city: cities
-        country: countries
+      - name: Dublin City University
+        department: Business School
+        city: Dublin
+        country: Ireland
 
-  - name: Author 2
-    email: email 2
+  - name: Edgar Morgenroth
+    email: edgar.morgenroth@dcu.ie
     affiliations:
-      - name: affiliations
-        department: departments
-        city: cities
-        country: countries
+      - name: Dublin City University
+        department: Business School
+        city: Dublin
+        country: Ireland
 
 present-address: |
-  present-address
+  DCU Business School, Dublin 9, Ireland.
 
@@
 ---
-
-```{r}
-#| label: setup
-#| include: false
-
-# libraries --------------------------------------------------------------------
-library(ggrepel)
-library(here)
-library(janitor)
-library(knitr)
-library(kableExtra)
-library(osfr)
-library(scales)
-library(tidyverse)
-
-# options ----------------------------------------------------------------------
-set.seed(123) # Seed for random number generation
-options(scipen = 999) # disable sci number format
-opts_chunk$set(
-  cache.extra = rand_seed, 
-  cache = FALSE,
-  message = FALSE, 
-  warning = FALSE, 
-  error = FALSE,
-  echo = FALSE, 
-  fig.retina = 3,
-  dev = "ragg_png"
-  )
-```
-
-```{r analyses, cache=FALSE}
-source(here("analysis/gam_analysis_europe.R"))
-source(here("analysis/te_analysis_europe.R"))
-```
-
-# Introduction
-
-The COVID-19 pandemic has had a profound impact on global health, with an estimate of 14.83 million excess deaths globally [@msemburi2023estimates].
-Beyond direct mortality, the pandemic has caused significant collateral damage, including losses of lives and livelihoods, necessitating a comprehensive approach to measuring its broader impacts.
-A wide range of non-pharmaceutical interventions (NPI) were enacted to control the spread of COVID-19, especially before the availability of effective vaccines.
-Due to their social mixing patterns, children have been identified as an age group that can drive the spread of respiratory infections such as influenza [@moser2018estimating], and school closures were found to be effective in mitigating the spread of influenza H1N1 in Japan in 2009 [@kawano2015substantial].
-It is therefore not surprising that school closures were one of the NPIs that were introduced in many countries to stop the spread of COVID-19.
-They were used particularly during the initial wave of the pandemic and during subsequent waves when case numbers were high.
-
-Recently, a body of literature has shown various negative consequences on child development and health due to school closures.
-School closures have been found to have negatively affected child mental health [@moulin2022longitudinal], nutrition/obesity [@sugimoto2023temporal], and education [@lerkkanen2023reading].
-Some governments were questioning their decisions to close schools [@de2021determines].
-For example, the German health minister Karl Lauterbach, in an interview with one of Germany’s publicly funded television stations, admitted that “in retrospect it had been wrong to keep schools and childcare closed for so long” [@ard2023lauterbach].
-Even if school closures had negative consequences for children, they might nevertheless have been effective at stopping the spread of COVID-19.
-However, analysis on the effectiveness of school closures on the spread of COVID-19 remains inconclusive.
-In a study of a panel of European countries, @alfano2022effects found that school closures were associated with a reduced COVID-19 incidence.
-In contrast, @walsh2021school in a review of 40 studies covering 150 countries concluded that the effectiveness of school closures was uncertain, with 60% having identified no impact and pointing to the potential for the analysis to be affected by confounding factors and collinearity.
-The latter shortcomings suggest that it is hard to draw firm inferences, as even a positive result may not imply causality.
-
-Once it was clear that COVID-19 spreads from person to person, it was natural that governments were advised to enact measures to reduce social contact in order to reduce the spread of the virus.
-Therefore, it is legitimate to believe that by closing schools, a reduction of the contaminations would be observed in the younger age groups.
-However, the efficiency of school closure on the reduction of COVID-19 cases is still questioned [@bayham2020impact; @esposito2021comprehensive].
-While some research has found that school closures contribute to limit or to reduce the growth rate of confirmed cases after implementation [@stage2021shut; @yoshiyuki2020effects], others did not observe a change in the evolution of COVID-19 cases [@chang2020modelling; @iwata2020school].
-For instance, a controlled comparison between similar localities in Japan with schools closed and schools open did not reveal any evidence that school closures reduced the spread of COVID-19 [@fukumoto2021no].
-If the school closure had a real impact on the evolution of confirmed COVID-19 cases, it should be possible to observe a decrease or at least an inflection in the trend of its evolution among younger age groups.
-
-A second implicit belief regarding the effect of school closure on the spread of COVID-19 is that school not only influences the spread of the virus among children and teenagers but also has a knock-on effect on the spread of the virus in older age groups, also called Secondary Attack Rate (SAR).
-The contaminated children and teenagers would bring the virus back home and would pass the virus on to their parents and relatives.
-For example, research investigating the contamination in the household network not only revealed an exceptionally high rate of secondary contamination but also that this contamination happened when the schools were closed [@sorianoarandes2021household].
-Despite being reported in several clinical and epidemiological studies [@siebach2021childhood; @zhendong2020clinical], multiple research have shown that the SAR from children to household members was lower than expected [@heavey2020evidence; @vanderhoek2020role; @kim2021role; @ludvigsson2020children].
-However, the SAR of children and teenagers to the household member is likely to be age-dependent, with differences between infants, primary and secondary school children, and college students [@grasleguen2021reopening].
-If a secondary transmission from children and teenagers to household member has a significant influence, then a temporal causality relationship between their evolution should be observed.
-
-A gap in these previous analyses is the insufficient granularity in examining age-specific effects and the complex dynamics of inter-age group transmission.
-This study aims to address these deficiencies by providing a more robust and nuanced assessment.
-First, Generalized Additive Models (GAMs) are utilized to flexibly model and quantify the potentially non-linear evolution of COVID-19 cases following school closures, specifically within these fine-grained age brackets.
-This allows us to move beyond simple pre-post comparisons and capture dynamic temporal patterns.
-Then, Transfer Entropy (TE), an information-theoretic measure, is applied to assess directional causality in time series data.
-TE is particularly adept at identifying how past states of one variable (e.g., cases in one age group) influence future states of another (e.g., cases in a different age group), even in complex, non-linear systems like epidemic spread.
-This dual methodology allows for a deeper understanding of both the direct age-specific consequences of school closures and the potential for altered intergenerational transmission dynamics.
-Our findings challenge simplistic narratives about school closure efficacy and highlight the importance of age-stratified analysis for future pandemic preparedness.
-
-# Methods
-
-## Study design
-
-A longitudinal study was conducted to observe changes in COVID-19 cases during school closures.
-Changes in COVID-19 from five age groups from the first to the 28th day of school closure are compared across 12 countries.
-Then, a causality analysis evaluate the impact of changes in these younger groups on older age groups.
-
-## Data collection
-
-Data on COVID-19 case counts by age group have been obtained from the COVerAGE-DB project [@riffe2021data].
-The COVerAGE-DB which collect numbers for every countries by groups of 5 or 10 years if available.
-In addition, spline approximations are used to deals with the heterogeneity of countries’ reporting formats by using when the data for this age bracket is not available for a country.
-Countries communicating data with groups of 5 years were chosen to match as much as possible the different school stages.
-
-To identify the impact of school closures on the number of cases, information regarding the school closures on a day-by-day basis is required, and this is taken from the “UNESCO global education coalition” (2022).
-For each day, in each country, the status of the schools is indicated as fully open, partially open, closed due to COVID-19, or closed due to an academic break.
-Because it would be difficult to measure the effect of school closures at a country level when schools are partially closed, only closures due to COVID-19 or due to an academic break are considered.
-Indeed, both are considered as closure at a country-wide level.
-Any COVID-19 case numbers beyond the 28-day period are not relevant for evaluating the impact of school closures.
-
-## Data validation
-
-In order to cross-validate the data obtained after spline approximations, a comparison with the data published by the @who2022https reveals perfect similarities.
-The original data consist of 14,089,320 observations of 10 variables (117 distinct countries, region within the country, a unique observation code, the date of the observation, the gender, which can be male, female, or both, the age bracket by 5 years from 0 to 100, a confirmation of the age interval for each bracket, the total number of cases so far, the total number of deaths, and the total number of tests performed) from `r format(min(df_global$date), "%B %d, %Y")` to `r format(max(df_global$date), "%B %d, %Y")`.
-After removing countries with missing and inconsistent values, only 22 are suitable for data analyses.
-However, to focus this analysis on geographically and culturally comparable countries, only 12 European countries are kept: Austria, Belgium, Bulgaria, Croatia, Estonia, France, Germany, Greece, Netherlands, Portugal, Slovakia, and Spain.
-The observations are reported in terms of the total number of COVID-19 cases per day from the start of the pandemic.
-The daily number of cases at a specific date $n\prime_{t}$ is calculated with the difference between the total cases at a date $t$ and the total cases at a date $t-1$ (i.e., derivative 1).
-In addition, the change in the daily number of cases $n\prime\prime_{t}$ between $n\prime_{t}$ and $n\prime_{t-1}$ has also been calculated (i.e., derivative 2).
-
-## Controlling for confounding Non-Pharmaceutical Interventions
-
-Recognizing that school closures rarely occurred in isolation, a critical component of this analysis was to account for other contemporaneous NPIs.
-To mitigate confounding from other NPIs, we incorporated four time-varying composite NPI index identified by the Oxford COVID-19 Government Response Tracker (OxCGRT) project [@hale2021global]: the average stringency index, the average government response index, the average containment health index, and the average economic support index.
-Specific data were collected from each country on various policy responses governments implemented to curb the spread of COVID-19, such as travel restrictions, workplace closures, public event cancellations, restrictions on gatherings, or public health messaging.
-These indicators were then aggregated to create indexes, reflecting the overall level of government restrictions.
-
-## Generalized additive model
-
-The effect of school closure on the trend of daily COVID-19 cases is analysed using a Generalized Additive Model (GAM).
-GAM is a flexible modelling approach that estimates non-linear relationships between variables.
-Compared to other methods such as Vector Autoregression, GAM can handle fixed and random smooth effects without making strict assumptions about linearity.
-
-The GAM is fitted on the daily COVID-19 cases to test the hypothesis of a significant non-linear evolution of cases among age groups from 0 to 4, from 5 to 9, from 10 to 14, from 15 to 19, and from 20 to 24 [@wood2017generalized].
-The model also estimates the overall non-linear effect by country, taking into account the interaction between age groups and countries as random intercepts and the interaction between time, countries, and the period of closure as random effects (Eq @eq-gam).
-
-By estimating the degree of smoothness of a Bayesian spline smoothing using restricted fast maximum likelihood estimation [@wood2011fast], GAM identifies dynamic patterns underlying the evolution of COVID-19 cases reported while including the random effect of different age groups and countries as follows:
-
-$$
-\begin{aligned}[t]
-  n\prime_t &\sim \text{Poisson}(\lambda_t) \\
-  \log(\lambda_t) &= \beta_0 + \beta_1\,wave_t + \beta_2\,country_t + \beta_3\,age\,group_t \\
-  &\quad + f_1(closure_t) + f_2(closure_t, country_t) \\
-  &\quad + f_3(closure_t, age\,group_t) + f_4(stringency\,index_t) \\
-  &\quad + f_5(government\,response\,index_t) + f_6(containment\,health\,index_t) \\
-  &\quad + f_7(economic\,support\,index_t) +\epsilon_t
-\end{aligned}
-$${#eq-gam}
-
-\noindent where $n\prime_t$ represents the confirmed COVID-19 cases, assuming a Poisson distribution for the fitting [@loader2006local], and $t$ is the date corresponding to the confirmed COVID-19 cases.
-The response variable includes a specific random effect taking into account variation within waves of school closure, countries, and age groups.
-The terms $f_1$ to $f_3$ are smooth functions of the time since closure, the time since closure for each country, and the time since closure for each age group.
-The terms $f_4$ to $f_7$ are random effects corresponding to the average stringency index, the average government response index, the average containment health index, and the average economic support index.
-The restricted maximum likelihood (REML) was used to avoid over fitting while estimating smoothing parameters.
-In order to accurately account for the autocorrelation arising from the time series data, the residuals are modelled using an AR1 error model such as $\epsilon_t = \phi \epsilon_{t-1} + \eta_t, \quad \eta_t \sim \mathcal{N}(0, \sigma^2)$ where $\phi$ is the autoregressive parameter.
-By incorporating the autoregressive component, the AR1 model acknowledges the dependence of each residual on its previous value, thus providing a comprehensive representation of the data’s temporal dynamics.
-
-Chi-square statistics were employed to assess if the degree of smoothness is significantly different from zero.
-
-## Transfer entropy
-
-Transfer entropy ($T$) is a measure of the directional information flow between two time series $X$ and $Y$, capturing how the past values of one variable can predict changes in another.
-Unlike correlation, transfer entropy accounts for the temporal order of events and non-linear relationships, making it particularly suited for studying dynamic systems such as epidemic data.
-In this analysis, transfer entropy quantifies how case counts in one age group $X$ influence subsequent case counts in another age group $Y$, providing insights into directional transmission patterns.
-If it does, $T$ is considered evidence of a causal effect from the age group $X$ to the age group $Y$ [@schreiber2000measuring].
-As such, Granger causality is a special case of transfer entropy applied to time series that are jointly Gaussian distributed [@barnett2009granger].
-Therefore, transfer entropy is a more robust analysis of time series, especially when applied to the impact of age cohorts on pandemic transmission [@kissler2020symbolic].
-
-The influence of the evolution in COVID-19 cases across all age groups is evaluated using Shannon’s transfer entropy, given by:
-
-$$
-\begin{aligned}[t]
-  T_{ag\,x \rightarrow ag\,y}(k,l) = \sum_{ag\,x_{t+1}, ag\,x_t^{(k)}, ag\,y_t^{(l)}} 
-  p\left(ag\,x_{t+1}, ag\,x_t^{(k)}, ag\,y_t^{(l)}\right) 
-  \log \left(\frac{p\left(ag\,x_{t+1} \mid ag\,x_t^{(k)}, ag\,y_t^{(l)}\right)}{p\left(ag\,x_{t+1} \mid ag\,x_t^{(k)}\right)}\right)
-\end{aligned}
-$$ {#eq-te}
-
-\noindent where $T_{ag\,x \rightarrow\,ag\,y}$ consequently measures the influence of the change dynamic from an age group $X$ (or $ag\,x$) to another age group $Y$ (or $ag\,y$) for every country (Eq @eq-te).
-
-The day-by-day difference in COVID-19 confirmed cases $n\prime\prime_t$ is used to satisfy the stationary requirement for the calculation of Shannon’s Transfer Entropy [@shannon1948mathematical; @behrendt2019rtransferentropy].
-
-# Results
-
-The data show that the trend of confirmed COVID-19 cases follows similar patterns across the selected European countries, with scales following the size of the population in these countries (@fig-overall).
-Thus, the bigger the country, the higher the total number of COVID-19 cases.
-
-```{r, fig.height=3, out.width="\\textwidth"}
-#| label: fig-overall
-#| fig.cap: "Cumulative COVID-19 cases number for selected European countries since the beginning of the pandemic. Source: COVerAGE-DB [@riffe2021data]."
-
-overall_plot <- df_global |> 
-  mutate(
-    country = fct_rev(country)
-  ) |> 
-  ggplot(aes(date, cumulative_cases_cal, group = country)) +
-  geom_line(na.rm = TRUE, color = "gray40", alpha = 0.4) +
-  geom_text_repel(
-    data = df_global |>
-      group_by(country) |>
-      slice_max(cumulative_cases_cal, with_ties = FALSE),
-    aes(label = country),
-    hjust = 0,
-    direction = "y",
-    segment.color = NA,
-    seed = 123,
-    size = 2,
-    force = 0.5,
-    family = "serif"
-  ) +
-  scale_x_date(
-    "Date", 
-    date_breaks = "3 months", 
-    date_labels = "%b %y", 
-    limits = c(as.Date("2020-02-01"), as.Date("2022-05-01"))
-  ) +
-  scale_y_continuous(
-    "Total COVID-19 Cases (Million)", 
-    labels = unit_format(unit = "M", scale = 1e-6)
-  ) +
-  theme_bw() +
-  theme(
-    legend.position = "none",
-    text = element_text(size = 10, family = "serif")
-  )
-
-overall_plot
-```
-
-The evolution of COVID-19 cases reveals some similarities across all age groups.
-However, the influence of each wave on individual age groups also has some particularities (@fig-descriptive).
-For example, the first wave was more important among the oldest age groups, whereas the third wave was more important among the youngest age groups.
-
-```{r, fig.height=3, out.width="\\textwidth"}
-#| label: fig-descriptive
-#| fig.cap: "Periods of school closure since the beginning of the COVID-19 pandemic for selected European countries and their reason: regular academic break versus closure due to government decisions.
-Source: @unesco2022https."
-
-descriptive_plot <- df_global |> 
-  filter(schools_closed == 1) |> 
-  mutate(
-    status = status |> 
-      str_remove("Closed due to ") |> 
-      fct_rev(),
-    country = country |> 
-      fct_rev()
-  ) |> 
-  ggplot(aes(date, country, color = status)) +
-  geom_point(na.rm = TRUE) +
-  scale_x_date("Date", date_breaks = "3 months", date_labels = "%b %y") +
-  scale_y_discrete("Selected European Countries") +
-  scale_colour_manual("Schools Closure Reason", values = c("gray10", "gray60")) +
-  theme_bw() +
-  theme(
-    legend.position = "bottom",
-    text = element_text(size = 10, family = "serif")
-  )
-
-descriptive_plot
-```
-
-As outlined above, to evaluate the shape of the trend in the numbers of COVID-19 cases reported after the three school closures longer than 21 consecutive days, a GAM was fitted as described above, taking into account the overall effect across all the selected European countries, as well as the effect for age groups: 0 to 4, 5 to 9, 10 to 14, 15 to 19, and 20 to 24 years old.
-The obtained results satisfy the requirements to fit this model, which explains `r percent(gam_summary$dev.expl, accuracy = 0.1)` of the variation in COVID-19 cases (@fig-age).
-
-```{r, fig.height=3, out.width="\\textwidth"}
-#| label: fig-age
-#| fig.cap: "Standardized effect of the smooth term in Generalized Additive Model by age group. Standardized effects are reported to compare the shape of the curve between age groups."
-
-gam_plot_age
-```
-
-Overall, the results revealed a decreasing non-linear effect of school closure at a country level for the selected European countries (Austria: `r gam_results["Austria"]`; Belgium: `r gam_results["Belgium"]`; Bulgaria: `r gam_results["Bulgaria"]`; Croatia: `r gam_results["Croatia"]`; Estonia: `r gam_results["Estonia"]`; France: `r gam_results["France"]`; Germany (`r gam_results["Germany"]`); Greece: `r gam_results["Greece"]`; The Netherlands: `r gam_results["Netherlands"]`; Portugal: `r gam_results["Portugal"]`; Slovakia (`r gam_results["Slovakia"]`; Spain: `r gam_results["Spain"]`, see @fig-country.
-
-```{r, fig.height=3, out.width="\\textwidth"}
-#| label: fig-country
-#| fig.cap: "Standardized effect of the smooth term in Generalized Additive Model by country. Standardized effects are reported to compare the shape of the curve between countries."
-
-gam_plot_country
-```
-
-While the general patterns described above were evident in the aggregated analysis, country-specific smooths indicated some national variations in the timing and magnitude of these age-specific effects, although the paradoxical increase in the 5-14 age cohorts was observed in a majority of countries.
-
-```{r}
-#| label: fig-te
-#| fig.cap: "Matrix of Transfer Entropy coefficients according to every age group combination for each of the selected European country.
-Age groups on the x-axis are influencing the age groups on the y-axis ($ag\\,x \\rightarrow\\,ag\\,y$).
-The significance of each Transfer Entropy coefficient is provided in Appendix 2."
-#| fig.height: 7
-
-te_plot
-```
-
-Considering all countries, the analysis of age groups reveals distinct patterns (@fig-te).
-For the 0 to 4 age group, i.e., the pre-school group, there was a downward trend during the initial three weeks following school closure, followed by an increase during the fourth week (`r gam_results["0to4"]`).
-A similar, but more moderate, trend is observer for the age group ranging from 5 to 9 (`r gam_results["5to9"]`), where a decrease is observed during the first two weeks, followed by an increase.
-while the trend for the age group ranging from 10 to 14 is flat during the first two weeks, a similar increase is shown at the beginning of the third week of school closure (`r gam_results["10to14"]`).
-For age groups between 15 and 24, there is a notable surge immediately after school closure, followed by a decrease after 14 days (15 to 19 age group: `r gam_results["15to19"]` and 20 to 24 age group: `r gam_results["20to24"]`).
-Thus, while the total case number declined following school closures, among younger age groups only the pre-school age group had a consistent reduction in cases, while for other age groups, there was even an increase in cases following school closures.
-
-Given that school closures were not only aimed at reducing COVID-19 cases among school-going children but also age groups, it is important to test the degree to which there was transmission across age groups, which is done using transfer entropy analysis.
-Before this analysis is carried out, an Augmented Dickey-Fuller has been applied to each age group to ensure that the daily changes in COVID-19 cases are stationary (see Appendix 1).
-
-The results of the transfer entropy calculations between age groups for each of the 12 selected European countries are reported in @fig-te.
-The absence of symmetry between influencing age groups (i.e., $ag\,x$) and influenced age groups (i.e., $ag\,y$) is found.
-Indeed, the change in COVID-19 cases in some age groups is influenced by other age groups, but they are not reciprocally influencing these age groups.
-F@fig-te shows how different age cohorts influence the COVID-19 case numbers of all age groups across countries.
-The upper left quadrant indicates how younger age cohorts are influencing older age cohorts, the bottom right quadrant indicates how older age cohorts are influencing younger age cohorts, finally, the lower left and upper right indicate how younger or older age cohorts are influencing themselves.
-By analyzing these quadrants, it is possible to identify similar patterns across multiple countries.
-Indeed, it appears that Austria, Germany, and The Netherlands have significantly higher $T$ coefficients in the upper left quadrant of the matrix, which indicates that the daily changes in COVID-19 cases number in younger cohorts are predicting the daily changes in COVID-19 cases number in older cohorts.
-Alternatively, it appears that Austria, the Netherlands, Portugal, and Spain have significantly higher $T$ coefficients in the lower right quadrant of the matrix, which indicates that the daily changes in COVID-19 cases number in older cohorts are predicting the daily changes in COVID-19 cases number in younger cohorts.
-
-# Discussion
-
-Knowledge about the transmission of the virus significantly improved over time as more studies have been published.
-While an early study found that children did not play an important role in the transmission of the virus [@li2020role], more recent results give a more nuanced position, stating that the spread of the virus in children is moderate.
-It is now suggested that the impact of school openings on infection rates was limited, with children and teachers more often contracting infections through household or community contacts rather than within schools [@soriano2023policies].
-This study, using robust statistical methods, considered the effect of school closures on COVID-19 cases across age groups.
-While there are commonalities in the evolution of COVID-19 cases across all European countries that were included in our data, school closures exhibit distinct impacts on different age groups.
-Notably, the analysis shows that the 0 to 4 age group experiences a downward trend in COVID-19 cases during the initial two weeks following school closure, followed by stabilization.
-In contrast, age groups ranging from 5 to 14 exhibit a stable profile after school closure.
-Additionally, age groups between 15 and 24 demonstrate a notable surge immediately after school closure, followed by a decrease after 14 days.
-Thus, the results do not support the hypothesis that school closures were effective for group ages older than 5 years old.
-One possibility is that school closures may lead to increased informal social gatherings among children and teenagers outside of controlled school environments, potentially facilitating virus transmission.
-It is also possible that transmission within households intensified as children spent more time at home.
-These results partially replicate observations from @alfano2022effects while providing a clearer picture of the structure of the effect of school closure.
-
-The Transfer Entropy calculations between age groups for each of the 12 selected European countries reveal an absence of symmetry between influencing age groups ($X$) and influenced age groups ($Y$).
-Some age groups influence the COVID-19 case numbers of other age groups, but there is no reciprocal influence.
-This finding suggests that changes in COVID-19 cases in certain age groups can predict the changes in other age groups but not vice versa, i.e., causality can be unidirectional.
-These findings highlight the importance of considering intergenerational interactions in designing effective control measures.
-
-The entropy analysis revealed intriguing directional patterns of case counts between age groups.
-In some countries, cases among younger individuals appear predictive of cases among older individuals, while the opposite pattern is observed elsewhere.
-These differences may reflect country-specific variations in social mixing patterns, healthcare responses, or policy measures.
-The absence of consistent patterns in infections among adjacent age groups raises questions about the mechanisms of transmission and suggests that factors beyond simple proximity, such as behavioural differences or differing susceptibility, may play a role.
-Future work should investigate these underlying mechanisms to better understand age-based transmission dynamics.
-
-Furthermore, the quadrant analysis reveals similar patterns across multiple countries.
-Austria, Germany, and the Netherlands exhibit significantly higher TE coefficients in the upper left quadrant, indicating that changes in COVID-19 cases in younger cohorts predict the changes in older cohorts.
-On the other hand, Austria, the Netherlands, Portugal, and Spain show higher TE coefficients in the lower right quadrant, indicating that changes in COVID-19 cases in older cohorts predict the changes in younger cohorts.
-However, these differences may also reflect country-specific variations in social mixing patterns, healthcare responses, or policy measures.
-The absence of consistent patterns in infections among adjacent age groups raises questions about the mechanisms of transmission and suggests that factors beyond simple proximity, such as behavioural differences or differing susceptibility, may play a role.
-
-This study, despite its methodological strengths, is subject to certain limitations that warrant consideration.
-Foremost among these is the inherent challenge of perfectly isolating the effect of school closures from other simultaneously implemented and dynamically changing NPIs.
-While we incorporated four composite NPI index derived from the Oxford COVID-19 Government Response Tracker (OxCGRT) project to account for the broader public health context, residual confounding from unmeasured aspects of NPIs or behavioural changes may still persist.
-Our sensitivity analyses showed consistent directional effects for school closures even under varying overall NPI indexes provide some confidence, but this remains an inherent complexity in observational studies of this nature.
-Another consideration is the ecological design of the study, which uses aggregated country-level data; this may mask important sub-national variations and precludes inferences about individual-level risk factors.
-In addition, the definition of "closure" which excluded "partially open" school statuses to simplify the intervention definition, means the impact of such intermediate states was not captured.
-Variations in testing strategies, case definitions, and reporting practices across countries and over time could also have influenced the observed case counts, despite efforts to use standardized data sources.
-Lastly, this study predominantly covers periods before widespread COVID-19 vaccination and the emergence of later variants such as Omicron; the effects of school closures might differ under those altered epidemiological conditions.
-
-Future studies should look more closely at why the virus spread between age groups as it did, and why TE patterns differed by country.
-This could involve using information on people's behaviour or more detailed data on other health rules.
-It would also be useful to study how different virus types and vaccination levels change these effects.
-Learning more about other influencing factors will help us better understand the real impact of school closures.
-
-# Conclusion
-
-This comprehensive analysis of COVID-19 school closures across 12 European nations reveals a complex and often counter-intuitive reality that challenges prevailing assumptions about their universal efficacy, particularly for children aged 5 and older.
-While national-level data showed an overall decrease in cases during closures, our age-disaggregated analysis using GAMs demonstrated that school-aged children (5-14 years) and young adults (15-24 years) frequently experienced periods of increased infections post-closure, a stark contrast to the consistent reductions seen only in pre-schoolers.
-Furthermore, Transfer Entropy analysis highlighted that the directional influence of transmission between age groups was asymmetric and varied by country, suggesting that the role of children in broader community transmission during such periods is not uniform.
-
-These findings underscore the critical importance of nuanced, age-specific analysis in evaluating public health interventions and caution against the broad application of measures like school closures without robust evidence of their benefit across all targeted demographics and a clear understanding of potential unintended consequences [@alfano2020efficacy; @molefi2021impact].
-For future pandemic preparedness, a more targeted, evidence-based approach is essential, prioritising measures that minimise societal disruption while effectively protecting public health across all age strata.
-
-# Author statements
-
-## Ethical approval
-
-This study is a longitudinal analysis on available data; thus, it does not require ethical approval.
-
-## Funding
-
-None declared.
-
-## Competing interests
-
-None declared.
-
-## Data Availability
-
-The data can be accessed from the COVerAGE-DB OSF repository (https://osf.io/mpwjq/), from the Oxford Covid-19 Government Response Tracker (OxCGRT) "covid-policy-dataset" GitHub repository (https://github.com/OxCGRT/covid-policy-dataset), and from the UNESCO servers (https://en.unesco.org/file/unesco-data-school-closures-february-2020-june-2022csv-zip).
-All preprocessing, analyses and code used to build the submitted paper are available at https://github.com/[hidden for anonymous peer-review].
-
-# References
-
-::: {#refs}
-:::

### 2025-06-05 manuscript_njd/title_page.tex
update

--- 2025-06-05 manuscript_njd/manuscript.tex
@@
 \maketitle
-\section{Introduction}\label{introduction}
-
-The COVID-19 pandemic has had a profound impact on global health, with
-an estimate of 14.83 million excess deaths globally
-\citep{msemburi2023estimates}. Beyond direct mortality, the pandemic has
-caused significant collateral damage, including losses of lives and
-livelihoods, necessitating a comprehensive approach to measuring its
-broader impacts. A wide range of non-pharmaceutical interventions (NPI)
-were enacted to control the spread of COVID-19, especially before the
-availability of effective vaccines. Due to their social mixing patterns,
-children have been identified as an age group that can drive the spread
-of respiratory infections such as influenza \citep{moser2018estimating},
-and school closures were found to be effective in mitigating the spread
-of influenza H1N1 in Japan in 2009 \citep{kawano2015substantial}. It is
-therefore not surprising that school closures were one of the NPIs that
-were introduced in many countries to stop the spread of COVID-19. They
-were used particularly during the initial wave of the pandemic and
-during subsequent waves when case numbers were high.
-
-Recently, a body of literature has shown various negative consequences
-on child development and health due to school closures. School closures
-have been found to have negatively affected child mental health
-\citep{moulin2022longitudinal}, nutrition/obesity
-\citep{sugimoto2023temporal}, and education
-\citep{lerkkanen2023reading}. Some governments were questioning their
-decisions to close schools \citep{de2021determines}. For example, the
-German health minister Karl Lauterbach, in an interview with one of
-Germany's publicly funded television stations, admitted that ``in
-retrospect it had been wrong to keep schools and childcare closed for so
-long'' \citep{ard2023lauterbach}. Even if school closures had negative
-consequences for children, they might nevertheless have been effective
-at stopping the spread of COVID-19. However, analysis on the
-effectiveness of school closures on the spread of COVID-19 remains
-inconclusive. In a study of a panel of European countries,
-\citet{alfano2022effects} found that school closures were associated
-with a reduced COVID-19 incidence. In contrast, \citet{walsh2021school}
-in a review of 40 studies covering 150 countries concluded that the
-effectiveness of school closures was uncertain, with 60\% having
-identified no impact and pointing to the potential for the analysis to
-be affected by confounding factors and collinearity. The latter
-shortcomings suggest that it is hard to draw firm inferences, as even a
-positive result may not imply causality.
-
-Once it was clear that COVID-19 spreads from person to person, it was
-natural that governments were advised to enact measures to reduce social
-contact in order to reduce the spread of the virus. Therefore, it is
-legitimate to believe that by closing schools, a reduction of the
-contaminations would be observed in the younger age groups. However, the
-efficiency of school closure on the reduction of COVID-19 cases is still
-questioned \citep{bayham2020impact, esposito2021comprehensive}. While
-some research has found that school closures contribute to limit or to
-reduce the growth rate of confirmed cases after implementation
-\citep{stage2021shut, yoshiyuki2020effects}, others did not observe a
-change in the evolution of COVID-19 cases
-\citep{chang2020modelling, iwata2020school}. For instance, a controlled
-comparison between similar localities in Japan with schools closed and
-schools open did not reveal any evidence that school closures reduced
-the spread of COVID-19 \citep{fukumoto2021no}. If the school closure had
-a real impact on the evolution of confirmed COVID-19 cases, it should be
-possible to observe a decrease or at least an inflection in the trend of
-its evolution among younger age groups.
-
-A second implicit belief regarding the effect of school closure on the
-spread of COVID-19 is that school not only influences the spread of the
-virus among children and teenagers but also has a knock-on effect on the
-spread of the virus in older age groups, also called Secondary Attack
-Rate (SAR). The contaminated children and teenagers would bring the
-virus back home and would pass the virus on to their parents and
-relatives. For example, research investigating the contamination in the
-household network not only revealed an exceptionally high rate of
-secondary contamination but also that this contamination happened when
-the schools were closed \citep{sorianoarandes2021household}. Despite
-being reported in several clinical and epidemiological studies
-\citep{siebach2021childhood, zhendong2020clinical}, multiple research
-have shown that the SAR from children to household members was lower
-than expected
-\citep{heavey2020evidence, vanderhoek2020role, kim2021role, ludvigsson2020children}.
-However, the SAR of children and teenagers to the household member is
-likely to be age-dependent, with differences between infants, primary
-and secondary school children, and college students
-\citep{grasleguen2021reopening}. If a secondary transmission from
-children and teenagers to household member has a significant influence,
-then a temporal causality relationship between their evolution should be
-observed.
-
-A gap in these previous analyses is the insufficient granularity in
-examining age-specific effects and the complex dynamics of inter-age
-group transmission. This study aims to address these deficiencies by
-providing a more robust and nuanced assessment. First, Generalized
-Additive Models (GAMs) are utilized to flexibly model and quantify the
-potentially non-linear evolution of COVID-19 cases following school
-closures, specifically within these fine-grained age brackets. This
-allows us to move beyond simple pre-post comparisons and capture dynamic
-temporal patterns. Then, Transfer Entropy (TE), an information-theoretic
-measure, is applied to assess directional causality in time series data.
-TE is particularly adept at identifying how past states of one variable
-(e.g., cases in one age group) influence future states of another (e.g.,
-cases in a different age group), even in complex, non-linear systems
-like epidemic spread. This dual methodology allows for a deeper
-understanding of both the direct age-specific consequences of school
-closures and the potential for altered intergenerational transmission
-dynamics. Our findings challenge simplistic narratives about school
-closure efficacy and highlight the importance of age-stratified analysis
-for future pandemic preparedness.
-
-\section{Methods}\label{methods}
-
-\subsection{Study design}\label{study-design}
-
-A longitudinal study was conducted to observe changes in COVID-19 cases
-during school closures. Changes in COVID-19 from five age groups from
-the first to the 28th day of school closure are compared across 12
-countries. Then, a causality analysis evaluate the impact of changes in
-these younger groups on older age groups.
-
-\subsection{Data collection}\label{data-collection}
-
-Data on COVID-19 case counts by age group have been obtained from the
-COVerAGE-DB project \citep{riffe2021data}. The COVerAGE-DB which collect
-numbers for every countries by groups of 5 or 10 years if available. In
-addition, spline approximations are used to deals with the heterogeneity
-of countries' reporting formats by using when the data for this age
-bracket is not available for a country. Countries communicating data
-with groups of 5 years were chosen to match as much as possible the
-different school stages.
-
-To identify the impact of school closures on the number of cases,
-information regarding the school closures on a day-by-day basis is
-required, and this is taken from the ``UNESCO global education
-coalition'' (2022). For each day, in each country, the status of the
-schools is indicated as fully open, partially open, closed due to
-COVID-19, or closed due to an academic break. Because it would be
-difficult to measure the effect of school closures at a country level
-when schools are partially closed, only closures due to COVID-19 or due
-to an academic break are considered. Indeed, both are considered as
-closure at a country-wide level. Any COVID-19 case numbers beyond the
-28-day period are not relevant for evaluating the impact of school
-closures.
-
-\subsection{Data validation}\label{data-validation}
-
-In order to cross-validate the data obtained after spline
-approximations, a comparison with the data published by the
-\citet{who2022https} reveals perfect similarities. The original data
-consist of 14,089,320 observations of 10 variables (117 distinct
-countries, region within the country, a unique observation code, the
-date of the observation, the gender, which can be male, female, or both,
-the age bracket by 5 years from 0 to 100, a confirmation of the age
-interval for each bracket, the total number of cases so far, the total
-number of deaths, and the total number of tests performed) from February
-16, 2020 to January 20, 2022. After removing countries with missing and
-inconsistent values, only 22 are suitable for data analyses. However, to
-focus this analysis on geographically and culturally comparable
-countries, only 12 European countries are kept: Austria, Belgium,
-Bulgaria, Croatia, Estonia, France, Germany, Greece, Netherlands,
-Portugal, Slovakia, and Spain. The observations are reported in terms of
-the total number of COVID-19 cases per day from the start of the
-pandemic. The daily number of cases at a specific date \(n\prime_{t}\)
-is calculated with the difference between the total cases at a date
-\(t\) and the total cases at a date \(t-1\) (i.e., derivative 1). In
-addition, the change in the daily number of cases \(n\prime\prime_{t}\)
-between \(n\prime_{t}\) and \(n\prime_{t-1}\) has also been calculated
-(i.e., derivative 2).
-
-\subsection{Controlling for confounding Non-Pharmaceutical
-Interventions}\label{controlling-for-confounding-non-pharmaceutical-interventions}
-
-Recognizing that school closures rarely occurred in isolation, a
-critical component of this analysis was to account for other
-contemporaneous NPIs. To mitigate confounding from other NPIs, we
-incorporated four time-varying composite NPI index identified by the
-Oxford COVID-19 Government Response Tracker (OxCGRT) project
-\citep{hale2021global}: the average stringency index, the average
-government response index, the average containment health index, and the
-average economic support index. Specific data were collected from each
-country on various policy responses governments implemented to curb the
-spread of COVID-19, such as travel restrictions, workplace closures,
-public event cancellations, restrictions on gatherings, or public health
-messaging. These indicators were then aggregated to create indexes,
-reflecting the overall level of government restrictions.
-
-\subsection{Generalized additive
-model}\label{generalized-additive-model}
-
-The effect of school closure on the trend of daily COVID-19 cases is
-analysed using a Generalized Additive Model (GAM). GAM is a flexible
-modelling approach that estimates non-linear relationships between
-variables. Compared to other methods such as Vector Autoregression, GAM
-can handle fixed and random smooth effects without making strict
-assumptions about linearity.
-
-The GAM is fitted on the daily COVID-19 cases to test the hypothesis of
-a significant non-linear evolution of cases among age groups from 0 to
-4, from 5 to 9, from 10 to 14, from 15 to 19, and from 20 to 24
-\citep{wood2017generalized}. The model also estimates the overall
-non-linear effect by country, taking into account the interaction
-between age groups and countries as random intercepts and the
-interaction between time, countries, and the period of closure as random
-effects (Eq \ref{eq-gam}).
-
-By estimating the degree of smoothness of a Bayesian spline smoothing
-using restricted fast maximum likelihood estimation
-\citep{wood2011fast}, GAM identifies dynamic patterns underlying the
-evolution of COVID-19 cases reported while including the random effect
-of different age groups and countries as follows:
-
-\begin{equation}\phantomsection\label{eq-gam}{
-\begin{aligned}[t]
-  n\prime_t &\sim \text{Poisson}(\lambda_t) \\
-  \log(\lambda_t) &= \beta_0 + \beta_1\,wave_t + \beta_2\,country_t + \beta_3\,age\,group_t \\
-  &\quad + f_1(closure_t) + f_2(closure_t, country_t) \\
-  &\quad + f_3(closure_t, age\,group_t) + f_4(stringency\,index_t) \\
-  &\quad + f_5(government\,response\,index_t) + f_6(containment\,health\,index_t) \\
-  &\quad + f_7(economic\,support\,index_t) +\epsilon_t
-\end{aligned}
-}\end{equation}
-
-\noindent where \(n\prime_t\) represents the confirmed COVID-19 cases,
-assuming a Poisson distribution for the fitting \citep{loader2006local},
-and \(t\) is the date corresponding to the confirmed COVID-19 cases. The
-response variable includes a specific random effect taking into account
-variation within waves of school closure, countries, and age groups. The
-terms \(f_1\) to \(f_3\) are smooth functions of the time since closure,
-the time since closure for each country, and the time since closure for
-each age group. The terms \(f_4\) to \(f_7\) are random effects
-corresponding to the average stringency index, the average government
-response index, the average containment health index, and the average
-economic support index. The restricted maximum likelihood (REML) was
-used to avoid over fitting while estimating smoothing parameters. In
-order to accurately account for the autocorrelation arising from the
-time series data, the residuals are modelled using an AR1 error model
-such as
-\(\epsilon_t = \phi \epsilon_{t-1} + \eta_t, \quad \eta_t \sim \mathcal{N}(0, \sigma^2)\)
-where \(\phi\) is the autoregressive parameter. By incorporating the
-autoregressive component, the AR1 model acknowledges the dependence of
-each residual on its previous value, thus providing a comprehensive
-representation of the data's temporal dynamics.
-
-Chi-square statistics were employed to assess if the degree of
-smoothness is significantly different from zero.
-
-\subsection{Transfer entropy}\label{transfer-entropy}
-
-Transfer entropy (\(T\)) is a measure of the directional information
-flow between two time series \(X\) and \(Y\), capturing how the past
-values of one variable can predict changes in another. Unlike
-correlation, transfer entropy accounts for the temporal order of events
-and non-linear relationships, making it particularly suited for studying
-dynamic systems such as epidemic data. In this analysis, transfer
-entropy quantifies how case counts in one age group \(X\) influence
-subsequent case counts in another age group \(Y\), providing insights
-into directional transmission patterns. If it does, \(T\) is considered
-evidence of a causal effect from the age group \(X\) to the age group
-\(Y\) \citep{schreiber2000measuring}. As such, Granger causality is a
-special case of transfer entropy applied to time series that are jointly
-Gaussian distributed \citep{barnett2009granger}. Therefore, transfer
-entropy is a more robust analysis of time series, especially when
-applied to the impact of age cohorts on pandemic transmission
-\citep{kissler2020symbolic}.
-
-The influence of the evolution in COVID-19 cases across all age groups
-is evaluated using Shannon's transfer entropy, given by:
-
-\begin{equation}\phantomsection\label{eq-te}{
-\begin{aligned}[t]
-  T_{ag\,x \rightarrow ag\,y}(k,l) = \sum_{ag\,x_{t+1}, ag\,x_t^{(k)}, ag\,y_t^{(l)}} 
-  p\left(ag\,x_{t+1}, ag\,x_t^{(k)}, ag\,y_t^{(l)}\right) 
-  \log \left(\frac{p\left(ag\,x_{t+1} \mid ag\,x_t^{(k)}, ag\,y_t^{(l)}\right)}{p\left(ag\,x_{t+1} \mid ag\,x_t^{(k)}\right)}\right)
-\end{aligned}
-}\end{equation}
-
-\noindent where \(T_{ag\,x \rightarrow\,ag\,y}\) consequently measures
-the influence of the change dynamic from an age group \(X\) (or
-\(ag\,x\)) to another age group \(Y\) (or \(ag\,y\)) for every country
-(Eq \ref{eq-te}).
-
-The day-by-day difference in COVID-19 confirmed cases
-\(n\prime\prime_t\) is used to satisfy the stationary requirement for
-the calculation of Shannon's Transfer Entropy
-\citep{shannon1948mathematical, behrendt2019rtransferentropy}.
-
-\section{Results}\label{results}
-
-The data show that the trend of confirmed COVID-19 cases follows similar
-patterns across the selected European countries, with scales following
-the size of the population in these countries
-(Figure~\ref{fig-overall}). Thus, the bigger the country, the higher the
-total number of COVID-19 cases.
-
-\begin{figure}[h]
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{manuscript_files/figure-pdf/fig-overall-1.png}}
-
-}
-
-\caption{\label{fig-overall}Cumulative COVID-19 cases number for
-selected European countries since the beginning of the pandemic. Source:
-COVerAGE-DB \citep{riffe2021data}.}
-
-\end{figure}%
-
-The evolution of COVID-19 cases reveals some similarities across all age
-groups. However, the influence of each wave on individual age groups
-also has some particularities (Figure~\ref{fig-descriptive}). For
-example, the first wave was more important among the oldest age groups,
-whereas the third wave was more important among the youngest age groups.
-
-\begin{figure}[h]
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{manuscript_files/figure-pdf/fig-descriptive-1.png}}
-
-}
-
-\caption{\label{fig-descriptive}Periods of school closure since the
-beginning of the COVID-19 pandemic for selected European countries and
-their reason: regular academic break versus closure due to government
-decisions. Source: \citet{unesco2022https}.}
-
-\end{figure}%
-
-As outlined above, to evaluate the shape of the trend in the numbers of
-COVID-19 cases reported after the three school closures longer than 21
-consecutive days, a GAM was fitted as described above, taking into
-account the overall effect across all the selected European countries,
-as well as the effect for age groups: 0 to 4, 5 to 9, 10 to 14, 15 to
-19, and 20 to 24 years old. The obtained results satisfy the
-requirements to fit this model, which explains 80.2\% of the variation
-in COVID-19 cases (Figure~\ref{fig-age}).
-
-\begin{figure}[h]
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{manuscript_files/figure-pdf/fig-age-1.png}}
-
-}
-
-\caption{\label{fig-age}Standardized effect of the smooth term in
-Generalized Additive Model by age group. Standardized effects are
-reported to compare the shape of the curve between age groups.}
-
-\end{figure}%
-
-Overall, the results revealed a decreasing non-linear effect of school
-closure at a country level for the selected European countries (Austria:
-\(\chi^2(5.82) = 3243.01\), \(p < 0.001\); Belgium:
-\(\chi^2(5.98) = 6206.45\), \(p < 0.001\); Bulgaria:
-\(\chi^2(5.89) = 258.19\), \(p < 0.001\); Croatia:
-\(\chi^2(5.1) = 740.24\), \(p < 0.001\); Estonia:
-\(\chi^2(5.93) = 1834.26\), \(p < 0.001\); France:
-\(\chi^2(1) = 2412.59\), \(p < 0.001\); Germany
-(\(\chi^2(5.98) = 3081.05\), \(p < 0.001\)); Greece:
-\(\chi^2(5.95) = 2423.67\), \(p < 0.001\); The Netherlands:
-\(\chi^2(4.97) = 5878.92\), \(p < 0.001\); Portugal:
-\(\chi^2(5.94) = 2692.56\), \(p < 0.001\); Slovakia
-(\(\chi^2(5.97) = 5335.74\), \(p < 0.001\); Spain:
-\(\chi^2(5.94) = 498.09\), \(p < 0.001\), see Figure~\ref{fig-country}.
-
-\begin{figure}[h]
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{manuscript_files/figure-pdf/fig-country-1.png}}
-
-}
-
-\caption{\label{fig-country}Standardized effect of the smooth term in
-Generalized Additive Model by country. Standardized effects are reported
-to compare the shape of the curve between countries.}
-
-\end{figure}%
-
-While the general patterns described above were evident in the
-aggregated analysis, country-specific smooths indicated some national
-variations in the timing and magnitude of these age-specific effects,
-although the paradoxical increase in the 5-14 age cohorts was observed
-in a majority of countries.
-
-\begin{figure}[h]
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{manuscript_files/figure-pdf/fig-te-1.png}}
-
-}
-
-\caption{\label{fig-te}Matrix of Transfer Entropy coefficients according
-to every age group combination for each of the selected European
-country. Age groups on the x-axis are influencing the age groups on the
-y-axis (\(ag\,x \rightarrow\,ag\,y\)). The significance of each Transfer
-Entropy coefficient is provided in Appendix 2.}
-
-\end{figure}%
-
-Considering all countries, the analysis of age groups reveals distinct
-patterns (Figure~\ref{fig-te}). For the 0 to 4 age group, i.e., the
-pre-school group, there was a downward trend during the initial three
-weeks following school closure, followed by an increase during the
-fourth week (\(\chi^2(5.96) = 410.48\), \(p < 0.001\)). A similar, but
-more moderate, trend is observer for the age group ranging from 5 to 9
-(\(\chi^2(5.95) = 878.29\), \(p < 0.001\)), where a decrease is observed
-during the first two weeks, followed by an increase. while the trend for
-the age group ranging from 10 to 14 is flat during the first two weeks,
-a similar increase is shown at the beginning of the third week of school
-closure (\(\chi^2(1) = 61.62\), \(p < 0.001\)). For age groups between
-15 and 24, there is a notable surge immediately after school closure,
-followed by a decrease after 14 days (15 to 19 age group:
-\(\chi^2(5.97) = 7929.88\), \(p < 0.001\) and 20 to 24 age group:
-\(\chi^2(4.98) = 18386.65\), \(p < 0.001\)). Thus, while the total case
-number declined following school closures, among younger age groups only
-the pre-school age group had a consistent reduction in cases, while for
-other age groups, there was even an increase in cases following school
-closures.
-
-Given that school closures were not only aimed at reducing COVID-19
-cases among school-going children but also age groups, it is important
-to test the degree to which there was transmission across age groups,
-which is done using transfer entropy analysis. Before this analysis is
-carried out, an Augmented Dickey-Fuller has been applied to each age
-group to ensure that the daily changes in COVID-19 cases are stationary
-(see Appendix 1).
-
-The results of the transfer entropy calculations between age groups for
-each of the 12 selected European countries are reported in
-Figure~\ref{fig-te}. The absence of symmetry between influencing age
-groups (i.e., \(ag\,x\)) and influenced age groups (i.e., \(ag\,y\)) is
-found. Indeed, the change in COVID-19 cases in some age groups is
-influenced by other age groups, but they are not reciprocally
-influencing these age groups. F@fig-te shows how different age cohorts
-influence the COVID-19 case numbers of all age groups across countries.
-The upper left quadrant indicates how younger age cohorts are
-influencing older age cohorts, the bottom right quadrant indicates how
-older age cohorts are influencing younger age cohorts, finally, the
-lower left and upper right indicate how younger or older age cohorts are
-influencing themselves. By analyzing these quadrants, it is possible to
-identify similar patterns across multiple countries. Indeed, it appears
-that Austria, Germany, and The Netherlands have significantly higher
-\(T\) coefficients in the upper left quadrant of the matrix, which
-indicates that the daily changes in COVID-19 cases number in younger
-cohorts are predicting the daily changes in COVID-19 cases number in
-older cohorts. Alternatively, it appears that Austria, the Netherlands,
-Portugal, and Spain have significantly higher \(T\) coefficients in the
-lower right quadrant of the matrix, which indicates that the daily
-changes in COVID-19 cases number in older cohorts are predicting the
-daily changes in COVID-19 cases number in younger cohorts.
-
-\section{Discussion}\label{discussion}
-
-Knowledge about the transmission of the virus significantly improved
-over time as more studies have been published. While an early study
-found that children did not play an important role in the transmission
-of the virus \citep{li2020role}, more recent results give a more nuanced
-position, stating that the spread of the virus in children is moderate.
-It is now suggested that the impact of school openings on infection
-rates was limited, with children and teachers more often contracting
-infections through household or community contacts rather than within
-schools \citep{soriano2023policies}. This study, using robust
-statistical methods, considered the effect of school closures on
-COVID-19 cases across age groups. While there are commonalities in the
-evolution of COVID-19 cases across all European countries that were
-included in our data, school closures exhibit distinct impacts on
-different age groups. Notably, the analysis shows that the 0 to 4 age
-group experiences a downward trend in COVID-19 cases during the initial
-two weeks following school closure, followed by stabilization. In
-contrast, age groups ranging from 5 to 14 exhibit a stable profile after
-school closure. Additionally, age groups between 15 and 24 demonstrate a
-notable surge immediately after school closure, followed by a decrease
-after 14 days. Thus, the results do not support the hypothesis that
-school closures were effective for group ages older than 5 years old.
-One possibility is that school closures may lead to increased informal
-social gatherings among children and teenagers outside of controlled
-school environments, potentially facilitating virus transmission. It is
-also possible that transmission within households intensified as
-children spent more time at home. These results partially replicate
-observations from \citet{alfano2022effects} while providing a clearer
-picture of the structure of the effect of school closure.
-
-The Transfer Entropy calculations between age groups for each of the 12
-selected European countries reveal an absence of symmetry between
-influencing age groups (\(X\)) and influenced age groups (\(Y\)). Some
-age groups influence the COVID-19 case numbers of other age groups, but
-there is no reciprocal influence. This finding suggests that changes in
-COVID-19 cases in certain age groups can predict the changes in other
-age groups but not vice versa, i.e., causality can be unidirectional.
-These findings highlight the importance of considering intergenerational
-interactions in designing effective control measures.
-
-The entropy analysis revealed intriguing directional patterns of case
-counts between age groups. In some countries, cases among younger
-individuals appear predictive of cases among older individuals, while
-the opposite pattern is observed elsewhere. These differences may
-reflect country-specific variations in social mixing patterns,
-healthcare responses, or policy measures. The absence of consistent
-patterns in infections among adjacent age groups raises questions about
-the mechanisms of transmission and suggests that factors beyond simple
-proximity, such as behavioural differences or differing susceptibility,
-may play a role. Future work should investigate these underlying
-mechanisms to better understand age-based transmission dynamics.
-
-Furthermore, the quadrant analysis reveals similar patterns across
-multiple countries. Austria, Germany, and the Netherlands exhibit
-significantly higher TE coefficients in the upper left quadrant,
-indicating that changes in COVID-19 cases in younger cohorts predict the
-changes in older cohorts. On the other hand, Austria, the Netherlands,
-Portugal, and Spain show higher TE coefficients in the lower right
-quadrant, indicating that changes in COVID-19 cases in older cohorts
-predict the changes in younger cohorts. However, these differences may
-also reflect country-specific variations in social mixing patterns,
-healthcare responses, or policy measures. The absence of consistent
-patterns in infections among adjacent age groups raises questions about
-the mechanisms of transmission and suggests that factors beyond simple
-proximity, such as behavioural differences or differing susceptibility,
-may play a role.
-
-This study, despite its methodological strengths, is subject to certain
-limitations that warrant consideration. Foremost among these is the
-inherent challenge of perfectly isolating the effect of school closures
-from other simultaneously implemented and dynamically changing NPIs.
-While we incorporated four composite NPI index derived from the Oxford
-COVID-19 Government Response Tracker (OxCGRT) project to account for the
-broader public health context, residual confounding from unmeasured
-aspects of NPIs or behavioural changes may still persist. Our
-sensitivity analyses showed consistent directional effects for school
-closures even under varying overall NPI indexes provide some confidence,
-but this remains an inherent complexity in observational studies of this
-nature. Another consideration is the ecological design of the study,
-which uses aggregated country-level data; this may mask important
-sub-national variations and precludes inferences about individual-level
-risk factors. In addition, the definition of ``closure'' which excluded
-``partially open'' school statuses to simplify the intervention
-definition, means the impact of such intermediate states was not
-captured. Variations in testing strategies, case definitions, and
-reporting practices across countries and over time could also have
-influenced the observed case counts, despite efforts to use standardized
-data sources. Lastly, this study predominantly covers periods before
-widespread COVID-19 vaccination and the emergence of later variants such
-as Omicron; the effects of school closures might differ under those
-altered epidemiological conditions.
-
-Future studies should look more closely at why the virus spread between
-age groups as it did, and why TE patterns differed by country. This
-could involve using information on people's behaviour or more detailed
-data on other health rules. It would also be useful to study how
-different virus types and vaccination levels change these effects.
-Learning more about other influencing factors will help us better
-understand the real impact of school closures.
-
-\section{Conclusion}\label{conclusion}
-
-This comprehensive analysis of COVID-19 school closures across 12
-European nations reveals a complex and often counter-intuitive reality
-that challenges prevailing assumptions about their universal efficacy,
-particularly for children aged 5 and older. While national-level data
-showed an overall decrease in cases during closures, our
-age-disaggregated analysis using GAMs demonstrated that school-aged
-children (5-14 years) and young adults (15-24 years) frequently
-experienced periods of increased infections post-closure, a stark
-contrast to the consistent reductions seen only in pre-schoolers.
-Furthermore, Transfer Entropy analysis highlighted that the directional
-influence of transmission between age groups was asymmetric and varied
-by country, suggesting that the role of children in broader community
-transmission during such periods is not uniform.
-
-These findings underscore the critical importance of nuanced,
-age-specific analysis in evaluating public health interventions and
-caution against the broad application of measures like school closures
-without robust evidence of their benefit across all targeted
-demographics and a clear understanding of potential unintended
-consequences \citep{alfano2020efficacy, molefi2021impact}. For future
-pandemic preparedness, a more targeted, evidence-based approach is
-essential, prioritising measures that minimise societal disruption while
-effectively protecting public health across all age strata.
-
-\section{Author statements}\label{author-statements}
-
-\subsection{Ethical approval}\label{ethical-approval}
-
-This study is a longitudinal analysis on available data; thus, it does
-not require ethical approval.
-
-\subsection{Funding}\label{funding}
-
-None declared.
-
-\subsection{Competing interests}\label{competing-interests}
-
-None declared.
-
-\subsection{Data Availability}\label{data-availability}
-
-The data can be accessed from the COVerAGE-DB OSF repository
-(https://osf.io/mpwjq/), from the Oxford Covid-19 Government Response
-Tracker (OxCGRT) ``covid-policy-dataset'' GitHub repository
-(https://github.com/OxCGRT/covid-policy-dataset), and from the UNESCO
-servers
-(https://en.unesco.org/file/unesco-data-school-closures-february-2020-june-2022csv-zip).
-All preprocessing, analyses and code used to build the submitted paper
-are available at https://github.com/damien-dupre/covid\_ts\_causality.
-
-\section{References}\label{references}
-
-\renewcommand{\bibsection}{}
-\bibliography{bibliography.bib}
 

### 2025-06-19 manuscript/appendix_1.Rmd
update

@@
-# Augmented Dickey-Fuller test for stationay timeseries
+---
+title: "Appendix 1: Dickey-Fuller test for stationay timeseries"
+toc: false
+
+output:
+  bookdown::pdf_book:
+    keep_tex: true
+
+header-includes:
+  - \usepackage{float}
+  - \usepackage{booktabs}
+  - \usepackage{longtable}
+  - \usepackage{threeparttable}
+  - \floatplacement{figure}{H}
+  - \floatplacement{table}{H}
+  - \usepackage{caption}
+---
+
+```{r setup, include = FALSE}
+# libraries --------------------------------------------------------------------
+library(here)
+library(tidyverse)
+library(knitr)
+library(kableExtra)
+
+# options ----------------------------------------------------------------------
+set.seed(123) # Seed for random number generation
+options(scipen = 999) # disable sci number format
+opts_chunk$set(
+  cache.extra = rand_seed, 
+  cache = FALSE,
+  message = FALSE, 
+  warning = FALSE, 
+  error = FALSE,
+  echo = FALSE, 
+  fig.retina = 3,
+  dev = "ragg_png"
+  )
+```
+
+Test for the null hypothesis of a unit root of a univarate time series (equivalently, non-stationary time series).
+
+The Augmented Dickey-Fuller test incorporates three types of linear regression models.
+The first type is a linear model with no drift and linear trend with respect to time.
+The second type is a linear model with drift but no linear trend.
+The third type is a linear model with both drift and linear trend.
+
+The p.value is calculated by interpolating the test statistics from the corresponding critical values tables (Fuller, 1996) for each type of linear models with given sample size.
 
 ```{r}
-adf_res <- df_long |>
-  group_by(age_group) |>
-  summarise(
-    aTSA::adf.test(covid_changes, output = FALSE) |>
-      map_df(as_tibble)
-  )
+df_global_age <- read_csv(here("data/df_global_age.csv"))
+
+adf_res_type1 <- df_global_age |>
+  group_by(age_grp, country) |>
+  group_modify(~ aTSA::adf.test(.x$daily_changes, output = FALSE, nlag = 2) |>
+                 bind_rows() |> 
+                 pluck("type1") |> 
+                 as_tibble()
+  ) |> 
+  rename(ADF_type1 = ADF, p.value_type1 = p.value)
+
+adf_res_type2 <- df_global_age |>
+  group_by(age_grp, country) |>
+  group_modify(~ aTSA::adf.test(.x$daily_changes, output = FALSE, nlag = 2) |>
+                 bind_rows() |> 
+                 pluck("type2") |> 
+                 as_tibble()
+  ) |> 
+  rename(ADF_type2 = ADF, p.value_type2 = p.value)
+
+adf_res_type3 <- df_global_age |>
+  group_by(age_grp, country) |>
+  group_modify(~ aTSA::adf.test(.x$daily_changes, output = FALSE, nlag = 2) |>
+                 bind_rows() |> 
+                 pluck("type3") |> 
+                 as_tibble()
+  ) |> 
+  rename(ADF_type3 = ADF, p.value_type3 = p.value)
+
+adf_res <- adf_res_type1 |> 
+  full_join(adf_res_type2) |> 
+  full_join(adf_res_type3)
 ```
+
+\newpage
+
+```{r}
+adf_res |>
+  filter(age_grp %in% c("aged0to4", "aged5to9")) |> 
+  kbl(
+    caption = "Dickey-Fuller test for all combinaisons of countries and age groups investigated",
+    col.names = c("Age Group", "Country", "Lag", "ADF", "p", "ADF", "p", "ADF", "p")
+  ) |> 
+  kable_classic(full_width = F) |> 
+  add_header_above(c(" " = 3, "Type 1" = 2, "Type 2" = 2, "type 3" = 2)) |>
+  collapse_rows(columns = 1:2, valign = "middle")
+```
+
+\newpage
+
+```{r}
+adf_res |>
+  filter(age_grp %in% c("aged10to14", "aged15to19")) |> 
+  kbl(
+    caption = "Dickey-Fuller test for all combinaisons of countries and age groups investigated",
+    col.names = c("Age Group", "Country", "Lag", "ADF", "p", "ADF", "p", "ADF", "p")
+  ) |> 
+  kable_classic(full_width = F) |> 
+  add_header_above(c(" " = 3, "Type 1" = 2, "Type 2" = 2, "type 3" = 2)) |>
+  collapse_rows(columns = 1:2, valign = "middle")
+```
+
+\newpage
+
+```{r}
+adf_res |>
+  filter(age_grp %in% c("aged20to24", "aged25to34")) |> 
+  kbl(
+    caption = "Dickey-Fuller test for all combinaisons of countries and age groups investigated",
+    col.names = c("Age Group", "Country", "Lag", "ADF", "p", "ADF", "p", "ADF", "p")
+  ) |> 
+  kable_classic(full_width = F) |> 
+  add_header_above(c(" " = 3, "Type 1" = 2, "Type 2" = 2, "type 3" = 2)) |>
+  collapse_rows(columns = 1:2, valign = "middle")
+```
+
+\newpage
+
+```{r}
+adf_res |>
+  filter(age_grp %in% c("aged35to44", "aged45to54")) |> 
+  kbl(
+    caption = "Dickey-Fuller test for all combinaisons of countries and age groups investigated",
+    col.names = c("Age Group", "Country", "Lag", "ADF", "p", "ADF", "p", "ADF", "p")
+  ) |> 
+  kable_classic(full_width = F) |> 
+  add_header_above(c(" " = 3, "Type 1" = 2, "Type 2" = 2, "type 3" = 2)) |>
+  collapse_rows(columns = 1:2, valign = "middle")
+```
+
+\newpage
+
+```{r}
+adf_res |>
+  filter(age_grp %in% c("aged55to64", "aged65+")) |> 
+  kbl(
+    caption = "Dickey-Fuller test for all combinaisons of countries and age groups investigated",
+    col.names = c("Age Group", "Country", "Lag", "ADF", "p", "ADF", "p", "ADF", "p")
+  ) |> 
+  kable_classic(full_width = F) |> 
+  add_header_above(c(" " = 3, "Type 1" = 2, "Type 2" = 2, "type 3" = 2)) |>
+  collapse_rows(columns = 1:2, valign = "middle")
+```

### 2025-06-19 manuscript/manuscript.qmd
update

--- 2025-06-05 manuscript_njd/manuscript_blind.qmd
@@
 The data can be accessed from the COVerAGE-DB OSF repository (https://osf.io/mpwjq/), from the Oxford Covid-19 Government Response Tracker (OxCGRT) "covid-policy-dataset" GitHub repository (https://github.com/OxCGRT/covid-policy-dataset), and from the UNESCO servers (https://en.unesco.org/file/unesco-data-school-closures-february-2020-june-2022csv-zip).
-All preprocessing, analyses and code used to build the submitted paper are available at https://github.com/[hidden for anonymous peer-review].
+All preprocessing, analyses and code used to build the submitted paper are available at https://github.com/damien-dupre/covid_ts_causality.
 

### 2025-07-07 manuscript/cover_letter.qmd
JHE

---
title: "Submission to the Journal of Health Economics"

format: pdf
---

Dear editors of the Journal of Health Economics,

Please find enclosed our manuscript for consideration for publication in the Journal of Health Economics, entitled “Divergent Impacts of COVID-19 School Closures on Youth Infection Dynamics: Evidence from 12 European Countries”.

This study offers an examination of how school closures influenced COVID-19 infection trends across different age groups in twelve European countries. By employing rigorous analytical methods, including Generalised Additive Models (GAM) and Transfer Entropy (TE), we aim to provide valuable insights into the role of age-specific public health interventions in shaping infection dynamics during the pandemic.

Our analysis reveals that the effectiveness of school closures varies not only across national contexts but also between age groups. We find that while the youngest children (ages 0 to 4) consistently experienced a reduction in infections following closures, the 5 to 14 age group saw a marked increase in cases. Among youth aged 15 to 24, a short-term surge was followed by a subsequent decline. Furthermore, our use of Transfer Entropy exposes asymmetrical relationships between age cohorts, highlighting how infection trends in one group can predict changes in another, thereby informing a more nuanced understanding of pandemic spread and policy impact.

We believe this research makes a timely and policy-relevant contribution to the field of public administration by offering evidence-based insights into how targeted interventions may yield differential effects. Our findings carry important implications for the design and evaluation of public health policies, particularly those aimed at mitigating the impact of communicable diseases in varied demographic groups.

All authors declare that there are no conflicts of interest, whether financial, personal or professional, that could be seen to influence the outcomes of this research. The manuscript has not been published, accepted or submitted elsewhere, and we have fully complied with ethical standards in data handling and manuscript preparation. We have not discussed this submission with any member of the editorial board prior to this submission.

The authors of this study are:

- Damien Dupré (corresponding author), Dublin City University, damien.dupre@dcu.ie
- Edgar Morgenroth, Dublin City University, edgar.morgenroth@dcu.ie

Should you have any questions regarding our submission, please do not hesitate to contact me directly. We appreciate your time and consideration, and we look forward to the opportunity to contribute to the ongoing scholarly dialogue through the Journal of Health Economics.

Yours sincerely,
Damien Dupré

### 2025-07-07 manuscript/manuscript.qmd
JHE

@@
 
-format:
-  wiley-njd-pdf:
-    article-type: "Article Type: Research Article"
-    # received: "10 November 2024"
-    # revised: "10 November 2024"
-    # accepted: "1 January 2025"
-    journal: "Journal of Policy Analysis and Management"
-    volume: "00"
-    copy-year: 2025
-    start-page: 1
+format: docx
+  # elsevier-pdf:
+  #   keep-tex: true
 
-    titlemark: "COVID-19 School Closures and Youth Infection Dynamics"
-    authormark: "Dupré and Morgenroth"
-
-    keywords:
-      - COVID-19
-      - School closures
-      - Intergenerational transmission
-      - Non-linear effects
-      - Infection dynamics.
+# author:
+#   - name: Damien Dupré
+#     email: damien.dupre@dcu.ie
+#     corresponding: true
+#     affiliations:
+#       - name: Dublin City University
+#         department: Business School
+#         city: Dublin
+#         country: Ireland
+# 
+#   - name: Edgar Morgenroth
+#     email: edgar.morgenroth@dcu.ie
+#     affiliations:
+#       - name: Dublin City University
+#         department: Business School
+#         city: Dublin
+#         country: Ireland
+# 
+# present-address: |
+#   DCU Business School, Dublin 9, Ireland.
 
 author:
-  - name: Damien Dupré
-    email: damien.dupre@dcu.ie
-    corresponding: true
-    affiliations:
-      - name: Dublin City University
-        department: Business School
-        city: Dublin
-        country: Ireland
-
-  - name: Edgar Morgenroth
-    email: edgar.morgenroth@dcu.ie
-    affiliations:
-      - name: Dublin City University
-        department: Business School
-        city: Dublin
-        country: Ireland
-
-present-address: |
-  DCU Business School, Dublin 9, Ireland.
+  - name: Anonymous Authors
 
@@
 These results challenge the assumption that school closures uniformly benefit younger populations and emphasise the need for age-specific analysis in pandemic planning and support the development of more nuanced, evidence-based public health policies.
-pdf-engine: lualatex
+
 keep-tex: true
 bibliography: bibliography.bib
+csl: https://www.zotero.org/styles/apa
+
+keywords:
+  - COVID-19
+  - School closures
+  - Intergenerational transmission
+  - Non-linear effects
+  - Infection dynamics.
 ---
@@
 The data can be accessed from the COVerAGE-DB OSF repository (https://osf.io/mpwjq/), from the Oxford Covid-19 Government Response Tracker (OxCGRT) "covid-policy-dataset" GitHub repository (https://github.com/OxCGRT/covid-policy-dataset), and from the UNESCO servers (https://en.unesco.org/file/unesco-data-school-closures-february-2020-june-2022csv-zip).
-All preprocessing, analyses and code used to build the submitted paper are available at https://github.com/damien-dupre/covid_ts_causality.
+All preprocessing, analyses and code used to build the submitted paper are available at [URL hidden for anonymous peer-review].
 

### 2025-07-07 manuscript/manuscript.tex
JHE

@@
-\documentclass[]{elsarticle} %review=doublespace preprint=single 5p=2 column
-%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
-\usepackage[hyphens]{url}
+% Options for packages loaded elsewhere
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{hyphens}{url}
+\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
+%
+\documentclass[
+  number]{elsarticle}
+\usepackage{xcolor}
+\usepackage{amsmath,amssymb}
+\setcounter{secnumdepth}{5}
+\usepackage{iftex}
+\ifPDFTeX
+  \usepackage[T1]{fontenc}
+  \usepackage[utf8]{inputenc}
+  \usepackage{textcomp} % provide euro and other symbols
+\else % if luatex or xetex
+  \usepackage{unicode-math} % this also loads fontspec
+  \defaultfontfeatures{Scale=MatchLowercase}
+  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\fi
+\usepackage{lmodern}
+\ifPDFTeX\else
+  % xetex/luatex font selection
+\fi
+% Use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+\IfFileExists{microtype.sty}{% use microtype if available
+  \usepackage[]{microtype}
+  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
+}{}
+\makeatletter
+\@ifundefined{KOMAClassName}{% if non-KOMA class
+  \IfFileExists{parskip.sty}{%
+    \usepackage{parskip}
+  }{% else
+    \setlength{\parindent}{0pt}
+    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
+}{% if KOMA class
+  \KOMAoptions{parskip=half}}
+\makeatother
+% Make \paragraph and \subparagraph free-standing
+\makeatletter
+\ifx\paragraph\undefined\else
+  \let\oldparagraph\paragraph
+  \renewcommand{\paragraph}{
+    \@ifstar
+      \xxxParagraphStar
+      \xxxParagraphNoStar
+  }
+  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
+  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
+\fi
+\ifx\subparagraph\undefined\else
+  \let\oldsubparagraph\subparagraph
+  \renewcommand{\subparagraph}{
+    \@ifstar
+      \xxxSubParagraphStar
+      \xxxSubParagraphNoStar
+  }
+  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
+  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
+\fi
+\makeatother
 
-  \journal{An awesome journal} % Sets Journal name
+\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
+\usepackage{graphicx}
+\makeatletter
+\newsavebox\pandoc@box
+\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
+  \sbox\pandoc@box{#1}%
+  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
+  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
+  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
+  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
+  \else\usebox{\pandoc@box}%
+  \fi%
+}
+% Set default figure placement to htbp
+\def\fps@figure{htbp}
+\makeatother
 
-\usepackage{lineno} % add
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+
 \providecommand{\tightlist}{%
@@
 
-\usepackage{graphicx}
-%%%%%%%%%%%%%%%% end my additions to header
+\usepackage[]{natbib}
+\bibliographystyle{elsarticle-num}
 
-\usepackage[T1]{fontenc}
-\usepackage{lmodern}
-\usepackage{amssymb,amsmath}
-\usepackage{ifxetex,ifluatex}
-\usepackage{fixltx2e} % provides \textsubscript
-% use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[utf8]{inputenc}
-\else % if luatex or xelatex
-  \usepackage{fontspec}
-  \ifxetex
-    \usepackage{xltxtra,xunicode}
+\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{wrapfig}
+\usepackage{float}
+\usepackage{colortbl}
+\usepackage{pdflscape}
+\usepackage{tabu}
+\usepackage{threeparttable}
+\usepackage{threeparttablex}
+\usepackage[normalem]{ulem}
+\usepackage{makecell}
+\usepackage{xcolor}
+\makeatletter
+\@ifpackageloaded{caption}{}{\usepackage{caption}}
+\AtBeginDocument{%
+\ifdefined\contentsname
+  \renewcommand*\contentsname{Table of contents}
+\else
+  \newcommand\contentsname{Table of contents}
 \fi
-  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
-  \newcommand{\euro}{€}
+\ifdefined\listfigurename
+  \renewcommand*\listfigurename{List of Figures}
+\else
+  \newcommand\listfigurename{List of Figures}
 \fi
-% use microtype if available
-\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
-\bibliographystyle{elsarticle-harv}
-\ifxetex
-  \usepackage[setpagesize=false, % page size defined by xetex
-              unicode=false, % unicode breaks when used with xetex
-              xetex]{hyperref}
+\ifdefined\listtablename
+  \renewcommand*\listtablename{List of Tables}
 \else
-  \usepackage[unicode=true]{hyperref}
+  \newcommand\listtablename{List of Tables}
 \fi
-\hypersetup{breaklinks=true,
-            bookmarks=true,
-            pdfauthor={},
-            pdftitle={Short Paper},
-            colorlinks=false,
-            urlcolor=blue,
-            linkcolor=magenta,
-            pdfborder={0 0 0}}
-\urlstyle{same}  % don't use monospace font for urls
+\ifdefined\figurename
+  \renewcommand*\figurename{Figure}
+\else
+  \newcommand\figurename{Figure}
+\fi
+\ifdefined\tablename
+  \renewcommand*\tablename{Table}
+\else
+  \newcommand\tablename{Table}
+\fi
+}
+\@ifpackageloaded{float}{}{\usepackage{float}}
+\floatstyle{ruled}
+\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
+\floatname{codelisting}{Listing}
+\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
+\makeatother
+\makeatletter
+\makeatother
+\makeatletter
+\@ifpackageloaded{caption}{}{\usepackage{caption}}
+\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
+\makeatother
+\usepackage{bookmark}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\urlstyle{same}
+\hypersetup{
+  pdftitle={Divergent Impacts of COVID-19 School Closures on Youth Infection Dynamics: Evidence from 12 European Countries},
+  pdfauthor={Anonymous Authors},
+  pdfkeywords={COVID-19, School closures, Intergenerational
+transmission, Non-linear effects, Infection dynamics.},
+  colorlinks=true,
+  linkcolor={blue},
+  filecolor={Maroon},
+  citecolor={Blue},
+  urlcolor={Blue},
+  pdfcreator={LaTeX via pandoc}}
 
-\setcounter{secnumdepth}{0}
-% Pandoc toggle for numbering sections (defaults to be off)
-\setcounter{secnumdepth}{0}
+\setlength{\parindent}{6pt}
+\begin{document}
 
-% Pandoc citation processing
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-% for Pandoc 2.8 to 2.10.1
-\newenvironment{cslreferences}%
-  {}%
-  {\par}
-% For Pandoc 2.11+
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
-  % set entry spacing
-  \ifnum #2 > 0
-  \setlength{\parskip}{#2\baselineskip}
-  \fi
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
+\begin{frontmatter}
+\title{Divergent Impacts of COVID-19 School Closures on Youth Infection
+Dynamics: Evidence from 12 European Countries}
+\author[]{Anonymous Authors%
+%
+}
 
-% Pandoc header
-
-\begin{document}
-\begin{frontmatter}
-
-  \title{Short Paper}
-    \author[Some Institute of Technology]{Alice Anonymous\corref{1}}
-   \ead{alice@example.com} 
-    \author[Another University]{Bob Security}
-   \ead{bob@example.com} 
-    \author[Another University]{Cat Memes\corref{2}}
-   \ead{cat@example.com} 
-    \author[Some Institute of Technology]{Derek Zoolander\corref{2}}
-   \ead{derek@example.com} 
-      \address[Some Institute of Technology]{Department, Street, City,
-State, Zip}
-    \address[Another University]{Department, Street, City, State, Zip}
-      \cortext[1]{Corresponding Author}
-    \cortext[2]{Equal contribution}
+\cortext[cor1]{Corresponding author}
 
 \begin{abstract}
-  This is the abstract.
-
-  It consists of two paragraphs.
+The effectiveness of school closures as a COVID-19 control measure
+remains a subject of debate. This paper critically reassesses their
+impact by examining infection trends across various age groups in 12
+European countries. We apply a dual analytical approach: Generalised
+Additive Models (GAMs) are used to capture complex, non-linear effects
+of closures on age-specific case numbers, while Transfer Entropy (TE) is
+employed to quantify directional transmission patterns between these age
+groups. Our findings reveal deviations from commonly held assumptions.
+While school closures were associated with a non-linear decline in
+overall national COVID-19 cases, the effects on children and young
+adults varied considerably. A consistent downward trend in infections
+was seen only in the pre-school age group. In contrast, school-aged
+children experienced a marked rise in COVID-19 cases following an
+initial period of stability or slight decrease after closures began. The
+Transfer Entropy analysis also identified asymmetric, directional
+patterns in transmission, revealing that infection trends in specific
+age groups could predict subsequent changes in others. These results
+challenge the assumption that school closures uniformly benefit younger
+populations and emphasise the need for age-specific analysis in pandemic
+planning and support the development of more nuanced, evidence-based
+public health policies.
 \end{abstract}
 
+\begin{keyword}
+    COVID-19 \sep School closures \sep Intergenerational
+transmission \sep Non-linear effects \sep 
+    Infection dynamics.
+\end{keyword}
 \end{frontmatter}
 
-\emph{Text based on elsarticle sample manuscript, see
-\url{http://www.elsevier.com/author-schemas/latex-instructions\#elsarticle}}
+\section{Introduction}\label{introduction}
 
-\hypertarget{the-elsevier-article-class}{%
-\section{The Elsevier article class}\label{the-elsevier-article-class}}
+The COVID-19 pandemic has had a profound impact on global health, with
+an estimate of 14.83 million excess deaths globally
+\citep{msemburi2023estimates}. Beyond direct mortality, the pandemic has
+caused significant collateral damage, including losses of lives and
+livelihoods, necessitating a comprehensive approach to measuring its
+broader impacts. A wide range of non-pharmaceutical interventions (NPI)
+were enacted to control the spread of COVID-19, especially before the
+availability of effective vaccines. Due to their social mixing patterns,
+children have been identified as an age group that can drive the spread
+of respiratory infections such as influenza \citep{moser2018estimating},
+and school closures were found to be effective in mitigating the spread
+of influenza H1N1 in Japan in 2009 \citep{kawano2015substantial}. It is
+therefore not surprising that school closures were one of the NPIs that
+were introduced in many countries to stop the spread of COVID-19. They
+were used particularly during the initial wave of the pandemic and
+during subsequent waves when case numbers were high.
 
-\hypertarget{installation}{%
-\paragraph{Installation}\label{installation}}
+Recently, a body of literature has shown various negative consequences
+on child development and health due to school closures. School closures
+have been found to have negatively affected child mental health
+\citep{moulin2022longitudinal}, nutrition/obesity
+\citep{sugimoto2023temporal}, and education
+\citep{lerkkanen2023reading}. Some governments were questioning their
+decisions to close schools \citep{de2021determines}. For example, the
+German health minister Karl Lauterbach, in an interview with one of
+Germany's publicly funded television stations, admitted that ``in
+retrospect it had been wrong to keep schools and childcare closed for so
+long'' \citep{ard2023lauterbach}. Even if school closures had negative
+consequences for children, they might nevertheless have been effective
+at stopping the spread of COVID-19. However, analysis on the
+effectiveness of school closures on the spread of COVID-19 remains
+inconclusive. In a study of a panel of European countries,
+\citep{alfano2022effects} found that school closures were associated
+with a reduced COVID-19 incidence. In contrast, \citep{walsh2021school}
+in a review of 40 studies covering 150 countries concluded that the
+effectiveness of school closures was uncertain, with 60\% having
+identified no impact and pointing to the potential for the analysis to
+be affected by confounding factors and collinearity. The latter
+shortcomings suggest that it is hard to draw firm inferences, as even a
+positive result may not imply causality.
 
-If the document class \emph{elsarticle} is not available on your
-computer, you can download and install the system package
-\emph{texlive-publishers} (Linux) or install the LaTeX package
-\emph{elsarticle} using the package manager of your TeX installation,
-which is typically TeX Live or MikTeX.
+Once it was clear that COVID-19 spreads from person to person, it was
+natural that governments were advised to enact measures to reduce social
+contact in order to reduce the spread of the virus. Therefore, it is
+legitimate to believe that by closing schools, a reduction of the
+contaminations would be observed in the younger age groups. However, the
+efficiency of school closure on the reduction of COVID-19 cases is still
+questioned \citep{bayham2020impact, esposito2021comprehensive}. While
+some research has found that school closures contribute to limit or to
+reduce the growth rate of confirmed cases after implementation
+\citep{stage2021shut, yoshiyuki2020effects}, others did not observe a
+change in the evolution of COVID-19 cases
+\citep{chang2020modelling, iwata2020school}. For instance, a controlled
+comparison between similar localities in Japan with schools closed and
+schools open did not reveal any evidence that school closures reduced
+the spread of COVID-19 \citep{fukumoto2021no}. If the school closure had
+a real impact on the evolution of confirmed COVID-19 cases, it should be
+possible to observe a decrease or at least an inflection in the trend of
+its evolution among younger age groups.
 
-\hypertarget{usage}{%
-\paragraph{Usage}\label{usage}}
+A second implicit belief regarding the effect of school closure on the
+spread of COVID-19 is that school not only influences the spread of the
+virus among children and teenagers but also has a knock-on effect on the
+spread of the virus in older age groups, also called Secondary Attack
+Rate (SAR). The contaminated children and teenagers would bring the
+virus back home and would pass the virus on to their parents and
+relatives. For example, research investigating the contamination in the
+household network not only revealed an exceptionally high rate of
+secondary contamination but also that this contamination happened when
+the schools were closed \citep{sorianoarandes2021household}. Despite
+being reported in several clinical and epidemiological studies
+\citep{siebach2021childhood, zhendong2020clinical}, multiple research
+have shown that the SAR from children to household members was lower
+than expected
+\citep{heavey2020evidence, vanderhoek2020role, kim2021role, ludvigsson2020children}.
+However, the SAR of children and teenagers to the household member is
+likely to be age-dependent, with differences between infants, primary
+and secondary school children, and college students
+\citep{grasleguen2021reopening}. If a secondary transmission from
+children and teenagers to household member has a significant influence,
+then a temporal causality relationship between their evolution should be
+observed.
 
-Once the package is properly installed, you can use the document class
-\emph{elsarticle} to create a manuscript. Please make sure that your
-manuscript follows the guidelines in the Guide for Authors of the
-relevant journal. It is not necessary to typeset your manuscript in
-exactly the same way as an article, unless you are submitting to a
-camera-ready copy (CRC) journal.
+A gap in these previous analyses is the insufficient granularity in
+examining age-specific effects and the complex dynamics of inter-age
+group transmission. This study aims to address these deficiencies by
+providing a more robust and nuanced assessment. First, Generalized
+Additive Models (GAMs) are utilized to flexibly model and quantify the
+potentially non-linear evolution of COVID-19 cases following school
+closures, specifically within these fine-grained age brackets. This
+allows us to move beyond simple pre-post comparisons and capture dynamic
+temporal patterns. Then, Transfer Entropy (TE), an information-theoretic
+measure, is applied to assess directional causality in time series data.
+TE is particularly adept at identifying how past states of one variable
+(e.g., cases in one age group) influence future states of another (e.g.,
+cases in a different age group), even in complex, non-linear systems
+like epidemic spread. This dual methodology allows for a deeper
+understanding of both the direct age-specific consequences of school
+closures and the potential for altered intergenerational transmission
+dynamics. Our findings challenge simplistic narratives about school
+closure efficacy and highlight the importance of age-stratified analysis
+for future pandemic preparedness.
 
-\hypertarget{functionality}{%
-\paragraph{Functionality}\label{functionality}}
+\section{Methods}\label{methods}
 
-The Elsevier article class is based on the standard article class and
-supports almost all of the functionality of that class. In addition, it
-features commands and options to format the
+\subsection{Study design}\label{study-design}
 
-\begin{itemize}
-\item
-  document style
-\item
-  baselineskip
-\item
-  front matter
-\item
-  keywords and MSC codes
-\item
-  theorems, definitions and proofs
-\item
-  lables of enumerations
-\item
-  citation style and labeling.
-\end{itemize}
+A longitudinal study was conducted to observe changes in COVID-19 cases
+during school closures. Changes in COVID-19 from five age groups from
+the first to the 28th day of school closure are compared across 12
+countries. Then, a causality analysis evaluate the impact of changes in
+these younger groups on older age groups.
 
-\hypertarget{front-matter}{%
-\section{Front matter}\label{front-matter}}
+\subsection{Data collection}\label{data-collection}
 
-The author names and affiliations could be formatted in two ways:
+Data on COVID-19 case counts by age group have been obtained from the
+COVerAGE-DB project \citep{riffe2021data}. The COVerAGE-DB which collect
+numbers for every countries by groups of 5 or 10 years if available. In
+addition, spline approximations are used to deals with the heterogeneity
+of countries' reporting formats by using when the data for this age
+bracket is not available for a country. Countries communicating data
+with groups of 5 years were chosen to match as much as possible the
+different school stages.
 
-\begin{enumerate}
-\def\labelenumi{(\arabic{enumi})}
-\item
-  Group the authors per affiliation.
-\item
-  Use footnotes to indicate the affiliations.
-\end{enumerate}
+To identify the impact of school closures on the number of cases,
+information regarding the school closures on a day-by-day basis is
+required, and this is taken from the ``UNESCO global education
+coalition'' (2022). For each day, in each country, the status of the
+schools is indicated as fully open, partially open, closed due to
+COVID-19, or closed due to an academic break. Because it would be
+difficult to measure the effect of school closures at a country level
+when schools are partially closed, only closures due to COVID-19 or due
+to an academic break are considered. Indeed, both are considered as
+closure at a country-wide level. Any COVID-19 case numbers beyond the
+28-day period are not relevant for evaluating the impact of school
+closures.
 
-See the front matter of this document for examples. You are recommended
-to conform your choice to the journal you are submitting to.
+\subsection{Data validation}\label{data-validation}
 
-\hypertarget{bibliography-styles}{%
-\section{Bibliography styles}\label{bibliography-styles}}
+In order to cross-validate the data obtained after spline
+approximations, a comparison with the data published by the
+\citep{who2022https} reveals perfect similarities. The original data
+consist of 14,089,320 observations of 10 variables (117 distinct
+countries, region within the country, a unique observation code, the
+date of the observation, the gender, which can be male, female, or both,
+the age bracket by 5 years from 0 to 100, a confirmation of the age
+interval for each bracket, the total number of cases so far, the total
+number of deaths, and the total number of tests performed) from February
+16, 2020 to January 20, 2022. After removing countries with missing and
+inconsistent values, only 22 are suitable for data analyses. However, to
+focus this analysis on geographically and culturally comparable
+countries, only 12 European countries are kept: Austria, Belgium,
+Bulgaria, Croatia, Estonia, France, Germany, Greece, Netherlands,
+Portugal, Slovakia, and Spain. The observations are reported in terms of
+the total number of COVID-19 cases per day from the start of the
+pandemic. The daily number of cases at a specific date \(n\prime_{t}\)
+is calculated with the difference between the total cases at a date
+\(t\) and the total cases at a date \(t-1\) (i.e., derivative 1). In
+addition, the change in the daily number of cases \(n\prime\prime_{t}\)
+between \(n\prime_{t}\) and \(n\prime_{t-1}\) has also been calculated
+(i.e., derivative 2).
 
-There are various bibliography styles available. You can select the
-style of your choice in the preamble of this document. These styles are
-Elsevier styles based on standard styles like Harvard and Vancouver.
-Please use BibTeX~to generate your bibliography and include DOIs
-whenever available.
+\subsection{Controlling for confounding Non-Pharmaceutical
+Interventions}\label{controlling-for-confounding-non-pharmaceutical-interventions}
 
-Here are two sample references: (Dirac, 1953; Feynman and Vernon Jr.;
-1963).
+Recognizing that school closures rarely occurred in isolation, a
+critical component of this analysis was to account for other
+contemporaneous NPIs. To mitigate confounding from other NPIs, we
+incorporated four time-varying composite NPI index identified by the
+Oxford COVID-19 Government Response Tracker (OxCGRT) project
+\citep{hale2021global}: the average stringency index, the average
+government response index, the average containment health index, and the
+average economic support index. Specific data were collected from each
+country on various policy responses governments implemented to curb the
+spread of COVID-19, such as travel restrictions, workplace closures,
+public event cancellations, restrictions on gatherings, or public health
+messaging. These indicators were then aggregated to create indexes,
+reflecting the overall level of government restrictions.
 
-\hypertarget{references}{%
-\section*{References}\label{references}}
-\addcontentsline{toc}{section}{References}
+\subsection{Generalized additive
+model}\label{generalized-additive-model}
 
-\hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\hypertarget{ref-Dirac1953888}{}%
-Dirac, P.A.M., 1953. The lorentz transformation and absolute time.
-Physica 19, 888--896.
-doi:\href{https://doi.org/10.1016/S0031-8914(53)80099-6}{10.1016/S0031-8914(53)80099-6}
+The effect of school closure on the trend of daily COVID-19 cases is
+analysed using a Generalized Additive Model (GAM). GAM is a flexible
+modelling approach that estimates non-linear relationships between
+variables. Compared to other methods such as Vector Autoregression, GAM
+can handle fixed and random smooth effects without making strict
+assumptions about linearity.
 
-\leavevmode\hypertarget{ref-Feynman1963118}{}%
-Feynman, R.P., Vernon Jr., F.L., 1963. The theory of a general quantum
-system interacting with a linear dissipative system. Annals of Physics
-24, 118--173.
-doi:\href{https://doi.org/10.1016/0003-4916(63)90068-X}{10.1016/0003-4916(63)90068-X}
+The GAM is fitted on the daily COVID-19 cases to test the hypothesis of
+a significant non-linear evolution of cases among age groups from 0 to
+4, from 5 to 9, from 10 to 14, from 15 to 19, and from 20 to 24
+\citep{wood2017generalized}. The model also estimates the overall
+non-linear effect by country, taking into account the interaction
+between age groups and countries as random intercepts and the
+interaction between time, countries, and the period of closure as random
+effects (Eq Equation~\ref{eq-gam}).
 
-\end{CSLReferences}
+By estimating the degree of smoothness of a Bayesian spline smoothing
+using restricted fast maximum likelihood estimation
+\citep{wood2011fast}, GAM identifies dynamic patterns underlying the
+evolution of COVID-19 cases reported while including the random effect
+of different age groups and countries as follows:
+
+\begin{equation}\phantomsection\label{eq-gam}{
+\begin{aligned}[t]
+  n\prime_t &\sim \text{Poisson}(\lambda_t) \\
+  \log(\lambda_t) &= \beta_0 + \beta_1\,wave_t + \beta_2\,country_t + \beta_3\,age\,group_t \\
+  &\quad + f_1(closure_t) + f_2(closure_t, country_t) \\
+  &\quad + f_3(closure_t, age\,group_t) + f_4(stringency\,index_t) \\
+  &\quad + f_5(government\,response\,index_t) + f_6(containment\,health\,index_t) \\
+  &\quad + f_7(economic\,support\,index_t) +\epsilon_t
+\end{aligned}
+}\end{equation}
+
+\noindent where \(n\prime_t\) represents the confirmed COVID-19 cases,
+assuming a Poisson distribution for the fitting \citep{loader2006local},
+and \(t\) is the date corresponding to the confirmed COVID-19 cases. The
+response variable includes a specific random effect taking into account
+variation within waves of school closure, countries, and age groups. The
+terms \(f_1\) to \(f_3\) are smooth functions of the time since closure,
+the time since closure for each country, and the time since closure for
+each age group. The terms \(f_4\) to \(f_7\) are random effects
+corresponding to the average stringency index, the average government
+response index, the average containment health index, and the average
+economic support index. The restricted maximum likelihood (REML) was
+used to avoid over fitting while estimating smoothing parameters. In
+order to accurately account for the autocorrelation arising from the
+time series data, the residuals are modelled using an AR1 error model
+such as
+\(\epsilon_t = \phi \epsilon_{t-1} + \eta_t, \quad \eta_t \sim \mathcal{N}(0, \sigma^2)\)
+where \(\phi\) is the autoregressive parameter. By incorporating the
+autoregressive component, the AR1 model acknowledges the dependence of
+each residual on its previous value, thus providing a comprehensive
+representation of the data's temporal dynamics.
+
+Chi-square statistics were employed to assess if the degree of
+smoothness is significantly different from zero.
+
+\subsection{Transfer entropy}\label{transfer-entropy}
+
+Transfer entropy (\(T\)) is a measure of the directional information
+flow between two time series \(X\) and \(Y\), capturing how the past
+values of one variable can predict changes in another. Unlike
+correlation, transfer entropy accounts for the temporal order of events
+and non-linear relationships, making it particularly suited for studying
+dynamic systems such as epidemic data. In this analysis, transfer
+entropy quantifies how case counts in one age group \(X\) influence
+subsequent case counts in another age group \(Y\), providing insights
+into directional transmission patterns. If it does, \(T\) is considered
+evidence of a causal effect from the age group \(X\) to the age group
+\(Y\) \citep{schreiber2000measuring}. As such, Granger causality is a
+special case of transfer entropy applied to time series that are jointly
+Gaussian distributed \citep{barnett2009granger}. Therefore, transfer
+entropy is a more robust analysis of time series, especially when
+applied to the impact of age cohorts on pandemic transmission
+\citep{kissler2020symbolic}.
+
+The influence of the evolution in COVID-19 cases across all age groups
+is evaluated using Shannon's transfer entropy, given by:
+
+\begin{equation}\phantomsection\label{eq-te}{
+\begin{aligned}[t]
+  T_{ag\,x \rightarrow ag\,y}(k,l) = \sum_{ag\,x_{t+1}, ag\,x_t^{(k)}, ag\,y_t^{(l)}} 
+  p\left(ag\,x_{t+1}, ag\,x_t^{(k)}, ag\,y_t^{(l)}\right) 
+  \log \left(\frac{p\left(ag\,x_{t+1} \mid ag\,x_t^{(k)}, ag\,y_t^{(l)}\right)}{p\left(ag\,x_{t+1} \mid ag\,x_t^{(k)}\right)}\right)
+\end{aligned}
+}\end{equation}
+
+\noindent where \(T_{ag\,x \rightarrow\,ag\,y}\) consequently measures
+the influence of the change dynamic from an age group \(X\) (or
+\(ag\,x\)) to another age group \(Y\) (or \(ag\,y\)) for every country
+(Eq Equation~\ref{eq-te}).
+
+The day-by-day difference in COVID-19 confirmed cases
+\(n\prime\prime_t\) is used to satisfy the stationary requirement for
+the calculation of Shannon's Transfer Entropy
+\citep{shannon1948mathematical, behrendt2019rtransferentropy}.
+
+\section{Results}\label{results}
+
+The data show that the trend of confirmed COVID-19 cases follows similar
+patterns across the selected European countries, with scales following
+the size of the population in these countries
+(Figure~\ref{fig-overall}). Thus, the bigger the country, the higher the
+total number of COVID-19 cases.
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{fig-overall-1.png}}
+
+}
+
+\caption{\label{fig-overall}Cumulative COVID-19 cases number for
+selected European countries since the beginning of the pandemic. Source:
+COVerAGE-DB \citep{riffe2021data}.}
+
+\end{figure}%
+
+The evolution of COVID-19 cases reveals some similarities across all age
+groups. However, the influence of each wave on individual age groups
+also has some particularities (Figure~\ref{fig-descriptive}). For
+example, the first wave was more important among the oldest age groups,
+whereas the third wave was more important among the youngest age groups.
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{fig-descriptive-1.png}}
+
+}
+
+\caption{\label{fig-descriptive}Periods of school closure since the
+beginning of the COVID-19 pandemic for selected European countries and
+their reason: regular academic break versus closure due to government
+decisions. Source: \citep{unesco2022https}.}
+
+\end{figure}%
+
+As outlined above, to evaluate the shape of the trend in the numbers of
+COVID-19 cases reported after the three school closures longer than 21
+consecutive days, a GAM was fitted as described above, taking into
+account the overall effect across all the selected European countries,
+as well as the effect for age groups: 0 to 4, 5 to 9, 10 to 14, 15 to
+19, and 20 to 24 years old. The obtained results satisfy the
+requirements to fit this model, which explains 80.2\% of the variation
+in COVID-19 cases (Figure~\ref{fig-age}).
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{fig-age-1.png}}
+
+}
+
+\caption{\label{fig-age}Standardized effect of the smooth term in
+Generalized Additive Model by age group. Standardized effects are
+reported to compare the shape of the curve between age groups.}
+
+\end{figure}%
+
+Overall, the results revealed a decreasing non-linear effect of school
+closure at a country level for the selected European countries (Austria:
+\(\chi^2(5.82) = 3243.01\), \(p < 0.001\); Belgium:
+\(\chi^2(5.98) = 6206.45\), \(p < 0.001\); Bulgaria:
+\(\chi^2(5.89) = 258.19\), \(p < 0.001\); Croatia:
+\(\chi^2(5.1) = 740.24\), \(p < 0.001\); Estonia:
+\(\chi^2(5.93) = 1834.26\), \(p < 0.001\); France:
+\(\chi^2(1) = 2412.59\), \(p < 0.001\); Germany
+(\(\chi^2(5.98) = 3081.05\), \(p < 0.001\)); Greece:
+\(\chi^2(5.95) = 2423.67\), \(p < 0.001\); The Netherlands:
+\(\chi^2(4.97) = 5878.92\), \(p < 0.001\); Portugal:
+\(\chi^2(5.94) = 2692.56\), \(p < 0.001\); Slovakia
+(\(\chi^2(5.97) = 5335.74\), \(p < 0.001\); Spain:
+\(\chi^2(5.94) = 498.09\), \(p < 0.001\), see Figure~\ref{fig-country}.
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{fig-country-1.png}}
+
+}
+
+\caption{\label{fig-country}Standardized effect of the smooth term in
+Generalized Additive Model by country. Standardized effects are reported
+to compare the shape of the curve between countries.}
+
+\end{figure}%
+
+While the general patterns described above were evident in the
+aggregated analysis, country-specific smooths indicated some national
+variations in the timing and magnitude of these age-specific effects,
+although the paradoxical increase in the 5-14 age cohorts was observed
+in a majority of countries.
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=17.1875in,height=\textheight,keepaspectratio]{fig-te-1.png}
+
+}
+
+\caption{\label{fig-te}Matrix of Transfer Entropy coefficients according
+to every age group combination for each of the selected European
+country. Age groups on the x-axis are influencing the age groups on the
+y-axis (\(ag\,x \rightarrow\,ag\,y\)). The significance of each Transfer
+Entropy coefficient is provided in Appendix 2.}
+
+\end{figure}%
+
+Considering all countries, the analysis of age groups reveals distinct
+patterns (Figure~\ref{fig-te}). For the 0 to 4 age group, i.e., the
+pre-school group, there was a downward trend during the initial three
+weeks following school closure, followed by an increase during the
+fourth week (\(\chi^2(5.96) = 410.48\), \(p < 0.001\)). A similar, but
+more moderate, trend is observer for the age group ranging from 5 to 9
+(\(\chi^2(5.95) = 878.29\), \(p < 0.001\)), where a decrease is observed
+during the first two weeks, followed by an increase. while the trend for
+the age group ranging from 10 to 14 is flat during the first two weeks,
+a similar increase is shown at the beginning of the third week of school
+closure (\(\chi^2(1) = 61.62\), \(p < 0.001\)). For age groups between
+15 and 24, there is a notable surge immediately after school closure,
+followed by a decrease after 14 days (15 to 19 age group:
+\(\chi^2(5.97) = 7929.88\), \(p < 0.001\) and 20 to 24 age group:
+\(\chi^2(4.98) = 18386.65\), \(p < 0.001\)). Thus, while the total case
+number declined following school closures, among younger age groups only
+the pre-school age group had a consistent reduction in cases, while for
+other age groups, there was even an increase in cases following school
+closures.
+
+Given that school closures were not only aimed at reducing COVID-19
+cases among school-going children but also age groups, it is important
+to test the degree to which there was transmission across age groups,
+which is done using transfer entropy analysis. Before this analysis is
+carried out, an Augmented Dickey-Fuller has been applied to each age
+group to ensure that the daily changes in COVID-19 cases are stationary
+(see Appendix 1).
+
+The results of the transfer entropy calculations between age groups for
+each of the 12 selected European countries are reported in
+Figure~\ref{fig-te}. The absence of symmetry between influencing age
+groups (i.e., \(ag\,x\)) and influenced age groups (i.e., \(ag\,y\)) is
+found. Indeed, the change in COVID-19 cases in some age groups is
+influenced by other age groups, but they are not reciprocally
+influencing these age groups. F@fig-te shows how different age cohorts
+influence the COVID-19 case numbers of all age groups across countries.
+The upper left quadrant indicates how younger age cohorts are
+influencing older age cohorts, the bottom right quadrant indicates how
+older age cohorts are influencing younger age cohorts, finally, the
+lower left and upper right indicate how younger or older age cohorts are
+influencing themselves. By analyzing these quadrants, it is possible to
+identify similar patterns across multiple countries. Indeed, it appears
+that Austria, Germany, and The Netherlands have significantly higher
+\(T\) coefficients in the upper left quadrant of the matrix, which
+indicates that the daily changes in COVID-19 cases number in younger
+cohorts are predicting the daily changes in COVID-19 cases number in
+older cohorts. Alternatively, it appears that Austria, the Netherlands,
+Portugal, and Spain have significantly higher \(T\) coefficients in the
+lower right quadrant of the matrix, which indicates that the daily
+changes in COVID-19 cases number in older cohorts are predicting the
+daily changes in COVID-19 cases number in younger cohorts.
+
+\section{Discussion}\label{discussion}
+
+Knowledge about the transmission of the virus significantly improved
+over time as more studies have been published. While an early study
+found that children did not play an important role in the transmission
+of the virus \citep{li2020role}, more recent results give a more nuanced
+position, stating that the spread of the virus in children is moderate.
+It is now suggested that the impact of school openings on infection
+rates was limited, with children and teachers more often contracting
+infections through household or community contacts rather than within
+schools \citep{soriano2023policies}. This study, using robust
+statistical methods, considered the effect of school closures on
+COVID-19 cases across age groups. While there are commonalities in the
+evolution of COVID-19 cases across all European countries that were
+included in our data, school closures exhibit distinct impacts on
+different age groups. Notably, the analysis shows that the 0 to 4 age
+group experiences a downward trend in COVID-19 cases during the initial
+two weeks following school closure, followed by stabilization. In
+contrast, age groups ranging from 5 to 14 exhibit a stable profile after
+school closure. Additionally, age groups between 15 and 24 demonstrate a
+notable surge immediately after school closure, followed by a decrease
+after 14 days. Thus, the results do not support the hypothesis that
+school closures were effective for group ages older than 5 years old.
+One possibility is that school closures may lead to increased informal
+social gatherings among children and teenagers outside of controlled
+school environments, potentially facilitating virus transmission. It is
+also possible that transmission within households intensified as
+children spent more time at home. These results partially replicate
+observations from \citep{alfano2022effects} while providing a clearer
+picture of the structure of the effect of school closure.
+
+The Transfer Entropy calculations between age groups for each of the 12
+selected European countries reveal an absence of symmetry between
+influencing age groups (\(X\)) and influenced age groups (\(Y\)). Some
+age groups influence the COVID-19 case numbers of other age groups, but
+there is no reciprocal influence. This finding suggests that changes in
+COVID-19 cases in certain age groups can predict the changes in other
+age groups but not vice versa, i.e., causality can be unidirectional.
+These findings highlight the importance of considering intergenerational
+interactions in designing effective control measures.
+
+The entropy analysis revealed intriguing directional patterns of case
+counts between age groups. In some countries, cases among younger
+individuals appear predictive of cases among older individuals, while
+the opposite pattern is observed elsewhere. These differences may
+reflect country-specific variations in social mixing patterns,
+healthcare responses, or policy measures. The absence of consistent
+patterns in infections among adjacent age groups raises questions about
+the mechanisms of transmission and suggests that factors beyond simple
+proximity, such as behavioural differences or differing susceptibility,
+may play a role. Future work should investigate these underlying
+mechanisms to better understand age-based transmission dynamics.
+
+Furthermore, the quadrant analysis reveals similar patterns across
+multiple countries. Austria, Germany, and the Netherlands exhibit
+significantly higher TE coefficients in the upper left quadrant,
+indicating that changes in COVID-19 cases in younger cohorts predict the
+changes in older cohorts. On the other hand, Austria, the Netherlands,
+Portugal, and Spain show higher TE coefficients in the lower right
+quadrant, indicating that changes in COVID-19 cases in older cohorts
+predict the changes in younger cohorts. However, these differences may
+also reflect country-specific variations in social mixing patterns,
+healthcare responses, or policy measures. The absence of consistent
+patterns in infections among adjacent age groups raises questions about
+the mechanisms of transmission and suggests that factors beyond simple
+proximity, such as behavioural differences or differing susceptibility,
+may play a role.
+
+This study, despite its methodological strengths, is subject to certain
+limitations that warrant consideration. Foremost among these is the
+inherent challenge of perfectly isolating the effect of school closures
+from other simultaneously implemented and dynamically changing NPIs.
+While we incorporated four composite NPI index derived from the Oxford
+COVID-19 Government Response Tracker (OxCGRT) project to account for the
+broader public health context, residual confounding from unmeasured
+aspects of NPIs or behavioural changes may still persist. Our
+sensitivity analyses showed consistent directional effects for school
+closures even under varying overall NPI indexes provide some confidence,
+but this remains an inherent complexity in observational studies of this
+nature. Another consideration is the ecological design of the study,
+which uses aggregated country-level data; this may mask important
+sub-national variations and precludes inferences about individual-level
+risk factors. In addition, the definition of ``closure'' which excluded
+``partially open'' school statuses to simplify the intervention
+definition, means the impact of such intermediate states was not
+captured. Variations in testing strategies, case definitions, and
+reporting practices across countries and over time could also have
+influenced the observed case counts, despite efforts to use standardized
+data sources. Lastly, this study predominantly covers periods before
+widespread COVID-19 vaccination and the emergence of later variants such
+as Omicron; the effects of school closures might differ under those
+altered epidemiological conditions.
+
+Future studies should look more closely at why the virus spread between
+age groups as it did, and why TE patterns differed by country. This
+could involve using information on people's behaviour or more detailed
+data on other health rules. It would also be useful to study how
+different virus types and vaccination levels change these effects.
+Learning more about other influencing factors will help us better
+understand the real impact of school closures.
+
+\section{Conclusion}\label{conclusion}
+
+This comprehensive analysis of COVID-19 school closures across 12
+European nations reveals a complex and often counter-intuitive reality
+that challenges prevailing assumptions about their universal efficacy,
+particularly for children aged 5 and older. While national-level data
+showed an overall decrease in cases during closures, our
+age-disaggregated analysis using GAMs demonstrated that school-aged
+children (5-14 years) and young adults (15-24 years) frequently
+experienced periods of increased infections post-closure, a stark
+contrast to the consistent reductions seen only in pre-schoolers.
+Furthermore, Transfer Entropy analysis highlighted that the directional
+influence of transmission between age groups was asymmetric and varied
+by country, suggesting that the role of children in broader community
+transmission during such periods is not uniform.
+
+These findings underscore the critical importance of nuanced,
+age-specific analysis in evaluating public health interventions and
+caution against the broad application of measures like school closures
+without robust evidence of their benefit across all targeted
+demographics and a clear understanding of potential unintended
+consequences \citep{alfano2020efficacy, molefi2021impact}. For future
+pandemic preparedness, a more targeted, evidence-based approach is
+essential, prioritising measures that minimise societal disruption while
+effectively protecting public health across all age strata.
+
+\section{Author statements}\label{author-statements}
+
+\subsection{Ethical approval}\label{ethical-approval}
+
+This study is a longitudinal analysis on available data; thus, it does
+not require ethical approval.
+
+\subsection{Funding}\label{funding}
+
+None declared.
+
+\subsection{Competing interests}\label{competing-interests}
+
+None declared.
+
+\subsection{Data Availability}\label{data-availability}
+
+The data can be accessed from the COVerAGE-DB OSF repository
+(https://osf.io/mpwjq/), from the Oxford Covid-19 Government Response
+Tracker (OxCGRT) ``covid-policy-dataset'' GitHub repository
+(https://github.com/OxCGRT/covid-policy-dataset), and from the UNESCO
+servers
+(https://en.unesco.org/file/unesco-data-school-closures-february-2020-june-2022csv-zip).
+All preprocessing, analyses and code used to build the submitted paper
+are available at {[}URL hidden for anonymous peer-review{]}.
+
+\section{References}\label{references}
+
+\renewcommand{\bibsection}{}
+\bibliography{bibliography.bib}
 
